### 2022-02-10 paper.tex
first commit after having started locally

\documentclass[twocolumn,10pt]{article}

\usepackage[a4paper,hmargin=1.5cm,vmargin=2.5cm,]{geometry}
\setlength{\columnsep}{0.7cm}
\usepackage{palatino}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\urlstyle{same} % no tt font for URLs

\usepackage{natbib}
\bibliographystyle{genome_research}
%\setcitestyle{aysep={}}

\usepackage[dvipsnames]{xcolor}
\hypersetup{
    colorlinks,
    linkcolor={blue!50!black},
    citecolor={blue!50!black},
    urlcolor={blue!50!black}
}
\renewcommand{\dbltopfraction}{0.7}
\renewcommand{\textfraction}{0.2}

\begin{document}

\setcounter{secnumdepth}{0}

\twocolumn[{%
\centering
\textbf{\Large Analysing single-cell bisulfite sequencing data with scbs}
\vspace{1.5ex}

Lukas P. M. Kremer, Leonie Küchenhoff, ..., Simon Anders\\
{\footnotesize University of Heidelberg, Germany}
\vspace{1.5ex}

November 2021
\vspace{6ex}
}]

\section{Abstract}

[...]

\section{Introduction}

Sequencing-based assays with single-cell resolution have offered new means to understand the differences between the cells making up a sample. Single-cell RNA sequencing techniques have matured at great pace in recent years, and methods to study chromatin are rapidly catching up. Of special interest here is DNA methylation, i.e., the methylation of bases in the DNA, because DNA methylation patterns are passed on to daughter cells in cell division. Especially in mammals, such methylation chiefly occurs at cytosine bases that are immediately after followed by guanine bases, so-called CpG sites.

To detect CpG methylation by sequencing, bisulfite conversion is commonly employed: treatment of DNA with bisulfite causes unmethylated guanine bases to be deaminated to uracil, which is paired with thymine in subsequent PCR. Therefore, unmethylated CpGs are read as TG in sequencing of bisulfite-converted DNA while only methylated CpG is correctly read as CG \citep{Frommer_1992}. Such so-called bisulfite sequencing has been used for bulk samples since long, and methods for bisulfite sequencing with single-cell resolution \citep{Smallwood_2014} have become commercially available recently and are increasingly used.

CpG dinocleotides are underrepresented in eukariotic genomes (due to their vulnerability to mutation by conversion to TpG due to accidental deanimation), and the majority of existing CpG sites are methylated. Presence or absence of the methyl group at CpG sites is believed to have effect on gene regulation, with methylation inhibiting transcription. Methylating DNA during differentiation is therefore conjectured to be an important mechanism to silence genome regions that are not needed in the target lineage. One important application of single-cell bisulfite sequencing (sc-BS-Seq) therefore lies in the study of differentiation processes.

In the present paper, we discuss strategies to analyse sc-BS-Seq data, suggest improvement to current approaches, show their value in a benchmark, and present software to perform the improved analysis methods.

\section{Standard approach}

We start by briefly reviewing the standard approach of analysing single-cell \emph{RNA}-Seq data, and how this approach is typically adapted to the case of DNA methylation data.

The starting point in scRNA-Seq is a matrix of UMI counts (i.e. of counts of distinct RNA molecules), one row for each cell and one column for each gene. As a first step towards assigning cell types or states to cells, one needs to establish which cells are similar to each other, and to this end, find a way to quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile. A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat or ScanPy, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm. In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-5}$) to the normalized fractions before taking the logarithm. Now, one could use Euclidean distances of these vectors  of logarithmized fractions as dissimilarity score. However, these scores would be exceedingly noisy due to the strong Poisson noise introduced by the many genes with very low counts. Therefore, one performs a principal component analysis (PCA), keeping only the top few (typically, 20 to 50) components. As Poisson noise is uncorrelated between genes, it will average out in the top principal components. Therefore, Euclidean distances between these "PCA space" vectors provide a robust dissimilarity score, which can subsequently be used as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data. The PCA space representation is furthermore used for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).

This procedure is commonly adapted when working with single-cell DNA-methylation data, so that all the tools used in scRNA-Seq downstream of the PCA step can be used. The question is therefore how to construct a suitable matrix as input for the PCA. A simple and popular approach is to tile the genome into windows of, say, 100 kB size, and calculate for each cell the average methylation of the DNA within each window. To this end, we identify all CpG sites in the window for which we have coverage with at least one read and can hence call the CpG to be either unmethylated or methylated in the cell. Then we denote as average DNA methylation of this window in a given cell the proportion of observed CpG sites in the window have been found to be methylated. This yields a matrix, with one row for each cell and one column for each window from the genome tiling, comprising numbers (methylation fractions) between 0 and 1. This matrix is now subjected to PCA.

As an example, we will us in this paper data from XX et al., who have performed sc-BS on [...] and analysed their data in the manner just described. Figure 1a shows a UMAP plot, derived from our re-analysis of the data, using a tiling of the genome into windows of size 100 kb. As one can see, the approach is suitable to see clear structure in the data, observe ...

Nevertheless, this analysis strategy has substantial room for improvement, as we will discuss next.

\begin{figure}
\includegraphics[width=\columnwidth]{need_for_relative_meth.png}
\caption{\textbf{Effect of read position:} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified. Two cells cover differing parts of the interval with one read each. If one simply counts for each cell which fraction of the covered CpG sites are methylated, one obtains very different values for the two cells.}
\end{figure}

\section{Read position affects average}

\end{document}

### 2022-02-10 paper.tex
added a section

@@
 Nevertheless, this analysis strategy has substantial room for improvement, as we will discuss next.
+Here, we discuss the proposed improvements only qualitatively.
+For the mathematical details, please see the Methods section.
 
@@
 
-\section{Read position affects average}
+\section{Reducing variance due to differing read positions}
+
+We first discuss the task of quantifying the level of methylation in given, fixed, genomic interval.
+In Figure \ref{}, such an interval is covered by a single read for each of the two cells shown.
+The read from the cell 2 shows much more methylation than the read from cell 1, and a simple analysis would therefore consider that cell as having stronger methylation in the interval.
+However, given that the two reads agree wherever they overlap, a more parsimonous interpretation would be that the cells do not show difference in methylation within the interval.
+Rather, both cells, and similarly maybe most other cells, might have stronger methylation in the left third of the interval than in the middle one.
+
+Therefore, we propose to not first obtain, for each CpG position, an average methylation across all cells and, then quantify for each individual cell its deviation from this average.
+In Figure \ref{}, the curved line depicts such an average over all cells, the red vertical lines show an individual cell's deviation from the ensemble average.
+We take the lengths of the red lines as signed values (``redidual''), positive for lines extending upwards from the curve (methylated CpG) and negative for lines extending downwards (unmethylated CpG), and take the average over the residual for all the CpGs in the interval that are covered by reads from this cell.
+The value thus obtained is what we use to quantify this cell's (relative) methylation in the interval.
+For a genome tiled into such intervals, we thus obtain a matrix, one row per interval, one column per cell, that can be used for downstream analysis, e.g., as input for a PCA.
+
+The signal to noise ratio in a matrix thus obtained will be much better than in a matrix obtained by simply averaging absolute methylation (0 or 1) over all the cell's covered CpG in a region.
+The reason for this is that we reduce the variation in situation as the one depicted in FIgure \ref{}, where absolute methylation might differ strongly even though there is no actual evidence for a difference between the two cells.
+
+How should one obtain the ensemble average (the curved line in Fig.\ \ref{})?
+A simple approach to get a value for a spcific CpG would be to take all cells with read coverage for the CpG and use the fraction of these that show the CpG as methylated.
+However, especially when only few cells offer coverage, these averages will be very noisy.
+Therefore, we propose to smoothen using a kernel smoother, i.e., by performing a kernel-weighted average over the CpG site's neighborhood.
+The kernel bandwidth (i.e., the size of the neighborhood to average over) is a tuning parameter; we got good results with XXX bp and used this value for the examples presented here.
+
+\section{PCA with imputation}
+
+...
+
+\section{Finding variably methylated regions}
 

### 2022-02-25 paper.tex
added two sections

@@
 
-\section{Standard approach}
+\subsection{Standard approach}
 
@@
 
-\section{Reducing variance due to differing read positions}
+\section{Three improvements to processing scBS data}
+
+\subsection{Smoothing residuals}
 
@@
 
-\section{PCA with imputation}
+\subsection{PCA with imputation}
+
+The standard PCA algorithm cannot deal with missing data.
+Therefore, one would need to make intervals wide enough to ensure that every interval is covered by at least one read for most windows and discard any interval that is not covered by any single cell.
+Hence, a naive analysis has to resort to a coarse tiling of the genome.
+To counter this, we propose a simple and straightforward way to deal with missing data in the input matrix to a PCA: In a first iteration, we replace each missing value in the centered input matrix with zero, then run the PCA.
+Then, these zeroes are replaced by the value predicted by the PCA and the PCA is rerun.
+Convergence is typically achieved after two to four iterations.
+For details, see Methods.
+
+The possibility to perform PCA with missing data allows us to use a shorter and thus more intervals and hence provide the PCA with richer input data.
+
+\subsection{Finding variably methylated regions}
+
+Typically, some regions in a chromosome will have very similar methylation status in all cells (most commonly, fully methylated in nearly all cells) while other regions show variability in methylation across cells.
+Only the latter regions are of value for our goal of quantitating dissimilarity between cells.
+We call these the variably methylated regions (VMRs).
+
+So far, we have discussed dividing up (tiling) each chromosme genome into non-overlapping, equal-sized intervals, and quantitating the methylation of each such window.
+Such rigid placing of interval boundaries is unlikely to be optimal: for example, a VMR might be much smaller than a tile and its signal will hence be drowned out by the larger part of uninformative CpGs that are equal in all cells, when averaging over all the CpG sites in the tile.
+
+Therefore, we propose the folloing approach: Divide up the chromosome into many \emph{overlapping} windows, that start at regular multiples of a fixed, small, step size.
+Quantify the methylation of each cell in each window by averaging the cell's methylation residuals over all CpGs in the window, as described above and depicted in Figure\ \ref{}.
+Then, calculate for each window the variance of these values over all cells.
+Select, say, the top XX\% windows with the highest variances and mark them as VMRs.
+Wherever thus marked windows overlap or are divided by only a small gap, merge them into one larger VMR.
+Then, calculate for each of these laregr VMRs the methylation signal, as before, by averaging for each cell over the residuals of all contained CpGs.
+
+In this manner, we obtain a methylation matrix, with one column per cell and one row per VMR, that is, in a sense, richer in information and has better signal-to-noise ratio than the matrix obtained by the simple analysis sketched at the very beginning.
+As we demonstrate below, a PCA performed on such a matrix provided a distance metric for the cells that contains more information on biological detail than one from a simpler analysis.
+
+Besides providing better input for the distance calculations, the identified VMRs can also be compared with genomic annotation, thus providing a useful starting point for examining the data with methods from functional genomics.
+
+\section{The scbs Python package}
+
+We have implemented the approach just described in a Python package, called ``scbs'', that offers the following functionality: ...
+
+The code is open source, the source-code and the package are available from ..., where detailled documentation on usage is provided as well.
+
+\section{Comparison}
+
+Now, we can compare the simple analysis described initially with an analysis using our method, in order to demonstrate that it improves results.
+
+\section{Discussion and Conclusion}
 
@@
 
-\section{Finding variably methylated regions}
-
 \end{document}

### 2022-02-28 paper.tex
added methods

@@
 \setlength{\columnsep}{0.7cm}
-\usepackage{palatino}
+\usepackage{amsmath}
 \usepackage{graphicx}
@@
 
-November 2021
+March 2022
 \vspace{6ex}
@@
 
+\appendix
+
+\section{Appendix: Methods in Detail}
+
+\subsection{Raw data}
+
+We write $x_{ij}$ for the methylation status of CpG $i$ in cell $j$.
+The index $i$ runs over all CpG positions present in the genome, the index $j$ over all cells in the assay.
+We write $x_{ij}=0$ if position $i$ was found to be unmethylated in cell $j$ by bisulfite sequencing , $x_{ij}=1$ if it was methylated, and $x_{ij}=\text{NA}$ if position $i$ was not covered by reads from cell $j$ and the methylation status is therefore not available ("NA"),
+
+These values can be readily obtained from single-cell bisulfite sequencing data using tools like Bismark.
+
+If multiple reads with the same cell barcode cover the positions, these will typically be PCR duplicates of each other and hence agree.
+Of course, the two alleles of a CpG may differ in their methylation status and while it is in principle possible that one obtains reads stemming the same position on both the paternal and the maternal chromosomes of the same cell, but this is so unlikely that we can ignore the case.
+Hence, whenever Bismarck reports multiple reads covering the same position in the same cell, we set $x_{ij}$ to 0 or 1 following the majority of reads and put $x_{ij}=\text{NA}$ in case of a tie.
+
+For later use: We write $C$ for the set of all cells in the assay (i.e., $C$ is the index set for the cell indices $j$). Moreover,
+we define $C_i\subset C$ as the set of all those cells $j$ that have reads covering position $i$,
+$$ C_i=\{j\in C: x_{ij}\neq\text{NA}\}.$$
+Conversely, we define $G_j$ as the set of all the CpG positions $i$ covered by reads from cell $j$ 
+$$ G_j=\{i: x_{ij}\neq\text{NA}\}.$$
+
+\subsection{Data preparation}
+
+The function `scbs prepare` reads a set of CpG coverage files (e.g.
+produced by Bismark) and produces a file that stores the matrix $x$ in a space-efficient format, as follows: $x$ is represented as a NumPy sparse matrix, encoding the actual values 0, 1, and NA as -1, 1, and 0, respectively.
+Coding NA (the most common value) as 0 leverages NumPy's sparse matrix storage.
+In all the follows here, any mention of $x$ will, however, always mean the encoding as $x_{ij}\in\{0,1,\text{NA}\}$.
+
+\subsection{Smoothing}
+
+For each CpG position $i$, we write 
+$$\overline{x}_i=\langle x_{ij} \rangle_{j\in C_i} = \frac{1}{|C_i|}\sum_{j\in C_i} x_{ij}$$ 
+for the average methylation at position $i$, where $\langle\cdot\rangle$ denotes averaging, and the average runs over all the cells $i\in C_j$, i.e.
+over those cells for which methylation data is available for position $i$.
+
+NOTE: Here, we could already do shrinkage and use $\frac{1}{|C_i|+1}\sum_i x_{ij}$ intead. But we don't do that, right?
+
+We then run a kernel smoother over these per-position averages to obtain the smoothed averages $\tilde x_i$. Specifically, we use
+
+\[ \tilde x_i = \frac{\sum_{i'} \overline x_{i'}\, k_h(d_{ii'})}{\sum_{i'} k_h(d_{ii'})},\]
+i.e., $\tilde x_i$ is the weighted average over the per-position averages $\overline{x}_{i'}$, taken over the CpG sites $i'$ in the neighborhood of $i$, and weighted using a smoothing kernel $k_h$ with bandwidth $h$.
+Here, $d_{ii'}$ is the distance between CpG positions $i$ and $i'$, measured in base pairs, $h$ is the smoothing bandwidth in base pairs (by default, $h=1000$), and $k_h$ is the tricube kernel,
+
+\[ k_h(d) = \left\{
+\begin{aligned}
+	&\left(1-|d/h|^3\right)^3 &\text{for } |d|<h \\
+	&\,0 &\text{otherwise}. 
+\end{aligned}
+\right.
+\]
+
+NOTE: Here, we could weight the positions by number of covered cells, i.e., use
+
+$$ \tilde x_i = \frac{\sum_{i'} \overline x_{i'}\, k_h(d_{ii'})\, n_{i'}}{\sum_{i'} k_h(d_{ii'})\, n_{i'}},$$
+
+but we don't do that, or do we?
+
+\subsection{Methylation for an interval}
+
+Next, we discuss averaging methylation over a range of CpG sites.
+
+Given an interval $I$ on the chromosome, we wish to quantify the average methylation $m_{Ij}$ of the CpG sites within the interval for cell $j$.
+If we interpret $I$ as the set of CpG positions $i$ in the interval, we may write
+
+$$ m_{Ij} = \left< x_{ij} \right>_{i\in I\cap G_j}.$$
+
+Here, the average runs over all those sites $i$ that lie within the interval $I$ and are covered by reads from cell $j$.
+
+If we wish to compare cells, it can be helpful to center this quantity by subtracting its average:
+
+$$ z_{Ij} = m_{Ij} - \langle m_{Ij'}\rangle_{j'\in C}.$$
+
+As an alternative, we suggest to consider the residuals of the individual CpG methylation values $x_{ij}$ from the smoothed average $\tilde x_i$,
+$$ r_{ij} = x_{ij} - \tilde x_i, $$
+and averaging over these, obtaining
+$$ r_{Ij} = \frac{\scriptsize{1}}{\scriptsize{|I\cap C_i|+1}}\sum_{i\in I\cap C_i}\left( x_{ij} - \tilde x_i \right).$$
+
+This is a shrunken average, with denominator $n+1$.
+This extra pseudocount has the effect of shrinking the value towards the "neutral" value 0, with the shrinkage becoming stronger if the data is "weak" because the number $|I\cap C_i|$ of positions covered by reads from cell $j$  is low.
+In the extreme case of none of the reads from cell $j$ covering $I$, the sum becomes 0 and the denominator 1, i.e.
+$r_{Ij}=0$ in this case.
+
+\subsection{Finding variably methylated regions (VMRs)}
+
+For any interval $I$, we denote by $v_I$ the sampling variance of its residual averages $r_{Ij}$:
+
+$$ v_I = \frac{\scriptsize{1}}{\scriptsize{|C_I|-1}}\left( r_{Ij} - \langle r_{Ij'}\rangle_{j'\in C_I} \right)^2,$$
+
+where the average runs only over the set $C_I=\bigcup_{i\in I}C_i$ of those cells $j$ which have reads covering interval I.
+
+To find VMRs, we define intervals $I_1, I_2, ...$, all of the same width, and with step-wise increasing starts, then calculate $v_1, v_2, ...$ for these intervals.
+We then mark the intervals with the 2% highest variances.
+We take the union of all these intervals, split the union into connected components, and call each component a VMR.
+Putting that last step in other words: We take all the intervals with variance in the top 2-percentile, fuse intervals that overlap and call the regions thus obtained the VMRs.
+
+\subsection{Calculating cell to cell distances}
+
+[...]
+
+\bibliography{scbs}
+
 \end{document}

### 2022-02-28 paper.tex
added smoothng pics

@@
 As a first step towards assigning cell types or states to cells, one needs to establish which cells are similar to each other, and to this end, find a way to quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile.
-A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat or ScanPy, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm.
+A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm.
 In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-5}$) to the normalized fractions before taking the logarithm.
@@
 The PCA space representation is furthermore used for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).
+
+\begin{figure}
+	\framebox[\columnwidth]{
+		\begin{minipage}[c][3.5cm]{\columnwidth}
+			\centering Figure
+		\end{minipage}
+	}
+	\caption{\small UMAP representations of the single-cell bisulfite data of XXX et al., calculated \textbf{(A)} using conventional preprocessing and \textbf{(B)} using the approach described here.}
+	\label{umaps}
+\end{figure}
 
@@
 As an example, we will us in this paper data from XX et al., who have performed sc-BS on [...] and analysed their data in the manner just described.
-Figure 1a shows a UMAP plot, derived from our re-analysis of the data, using a tiling of the genome into windows of size 100 kb.
+Figure \ref{umaps}A shows a UMAP plot, derived from our re-analysis of the data, using a tiling of the genome into windows of size 100 kb.
 As one can see, the approach is suitable to see clear structure in the data, observe ...
@@
 \begin{figure}
+	A\\
 	\includegraphics[width=\columnwidth]{need_for_relative_meth.png}
-\caption{\textbf{Effect of read position:} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified.
+	B\\
+	.\hspace{.39cm}\includegraphics[width=.83\columnwidth]{smoothing_residuals.png}
+\caption{\small \textbf{Effect of read position:} \textbf{(A)} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified.
 Two cells cover differing parts of the interval with one read each.
-If one simply counts for each cell which fraction of the covered CpG sites are methylated, one obtains very different values for the two cells.}
+If one simply counts for each cell which fraction of the covered CpG sites are methylated, one obtains very different values for the two cells.
+\textbf{(B)} By averaging each CpG site's methylation over all cells and subsequent smoothing, the thick black ``average methylation curve'' is obtained.
+To quantify the methylation of cell 1 from (A) relative to this average over all cells, we propose to use the cell's residuals to the smoothed curve (lengths of the vertical red lines) and take their average, counting residuals of methylated CpGs positive and residuals of unmethylated CpGs negative.}
+\label{smoothres}
 \end{figure}
@@
 We first discuss the task of quantifying the level of methylation in given, fixed, genomic interval.
-In Figure \ref{}, such an interval is covered by a single read for each of the two cells shown.
+In Figure \ref{smoothres}A, such an interval is covered by a single read for each of the two cells shown.
 The read from the cell 2 shows much more methylation than the read from cell 1, and a simple analysis would therefore consider that cell as having stronger methylation in the interval.
@@
 The signal to noise ratio in a matrix thus obtained will be much better than in a matrix obtained by simply averaging absolute methylation (0 or 1) over all the cell's covered CpG in a region.
-The reason for this is that we reduce the variation in situation as the one depicted in FIgure \ref{}, where absolute methylation might differ strongly even though there is no actual evidence for a difference between the two cells.
+The reason for this is that we reduce the variation in situation as the one depicted in Figure \ref{smotthres}A, where absolute methylation might differ strongly even though there is no actual evidence for a difference between the two cells.
 
-How should one obtain the ensemble average (the curved line in Fig.\ \ref{})?
+How should one obtain the ensemble average (the curved line in Fig.\ \ref{smoothres}B)?
 A simple approach to get a value for a spcific CpG would be to take all cells with read coverage for the CpG and use the fraction of these that show the CpG as methylated.
@@
 
+\begin{figure}
+	\includegraphics[width=\columnwidth]{vmr.png}
+	\caption{\small \textbf{Finding variably methylated regions (VMR):} The chromosomes are divided up into overlapping windows (first four shown at the bottom), and for each window, the cells' methylation values are calculated as described and as depicted in Fig.\ \ref{smoothres}B.
+Then, the variance of these values are calculated (each point represents one of the overlapping windows), a threshold (red line) is chosen such that a chosen quantile of windows have a variance exceeding the threshold.
+Windows with above-threshold variance are merged if they overlap, yielding the ``variably methylated regions''.}
+	\label{vmr}
+\end{figure}
+
 Typically, some regions in a chromosome will have very similar methylation status in all cells (most commonly, fully methylated in nearly all cells) while other regions show variability in methylation across cells.
@@
 
-Therefore, we propose the folloing approach: Divide up the chromosome into many \emph{overlapping} windows, that start at regular multiples of a fixed, small, step size.
-Quantify the methylation of each cell in each window by averaging the cell's methylation residuals over all CpGs in the window, as described above and depicted in Figure\ \ref{}.
+Therefore, we propose the following approach (Fig.
+\ref{vmr}): Divide up the chromosome into many \emph{overlapping} windows, that start at regular multiples of a fixed, small, step size.
+Quantify the methylation of each cell in each window by averaging the cell's methylation residuals over all CpGs in the window, as described above and depicted in Figure\ \ref{smoothres}B.
 Then, calculate for each window the variance of these values over all cells.
@@
 
-\appendix
-
-\section{Appendix: Methods in Detail}
+\section{Methods}
 
@@
 
-\bibliography{scbs}
+{\small \bibliography{scbs}}
 

### 2022-02-28 paper.tex
minor

@@
 and averaging over these, obtaining
-$$ r_{Ij} = \frac{\scriptsize{1}}{\scriptsize{|I\cap C_i|+1}}\sum_{i\in I\cap C_i}\left( x_{ij} - \tilde x_i \right).$$
+\begin{equation} 
+r_{Ij} = \frac{\scriptsize{1}}{\scriptsize{|I\cap C_i|+1}}\sum_{i\in I\cap C_i}\left( x_{ij} - \tilde x_i \right). \label{avgres}
+\end{equation}
 
@@
 To find VMRs, we define intervals $I_1, I_2, ...$, all of the same width, and with step-wise increasing starts, then calculate $v_1, v_2, ...$ for these intervals.
-We then mark the intervals with the 2% highest variances.
+We then mark the intervals with the 2\% highest variances.
 We take the union of all these intervals, split the union into connected components, and call each component a VMR.
@@
 \subsection{Calculating cell to cell distances}
+
+Given a set, $\mathcal{V}=\{I^\text{v}_1,I^\text{v}_2,\dots\}$, of intervals corresponding to VMRs, we get a relative methylation fraction $r_{ij}$ for each VMR $I^\text{v}_i$ and each cell $j$ from Eq.\ ()\ref{avgres}).
+The matrix thus obtained can then be centered and used as input for a PCA.
+If we calculate the top $R$ principal components, we thus obtain for each cell $j$ an $R$-dimensional principal component vector $\mathbf{x}^\text{P}_j$.
+For any two cells $j$, $j'$, we use the Euclidean distance $\|\mathbf{x}^\text{P}_j - \mathbf{x}^\text{P}_{j'}\|$ as the measure of dissimilarity of the two cells.
+Thus, the matrix of PC scores can be used as input to dimension reduction methods like t-SNE or UMAP, and to clustering methods like the Louvain or Leiden algorithm, which requirte such a matrix as input to the approximate mnearest neighbor (ANN) finding algorithm that is their first step.
+
+\subsection{PCA with imputation}
 

### 2022-03-16 part_leonie_git/part_Leonie.tex
Add files via upload
 In order to keep a good structure of the paper, I decided to make a separate file with a short description of the plots I made. (And the plots).

\documentclass[twocolumn,10pt]{article}

\usepackage[a4paper,hmargin=1.5cm,vmargin=2.5cm,]{geometry}
\setlength{\columnsep}{0.7cm}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\urlstyle{same} % no tt font for URLs

\usepackage{natbib}
\bibliographystyle{genome_research}
%\setcitestyle{aysep={}}

\usepackage[dvipsnames]{xcolor}
\hypersetup{
    colorlinks,
    linkcolor={blue!50!black},
    citecolor={blue!50!black},
    urlcolor={blue!50!black}
}
\renewcommand{\dbltopfraction}{0.7}
\renewcommand{\textfraction}{0.2}

\begin{document}

\setcounter{secnumdepth}{0}

\twocolumn[{%
\centering
\textbf{\Large Analysing single-cell bisulfite sequencing data with scbs}
\vspace{1.5ex}

Lukas P. M. Kremer, Leonie Küchenhoff, ..., Simon Anders\\
{\footnotesize University of Heidelberg, Germany}
\vspace{1.5ex}

March 2022
\vspace{6ex}
}]

\section{Comparison}

To demonstrate an improvement of results, we compare previously published analyses described above with our analysis method.
First, we use a dataset by \citealp{Argelaguet_2019}. In their work, they generated a single cell methylome dataset of mouse embryo cells across different stages of development (E4.5 – E7.5). They calculated the correlation between RNA expression intensity and corresponding DNA methylation at the promoter region to elucidate the implications of the epigenome on RNA expression during development. As promoters, they used DNA regions of $\pm$ 2kb around transcriptions start sites (TSS).

--The following paragraph potentially needs to be changed when new plots are made--

In our analysis, we compare correlations between RNA expression intensity and DNA methylation of these promotor regions to correlations between RNA expression intensity and DNA methylation of the closest variable methylation regions that we identified with scbs. In figure \ref{figure:correlation}, we show that correlation of RNA expression intensity to gene methylation around the TSS $\pm$ 2kb is less strong compared to the correlation at the closest variable region. This effect can be observed with mehtylation fractions (figure \ref{figure:correlation}A) or shrunken residuals (figure \ref{figure:correlation}B) as data inputs.

\begin{figure}[!htp]
	A\\
	\hspace{.2cm}\includegraphics[width=.7\columnwidth]{leonie_plots/corr_methylation_fraction_4kbwindow.pdf} \\
	B\\
	\hspace{.2cm}\includegraphics[width=.7\columnwidth]{leonie_plots/corr_shrunken_residual_4kbwindow.pdf} \\
	C\\
	\hspace{.39cm}\includegraphics[width=\columnwidth]{leonie_plots/comparison_meth_RNA.pdf}

	\caption{\small \textbf{Correlation of DNA methylation and RNA expression intensity}. All plots are based on the data by \citealp{Argelaguet_2019}.
	\textbf{(A)} Correlation of each gene's RNA expression intensity to methylation fraction of TSS $\pm$ 2kb (y-axis) or to methylation fraction of closest variable gene.
	\textbf{(B)} Correlation of each gene's RNA expression intensity to shrunken residual of methylation of TSS $\pm$ 2kb (y-axis) or to shrunken residual of closest variable gene.
	\textbf{(C)} Mean gene methylation of Ube2l6 gene (ENSMUSG00000027078, marked in yellow in A \& B). Cells were into 4 groups based on their RNA expression (group 0: cells with no expression of gene, group 1-3: cells that express gene divided into three equally large groups with group 3 having the highest expression). Mean methylation level of each group was smoothed with a tricube kernel of bandwidth = 2000.}
	\label{figure:correlation}
\end{figure}

As an example to illustrate this effect more clearly, we picked a gene with significantly stronger correlation at the closest variable region than at the TSS. This gene is marked on figure \ref{figure:correlation}A and B with a yellow dot. We divided all cells into four groups of cells with no, low, medium and high expression of this exemplary gene Ube216 and plotted a smoothed methylation level per group. As shown in figure \ref{figure:correlation}C, the promoter region is sparsely methylated in all four RNA expression groups and is therefore a poor region to make conclusions about the relation of DNA methylation and RNA expression intensity. However, a region close to the TSS, identified as the closest highly variable methylation region with scbs, shows a much clearer negative correlation between RNA expression intensity and DNA methylation.

To further demonstrate the benefit of detecting variable methylation regions, we used a second dataset, published by \citealp{Luo_2017}. In their work, \citealp{Luo_2017} generated methylomes from single cell neuronal nuclei in mice to identify genomic regulatory elements across neuron types. To compare how well cell types are separated by different analysis techniques, we processed the same dataset into 2-dimensional UMAPs several times. Subsequently, we compared the results by assigning a score to each output based on cell type separation in the final 15 principal components. We see an improvement of cell separation when comparing the analysis of variable methylation regions to 100 kb regions (\ref{figure:score}A and C). The difference of using shrunken residuals instead of methylation fractions to assess the methylation per region is not as significant. However, when emulating a sparse dataset with less cells by randomly reducing the dataset to 500 cells, it becomes clear that using shrunken residuals over methylation fractions also improves the outcome of our analysis (\ref{figure:score}B and D).

\begin{figure}[!htp]
	A\\
	\hspace{.3cm}\includegraphics[width=.7\columnwidth]{leonie_plots/complete_135k_12cm_log.pdf} \\
	B\\
	\hspace{.3cm}\includegraphics[width=.7\columnwidth]{leonie_plots/cell500_20k_12cm_log.pdf} \\
	C\\
	\hspace{.39cm}\includegraphics[width=\columnwidth]{leonie_plots/UMAP_fulldataset.pdf}
	D\\
	\hspace{.39cm}\includegraphics[width=\columnwidth]{leonie_plots/UMAP_reduceddataset.pdf}
	\caption{\small \textbf{Analysis comparison}. All plots are based on the data by \citealp{Luo_2017}. \textbf{(A)} Nearest neighbor count score for complete dataset using four different analysis techniques and a variable number of regions as input. Exemplary UMAPs for encircled scores are shown in (C). \textbf{(B)} Nearest neighbor count score for the same dataset reduced to 500 cells using four different analysis techniques and a variable number of regions as input. Exemplary UMAPs for encircled scores are shown in (D). \textbf{(C)} Two exemplary UMAPs using 1000 regions of the full dataset and methylation fraction of 100 kb windows (left) or shrunken residuals of variable methylation regions (right) \textbf{(D)} Two exemplary UMAPs using 1000 regions of the reduced dataset and methylation fraction of 100 kb windows (left) or shrunken residuals of variable methylation regions (right).}
	\label{figure:score}
\end{figure}

\section{Discussion and Conclusion}

...

\section{Methods}

\subsection{Data processing}
To analyze CpG methylation data by \citealp{Luo_2017}, the mouse genome was first separated into regions. Regions were either obtained by tiling the genome into non-overlapping 100 kb bins or by acquiring the most variable methylation regions with the scbs script as described above. For each region, the mean methylation fraction or methylation residual was calculated. Each region that contained less than four measured CpGs or that was measured in less than 10\% of cells was removed from the dataset. To select $n$ regions for further analysis, top $n$ regions were picked based on the cell number the region was measured in. 
For dimensionality reduction, missing values were imputed with an iterative PCA as described below. To visualize data, a UMAP was performed on the final 15 principal components.

\subsection{Scoring}
To assess data processing performance of several DNA methylation analysis techniques, we processed single cell DNA methylation data with several approaches and scored each result. Scores evaluate the ability of each method to separate cell types in a 15-dimensional space of the first 15 principal components based on their CpG methylation. Each score varies between 0 and 1, where higher scores reflect better data processing performance.
\subsubsection{Nearest Neighbor Count Score}
The nearest neighbor count score $nnc(k)$ is based on the $\Gamma -Score$ published by \citealp{Kireeva_2014}. Each cell $c_j^i$ belongs to a cell type $j$. For each cell, $k$ nearest neighbors $nn_j^k$ were computed based on their Euclidean Distance. Based on the cell type $j$ of the $k$ nearest neighbors of each cell, the nnc score $nnc(k)$ was computed with fomula (1), in which $d(c_j^i, k_j^i)=1$ if at least 90\% of neighbor classes match the class of cell $c_j^i$ and $d(c_j^i, k_j^i)=0$ if less than 90\% of classes match, while $n$ are the total number of cells.
\begin{equation}
	nnc(k) = \frac{1}{n}d(c_j^i, k_j^i)
\end{equation}

{\small \bibliography{scbs}}

\end{document}

### 2022-05-30 paper.tex
add Leonies part, unchanged for now

@@
 Methylating DNA during differentiation is therefore conjectured to be an important mechanism to silence genome regions that are not needed in the target lineage.
-One important application of single-cell bisulfite sequencing (sc-BS-Seq) therefore lies in the study of differentiation processes.
+One important application of single-cell bisulfite sequencing (scBS) therefore lies in the study of differentiation processes.
 
-In the present paper, we discuss strategies to analyse sc-BS-Seq data, suggest improvement to current approaches, show their value in a benchmark, and present software to perform the improved analysis methods.
+In the present paper, we discuss strategies to analyse scBS data, suggest improvement to current approaches, show their value in a benchmark, and present software to perform the improved analysis methods.
 
@@
 
-As an example, we will us in this paper data from XX et al., who have performed sc-BS on [...] and analysed their data in the manner just described.
+As an example, we will us in this paper data from XX et al., who have performed scBS on [...] and analysed their data in the manner just described.
 Figure \ref{umaps}A shows a UMAP plot, derived from our re-analysis of the data, using a tiling of the genome into windows of size 100 kb.
@@
 	A\\ 
-	\includegraphics[width=\columnwidth]{need_for_relative_meth.png}
+	\includegraphics[width=\columnwidth]{figures/Fig_residuals_A.png}
 	B\\
-	.\hspace{.39cm}\includegraphics[width=.83\columnwidth]{smoothing_residuals.png}
+	\includegraphics[width=\columnwidth]{figures/Fig_residuals_B.png}
 \caption{\small \textbf{Effect of read position:} \textbf{(A)} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified.
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{vmr.png}
-	\caption{\small \textbf{Finding variably methylated regions (VMR):} The chromosomes are divided up into overlapping windows (first four shown at the bottom), and for each window, the cells' methylation values are calculated as described and as depicted in Fig.\ \ref{smoothres}B.
+	\includegraphics[width=\columnwidth]{figures/Fig_sliding.png}
+	\caption{\small \textbf{Finding variably methylated regions (VMR):} The chromosomes are divided up into overlapping windows (first five shown at the bottom), and for each window, the cells' methylation values are calculated as described and as depicted in Fig.\ \ref{smoothres}B.
 Then, the variance of these values are calculated (each point represents one of the overlapping windows), a threshold (red line) is chosen such that a chosen quantile of windows have a variance exceeding the threshold.
@@
 
-We have implemented the approach just described in a Python package, called ``scbs'', that offers the following functionality: ...
+We have implemented the approach just described in a Python package, called ``scbs'', that offers the following functionality:
+
+\subsection{Efficient storage and access of whole-genome scBS data}
+
+\subsection{Identification and filtering of low-quality cells}
+
+\subsection{Finding variably methylated regions}
+
+\subsection{Quantification of methylation in genomic intervals}
 
@@
 
-Now, we can compare the simple analysis described initially with an analysis using our method, in order to demonstrate that it improves results.
+To demonstrate an improvement of results, we compare previously published analyses described above with our analysis method.
+First, we use a dataset by \citealp{Argelaguet_2019}.
+In their work, they generated a single cell methylome dataset of mouse embryo cells across different stages of development (E4.5 – E7.5).
+They calculated the correlation between RNA expression intensity and corresponding DNA methylation at the promoter region to elucidate the implications of the epigenome on RNA expression during development.
+As promoters, they used DNA regions of $\pm$ 2kb around transcriptions start sites (TSS).
+
+--The following paragraph potentially needs to be changed when new plots are made--
+
+In our analysis, we compare correlations between RNA expression intensity and DNA methylation of these promotor regions to correlations between RNA expression intensity and DNA methylation of the closest variable methylation regions that we identified with scbs.
+In figure \ref{figure:correlation}, we show that correlation of RNA expression intensity to gene methylation around the TSS $\pm$ 2kb is less strong compared to the correlation at the closest variable region.
+This effect can be observed with mehtylation fractions (figure \ref{figure:correlation}A) or shrunken residuals (figure \ref{figure:correlation}B) as data inputs.
+
+\begin{figure}[!htp]
+	A\\
+	\hspace{.2cm}\includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/corr_methylation_fraction_4kbwindow.pdf} \\
+	B\\
+	\hspace{.2cm}\includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/corr_shrunken_residual_4kbwindow.pdf} \\
+	C\\
+	\hspace{.39cm}\includegraphics[width=\columnwidth]{part_leonie_git/leonie_plots/comparison_meth_RNA.pdf}
+
+	\caption{\small \textbf{Correlation of DNA methylation and RNA expression intensity}. All plots are based on the data by \citealp{Argelaguet_2019}.
+	\textbf{(A)} Correlation of each gene's RNA expression intensity to methylation fraction of TSS $\pm$ 2kb (y-axis) or to methylation fraction of closest variable gene.
+	\textbf{(B)} Correlation of each gene's RNA expression intensity to shrunken residual of methylation of TSS $\pm$ 2kb (y-axis) or to shrunken residual of closest variable gene.
+	\textbf{(C)} Mean gene methylation of Ube2l6 gene (ENSMUSG00000027078, marked in yellow in A \& B).
+Cells were into 4 groups based on their RNA expression (group 0: cells with no expression of gene, group 1-3: cells that express gene divided into three equally large groups with group 3 having the highest expression).
+Mean methylation level of each group was smoothed with a tricube kernel of bandwidth = 2000.}
+	\label{figure:correlation}
+\end{figure}
+
+As an example to illustrate this effect more clearly, we picked a gene with significantly stronger correlation at the closest variable region than at the TSS.
+This gene is marked on figure \ref{figure:correlation}A and B with a yellow dot.
+We divided all cells into four groups of cells with no, low, medium and high expression of this exemplary gene Ube216 and plotted a smoothed methylation level per group.
+As shown in figure \ref{figure:correlation}C, the promoter region is sparsely methylated in all four RNA expression groups and is therefore a poor region to make conclusions about the relation of DNA methylation and RNA expression intensity.
+However, a region close to the TSS, identified as the closest highly variable methylation region with scbs, shows a much clearer negative correlation between RNA expression intensity and DNA methylation.
+
+To further demonstrate the benefit of detecting variable methylation regions, we used a second dataset, published by \citealp{Luo_2017}.
+In their work, \citealp{Luo_2017} generated methylomes from single cell neuronal nuclei in mice to identify genomic regulatory elements across neuron types.
+To compare how well cell types are separated by different analysis techniques, we processed the same dataset into 2-dimensional UMAPs several times.
+Subsequently, we compared the results by assigning a score to each output based on cell type separation in the final 15 principal components.
+We see an improvement of cell separation when comparing the analysis of variable methylation regions to 100 kb regions (\ref{figure:score}A and C).
+The difference of using shrunken residuals instead of methylation fractions to assess the methylation per region is not as significant.
+However, when emulating a sparse dataset with less cells by randomly reducing the dataset to 500 cells, it becomes clear that using shrunken residuals over methylation fractions also improves the outcome of our analysis (\ref{figure:score}B and D).
+
+\begin{figure}[!htp]
+	A\\
+	\hspace{.3cm}\includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/complete_135k_12cm_log.pdf} \\
+	B\\
+	\hspace{.3cm}\includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/cell500_20k_12cm_log.pdf} \\
+	C\\
+	\hspace{.39cm}\includegraphics[width=\columnwidth]{part_leonie_git/leonie_plots/UMAP_fulldataset.pdf}
+	D\\
+	\hspace{.39cm}\includegraphics[width=\columnwidth]{part_leonie_git/leonie_plots/UMAP_reduceddataset.pdf}
+	\caption{\small \textbf{Analysis comparison}.
+All plots are based on the data by \citealp{Luo_2017}.
+\textbf{(A)} Nearest neighbor count score for complete dataset using four different analysis techniques and a variable number of regions as input.
+Exemplary UMAPs for encircled scores are shown in (C).
+\textbf{(B)} Nearest neighbor count score for the same dataset reduced to 500 cells using four different analysis techniques and a variable number of regions as input.
+Exemplary UMAPs for encircled scores are shown in (D).
+\textbf{(C)} Two exemplary UMAPs using 1000 regions of the full dataset and methylation fraction of 100 kb windows (left) or shrunken residuals of variable methylation regions (right) \textbf{(D)} Two exemplary UMAPs using 1000 regions of the reduced dataset and methylation fraction of 100 kb windows (left) or shrunken residuals of variable methylation regions (right).}
+	\label{figure:score}
+\end{figure}
 

### 2022-06-01 paper.tex
add workflow figure

@@
 
-Lukas P. M. Kremer, Leonie Küchenhoff, ..., Simon Anders\\
+Lukas P.~M. Kremer, Leonie Küchenhoff, Santiago Cerrizuela, ..., Ana Martin-Villalba, Simon Anders\\
 {\footnotesize University of Heidelberg, Germany}
@@
 	\caption{\small \textbf{Finding variably methylated regions (VMR):} The chromosomes are divided up into overlapping windows (first five shown at the bottom), and for each window, the cells' methylation values are calculated as described and as depicted in Fig.\ \ref{smoothres}B.
-Then, the variance of these values are calculated (each point represents one of the overlapping windows), a threshold (red line) is chosen such that a chosen quantile of windows have a variance exceeding the threshold.
+Then, the variance of these values is calculated (each point represents one of the overlapping windows), a threshold (dashed line) is chosen such that a chosen quantile of windows have a variance exceeding the threshold.
 Windows with above-threshold variance are merged if they overlap, yielding the ``variably methylated regions''.}
@@
 
-We have implemented the approach just described in a Python package, called ``scbs'', that offers the following functionality:
+\begin{figure}
+	\includegraphics[width=\columnwidth]{figures/Fig_workflow.png}
+	\caption{\small \textbf{Overview of the functionalities implemented in the scbs package} Bla bla.}
+	\label{workflow}
+\end{figure}
+
+We have implemented the approach just described in a Python package, called ``scbs'', that offers the following functionality \ref{workflow}:
 
@@
 The function `scbs prepare` reads a set of CpG coverage files (e.g.
-produced by Bismark) and produces a file that stores the matrix $x$ in a space-efficient format, as follows: $x$ is represented as a NumPy sparse matrix, encoding the actual values 0, 1, and NA as -1, 1, and 0, respectively.
-Coding NA (the most common value) as 0 leverages NumPy's sparse matrix storage.
+produced by Bismark) and produces a file that stores the matrix $x$ in a space-efficient format, as follows: $x$ is represented as a SciPy sparse matrix, encoding the actual values 0, 1, and NA as -1, 1, and 0, respectively.
+Coding NA (the most common value) as 0 leverages SciPy's sparse matrix storage.
 In all the follows here, any mention of $x$ will, however, always mean the encoding as $x_{ij}\in\{0,1,\text{NA}\}$.

### 2022-06-09 paper.tex
add section about capabilities of the python package

@@
 \textbf{\Large Analysing single-cell bisulfite sequencing data with scbs}
-\vspace{1.5ex}
 
-Lukas P.~M. Kremer, Leonie Küchenhoff, Santiago Cerrizuela, ..., Ana Martin-Villalba, Simon Anders\\
-{\footnotesize University of Heidelberg, Germany}
+Lukas P.~M. Kremer\textsuperscript{1,2}, Leonie Küchenhoff\textsuperscript{1,2}, Santiago Cerrizuela\textsuperscript{2}, ..., Ana Martin-Villalba\textsuperscript{2}, Simon Anders\textsuperscript{1}\\
+{\footnotesize \textsuperscript{1} BioQuant, University of Heidelberg, Germany}\\
+{\footnotesize \textsuperscript{2} Molecular Neurobiology, German Cancer Research Center (DKFZ), Heidelberg, Germany}
 \vspace{1.5ex}
@@
 Sequencing-based assays with single-cell resolution have offered new means to understand the differences between the cells making up a sample.
-Single-cell RNA sequencing techniques have matured at great pace in recent years, and methods to study chromatin are rapidly catching up.
+Single-cell RNA sequencing (scRNA-seq) techniques have matured at great pace in recent years, and methods to study chromatin are rapidly catching up.
 Of special interest here is DNA methylation, i.e., the methylation of bases in the DNA, because DNA methylation patterns are passed on to daughter cells in cell division.
@@
 
-To detect CpG methylation by sequencing, bisulfite conversion is commonly employed: treatment of DNA with bisulfite causes unmethylated guanine bases to be deaminated to uracil, which is paired with thymine in subsequent PCR.
+To detect CpG methylation by sequencing, bisulfite conversion is commonly employed: treatment of DNA with bisulfite causes unmethylated cytosine bases to be deaminated to uracil, which is paired with thymine in subsequent PCR.
 Therefore, unmethylated CpGs are read as TG in sequencing of bisulfite-converted DNA while only methylated CpG is correctly read as CG \citep{Frommer_1992}.
-Such so-called bisulfite sequencing has been used for bulk samples since long, and methods for bisulfite sequencing with single-cell resolution \citep{Smallwood_2014} have become commercially available recently and are increasingly used.
+Such so-called bisulfite sequencing has been used for bulk samples since long, and methods for bisulfite sequencing with single-cell resolution (scBS) were recently developed \citep{Smallwood_2014} and are increasingly used.
 
-CpG dinocleotides are underrepresented in eukariotic genomes (due to their vulnerability to mutation by conversion to TpG due to accidental deanimation), and the majority of existing CpG sites are methylated.
+CpG dinocleotides are underrepresented in eukaryotic genomes (due to their vulnerability to mutation by conversion to TpG due to accidental deanimation), and the majority of existing CpG sites are methylated.
 Presence or absence of the methyl group at CpG sites is believed to have effect on gene regulation, with methylation inhibiting transcription.
@@
 
-We start by briefly reviewing the standard approach of analysing single-cell \emph{RNA}-Seq data, and how this approach is typically adapted to the case of DNA methylation data.
+We start by briefly reviewing the standard approach of analysing single-cell \emph{RNA}-Seq data, and how this approach is typically adapted to the case of single-cell DNA methylation data.
 
@@
 As Poisson noise is uncorrelated between genes, it will average out in the top principal components.
-Therefore, Euclidean distances between these "PCA space" vectors provide a robust dissimilarity score, which can subsequently be used as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data.
+Therefore, Euclidean distances between these ``PCA space'' vectors provide a robust dissimilarity score, which can subsequently be used as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data.
 The PCA space representation is furthermore used for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).
@@
 
-\subsection{Smoothing residuals}
+\subsection{Smoothing residuals} \label{residuals}
 
@@
 	\includegraphics[width=\columnwidth]{figures/Fig_workflow.png}
-	\caption{\small \textbf{Overview of the functionalities implemented in the scbs package} Bla bla.}
+	\caption{\small \textbf{Overview of the functionalities implemented in the scbs package}
+	A 
+	}
 	\label{workflow}
@@
 
-We have implemented the approach just described in a Python package, called ``scbs'', that offers the following functionality \ref{workflow}:
+We have implemented the approach just described in a Python package with a command line interface, called \texttt{scbs}, which also offers a number of other functionalities for the analysis of scBS data Figure \ref{workflow}.
+The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark} or methylpy \citep{methylpy}.
+For each cell, these tools produce a text file that lists the methylation status of all CpG sites.
+Since it is inconvenient to work with hundreds or thousands of text files, \texttt{scbs} provides the \texttt{prepare} command which parses methylation files and stores their content in a compressed format that enables efficient access to all CpG sites of the genome.
 
-\subsection{Efficient storage and access of whole-genome scBS data}
+\texttt{scbs prepare} also computes a number of summary statistics for each cell, including the mean genome-wide methylation level and the number of observed CpG sites, i.~e.
+the number of CpG sites that have sequencing coverage.
+These summary statistics can be used to detect cells with poor quality.
+The quality of each single cell methylome can furthermore be inspected with \texttt{scbs profile}, which computes the average methylation profile of a set of user-defined genomic regions such as transcription start sites (TSS).
+The TSS profile is a useful quality control plot since methylation shows a characteristic dip roughly \textpm1~kb around the transcription start site in mammalian genomes.
+Cells that do not show this pattern, or cells with few observed CpG sites, may then be filtered from the data set with \texttt{scbs filter}.
 
-\subsection{Identification and filtering of low-quality cells}
+After quality control, the user has access to the previously described genome-wide VMR detection approach via \texttt{scbs scan}.
+This produces a BED-file that lists the genome coordinates of VMRs.
+To finally obtain a methylation matrix analogous to an scRNA-seq count matrix, this BED-file can be used as input for \texttt{scbs matrix}, which quantifies methylation at genomic intervals in all cells.
+The command produces both the percentage of methylated CpG sites, as well as our proposed methylation measure (shrunken mean of the residuals) that is more robust to variation in sequencing coverage or stochastic differences in read position between cells.
+We note that both \texttt{matrix} and \texttt{profile} accept any valid BED-file as input, which means that the user can quantify and visualize methylation at any genomic feature of interest, such as promoters, enhancers or transcription factor binding sites.
+The obtained methylation matrix can then serve as input for common single-cell analysis methods such as dimensionality reduction and cell clustering.
 
-\subsection{Finding variably methylated regions}
-
-\subsection{Quantification of methylation in genomic intervals}
-
-The code is open source, the source-code and the package are available from ..., where detailled documentation on usage is provided as well.
+\texttt{scbs} is free and open source, and the package can be installed via the Python package manager.
+The source code and the package are available at \href{https://github.com/LKremer/scbs}, where detailed documentation on usage, as well as a tutorial using example data, is provided as well.
 
@@
 
-NOTE: Here, we could weight the positions by number of covered cells, i.e., use
+We also provide Here, we could weight the positions by number of covered cells, i.e., use
 
@@
 
-For any interval $I$, we denote by $v_I$ the sampling variance of its residual averages $r_{Ij}$:
+For any interval $I$, we denote by $v_I$ the variance of its residual averages $r_{Ij}$:
 
-$$ v_I = \frac{\scriptsize{1}}{\scriptsize{|C_I|-1}}\left( r_{Ij} - \langle r_{Ij'}\rangle_{j'\in C_I} \right)^2,$$
+$$ v_I = \frac{\scriptsize{1}}{\scriptsize{|C_I|}}\left( r_{Ij} - \langle r_{Ij'}\rangle_{j'\in C_I} \right)^2,$$
 

### 2022-06-10 paper.tex
fix most latex warnings

@@
 
-CpG dinocleotides are underrepresented in eukaryotic genomes (due to their vulnerability to mutation by conversion to TpG due to accidental deanimation), and the majority of existing CpG sites are methylated.
+CpG dinocleotides are underrepresented in eukaryotic genomes (due to their vulnerability to mutation by conversion to TpG due to accidental deamination), and the majority of existing CpG sites are methylated.
 Presence or absence of the methyl group at CpG sites is believed to have effect on gene regulation, with methylation inhibiting transcription.
@@
 
-In the present paper, we discuss strategies to analyse scBS data, suggest improvement to current approaches, show their value in a benchmark, and present software to perform the improved analysis methods.
+In the present paper, we discuss strategies to analyze scBS data, suggest improvement to current approaches, show their value in a benchmark, and present software to perform the improved analysis methods.
 
@@
 
-We start by briefly reviewing the standard approach of analysing single-cell \emph{RNA}-Seq data, and how this approach is typically adapted to the case of single-cell DNA methylation data.
+We start by briefly reviewing the standard approach of analyzing single-cell \emph{RNA}-Seq data, and how this approach is typically adapted to the case of single-cell DNA methylation data.
 
@@
 
-\begin{figure}
+\begin{figure*}
 	A\\
-	\includegraphics[width=\columnwidth]{figures/Fig_residuals_A.png}
+    \begin{center}
+        \includegraphics[width=0.75\textwidth]{figures/Fig_residuals_A.png}\\
+    \end{center}
 	B\\
-	\includegraphics[width=\columnwidth]{figures/Fig_residuals_B.png}
+    \begin{center}
+        \includegraphics[width=0.75\textwidth]{figures/Fig_residuals_B.png}
+    \end{center}
 \caption{\small \textbf{Effect of read position:} \textbf{(A)} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified.
@@
 To quantify the methylation of cell 1 from (A) relative to this average over all cells, we propose to use the cell's residuals to the smoothed curve (lengths of the vertical red lines) and take their average, counting residuals of methylated CpGs positive and residuals of unmethylated CpGs negative.}
-\label{smoothres}
-\end{figure}
+\label{fig:smoothres}
+\end{figure*}
 
@@
 We first discuss the task of quantifying the level of methylation in given, fixed, genomic interval.
-In Figure \ref{smoothres}A, such an interval is covered by a single read for each of the two cells shown.
+In Figure \ref{fig:smoothres}A, such an interval is covered by a single read for each of the two cells shown.
 The read from the cell 2 shows much more methylation than the read from cell 1, and a simple analysis would therefore consider that cell as having stronger methylation in the interval.
-However, given that the two reads agree wherever they overlap, a more parsimonous interpretation would be that the cells do not show difference in methylation within the interval.
+However, given that the two reads agree wherever they overlap, a more parsimonious interpretation would be that the cells do not show difference in methylation within the interval.
 Rather, both cells, and similarly maybe most other cells, might have stronger methylation in the left third of the interval than in the middle one.
@@
 Therefore, we propose to not first obtain, for each CpG position, an average methylation across all cells and, then quantify for each individual cell its deviation from this average.
-In Figure \ref{}, the curved line depicts such an average over all cells, the red vertical lines show an individual cell's deviation from the ensemble average.
-We take the lengths of the red lines as signed values (``redidual''), positive for lines extending upwards from the curve (methylated CpG) and negative for lines extending downwards (unmethylated CpG), and take the average over the residual for all the CpGs in the interval that are covered by reads from this cell.
+In Figure \ref{fig:smoothres}B, the curved line depicts such an average over all cells, the red vertical lines show an individual cell's deviation from the ensemble average.
+We take the lengths of the red lines as signed values (``residual''), positive for lines extending upwards from the curve (methylated CpG) and negative for lines extending downwards (unmethylated CpG), and take the average over the residual for all the CpGs in the interval that are covered by reads from this cell.
 The value thus obtained is what we use to quantify this cell's (relative) methylation in the interval.
@@
 The signal to noise ratio in a matrix thus obtained will be much better than in a matrix obtained by simply averaging absolute methylation (0 or 1) over all the cell's covered CpG in a region.
-The reason for this is that we reduce the variation in situation as the one depicted in Figure \ref{smotthres}A, where absolute methylation might differ strongly even though there is no actual evidence for a difference between the two cells.
+The reason for this is that we reduce the variation in situation as the one depicted in Figure \ref{fig:smoothres}A, where absolute methylation might differ strongly even though there is no actual evidence for a difference between the two cells.
 
-How should one obtain the ensemble average (the curved line in Fig.\ \ref{smoothres}B)?
-A simple approach to get a value for a spcific CpG would be to take all cells with read coverage for the CpG and use the fraction of these that show the CpG as methylated.
+How should one obtain the ensemble average (the curved line in Fig.\ \ref{fig:smoothres}B)?
+A simple approach to get a value for a specific CpG would be to take all cells with read coverage for the CpG and use the fraction of these that show the CpG as methylated.
 However, especially when only few cells offer coverage, these averages will be very noisy.
 Therefore, we propose to smoothen using a kernel smoother, i.e., by performing a kernel-weighted average over the CpG site's neighborhood.
-The kernel bandwidth (i.e., the size of the neighborhood to average over) is a tuning parameter; we got good results with XXX bp and used this value for the examples presented here.
+The kernel bandwidth (i.e., the size of the neighborhood to average over) is a tuning parameter; we got good results with 1000~bp and used this value for the examples presented here.
 
@@
 	\includegraphics[width=\columnwidth]{figures/Fig_sliding.png}
-	\caption{\small \textbf{Finding variably methylated regions (VMR):} The chromosomes are divided up into overlapping windows (first five shown at the bottom), and for each window, the cells' methylation values are calculated as described and as depicted in Fig.\ \ref{smoothres}B.
+	\caption{\small \textbf{Finding variably methylated regions (VMR):} The chromosomes are divided up into overlapping windows (first five shown at the bottom), and for each window, the cells' methylation values are calculated as described and as depicted in Fig.\ \ref{fig:smoothres}B.
 Then, the variance of these values is calculated (each point represents one of the overlapping windows), a threshold (dashed line) is chosen such that a chosen quantile of windows have a variance exceeding the threshold.
@@
 
-So far, we have discussed dividing up (tiling) each chromosme genome into non-overlapping, equal-sized intervals, and quantitating the methylation of each such window.
+So far, we have discussed dividing up (tiling) each chromosome genome into non-overlapping, equal-sized intervals, and quantitating the methylation of each such window.
 Such rigid placing of interval boundaries is unlikely to be optimal: for example, a VMR might be much smaller than a tile and its signal will hence be drowned out by the larger part of uninformative CpGs that are equal in all cells, when averaging over all the CpG sites in the tile.
@@
 \ref{vmr}): Divide up the chromosome into many \emph{overlapping} windows, that start at regular multiples of a fixed, small, step size.
-Quantify the methylation of each cell in each window by averaging the cell's methylation residuals over all CpGs in the window, as described above and depicted in Figure\ \ref{smoothres}B.
+Quantify the methylation of each cell in each window by averaging the cell's methylation residuals over all CpGs in the window, as described above and depicted in Figure\ \ref{fig:smoothres}B.
 Then, calculate for each window the variance of these values over all cells.
@@
 Wherever thus marked windows overlap or are divided by only a small gap, merge them into one larger VMR.
-Then, calculate for each of these laregr VMRs the methylation signal, as before, by averaging for each cell over the residuals of all contained CpGs.
+Then, calculate for each of these larger VMRs the methylation signal, as before, by averaging for each cell over the residuals of all contained CpGs.
 
@@
 
-\begin{figure}
-	\includegraphics[width=\columnwidth]{figures/Fig_workflow.png}
+\begin{figure*}
+    \begin{center}
+	    \includegraphics[width=.8\textwidth]{figures/Fig_workflow.png}
+    \end{center}
 	\caption{\small \textbf{Overview of the functionalities implemented in the scbs package}
@@
 	\label{workflow}
-\end{figure}
+\end{figure*}
 
@@
 \texttt{scbs} is free and open source, and the package can be installed via the Python package manager.
-The source code and the package are available at \href{https://github.com/LKremer/scbs}, where detailed documentation on usage, as well as a tutorial using example data, is provided as well.
+The source code and the package are available at \href{https://github.com/LKremer/scbs}{https://github.com/LKremer/scbs}, where we also provide detailed documentation including a tutorial that demonstrates a complete \texttt{scbs} analysis on an example data set.
 
@@
 To demonstrate an improvement of results, we compare previously published analyses described above with our analysis method.
-First, we use a dataset by \citealp{Argelaguet_2019}.
-In their work, they generated a single cell methylome dataset of mouse embryo cells across different stages of development (E4.5 – E7.5).
+First, we use a data set by \citet{argelaguet2019gastru}.
+In their work, they generated a single cell methylome data set of mouse embryo cells across different stages of development (E4.5 – E7.5).
 They calculated the correlation between RNA expression intensity and corresponding DNA methylation at the promoter region to elucidate the implications of the epigenome on RNA expression during development.
-As promoters, they used DNA regions of $\pm$ 2kb around transcriptions start sites (TSS).
+As promoters, they used DNA regions of \textpm2~kb around transcriptions start sites (TSS).
 
@@
 
-In our analysis, we compare correlations between RNA expression intensity and DNA methylation of these promotor regions to correlations between RNA expression intensity and DNA methylation of the closest variable methylation regions that we identified with scbs.
-In figure \ref{figure:correlation}, we show that correlation of RNA expression intensity to gene methylation around the TSS $\pm$ 2kb is less strong compared to the correlation at the closest variable region.
-This effect can be observed with mehtylation fractions (figure \ref{figure:correlation}A) or shrunken residuals (figure \ref{figure:correlation}B) as data inputs.
+In our analysis, we compare correlations between RNA expression intensity and DNA methylation of these promotor regions to correlations between RNA expression intensity and DNA methylation of the closest variable methylation regions that we identified with \texttt{scbs}.
+In Figure \ref{figure:correlation}, we show that correlation of RNA expression intensity to gene methylation around the TSS \textpm2~kb is less strong compared to the correlation at the closest variable region.
+This effect can be observed with methylation fractions (Figure \ref{figure:correlation}A) or shrunken residuals (Figure \ref{figure:correlation}B) as data inputs.
 
@@
 
-	\caption{\small \textbf{Correlation of DNA methylation and RNA expression intensity}. All plots are based on the data by \citealp{Argelaguet_2019}.
+	\caption{\small \textbf{Correlation of DNA methylation and RNA expression intensity}. All plots are based on the data by \citep{luo2017single}.
 	\textbf{(A)} Correlation of each gene's RNA expression intensity to methylation fraction of TSS $\pm$ 2kb (y-axis) or to methylation fraction of closest variable gene.
@@
 As an example to illustrate this effect more clearly, we picked a gene with significantly stronger correlation at the closest variable region than at the TSS.
-This gene is marked on figure \ref{figure:correlation}A and B with a yellow dot.
+This gene is marked on Figure \ref{figure:correlation}A and B with a yellow dot.
 We divided all cells into four groups of cells with no, low, medium and high expression of this exemplary gene Ube216 and plotted a smoothed methylation level per group.
-As shown in figure \ref{figure:correlation}C, the promoter region is sparsely methylated in all four RNA expression groups and is therefore a poor region to make conclusions about the relation of DNA methylation and RNA expression intensity.
+As shown in Figure \ref{figure:correlation}C, the promoter region is sparsely methylated in all four RNA expression groups and is therefore a poor region to make conclusions about the relation of DNA methylation and RNA expression intensity.
 However, a region close to the TSS, identified as the closest highly variable methylation region with scbs, shows a much clearer negative correlation between RNA expression intensity and DNA methylation.
 
-To further demonstrate the benefit of detecting variable methylation regions, we used a second dataset, published by \citealp{Luo_2017}.
-In their work, \citealp{Luo_2017} generated methylomes from single cell neuronal nuclei in mice to identify genomic regulatory elements across neuron types.
+To further demonstrate the benefit of detecting variable methylation regions, we used a second dataset, published by \citet{luo2017single}.
+In their work, \citet{luo2017single} generated methylomes from single cell neuronal nuclei in mice to identify genomic regulatory elements across neuron types.
 To compare how well cell types are separated by different analysis techniques, we processed the same dataset into 2-dimensional UMAPs several times.
@@
 	A\\
-	\hspace{.3cm}\includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/complete_135k_12cm_log.pdf} \\
-	B\\
-	\hspace{.3cm}\includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/cell500_20k_12cm_log.pdf} \\
+	\hspace{.3cm}\includegraphics[width=.45\columnwidth]{part_leonie_git/leonie_plots/complete_135k_12cm_log.pdf} 
+	B
+	\hspace{.3cm}\includegraphics[width=.45\columnwidth]{part_leonie_git/leonie_plots/cell500_20k_12cm_log.pdf} \\
 	C\\
@@
 	\caption{\small \textbf{Analysis comparison}.
-All plots are based on the data by \citealp{Luo_2017}.
+All plots are based on the data by \citep{luo2017single}.
 \textbf{(A)} Nearest neighbor count score for complete dataset using four different analysis techniques and a variable number of regions as input.

### 2022-06-10 paper.tex
fixed two citations

--- 2022-06-09 paper.tex
@@
 To demonstrate an improvement of results, we compare previously published analyses described above with our analysis method.
-First, we use a dataset by \citealp{Argelaguet_2019}.
+First, we use a dataset by \citet{Argelaguet_2019}.
 In their work, they generated a single cell methylome dataset of mouse embryo cells across different stages of development (E4.5 – E7.5).
@@
 
-To further demonstrate the benefit of detecting variable methylation regions, we used a second dataset, published by \citealp{Luo_2017}.
-In their work, \citealp{Luo_2017} generated methylomes from single cell neuronal nuclei in mice to identify genomic regulatory elements across neuron types.
+To further demonstrate the benefit of detecting variable methylation regions, we used a second dataset, published by \citealp{luo2017single}.
+In their work, \citealp{luo2017single} generated methylomes from single cell neuronal nuclei in mice to identify genomic regulatory elements across neuron types.
 To compare how well cell types are separated by different analysis techniques, we processed the same dataset into 2-dimensional UMAPs several times.
@@
 	\caption{\small \textbf{Analysis comparison}.
-All plots are based on the data by \citealp{Luo_2017}.
+All plots are based on the data by \citealp{luo2017single}.
 \textbf{(A)} Nearest neighbor count score for complete dataset using four different analysis techniques and a variable number of regions as input.

### 2022-06-10 paper.tex
a bit of rewording in the intro

@@
 Sequencing-based assays with single-cell resolution have offered new means to understand the differences between the cells making up a sample.
-Single-cell RNA sequencing (scRNA-seq) techniques have matured at great pace in recent years, and methods to study chromatin are rapidly catching up.
+Single-cell RNA sequencing (scRNA-seq) techniques have matured at great pace in recent years.
+Now, methods to study chromatin are rapidly catching up.
 Of special interest here is DNA methylation, i.e., the methylation of bases in the DNA, because DNA methylation patterns are passed on to daughter cells in cell division.
@@
 CpG dinocleotides are underrepresented in eukaryotic genomes (due to their vulnerability to mutation by conversion to TpG due to accidental deanimation), and the majority of existing CpG sites are methylated.
-Presence or absence of the methyl group at CpG sites is believed to have effect on gene regulation, with methylation inhibiting transcription.
-Methylating DNA during differentiation is therefore conjectured to be an important mechanism to silence genome regions that are not needed in the target lineage.
+Presence or absence of the methyl group at CpG sites is believed to affect gene regulation, with methylation inhibiting transcription.
+In differentiation, methylation of DNA is conjectured to be an important mechanism to silence genome regions that are not needed in the target lineage.
 One important application of single-cell bisulfite sequencing (scBS) therefore lies in the study of differentiation processes.
@@
 
-We start by briefly reviewing the standard approach of analysing single-cell \emph{RNA}-Seq data, and how this approach is typically adapted to the case of single-cell DNA methylation data.
+The standard approach to analyse scBS data is based on methodology developed for the analysis of single-cell RNA-Seq data.
+Therefore, we will start by briefly reviewing how single-cell \emph{RNA}-Seq data is commonly analysed, before we discuss scBS data analysis.
 
-The starting point in scRNA-Seq is a matrix of UMI counts (i.e.
+The starting point in most scRNA-Seq analyses is a matrix of UMI counts (i.e.
 of counts of distinct RNA molecules), one row for each cell and one column for each gene.
-As a first step towards assigning cell types or states to cells, one needs to establish which cells are similar to each other, and to this end, find a way to quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile.
+A first goal is then to assigning cell types or states to cells, and to this end, one needs to establish which cells are similar to each other, i.e., find a way to quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile.
 A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm.
@@
 Therefore, one performs a principal component analysis (PCA), keeping only the top few (typically, 20 to 50) components.
-As Poisson noise is uncorrelated between genes, it will average out in the top principal components.
-Therefore, Euclidean distances between these ``PCA space'' vectors provide a robust dissimilarity score, which can subsequently be used as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data.
-The PCA space representation is furthermore used for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).
+As Poisson noise is uncorrelated between genes, it will average out in the top principal components, as these are all linear combinations with weight on very many genes.
+Therefore, Euclidean distances between these ``PCA space'' vectors provide a robust dissimilarity score.
+Hence, the PCA space representation y are suitable as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data, or to methods for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).
 

### 2022-06-10 paper.tex

--- 2022-06-10 paper.tex
@@
 CpG dinocleotides are underrepresented in eukaryotic genomes (due to their vulnerability to mutation by conversion to TpG due to accidental deamination), and the majority of existing CpG sites are methylated.
-Presence or absence of the methyl group at CpG sites is believed to have effect on gene regulation, with methylation inhibiting transcription.
-Methylating DNA during differentiation is therefore conjectured to be an important mechanism to silence genome regions that are not needed in the target lineage.
+Presence or absence of the methyl group at CpG sites is believed to affect gene regulation, with methylation inhibiting transcription.
+In differentiation, methylation of DNA is conjectured to be an important mechanism to silence genome regions that are not needed in the target lineage.
 One important application of single-cell bisulfite sequencing (scBS) therefore lies in the study of differentiation processes.
@@
 
-We start by briefly reviewing the standard approach of analyzing single-cell \emph{RNA}-Seq data, and how this approach is typically adapted to the case of single-cell DNA methylation data.
+The standard approach to analyse scBS data is based on methodology developed for the analysis of single-cell RNA-Seq data.
+Therefore, we will start by briefly reviewing how single-cell \emph{RNA}-Seq data is commonly analyzed, before we discuss scBS data analysis.
 
-The starting point in scRNA-Seq is a matrix of UMI counts (i.e.
+The starting point in most scRNA-Seq analyses is a matrix of UMI counts (i.e.
 of counts of distinct RNA molecules), one row for each cell and one column for each gene.
-As a first step towards assigning cell types or states to cells, one needs to establish which cells are similar to each other, and to this end, find a way to quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile.
+A first goal is then to assigning cell types or states to cells, and to this end, one needs to establish which cells are similar to each other, i.e., find a way to quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile.
 A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm.

### 2022-06-10 paper.tex
minor change

@@
 Sequencing-based assays with single-cell resolution have offered new means to understand the differences between the cells making up a sample.
-Single-cell RNA sequencing (scRNA-seq) techniques have matured at great pace in recent years.
-Now, methods to study chromatin are rapidly catching up.
-Of special interest here is DNA methylation, i.e., the methylation of bases in the DNA, because DNA methylation patterns are passed on to daughter cells in cell division.
-Especially in mammals, such methylation chiefly occurs at cytosine bases that are immediately after followed by guanine bases, so-called CpG sites.
+Single-cell RNA sequencing (scRNA-seq) techniques have matured at great pace in recent years, and methods to study epigenetics at single-cell resolution are rapidly catching up.
+Of special interest here is DNA methylation, i.e., the methylation of bases in the DNA, which is passed on to daughter cells in cell division.
+Especially in mammals, such methylation chiefly occurs at cytosine bases that are immediately after followed by guanine bases, so-called CpG \mbox{(5'--C--phosphate--G--3')} sites.
 
@@
 CpG dinocleotides are underrepresented in eukaryotic genomes (due to their vulnerability to mutation by conversion to TpG due to accidental deamination), and the majority of existing CpG sites are methylated.
-Presence or absence of the methyl group at CpG sites is believed to affect gene regulation, with methylation inhibiting transcription.
-In differentiation, methylation of DNA is conjectured to be an important mechanism to silence genome regions that are not needed in the target lineage.
+Presence or absence of the methyl group at CpG sites is believed to have effect on gene regulation, with methylation inhibiting transcription.
+Methylating DNA during differentiation is therefore conjectured to be an important mechanism to silence genome regions that are not needed in the target lineage.
 One important application of single-cell bisulfite sequencing (scBS) therefore lies in the study of differentiation processes.
@@
 
-The standard approach to analyse scBS data is based on methodology developed for the analysis of single-cell RNA-Seq data.
-Therefore, we will start by briefly reviewing how single-cell \emph{RNA}-Seq data is commonly analyzed, before we discuss scBS data analysis.
+We start by briefly reviewing the standard approach of analyzing single-cell \emph{RNA}-Seq data, and how this approach is typically adapted to the case of single-cell DNA methylation data.
 
-The starting point in most scRNA-Seq analyses is a matrix of UMI counts (i.e.
+The starting point in scRNA-Seq is a matrix of UMI counts (i.e.
 of counts of distinct RNA molecules), one row for each cell and one column for each gene.
-A first goal is then to assigning cell types or states to cells, and to this end, one needs to establish which cells are similar to each other, i.e., find a way to quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile.
+As a first step towards assigning cell types or states to cells, one needs to establish which cells are similar to each other, and to this end, find a way to quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile.
 A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm.
@@
 Therefore, one performs a principal component analysis (PCA), keeping only the top few (typically, 20 to 50) components.
-As Poisson noise is uncorrelated between genes, it will average out in the top principal components, as these are all linear combinations with weight on very many genes.
-Therefore, Euclidean distances between these ``PCA space'' vectors provide a robust dissimilarity score.
-Hence, the PCA space representation y are suitable as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data, or to methods for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).
+As Poisson noise is uncorrelated between genes, it will average out in the top principal components.
+Therefore, Euclidean distances between these ``PCA space'' vectors provide a robust dissimilarity score, which can subsequently be used as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data.
+The PCA space representation is furthermore used for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).
 

### 2022-06-10 paper.tex

--- 2022-06-10 paper.tex
@@
 Single-cell RNA sequencing (scRNA-seq) techniques have matured at great pace in recent years.
-Now, methods to study chromatin are rapidly catching up.
-Of special interest here is DNA methylation, i.e., the methylation of bases in the DNA, because DNA methylation patterns are passed on to daughter cells in cell division.
-Especially in mammals, such methylation chiefly occurs at cytosine bases that are immediately after followed by guanine bases, so-called CpG sites.
+Now, methods to study epigenetics at single-cell resolution are rapidly catching up.
+Of special interest here is DNA methylation, i.e., the methylation of bases in the DNA, which is passed on to daughter cells in cell division.
+Especially in mammals, such methylation chiefly occurs at cytosine bases that are immediately after followed by guanine bases, so-called CpG \mbox{(5'--C--phosphate--G--3')} sites.
 
@@
 The standard approach to analyse scBS data is based on methodology developed for the analysis of single-cell RNA-Seq data.
-Therefore, we will start by briefly reviewing how single-cell \emph{RNA}-Seq data is commonly analyzed, before we discuss scBS data analysis.
+Therefore, we will start by briefly reviewing how single-cell \emph{RNA}-Seq data is commonly analysed, before we discuss scBS data analysis.
 

### 2022-06-10 paper.tex
minor intro adjustments

@@
 
-To detect CpG methylation by sequencing, bisulfite conversion is commonly employed: treatment of DNA with bisulfite causes unmethylated cytosine bases to be deaminated to uracil, which is paired with thymine in subsequent PCR.
-Therefore, unmethylated CpGs are read as TG in sequencing of bisulfite-converted DNA while only methylated CpG is correctly read as CG \citep{Frommer_1992}.
-Such so-called bisulfite sequencing has been used for bulk samples since long, and methods for bisulfite sequencing with single-cell resolution (scBS) were recently developed \citep{Smallwood_2014} and are increasingly used.
+To detect CpG methylation by sequencing, bisulfite conversion is commonly employed: treatment of DNA with bisulfite causes unmethylated cytosine bases to be deaminated to uracil, which is replaced by thymine in subsequent PCR.
+Therefore, unmethylated CpGs are read as TG in sequencing of bisulfite-converted DNA while only methylated CpG sites are correctly read as CG \citep{Frommer_1992}.
+Such so-called bisulfite sequencing has been used for bulk samples since long, and methods for bisulfite sequencing with single-cell resolution were recently developed \citep{Smallwood_2014} and are increasingly used.
 
-CpG dinocleotides are underrepresented in eukaryotic genomes (due to their vulnerability to mutation by conversion to TpG due to accidental deamination), and the majority of existing CpG sites are methylated.
+CpG dinocleotides are underrepresented in eukaryotic genomes (due to their vulnerability to mutation by conversion to TpG due to spontaneous deamination), and the majority of existing CpG sites are methylated.
 Presence or absence of the methyl group at CpG sites is believed to have effect on gene regulation, with methylation inhibiting transcription.
-Methylating DNA during differentiation is therefore conjectured to be an important mechanism to silence genome regions that are not needed in the target lineage.
-One important application of single-cell bisulfite sequencing (scBS) therefore lies in the study of differentiation processes.
+The removal of DNA methylation during cell differentiation is therefore conjectured to be an important mechanism to activate genome regions that are required in the target lineage and define cell identity \citet{cedar2022role}.
+One important application of single-cell bisulfite sequencing (scBS) therefore lies in the study of differentiation processes, for instance in embryonic development \citet{argelaguet2019gastru}.
 
@@
 
-This procedure is commonly adapted when working with single-cell DNA-methylation data, so that all the tools used in scRNA-Seq downstream of the PCA step can be used.
+This procedure is commonly adapted when working with single-cell DNA-methylation data, so that established methods used in scRNA-Seq can be used.
 The question is therefore how to construct a suitable matrix as input for the PCA.
-A simple and popular approach is to tile the genome into windows of, say, 100 kB size, and calculate for each cell the average methylation of the DNA within each window.
+A simple and common approach, used for instance by \citet{luo2017single}, is to tile the genome into windows of, say, 100~kb size, and calculate for each cell the average methylation of the DNA within each window.
 To this end, we identify all CpG sites in the window for which we have coverage with at least one read and can hence call the CpG to be either unmethylated or methylated in the cell.
-Then we denote as average DNA methylation of this window in a given cell the proportion of observed CpG sites in the window have been found to be methylated.
+Then we denote as average DNA methylation of this window in a given cell the proportion of observed CpG sites in the window that were found to be methylated (Figure \ref{fig:smoothres}A).
 This yields a matrix, with one row for each cell and one column for each window from the genome tiling, comprising numbers (methylation fractions) between 0 and 1.
@@
 
-As an example, we will us in this paper data from XX et al., who have performed scBS on [...] and analysed their data in the manner just described.
-Figure \ref{umaps}A shows a UMAP plot, derived from our re-analysis of the data, using a tiling of the genome into windows of size 100 kb.
+As an example, we will use in this paper data from \citet{luo2017single}, who performed scBS on 16 neuronal subtypes of the mouse frontal cortex, and analysed their data in the manner just described.
+Figure \ref{umaps}A shows a UMAP plot, derived from our re-analysis of the data, using a tiling of the genome into windows of size 100~kb.
 As one can see, the approach is suitable to see clear structure in the data, observe ...

### 2022-06-10 paper.tex
adjust main section

@@
 
-We first discuss the task of quantifying the level of methylation in given, fixed, genomic interval.
+We first discuss the task of quantifying the level of methylation in a given, fixed, genomic interval.
 In Figure \ref{fig:smoothres}A, such an interval is covered by a single read for each of the two cells shown.
-The read from the cell 2 shows much more methylation than the read from cell 1, and a simple analysis would therefore consider that cell as having stronger methylation in the interval.
+The read from cell 2 shows much more methylation than the read from cell 1, and a simple analysis would therefore consider that cell as having stronger methylation in the interval.
 However, given that the two reads agree wherever they overlap, a more parsimonious interpretation would be that the cells do not show difference in methylation within the interval.
@@
 
-Therefore, we propose to not first obtain, for each CpG position, an average methylation across all cells and, then quantify for each individual cell its deviation from this average.
+Therefore, we propose to first obtain, for each CpG position, an average methylation across all cells and, then quantify for each individual cell its deviation from this average.
 In Figure \ref{fig:smoothres}B, the curved line depicts such an average over all cells, the red vertical lines show an individual cell's deviation from the ensemble average.
@@
 The value thus obtained is what we use to quantify this cell's (relative) methylation in the interval.
-For a genome tiled into such intervals, we thus obtain a matrix, one row per interval, one column per cell, that can be used for downstream analysis, e.g., as input for a PCA.
+For a genome tiled into such intervals, we thus obtain a matrix, one row per cell, one column per interval, that can be used for downstream analysis, e.g., as input for a PCA.
 
-The signal to noise ratio in a matrix thus obtained will be much better than in a matrix obtained by simply averaging absolute methylation (0 or 1) over all the cell's covered CpG in a region.
+The signal to noise ratio in a matrix thus obtained will be much better than in a matrix obtained by simply averaging absolute methylation (0 or 1) over all the cell's covered CpG sites in a region.
 The reason for this is that we reduce the variation in situation as the one depicted in Figure \ref{fig:smoothres}A, where absolute methylation might differ strongly even though there is no actual evidence for a difference between the two cells.
@@
 To counter this, we propose a simple and straightforward way to deal with missing data in the input matrix to a PCA: In a first iteration, we replace each missing value in the centered input matrix with zero, then run the PCA.
-Then, these zeroes are replaced by the value predicted by the PCA and the PCA is rerun.
-Convergence is typically achieved after two to four iterations.
+Then, only these zeroes are replaced by the value predicted by the PCA and the PCA is rerun.
+Convergence is typically achieved after two to five iterations.
 For details, see Methods.
 
-The possibility to perform PCA with missing data allows us to use a shorter and thus more intervals and hence provide the PCA with richer input data.
+The possibility to perform PCA with missing data allows us to use shorter, and thus more intervals and hence provide the PCA with richer input data.
 
@@
 
-Typically, some regions in a chromosome will have very similar methylation status in all cells (most commonly, fully methylated in nearly all cells) while other regions show variability in methylation across cells.
-Only the latter regions are of value for our goal of quantitating dissimilarity between cells.
-We call these the variably methylated regions (VMRs).
+Typically, some regions in a chromosome will have very similar methylation status in all cells, while other regions show variability in methylation across cells.
+For instance, it is long known that CpG-rich promoters of housekeeping genes are unmethylated, and that a large proportion of the remaining genome is highly methylated regardless of cell type \citet{bird1986cpg}.
+In contrast, DNA methylation at certain genomic features such as enhancers is more dynamic \citep{argelaguet2019gastru}, and thus more variable across cells.
+Only the latter regions are of value for our goal of quantitating dissimilarity between cells. We call these the variably methylated regions (VMRs).
 
-So far, we have discussed dividing up (tiling) each chromosome genome into non-overlapping, equal-sized intervals, and quantitating the methylation of each such window.
+So far, we have discussed dividing up (tiling) each chromosome into non-overlapping, equal-sized intervals, and quantitating the methylation of each such window.
 Such rigid placing of interval boundaries is unlikely to be optimal: for example, a VMR might be much smaller than a tile and its signal will hence be drowned out by the larger part of uninformative CpGs that are equal in all cells, when averaging over all the CpG sites in the tile.
@@
 Then, calculate for each window the variance of these values over all cells.
-Select, say, the top XX\% windows with the highest variances and mark them as VMRs.
+Select, say, the top 2\% windows with the highest variances and mark them as VMRs.
 Wherever thus marked windows overlap or are divided by only a small gap, merge them into one larger VMR.
-Then, calculate for each of these larger VMRs the methylation signal, as before, by averaging for each cell over the residuals of all contained CpGs.
+Then, calculate for each of these larger VMRs the methylation signal, as before, by averaging for each cell over the residuals of all contained CpG sites.
 
-In this manner, we obtain a methylation matrix, with one column per cell and one row per VMR, that is, in a sense, richer in information and has better signal-to-noise ratio than the matrix obtained by the simple analysis sketched at the very beginning.
+In this manner, we obtain a methylation matrix, with one row per cell and one column per VMR, that is, in a sense, richer in information and has better signal-to-noise ratio than the matrix obtained by the simple analysis sketched at the very beginning.
 As we demonstrate below, a PCA performed on such a matrix provided a distance metric for the cells that contains more information on biological detail than one from a simpler analysis.
 
-Besides providing better input for the distance calculations, the identified VMRs can also be compared with genomic annotation, thus providing a useful starting point for examining the data with methods from functional genomics.
+Besides providing better input for the distance calculations, the small size of identified VMRs also means that they can be readily compared with genomic annotation, which aids in their functional interpretation.
+While a large 100~kb bin will often contain multiple genes with distinct expression patterns, small VMRs located in known regulatory elements or gene bodies are more easily associated with individual genes and their associated function.
+Thus, VMRs provide a useful starting point for examining the data with methods from functional genomics.
 
-\section{The scbs Python package}
+\section{The \texttt{scbs} Python package}
 

### 2022-06-10 paper.tex
going through Leonies part

@@
 Typically, some regions in a chromosome will have very similar methylation status in all cells, while other regions show variability in methylation across cells.
-For instance, it is long known that CpG-rich promoters of housekeeping genes are unmethylated, and that a large proportion of the remaining genome is highly methylated regardless of cell type \citet{bird1986cpg}.
+For instance, it is long known that CpG-rich promoters of housekeeping genes are unmethylated, and that a large proportion of the remaining genome is highly methylated regardless of cell type \citep{bird1986cpg}.
 In contrast, DNA methylation at certain genomic features such as enhancers is more dynamic \citep{argelaguet2019gastru}, and thus more variable across cells.
@@
 
-We have implemented the approach just described in a Python package with a command line interface, called \texttt{scbs}, which also offers a number of other functionalities for the analysis of scBS data Figure \ref{workflow}.
+We have implemented the approach just described in a Python package with a command line interface, called \texttt{scbs}, which also offers a number of other functionalities for the analysis of scBS data (Figure \ref{workflow}).
 The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark} or methylpy \citep{methylpy}.
@@
 
+In brief, we store methylation of each chromosome in a binary matrix where each column represents a cell and each row represents a base pair.
+Since the vast majority of values in this matrix are missing due to the sparsity of scBS data, we encode missing values as zero and then store the data in Compressed Sparse Row (CSR) format.
+This format does not explicitly store zeroes (here: missing values) and is optimized for row-wise access, which results in significant compression and allows fast access to the methylation status of genomic intervals.
+
 \texttt{scbs prepare} also computes a number of summary statistics for each cell, including the mean genome-wide methylation level and the number of observed CpG sites, i.~e.
@@
 These summary statistics can be used to detect cells with poor quality.
-The quality of each single cell methylome can furthermore be inspected with \texttt{scbs profile}, which computes the average methylation profile of a set of user-defined genomic regions such as transcription start sites (TSS).
+The quality of each single cell methylome can furthermore be inspected with \texttt{scbs profile}, which computes the average methylation profile of a set of user-defined genomic regions such as transcription start sites (TSS) at single-cell resolution.
 The TSS profile is a useful quality control plot since methylation shows a characteristic dip roughly \textpm1~kb around the transcription start site in mammalian genomes.
-Cells that do not show this pattern, or cells with few observed CpG sites, may then be filtered from the data set with \texttt{scbs filter}.
+Cells that do not show this pattern, or cells with few observed CpG sites, may then be discarded with \texttt{scbs filter}.
 
@@
 This produces a BED-file that lists the genome coordinates of VMRs.
-To finally obtain a methylation matrix analogous to an scRNA-seq count matrix, this BED-file can be used as input for \texttt{scbs matrix}, which quantifies methylation at genomic intervals in all cells.
+To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED-file can be used as input for \texttt{scbs matrix}, which quantifies methylation at genomic intervals in all cells.
 The command produces both the percentage of methylated CpG sites, as well as our proposed methylation measure (shrunken mean of the residuals) that is more robust to variation in sequencing coverage or stochastic differences in read position between cells.
-We note that both \texttt{matrix} and \texttt{profile} accept any valid BED-file as input, which means that the user can quantify and visualize methylation at any genomic feature of interest, such as promoters, enhancers or transcription factor binding sites.
+We note that both \texttt{matrix} and \texttt{profile} accept any valid BED-file as input, which means that the user can quantify and visualize methylation at any genomic feature of interest, such as promoters, enhancers or transcription factor binding sites, in single cells.
 The obtained methylation matrix can then serve as input for common single-cell analysis methods such as dimensionality reduction and cell clustering.
@@
 \section{Comparison}
+To assess performance of our newly devised methods, we obtained two previously published scBS data sets and re-analysed them with \texttt{scbs}.
 
-To demonstrate an improvement of results, we compare previously published analyses described above with our analysis method.
-First, we use a data set by \citet{argelaguet2019gastru}.
-In their work, they generated a single cell methylome data set of mouse embryo cells across different stages of development (E4.5 – E7.5).
-They calculated the correlation between RNA expression intensity and corresponding DNA methylation at the promoter region to elucidate the implications of the epigenome on RNA expression during development.
-As promoters, they used DNA regions of \textpm2~kb around transcriptions start sites (TSS).
-
---The following paragraph potentially needs to be changed when new plots are made--
+The first data set is a single-cell multi-omic data set generated by \citet{argelaguet2019gastru}.
+In their work, the authors obtained single-cell methylomes of mouse embryo cells across different stages of development (E4.5 – E7.5), as well as single-cell transcriptomes from the same cells.
+They calculated the correlation between RNA expression and DNA methylation at the corresponding promoter region to elucidate the implications of the epigenome on gene expression in development.
+\citet{argelaguet2019gastru} defined promoters as genomic intervals \textpm2~kb around the TSS of each gene.
 

### 2022-06-11 paper.tex
incorporate all of Leonies benchmarking stuff

@@
 \setlength{\columnsep}{0.7cm}
+\usepackage{microtype} % nice typesetting
 \usepackage{amsmath}
 \usepackage{graphicx}
+
 \usepackage[utf8]{inputenc}
@@
 \urlstyle{same} % no tt font for URLs
-
 \usepackage{natbib}
@@
 %\setcitestyle{aysep={}} 
-
 \usepackage[dvipsnames]{xcolor}
@@
 \centering
-\textbf{\Large Analysing single-cell bisulfite sequencing data with scbs}
+\textbf{\Large Analysing single-cell bisulfite sequencing data with \texttt{scbs}}
 
@@
 
-\begin{figure}
-    \framebox[\columnwidth]{
-        \begin{minipage}[c][3.5cm]{\columnwidth}
-            \centering Figure
-        \end{minipage}
-    }
-    \caption{\small UMAP representations of the single-cell bisulfite data of XXX et al., calculated \textbf{(A)} using conventional preprocessing and \textbf{(B)} using the approach described here.}
-    \label{umaps}
-\end{figure}
+%\begin{figure}
+%   \framebox[\columnwidth]{
+%       \begin{minipage}[c][3.5cm]{\columnwidth}
+%           \centering Figure
+%       \end{minipage}
+%   }
+%   \caption{\small UMAP representations of the single-cell bisulfite data of XXX et al., calculated \textbf{(A)} using conventional preprocessing and \textbf{(B)} using the approach described here.}
+%   \label{umaps}
+%\end{figure}
 
@@
 
-As an example, we will use in this paper data from \citet{luo2017single}, who performed scBS on 16 neuronal subtypes of the mouse frontal cortex, and analysed their data in the manner just described.
-Figure \ref{umaps}A shows a UMAP plot, derived from our re-analysis of the data, using a tiling of the genome into windows of size 100~kb.
+%As an example, we will use in this paper data from \citet{luo2017single}, who performed scBS on 16 neuronal subtypes of the mouse frontal cortex, and analysed their data in the manner just described.
+%Figure \ref{umaps}A shows a UMAP plot, derived from our re-analysis of the data, using a tiling of the genome into windows of size 100~kb.
 As one can see, the approach is suitable to see clear structure in the data, observe ...
 
-Nevertheless, this analysis strategy has substantial room for improvement, as we will discuss next.
+We think that this analysis strategy has room for improvement, as we will discuss next.
 Here, we discuss the proposed improvements only qualitatively.
@@
     \begin{center}
-        \includegraphics[width=0.75\textwidth]{figures/Fig_residuals_A.png}\\
+        \includegraphics[width=0.7\textwidth]{figures/Fig_residuals_A.png}\\
     \end{center}
@@
     \begin{center}
-        \includegraphics[width=0.75\textwidth]{figures/Fig_residuals_B.png}
+        \includegraphics[width=0.7\textwidth]{figures/Fig_residuals_B.png}
     \end{center}
@@
     \caption{\small \textbf{Overview of the functionalities implemented in the scbs package}
-    A 
     }
@@
 
-\section{Comparison}
+\section{Benchmarks}
 To assess performance of our newly devised methods, we obtained two previously published scBS data sets and re-analysed them with \texttt{scbs}.
@@
 The first data set is a single-cell multi-omic data set generated by \citet{argelaguet2019gastru}.
-In their work, the authors obtained single-cell methylomes of mouse embryo cells across different stages of development (E4.5 – E7.5), as well as single-cell transcriptomes from the same cells.
-They calculated the correlation between RNA expression and DNA methylation at the corresponding promoter region to elucidate the implications of the epigenome on gene expression in development.
-\citet{argelaguet2019gastru} defined promoters as genomic intervals \textpm2~kb around the TSS of each gene.
+In their work, the authors obtained single-cell methylomes of mouse embryo cells across different stages of development (E4.5~–~E7.5), as well as single-cell transcriptomes from the same cells.
+They calculated the correlation between gene expression and promoter methylation to elucidate the implications of the epigenome on gene expression in development.
+The authors identified several genes where promoter methylation predicts gene expression, but obtained low correlations for the vast majority of genes.
 
-In our analysis, we compare correlations between RNA expression intensity and DNA methylation of these promotor regions to correlations between RNA expression intensity and DNA methylation of the closest variable methylation regions that we identified with \texttt{scbs}.
-In Figure \ref{figure:correlation}, we show that correlation of RNA expression intensity to gene methylation around the TSS \textpm2~kb is less strong compared to the correlation at the closest variable region.
-This effect can be observed with methylation fractions (Figure \ref{figure:correlation}A) or shrunken residuals (Figure \ref{figure:correlation}B) as data inputs.
+We reasoned that VMRs obtained with the previously described sliding window approach might be a better predictor of gene expression. 
+In Figure \ref{figure:correlation}A, we compare correlations of gene expression and promoter methylation to correlations of gene expression and methylation of the closest VMR identified with \texttt{scbs scan}.
+This analysis revealed a number of genes whose expression does not correlate with promoter methylation, but correlates with VMR methylation.
+This effect is also observed when using our proposed measure of methylation, the shrunken mean of residuals (Figure \ref{figure:correlation}B).
 
-\begin{figure}[!htp]
+\begin{figure}
     A\\
-    \hspace{.2cm}\includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/corr_methylation_fraction_4kbwindow.pdf} \\
+    \includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/corr_methylation_fraction_4kbwindow.pdf} \\
     B\\
-    \hspace{.2cm}\includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/corr_shrunken_residual_4kbwindow.pdf} \\
+    \includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/corr_shrunken_residual_4kbwindow.pdf} \\
     C\\
-    \hspace{.39cm}\includegraphics[width=\columnwidth]{part_leonie_git/leonie_plots/comparison_meth_RNA.pdf}
-
-    \caption{\small \textbf{Correlation of DNA methylation and RNA expression intensity}. All plots are based on the data by \citep{luo2017single}.
-    \textbf{(A)} Correlation of each gene's RNA expression intensity to methylation fraction of TSS $\pm$ 2kb (y-axis) or to methylation fraction of closest variable gene.
-    \textbf{(B)} Correlation of each gene's RNA expression intensity to shrunken residual of methylation of TSS $\pm$ 2kb (y-axis) or to shrunken residual of closest variable gene.
-    \textbf{(C)} Mean gene methylation of Ube2l6 gene (ENSMUSG00000027078, marked in yellow in A \& B).
-Cells were into 4 groups based on their RNA expression (group 0: cells with no expression of gene, group 1-3: cells that express gene divided into three equally large groups with group 3 having the highest expression).
-Mean methylation level of each group was smoothed with a tricube kernel of bandwidth = 2000.}
+    \includegraphics[width=.95\columnwidth]{part_leonie_git/leonie_plots/comparison_meth_RNA.pdf}
+    \caption{\small \textbf{Correlation of DNA methylation and gene expression}.
+    \textbf{(A)} Correlation of each gene's RNA expression values to the the methylation fractions that were measured at promoters (y-axis), or to methylation fractions of the nearest VMR (x-axis).
+Promoters are defined as intervals \textpm2~kb around the TSS.
+    \textbf{(B)} Same as in (A), but using mean shrunken residuals as a measure of methylation.
+    \textbf{(C)} Mean smoothed methylation near the gene Ube2l6 (ENSMUSG00000027078, marked in yellow in A \& B).
+Cells are assigned to 4 groups based on Ube2l6 expression (group 0: cells with no Ube2l6 expression, groups 1-3: cells that express Ube2l6, divided into three equally large groups with group 3 having the highest expression).
+The mean methylation level of each group was smoothed with a tricube kernel of bandwidth 2000.
+    Data from \citet{argelaguet2019gastru}.}
     \label{figure:correlation}
@@
 
-As an example to illustrate this effect more clearly, we picked a gene with significantly stronger correlation at the closest variable region than at the TSS.
-This gene is marked on Figure \ref{figure:correlation}A and B with a yellow dot.
-We divided all cells into four groups of cells with no, low, medium and high expression of this exemplary gene Ube216 and plotted a smoothed methylation level per group.
-As shown in Figure \ref{figure:correlation}C, the promoter region is sparsely methylated in all four RNA expression groups and is therefore a poor region to make conclusions about the relation of DNA methylation and RNA expression intensity.
-However, a region close to the TSS, identified as the closest highly variable methylation region with scbs, shows a much clearer negative correlation between RNA expression intensity and DNA methylation.
+To illustrate this effect with an example, we picked a gene where VMR methylation is a much better predictor of gene expression than promoter methylation (Ube2l6, highlighted in orange in Figure \ref{figure:correlation}).
+Grouping all cells into four groups according to their level of Ube2l6 expression, it is clearly visible that the promoter of this gene is lowly methylated regardless of gene expression, while methylation at the nearest VMR is greatly decreased in cells with high Ube2l6 expression.
 
-To further demonstrate the benefit of detecting variable methylation regions, we used a second dataset, published by \citet{luo2017single}.
-In their work, \citet{luo2017single} generated methylomes from single cell neuronal nuclei in mice to identify genomic regulatory elements across neuron types.
-To compare how well cell types are separated by different analysis techniques, we processed the same dataset into 2-dimensional UMAPs several times.
-Subsequently, we compared the results by assigning a score to each output based on cell type separation in the final 15 principal components.
-We see an improvement of cell separation when comparing the analysis of variable methylation regions to 100 kb regions (\ref{figure:score}A and C).
-The difference of using shrunken residuals instead of methylation fractions to assess the methylation per region is not as significant.
-However, when emulating a sparse dataset with less cells by randomly reducing the dataset to 500 cells, it becomes clear that using shrunken residuals over methylation fractions also improves the outcome of our analysis (\ref{figure:score}B and D).
+To assess the impact of our methods on dimensionality reduction and the ability to distinguish different (sub)types of cells, we obtained a second data set from \citet{luo2017single}.
+This data set consists of over three thousand single-cell CpG methylomes from murine frontal cortex, comprising 16 neuronal subtypes.
+We subjected this data set to the standard scBS workflow: defining a set of genomic intervals, quantifying methylation in these intervals, and subjecting the resulting matrix to PCA.
+We repeated this analysis several times, either using 100~kb tiles or VMRs, and either using methylation fractions or mean shrunken residuals as a measure of methylation.
 
-\begin{figure}[!htp]
-    A\\
-    \hspace{.3cm}\includegraphics[width=.45\columnwidth]{part_leonie_git/leonie_plots/complete_135k_12cm_log.pdf} 
-    B
-    \hspace{.3cm}\includegraphics[width=.45\columnwidth]{part_leonie_git/leonie_plots/cell500_20k_12cm_log.pdf} \\
+To assess performance of each combination of methods, we used a nearest neighbor count score that quantifies how well the different neuronal subtypes are separated in 15 dimensional PC space (see Methods).
+The results demonstrate that the 16 neuronal subtypes are more cleanly separated when using our proposed methods (Figure \ref{figure:score}A).
+Most of this gain in performance can be attributed to the use of VMRs over genomic tiles. The use of mean shrunken residuals over methylation fractions provides an additional small improvement.
+Of note, our results show that even few VMRs, e.g. the 1000 VMRs that have read coverage in the highest number of cells, already contain sufficient information to distinguish neuronal subtypes.
+In contrast, the neighbor count score for the top 1000 genomic tiles is much lower.
+To visualize the different degrees of separation, we further reduced the PC space from both analyses with UMAP, showing that some neuronal subtypes such as mL5-1 and mDL-3 are separated from the other subtypes when using 1000 VMRs (Figure \ref{figure:score}C).
+% why not just plot PC1 vs. PC2 here?
+
+Since some scBS protocols are based on the use of 96-well-plates, and since scBS is costly and labor-intensive, only few laboratories are currently able to obtain thousands of single cell methylomes.
+To simulate a smaller data set, we subsampled the frontal cortex data set to 500 cells and repeated our analysis.
+In this smaller data set, the performance gains were even more pronounced, indicating that our methods help extract information even from challenging data sets (Figure \ref{figure:score}D).
+
+\begin{figure*}
+    \begin{center}
+    A\hspace{.35\textwidth}B\\
+    \hspace{.3cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/complete_135k_12cm_log.pdf}
+    \hspace{.3cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/cell500_20k_12cm_log.pdf}\\
     C\\
-    \hspace{.39cm}\includegraphics[width=\columnwidth]{part_leonie_git/leonie_plots/UMAP_fulldataset.pdf}
+    \hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_fulldataset.pdf}\\
     D\\
-    \hspace{.39cm}\includegraphics[width=\columnwidth]{part_leonie_git/leonie_plots/UMAP_reduceddataset.pdf}
-    \caption{\small \textbf{Analysis comparison}.
-All plots are based on the data by \citep{luo2017single}.
-\textbf{(A)} Nearest neighbor count score for complete dataset using four different analysis techniques and a variable number of regions as input.
-Exemplary UMAPs for encircled scores are shown in (C).
-\textbf{(B)} Nearest neighbor count score for the same dataset reduced to 500 cells using four different analysis techniques and a variable number of regions as input.
-Exemplary UMAPs for encircled scores are shown in (D).
-\textbf{(C)} Two exemplary UMAPs using 1000 regions of the full dataset and methylation fraction of 100 kb windows (left) or shrunken residuals of variable methylation regions (right) \textbf{(D)} Two exemplary UMAPs using 1000 regions of the reduced dataset and methylation fraction of 100 kb windows (left) or shrunken residuals of variable methylation regions (right).}
+    \hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_reduceddataset.pdf}
+    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of 16 neuronal subtypes}.
+\textbf{(A)} Nearest neighbor count score using four different method combinations and a variable number of genomic intervals (regions) as input to PCA.
+Exemplary UMAPs for encircled analyses are shown in (C).
+\textbf{(B)} Same benchmark as in (A), but for a smaller data set consisting of 500 randomly subsampled cells.
+Exemplary UMAPs for analyses are shown in (D).
+\textbf{(C)} Two exemplary UMAPs using 1000 high-coverage intervals of the full data set and methylation fractions of 100~kb windows (left) or shrunken residuals of variable methylation intervals (right) \textbf{(D)} Two exemplary UMAPs using 1000 intervals of the smaller data set and methylation fraction of 100 kb windows (left) or shrunken residuals of variable methylation intervals (right).
+Data from \citet{luo2017single}.}
     \label{figure:score}
-\end{figure}
+    \end{center}
+\end{figure*}
 
@@
 
-\subsection{Data preparation}
+\subsection{Data storage}
 
-The function `scbs prepare` reads a set of CpG coverage files (e.g.
+The function \texttt{scbs prepare} reads a set of methylation files (e.g.
 produced by Bismark) and produces a file that stores the matrix $x$ in a space-efficient format, as follows: $x$ is represented as a SciPy sparse matrix, encoding the actual values 0, 1, and NA as -1, 1, and 0, respectively.
@@
 \]
-
-We also provide Here, we could weight the positions by number of covered cells, i.e., use
-
-$$ \tilde x_i = \frac{\sum_{i'} \overline x_{i'}\, k_h(d_{ii'})\, n_{i'}}{\sum_{i'} k_h(d_{ii'})\, n_{i'}},$$
-
-but we don't do that, or do we?
 
@@
 \subsection{PCA with imputation}
+[...]
 
-[...]
+\subsection{Re-analysis of published scBS datasets}
+To re-analyze scBS data sets by \citet{luo2017single} and \citet{argelaguet2019gastru}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb bins, or by acquiring VMRs with \texttt{scbs scan} using default options.
+For all intervals (VMRs and tiles), the mean methylation fractions and mean shrunken residuals were obtained with \texttt{scbs matrix}.
+Intervals containing less than four measured CpGs, or that were measured in less than 10\% of cells, were discarded.
+To select $n$ intervals for further analysis, the top $n$ intervals were picked based on the number of cells that the region was measured in.
+For dimensionality reduction, missing values were imputed with an iterative PCA as described above. To visualize data, UMAP was performed on the final 15 principal components.
+
+\subsection{Nearest neighbor count score}
+To assess performance of our methods, we processed mouse CpG-methylation data from \citet{luo2017single} with several method combinations and scored each result.
+The score evaluates the ability of each method combination to separate cell types in 15-dimensional PCA space, based on their CpG methylation.
+Each score varies between 0 and 1, where higher scores reflect better separation.
+
+The nearest neighbor count score $nnc(k)$ is based on the $\Gamma$-score \citep{Kireeva_2014}.
+Each cell $c_j^i$ belongs to a cell type $j$.
+For each cell, $k$ nearest neighbors $nn_j^k$ were computed based on their Euclidean distance.
+Based on the cell type $j$ of the $k$ nearest neighbors of each cell, the nnc score $nnc(k)$ was computed with formula (1), in which $d(c_j^i, k_j^i)=1$ if at least 90\% of neighbor classes match the class of cell $c_j^i$ and $d(c_j^i, k_j^i)=0$ if less than 90\% of classes match, while $n$ are the total number of cells.
+\begin{equation}
+    nnc(k) = \frac{1}{n}d(c_j^i, k_j^i)
+\end{equation}
 

### 2022-06-12 paper.tex
added e-mail addresses

@@
 \setlength{\columnsep}{0.7cm}
-\usepackage{microtype} % nice typesetting
+%\usepackage{microtype} % nice typesetting  % doesn't work for me :-( -S
 \usepackage{amsmath}
@@
 
+\newcommand\blfootnote[1]{%
+	\begingroup
+	\renewcommand\thefootnote{}\footnote{#1}%
+	\addtocounter{footnote}{-1}%
+	\endgroup
+}
+
 \begin{document}
@@
 \centering
-\textbf{\Large Analysing single-cell bisulfite sequencing data with \texttt{scbs}}
+\textbf{\Large Analysing single-cell bisulfite sequencing data with \texttt{scbs}}\\[1.5ex]
 
-Lukas P.~M. Kremer\textsuperscript{1,2}, Leonie Küchenhoff\textsuperscript{1,2}, Santiago Cerrizuela\textsuperscript{2}, ..., Ana Martin-Villalba\textsuperscript{2}, Simon Anders\textsuperscript{1}\\
-{\footnotesize \textsuperscript{1} BioQuant, University of Heidelberg, Germany}\\
-{\footnotesize \textsuperscript{2} Molecular Neurobiology, German Cancer Research Center (DKFZ), Heidelberg, Germany}
+Lukas P.~M.
+Kremer\textsuperscript{1,2,*}, Leonie Küchenhoff\textsuperscript{1,2}, Santiago Cerrizuela\textsuperscript{2}, ..., Ana Martin-Villalba\textsuperscript{2}, Simon Anders\textsuperscript{1,*}\\[1ex]
+{\footnotesize \textsuperscript{1} BioQuant Centre, University of Heidelberg, Heidelberg, Germany\\[-.5ex]
+\textsuperscript{2} Molecular Neurobiology Dept., German Cancer Research Center (DKFZ), Heidelberg, Germany}
 \vspace{1.5ex}
@@
 }]
-
+\blfootnote{\hspace{-.5cm}\raggedright\textsuperscript{*} e-mail:  \texttt{l.kremer@dkfz-heidelberg.de} and\\ \hspace{1cm}\texttt{simon.anders@bioquant.uni-heidelberg.de}}
 \section{Abstract}
@@
 
-The first data set is a single-cell multi-omic data set generated by \citet{argelaguet2019gastru}.
-In their work, the authors obtained single-cell methylomes of mouse embryo cells across different stages of development (E4.5~–~E7.5), as well as single-cell transcriptomes from the same cells.
+The first data set is a single-cell multi-omics data set generated by \citet{argelaguet2019gastru}.
+In their work, the authors obtained single-cell methylomes of mouse embryo cells across different stages of development (E4.5--E7.5) as well as single-cell transcriptomes from the same cells.
 They calculated the correlation between gene expression and promoter methylation to elucidate the implications of the epigenome on gene expression in development.
 The authors identified several genes where promoter methylation predicts gene expression, but obtained low correlations for the vast majority of genes.
+We reasoned that VMRs obtained with the previously described sliding window approach might be a better predictor of gene expression.
 
-We reasoned that VMRs obtained with the previously described sliding window approach might be a better predictor of gene expression. 
 In Figure \ref{figure:correlation}A, we compare correlations of gene expression and promoter methylation to correlations of gene expression and methylation of the closest VMR identified with \texttt{scbs scan}.

### 2022-06-12 paper.tex
a few changes to the benchmark section

@@
 }
-\renewcommand{\dbltopfraction}{0.7}
-\renewcommand{\textfraction}{0.2}
+\renewcommand{\dbltopfraction}{0.9}
+\renewcommand{\textfraction}{0.15}
 
@@
 }
+
+\newcommand{\todo}[1]{[\textcolor{orange}{#1}]}
 
@@
 CpG dinocleotides are underrepresented in eukaryotic genomes (due to their vulnerability to mutation by conversion to TpG due to spontaneous deamination), and the majority of existing CpG sites are methylated.
-Presence or absence of the methyl group at CpG sites is believed to have effect on gene regulation, with methylation inhibiting transcription.
-The removal of DNA methylation during cell differentiation is therefore conjectured to be an important mechanism to activate genome regions that are required in the target lineage and define cell identity \citet{cedar2022role}.
-One important application of single-cell bisulfite sequencing (scBS) therefore lies in the study of differentiation processes, for instance in embryonic development \citet{argelaguet2019gastru}.
+Presence or absence of the methyl group at CpG sites is believed to affect gene regulation, with methylation inhibiting transcription.
+The removal of DNA methylation during cell differentiation is therefore conjectured to be an important mechanism to activate genome regions that are required in the target lineage and define cell identity \citep{cedar2022role}.
+One important application of single-cell bisulfite sequencing (scBS) therefore lies in the study of differentiation processes, for instance in embryonic development \citep{argelaguet2019gastru}.
 
@@
 
-The standard approach to analyse scBS data is based on methodology developed for the analysis of single-cell RNA-Seq data.
+The standard approach to analyse scBS-Seq data is based on methodology developed for the analysis of single-cell RNA-Seq data.
 Therefore, we will start by briefly reviewing how single-cell \emph{RNA}-Seq data is commonly analysed, before we discuss scBS data analysis.
@@
     \begin{center}
-        \includegraphics[width=.8\textwidth]{figures/Fig_workflow.png}
+        \includegraphics[width=.85\textwidth]{figures/Fig_workflow.png}
     \end{center}
@@
 
-In brief, we store methylation of each chromosome in a binary matrix where each column represents a cell and each row represents a base pair.
-Since the vast majority of values in this matrix are missing due to the sparsity of scBS data, we encode missing values as zero and then store the data in Compressed Sparse Row (CSR) format.
+In brief, we store methylation of each chromosome in a matrix where each column represents a cell and each row represents a base pair.
+Methylated and unmethylated sites are coded as $+1$ and $-1$, respectivelty.
+Since the vast majority of values in this matrix are missing due to the sparsity of scBS data (and because rows for base-pairs not corresponding to a CpG site contain no data), we encode missing values as zero and then store the data in Compressed Sparse Row (CSR) format.
 This format does not explicitly store zeroes (here: missing values) and is optimized for row-wise access, which results in significant compression and allows fast access to the methylation status of genomic intervals.
@@
 
-After quality control, the user has access to the previously described genome-wide VMR detection approach via \texttt{scbs scan}.
-This produces a BED-file that lists the genome coordinates of VMRs.
-To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED-file can be used as input for \texttt{scbs matrix}, which quantifies methylation at genomic intervals in all cells.
+After quality control, the user has access to the genome-wide VMR detection approach described above via \texttt{scbs scan}.
+This produces a BED file that lists the genome coordinates of VMRs.
+To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED file can be used as input for \texttt{scbs matrix}, which quantifies methylation at genomic intervals in all cells.
 The command produces both the percentage of methylated CpG sites, as well as our proposed methylation measure (shrunken mean of the residuals) that is more robust to variation in sequencing coverage or stochastic differences in read position between cells.
@@
 \section{Benchmarks}
-To assess performance of our newly devised methods, we obtained two previously published scBS data sets and re-analysed them with \texttt{scbs}.
+To assess performance of our methods, we obtained two previously published scBS-Seq data sets and re-analysed them with \texttt{scbs}.
+
+\subsection{Correlation with gene expression}
 
@@
 In their work, the authors obtained single-cell methylomes of mouse embryo cells across different stages of development (E4.5--E7.5) as well as single-cell transcriptomes from the same cells.
-They calculated the correlation between gene expression and promoter methylation to elucidate the implications of the epigenome on gene expression in development.
+In one of their analyses, the authors calculated the correlation between gene expression and promoter methylation to elucidate the implications of the epigenome on gene expression in development.
 The authors identified several genes where promoter methylation predicts gene expression, but obtained low correlations for the vast majority of genes.
@@
 
-To illustrate this effect with an example, we picked a gene where VMR methylation is a much better predictor of gene expression than promoter methylation (Ube2l6, highlighted in orange in Figure \ref{figure:correlation}).
+To illustrate this effect with an example, we picked a gene where VMR methylation is a much better predictor of gene expression than promoter methylation (Ube2l6, highlighted in orange in Figure \ref{figure:correlation}A,B).
 Grouping all cells into four groups according to their level of Ube2l6 expression, it is clearly visible that the promoter of this gene is lowly methylated regardless of gene expression, while methylation at the nearest VMR is greatly decreased in cells with high Ube2l6 expression.
 
-To assess the impact of our methods on dimensionality reduction and the ability to distinguish different (sub)types of cells, we obtained a second data set from \citet{luo2017single}.
-This data set consists of over three thousand single-cell CpG methylomes from murine frontal cortex, comprising 16 neuronal subtypes.
+\subsection{Identification of cellular sub-types}
+
+To assess the impact of our methods on dimensionality reduction and the ability to distinguish different (sub)types of cells, we used the data set from \citet{luo2017single}.
+This data set consists of over three thousand \todo{Give exact number.} single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal subtypes.
 We subjected this data set to the standard scBS workflow: defining a set of genomic intervals, quantifying methylation in these intervals, and subjecting the resulting matrix to PCA.
 We repeated this analysis several times, either using 100~kb tiles or VMRs, and either using methylation fractions or mean shrunken residuals as a measure of methylation.
+\todo{Is the following correct?} For both types of regions, the 100~kb tiles and the VMRs, we ranked the regions by the number of covering cells (i.e., cells with at least on read in the region) and kep only the top 1000 regions.
 
@@
 Most of this gain in performance can be attributed to the use of VMRs over genomic tiles. The use of mean shrunken residuals over methylation fractions provides an additional small improvement.
-Of note, our results show that even few VMRs, e.g. the 1000 VMRs that have read coverage in the highest number of cells, already contain sufficient information to distinguish neuronal subtypes.
-In contrast, the neighbor count score for the top 1000 genomic tiles is much lower.
-To visualize the different degrees of separation, we further reduced the PC space from both analyses with UMAP, showing that some neuronal subtypes such as mL5-1 and mDL-3 are separated from the other subtypes when using 1000 VMRs (Figure \ref{figure:score}C).
-% why not just plot PC1 vs. PC2 here?
 
@@
 To simulate a smaller data set, we subsampled the frontal cortex data set to 500 cells and repeated our analysis.
-In this smaller data set, the performance gains were even more pronounced, indicating that our methods help extract information even from challenging data sets (Figure \ref{figure:score}D).
+Here, our approach keeps its good score, while the simple approach falls behind (Figure \ref{figure:score}B).
+
+Of note, our results show that even few VMRs (here, the 1000 VMRs that have read coverage in the highest number of cells) already contain sufficient information to distinguish neuronal subtypes.
+In contrast, the neighbor count score for the top 1000 genomic tiles is much lower.
+To visualize the different degrees of separation, we further reduced the PC space from both analyses with UMAP, showing that some neuronal subtypes such as mL5-1 and mDL-3 are separated from the other subtypes when using 1000 VMRs (Figure \ref{figure:score}C).
+% why not just plot PC1 vs. PC2 here? % because people will want to see the UMAP. -S
+Again, the difference is even more pronounced in the sub-sampled analysis with only 500 cells (Figure \ref{figure:score}D).
 
@@
     \begin{center}
-    A\hspace{.35\textwidth}B\\
+    A\hspace{.4\textwidth}B\hspace*{.3\textwidth}~\\
     \hspace{.3cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/complete_135k_12cm_log.pdf}
-    \hspace{.3cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/cell500_20k_12cm_log.pdf}\\
-    C\\
+    \hspace{1cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/cell500_20k_12cm_log.pdf}\\
+    C\hspace*{.7\textwidth}~\\
     \hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_fulldataset.pdf}\\
-    D\\
+    D\hspace*{.7\textwidth}~\\
     \hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_reduceddataset.pdf}
@@
 Exemplary UMAPs for analyses are shown in (D).
-\textbf{(C)} Two exemplary UMAPs using 1000 high-coverage intervals of the full data set and methylation fractions of 100~kb windows (left) or shrunken residuals of variable methylation intervals (right) \textbf{(D)} Two exemplary UMAPs using 1000 intervals of the smaller data set and methylation fraction of 100 kb windows (left) or shrunken residuals of variable methylation intervals (right).
+\textbf{(C)} Two exemplary UMAPs of the full data set and methylation fractions of 100~kb windows (left) or shrunken residuals of variable methylation intervals (right) \textbf{(D)} Two exemplary UMAPs of the smaller data set and methylation fraction of 100 kb windows (left) or shrunken residuals of variable methylation intervals (right).
 Data from \citet{luo2017single}.}

### 2022-06-12 paper.tex
changed date
fixed capitalization in bibliography

@@
 
-March 2022
+June 2022
 \vspace{6ex}
@@
 To re-analyze scBS data sets by \citet{luo2017single} and \citet{argelaguet2019gastru}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb bins, or by acquiring VMRs with \texttt{scbs scan} using default options.
-For all intervals (VMRs and tiles), the mean methylation fractions and mean shrunken residuals were obtained with \texttt{scbs matrix}.
+\todo{Mention choice of top 1000 regions.} For all intervals (VMRs and tiles), the mean methylation fractions and mean shrunken residuals were obtained with \texttt{scbs matrix}.
 Intervals containing less than four measured CpGs, or that were measured in less than 10\% of cells, were discarded.
@@
 
-The nearest neighbor count score $nnc(k)$ is based on the $\Gamma$-score \citep{Kireeva_2014}.
+\todo{Improve wording} The nearest neighbor count score $nnc(k)$ is based on the $\Gamma$-score \citep{Kireeva_2014}.
 Each cell $c_j^i$ belongs to a cell type $j$.
 For each cell, $k$ nearest neighbors $nn_j^k$ were computed based on their Euclidean distance.
-Based on the cell type $j$ of the $k$ nearest neighbors of each cell, the nnc score $nnc(k)$ was computed with formula (1), in which $d(c_j^i, k_j^i)=1$ if at least 90\% of neighbor classes match the class of cell $c_j^i$ and $d(c_j^i, k_j^i)=0$ if less than 90\% of classes match, while $n$ are the total number of cells.
+Based on the cell type $j$ of the $k$ nearest neighbors of each cell, the nnc score $nnc(k)$ was computed with formula (\todo{ref}), in which $d(c_j^i, k_j^i)=1$ if at least 90\% of neighbor classes match the class of cell $c_j^i$ and $d(c_j^i, k_j^i)=0$ if less than 90\% of classes match, while $n$ are the total number of cells.
 \begin{equation}
-    nnc(k) = \frac{1}{n}d(c_j^i, k_j^i)
+    \text{nnc}(k) = \frac{1}{n}d(c_j^i, k_j^i)
 \end{equation}

### 2022-06-12 paper.tex
added citation to SciPy

@@
 The function \texttt{scbs prepare} reads a set of methylation files (e.g.
-produced by Bismark) and produces a file that stores the matrix $x$ in a space-efficient format, as follows: $x$ is represented as a SciPy sparse matrix, encoding the actual values 0, 1, and NA as -1, 1, and 0, respectively.
+produced by Bismark) and produces a file that stores the matrix $x$ in a space-efficient format, as follows: $x$ is represented as a SciPy sparse matrix \citep{SciPy}, encoding the actual values 0, 1, and NA as -1, 1, and 0, respectively.
 Coding NA (the most common value) as 0 leverages SciPy's sparse matrix storage.

### 2022-06-12 paper.tex
marked notes in orange

@@
 A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm.
-In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-5}$) to the normalized fractions before taking the logarithm.
+In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-4}$) to the normalized fractions before taking the logarithm.
 Now, one could use Euclidean distances of these vectors of logarithmized fractions as dissimilarity score.
@@
 
-NOTE: Here, we could already do shrinkage and use $\frac{1}{|C_i|+1}\sum_i x_{ij}$ intead. But we don't do that, right?
+\todo{NOTE: Here, we could already do shrinkage and use $\frac{1}{|C_i|+1}\sum_i x_{ij}$ intead. But we don't do that, right?}
 

### 2022-06-13 paper.tex
add workflow fig caption

@@
     \begin{center}
-        \includegraphics[width=.85\textwidth]{figures/Fig_workflow.png}
+        \includegraphics[width=.8\textwidth]{figures/Fig_workflow.png}
     \end{center}
-    \caption{\small \textbf{Overview of the functionalities implemented in the scbs package}
+    \caption{\small \textbf{Overview of the functionalities implemented in the \texttt{scbs} package.}
+    \texttt{scbs prepare} parses methylation files produced by common bisulfite sequencing mappers and stores their contents in a compressed format optimised for efficient access to genomic intervals.
+    \texttt{scbs} also produces cell-wise summary statistics and quality plots (here: average methylation around the transcription start site) that are used to detect low-quality cells.
+    These cells can be discarded with \texttt{scbs filter}.
+    To obtain a methylation matrix, similar to the count matrices used in scRNA-seq, the user must first decide in which genomic intervals methylation should be quantified.
+    The user may either provide genomic regions of \emph{a priori} interest, or they may choose to discover VMRs (variably-methylated regions) in the data with \texttt{scbs scan}.
+    The resulting methylation matrix can then be used for downstream analysis such as cell clustering and dimensionality reduction.
     }
@@
 For each cell, these tools produce a text file that lists the methylation status of all CpG sites.
-Since it is inconvenient to work with hundreds or thousands of text files, \texttt{scbs} provides the \texttt{prepare} command which parses methylation files and stores their content in a compressed format that enables efficient access to all CpG sites of the genome.
+Since it is inconvenient to work with hundreds or thousands of text files, \texttt{scbs} provides the \texttt{prepare} command which parses these methylation files and stores their content in a compressed format that enables efficient access to all CpG sites of the genome.
+\texttt{scbs prepare} is flexible and accepts all tabular input formats including Bismark .cov-files, methylpy .allc-files, or even custom user-defined formats.
 
 In brief, we store methylation of each chromosome in a matrix where each column represents a cell and each row represents a base pair.
-Methylated and unmethylated sites are coded as $+1$ and $-1$, respectivelty.
+Methylated and unmethylated sites are coded as $+1$ and $-1$, respectively.
 Since the vast majority of values in this matrix are missing due to the sparsity of scBS data (and because rows for base-pairs not corresponding to a CpG site contain no data), we encode missing values as zero and then store the data in Compressed Sparse Row (CSR) format.

### 2022-06-13 paper.tex
moved panel letters

--- 2022-06-12 paper.tex
@@
 Lukas P.~M.
-Kremer\textsuperscript{1,2,*}, Leonie Küchenhoff\textsuperscript{1,2}, Santiago Cerrizuela\textsuperscript{2}, ..., Ana Martin-Villalba\textsuperscript{2}, Simon Anders\textsuperscript{1,*}\\[1ex]
+Kremer\textsuperscript{1,2,*}, Leonie Küchenhoff\textsuperscript{1,2}, Santiago Cerrizuela\textsuperscript{2}, \todo{...}, Ana Martin-Villalba\textsuperscript{2}, Simon Anders\textsuperscript{1,*}\\[1ex]
 {\footnotesize \textsuperscript{1} BioQuant Centre, University of Heidelberg, Heidelberg, Germany\\[-.5ex]
@@
 
+\begin{figure*}
+	\hspace{2.5cm}A\\[-7ex]
+	\begin{center}
+		\includegraphics[width=0.7\textwidth]{figures/Fig_residuals_A.png}\\
+	\end{center}
+	\hspace{2.5cm}B\\[-3ex]
+	\begin{center}
+		\includegraphics[width=0.7\textwidth]{figures/Fig_residuals_B.png}
+	\end{center}
+	\caption{\small \textbf{Effect of read position:} \textbf{(A)} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified.
+Two cells cover differing parts of the interval with one read each.
+If one simply counts for each cell which fraction of the covered CpG sites are methylated, one obtains very different values for the two cells.
+\textbf{(B)} By averaging each CpG site's methylation over all cells and subsequent smoothing, the thick black ``average methylation curve'' is obtained.
+To quantify the methylation of cell 1 from (A) relative to this average over all cells, we propose to use the cell's residuals to the smoothed curve (lengths of the vertical red lines) and take their average, counting residuals of methylated CpGs positive and residuals of unmethylated CpGs negative.}
+	\label{fig:smoothres}
+\end{figure*}
+
 The standard approach to analyse scBS-Seq data is based on methodology developed for the analysis of single-cell RNA-Seq data.
@@
 
-%\begin{figure}
-%   \framebox[\columnwidth]{
-%       \begin{minipage}[c][3.5cm]{\columnwidth}
-%           \centering Figure
-%       \end{minipage}
-%   }
-%   \caption{\small UMAP representations of the single-cell bisulfite data of XXX et al., calculated \textbf{(A)} using conventional preprocessing and \textbf{(B)} using the approach described here.}
-%   \label{umaps}
-%\end{figure}
-
 This procedure is commonly adapted when working with single-cell DNA-methylation data, so that established methods used in scRNA-Seq can be used.
@@
 
-%As an example, we will use in this paper data from \citet{luo2017single}, who performed scBS on 16 neuronal subtypes of the mouse frontal cortex, and analysed their data in the manner just described.
-%Figure \ref{umaps}A shows a UMAP plot, derived from our re-analysis of the data, using a tiling of the genome into windows of size 100~kb.
-As one can see, the approach is suitable to see clear structure in the data, observe ...
-
 We think that this analysis strategy has room for improvement, as we will discuss next.
@@
 For the mathematical details, please see the Methods section.
-
-\begin{figure*}
-    A\\
-    \begin{center}
-        \includegraphics[width=0.7\textwidth]{figures/Fig_residuals_A.png}\\
-    \end{center}
-    B\\
-    \begin{center}
-        \includegraphics[width=0.7\textwidth]{figures/Fig_residuals_B.png}
-    \end{center}
-\caption{\small \textbf{Effect of read position:} \textbf{(A)} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified.
-Two cells cover differing parts of the interval with one read each.
-If one simply counts for each cell which fraction of the covered CpG sites are methylated, one obtains very different values for the two cells.
-\textbf{(B)} By averaging each CpG site's methylation over all cells and subsequent smoothing, the thick black ``average methylation curve'' is obtained.
-To quantify the methylation of cell 1 from (A) relative to this average over all cells, we propose to use the cell's residuals to the smoothed curve (lengths of the vertical red lines) and take their average, counting residuals of methylated CpGs positive and residuals of unmethylated CpGs negative.}
-\label{fig:smoothres}
-\end{figure*}
 
@@
 We then run a kernel smoother over these per-position averages to obtain the smoothed averages $\tilde x_i$. Specifically, we use
-
 \[ \tilde x_i = \frac{\sum_{i'} \overline x_{i'}\, k_h(d_{ii'})}{\sum_{i'} k_h(d_{ii'})},\]

### 2022-06-13 paper.tex
minor stuff + discussion draft

@@
 \begin{figure}
-    A\\
+    \begin{center}
+    A\hspace{.9\columnwidth}~\\
     \includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/corr_methylation_fraction_4kbwindow.pdf} \\
-    B\\
+    B\hspace{.9\columnwidth}~\\
     \includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/corr_shrunken_residual_4kbwindow.pdf} \\
-    C\\
+    C\hspace{.9\columnwidth}~\\
     \includegraphics[width=.95\columnwidth]{part_leonie_git/leonie_plots/comparison_meth_RNA.pdf}
+    \end{center}
     \caption{\small \textbf{Correlation of DNA methylation and gene expression}.
@@
 
-To assess the impact of our methods on dimensionality reduction and the ability to distinguish different (sub)types of cells, we used the data set from \citet{luo2017single}.
-This data set consists of over three thousand \todo{Give exact number.} single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal subtypes.
+To assess the impact of our methods on dimensionality reduction and the ability to distinguish different (sub)types of cells, we used a data set from \citet{luo2017single}.
+This data set consists of 3377 single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal subtypes.
 We subjected this data set to the standard scBS workflow: defining a set of genomic intervals, quantifying methylation in these intervals, and subjecting the resulting matrix to PCA.
 We repeated this analysis several times, either using 100~kb tiles or VMRs, and either using methylation fractions or mean shrunken residuals as a measure of methylation.
-\todo{Is the following correct?} For both types of regions, the 100~kb tiles and the VMRs, we ranked the regions by the number of covering cells (i.e., cells with at least on read in the region) and kep only the top 1000 regions.
+\todo{Is the following correct?} For both types of regions, the 100~kb tiles and the VMRs, we ranked the regions by the number of covering cells (i.e., cells with at least on read in the region) and keep only the top 1000 regions.
 
@@
 
-...
+%Here we propose a number of new methods for the analysis of single cell methylation data and provide them to the research community as a command line tool called \texttt{scbs}.
+%Instead of dividing the genome into coarse-grained tiles, we propose to discover variably-methylated regions (VMRs) in the data itself, which improves performance of common downstream applications.
+%We also think that VMRs, due to their smaller size, are easier to interpret than  and furthermore provides a list of candidate regions.
+%These regions may then be linked to individual genes, tested for celltype-specific methylation differences, or subjected to gene ontology or motif enrichment, depending on the research question at hand.
+%We also propose a number of methods specifically tailored to the sparse nature of scBS data.
+%First, we observe
+
+%- propose a number of new methods for the analysis of scBS data
+%- developed methods to analyse scBS data that go beyond simple genome tiling
+% - instead we discover potential genomic regions of interest on the data itself, which improves performance of common downstream applications and aids interpretability.
+% - VMRs can also be used as input for e.g. motif enrichment, go enrichment etc
+%- devised a method to store genome-wide scBS data efficiently
+%- propose a new measure of DNA methylation in a genomic interval (~ relative methylation)
+
+%- implement them in Python and release them to the community as a Python package with a CLI
+%-> no more custom scripts and manual data wrangling of 1000 text files required, instead every user can quickly go from raw data to count matrix
+%- lowers the entry barrier for scBS data analysis and makes this new and exciting data type available to a greater audience
+
+%whats missing?
+%-
 
@@
 
-\todo{NOTE: Here, we could already do shrinkage and use $\frac{1}{|C_i|+1}\sum_i x_{ij}$ intead. But we don't do that, right?}
+%\todo{NOTE: Here, we could already do shrinkage and use $\frac{1}{|C_i|+1}\sum_i x_{ij}$ instead. But we don't do that, right?}
 
@@
 For any two cells $j$, $j'$, we use the Euclidean distance $\|\mathbf{x}^\text{P}_j - \mathbf{x}^\text{P}_{j'}\|$ as the measure of dissimilarity of the two cells.
-Thus, the matrix of PC scores can be used as input to dimension reduction methods like t-SNE or UMAP, and to clustering methods like the Louvain or Leiden algorithm, which requirte such a matrix as input to the approximate mnearest neighbor (ANN) finding algorithm that is their first step.
+Thus, the matrix of PC scores can be used as input to dimension reduction methods like t-SNE or UMAP, and to clustering methods like the Louvain or Leiden algorithm, which require such a matrix as input to the approximate mnearest neighbor (ANN) finding algorithm that is their first step.
 
@@
 \subsection{Nearest neighbor count score}
-To assess performance of our methods, we processed mouse CpG-methylation data from \citet{luo2017single} with several method combinations and scored each result.
+To assess performance of our methods, we processed mouse CpG methylation data from \citet{luo2017single} with several method combinations and scored each result.
 The score evaluates the ability of each method combination to separate cell types in 15-dimensional PCA space, based on their CpG methylation.

### 2022-06-13 paper.tex
shortened introduction

@@
 Sequencing-based assays with single-cell resolution have offered new means to understand the differences between the cells making up a sample.
-Single-cell RNA sequencing (scRNA-seq) techniques have matured at great pace in recent years.
-Now, methods to study epigenetics at single-cell resolution are rapidly catching up.
-Of special interest here is DNA methylation, i.e., the methylation of bases in the DNA, which is passed on to daughter cells in cell division.
-Especially in mammals, such methylation chiefly occurs at cytosine bases that are immediately after followed by guanine bases, so-called CpG \mbox{(5'--C--phosphate--G--3')} sites.
+Single-cell RNA sequencing (sc-RNA-seq) techniques have matured at great pace in recent years, with well developed analysis methodology, and methods to study epigenetics at single-cell resolution are rapidly catching up.
 
-To detect CpG methylation by sequencing, bisulfite conversion is commonly employed: treatment of DNA with bisulfite causes unmethylated cytosine bases to be deaminated to uracil, which is replaced by thymine in subsequent PCR.
-Therefore, unmethylated CpGs are read as TG in sequencing of bisulfite-converted DNA while only methylated CpG sites are correctly read as CG \citep{Frommer_1992}.
-Such so-called bisulfite sequencing has been used for bulk samples since long, and methods for bisulfite sequencing with single-cell resolution were recently developed \citep{Smallwood_2014} and are increasingly used.
+Here, we discuss bisulfite sequencing (sc-BS-Seq), a method to study DNA methylation.
+In brief, DNA is treated with bisulfite, which converts unmethylated cytosines to uracils which are read as thymine in subsequent PCR, while methylated cytosines are protected from conversion.
+After sequencing, these conversions allow to determine the methylation status of all cytosines covered by reads \citep{Frommer_1992}.
+Bisulfite sequencing can also be performed with single-cell resolution \citep{Smallwood_2014} and even in parallel with sc-RNA-Seq \citep{Clark2018}.
 
-CpG dinocleotides are underrepresented in eukaryotic genomes (due to their vulnerability to mutation by conversion to TpG due to spontaneous deamination), and the majority of existing CpG sites are methylated.
-Presence or absence of the methyl group at CpG sites is believed to affect gene regulation, with methylation inhibiting transcription.
-The removal of DNA methylation during cell differentiation is therefore conjectured to be an important mechanism to activate genome regions that are required in the target lineage and define cell identity \citep{cedar2022role}.
-One important application of single-cell bisulfite sequencing (scBS) therefore lies in the study of differentiation processes, for instance in embryonic development \citep{argelaguet2019gastru}.
-
-In the present paper, we discuss strategies to analyze scBS data, suggest improvement to current approaches, show their value in a benchmark, and present software to perform the improved analysis methods.
+In the present paper, we discuss strategies to analyze sc-BS-Seq data, suggest improvement to current approaches, show their value in a benchmark, and present software to perform the improved analysis methods.
 
@@
 The standard approach to analyse scBS-Seq data is based on methodology developed for the analysis of single-cell RNA-Seq data.
-Therefore, we will start by briefly reviewing how single-cell \emph{RNA}-Seq data is commonly analysed, before we discuss scBS data analysis.
+Therefore, we start by briefly reviewing how single-cell \emph{RNA}-Seq data is commonly analysed, before we discuss scBS data analysis.
 
@@
 of counts of distinct RNA molecules), one row for each cell and one column for each gene.
-A first goal is then to assigning cell types or states to cells, and to this end, one needs to establish which cells are similar to each other, i.e., find a way to quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile.
+A first goal is usually to assigning cell types or states to cells.
+To this end, one needs to establish which cells are similar to each other, i.e., quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile.
 A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm.
@@
 Therefore, Euclidean distances between these ``PCA space'' vectors provide a robust dissimilarity score.
-Hence, the PCA space representation y are suitable as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data, or to methods for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).
+Hence, the PCA space representation is suitable as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data, or to methods for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).
 
 This procedure is commonly adapted when working with single-cell DNA-methylation data, so that established methods used in scRNA-Seq can be used.
-The question is therefore how to construct a suitable matrix as input for the PCA.
+The task is therefore how to construct a suitable matrix as input for the PCA.
 A simple and common approach, used for instance by \citet{luo2017single}, is to tile the genome into windows of, say, 100~kb size, and calculate for each cell the average methylation of the DNA within each window.
@@
 This matrix is now subjected to PCA.
+After the PCA, one can proceed using approaches known from sc-RNA-Seq.
 
-We think that this analysis strategy has room for improvement, as we will discuss next.
-Here, we discuss the proposed improvements only qualitatively.
-For the mathematical details, please see the Methods section.
+While this simple procedure is straight-forward and produces satisfactory results, there is room for improvement.
+Here, we propose three improvements that, as we show afterwards, considerably increase the information content of the processed data.
+We will first explain the proposed improvement in a qualitative manner, while mathematical details are given later in the Methods section.
+Then, we will demonstrate the improvement using benchmarks.
 

### 2022-06-13 paper.tex
edited  approach description

@@
 
-\section{Three improvements to processing scBS data}
+\section{Improved processing of scBS data}
 
@@
 In Figure \ref{fig:smoothres}A, such an interval is covered by a single read for each of the two cells shown.
-The read from cell 2 shows much more methylation than the read from cell 1, and a simple analysis would therefore consider that cell as having stronger methylation in the interval.
+The read from cell 2 shows much more methylation than the read from cell 1, and the standard analysis would therefore consider that cell as having stronger methylation in the interval.
 However, given that the two reads agree wherever they overlap, a more parsimonious interpretation would be that the cells do not show difference in methylation within the interval.
@@
 
-Therefore, we propose to first obtain, for each CpG position, an average methylation across all cells and, then quantify for each individual cell its deviation from this average.
-In Figure \ref{fig:smoothres}B, the curved line depicts such an average over all cells, the red vertical lines show an individual cell's deviation from the ensemble average.
+Therefore, we propose to first obtain, for each CpG position, an average of the methylation across all cells and, then quantify for each individual cell its deviation from this average.
+In Figure \ref{fig:smoothres}B, the curved line depicts such an average over all cells, and the red vertical lines show an individual cell's deviation from the ensemble average.
 We take the lengths of the red lines as signed values (``residual''), positive for lines extending upwards from the curve (methylated CpG) and negative for lines extending downwards (unmethylated CpG), and take the average over the residual for all the CpGs in the interval that are covered by reads from this cell.
@@
 
-The signal to noise ratio in a matrix thus obtained will be much better than in a matrix obtained by simply averaging absolute methylation (0 or 1) over all the cell's covered CpG sites in a region.
-The reason for this is that we reduce the variation in situation as the one depicted in Figure \ref{fig:smoothres}A, where absolute methylation might differ strongly even though there is no actual evidence for a difference between the two cells.
+The signal to noise ratio in a matrix thus obtained will be better than in a matrix obtained by simply averaging absolute methylation (0 or 1) over all the cells' covered CpG sites in a region.
+The reason for this is that we reduce the variation in situation as the one depicted in Figure \ref{fig:smoothres}A, where the methylation of the reads might differ strongly even though there is no actual evidence for a difference between the two cells.
 
@@
 
-So far, we have discussed dividing up (tiling) each chromosome into non-overlapping, equal-sized intervals, and quantitating the methylation of each such window.
+In the standard approach, one divides up (tiles) each chromosome into non-overlapping, equal-sized intervals, and quantitates the methylation of each such window.
 Such rigid placing of interval boundaries is unlikely to be optimal: for example, a VMR might be much smaller than a tile and its signal will hence be drowned out by the larger part of uninformative CpGs that are equal in all cells, when averaging over all the CpG sites in the tile.
@@
 Since it is inconvenient to work with hundreds or thousands of text files, \texttt{scbs} provides the \texttt{prepare} command which parses these methylation files and stores their content in a compressed format that enables efficient access to all CpG sites of the genome.
-\texttt{scbs prepare} is flexible and accepts all tabular input formats including Bismark .cov-files, methylpy .allc-files, or even custom user-defined formats.
+\texttt{scbs prepare} is flexible and accepts all tabular input formats including Bismark .cov-files, methylpy .allc-files, or custom user-defined formats.
 
@@
 Methylated and unmethylated sites are coded as $+1$ and $-1$, respectively.
-Since the vast majority of values in this matrix are missing due to the sparsity of scBS data (and because rows for base-pairs not corresponding to a CpG site contain no data), we encode missing values as zero and then store the data in Compressed Sparse Row (CSR) format.
+Since the vast majority of values in this matrix are missing due to the sparsity of sc-BS-Seq data (and because rows for base-pairs not corresponding to a CpG site contain no data), we encode missing values as zero and then store the data in Compressed Sparse Row (CSR) format.
 This format does not explicitly store zeroes (here: missing values) and is optimized for row-wise access, which results in significant compression and allows fast access to the methylation status of genomic intervals.

### 2022-06-13 paper.tex
moved package description behind benchmark
a few more changes

@@
 
+\section{Benchmarks}
+
+To demonstrate the value of our proposed improvements, we obtained two previously published sc-BS-Seq data sets and re-analysed them.
+
+\subsection{Correlation with gene expression}
+
+The first data set is a single-cell multi-omics data set generated by \citet{argelaguet2019gastru}.
+In their work, the authors obtained single-cell methylomes of mouse embryo cells across different stages of development (E4.5--E7.5) as well as single-cell transcriptomes from the same cells.
+In one of their analyses, the authors calculated the correlation between gene expression and promoter methylation to elucidate the implications of the epigenome on gene expression in development.
+The authors identified several genes where promoter methylation predicts gene expression, but obtained low correlations for the vast majority of genes.
+We reasoned that VMRs obtained with the previously described sliding window approach might be a better predictor of gene expression.
+
+In Figure \ref{figure:correlation}A, we compare correlations of gene expression and promoter methylation to correlations of gene expression and methylation of the closest VMR identified with \texttt{scbs scan}.
+This analysis revealed a number of genes whose expression does not correlate with promoter methylation, but correlates with VMR methylation.
+This effect is also observed when using our proposed measure of methylation, the shrunken mean of residuals (Figure \ref{figure:correlation}B).
+
+\begin{figure}
+    \begin{center}
+    A\hspace{.9\columnwidth}~\\
+    \includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/corr_methylation_fraction_4kbwindow.pdf} \\
+    B\hspace{.9\columnwidth}~\\
+    \includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/corr_shrunken_residual_4kbwindow.pdf} \\
+    C\hspace{.9\columnwidth}~\\
+    \includegraphics[width=.95\columnwidth]{part_leonie_git/leonie_plots/comparison_meth_RNA.pdf}
+    \end{center}
+    \caption{\small \textbf{Correlation of DNA methylation and gene expression}.
+    \textbf{(A)} Correlation of each gene's RNA expression values to the the methylation fractions that were measured at promoters (y-axis), or to methylation fractions of the nearest VMR (x-axis).
+Promoters are defined as intervals \textpm2~kb around the TSS.
+    \textbf{(B)} Same as in (A), but using mean shrunken residuals as a measure of methylation.
+    \textbf{(C)} Mean smoothed methylation near the gene Ube2l6 (ENSMUSG00000027078, marked in yellow in A \& B).
+Cells are assigned to 4 groups based on Ube2l6 expression (group 0: cells with no Ube2l6 expression, groups 1-3: cells that express Ube2l6, divided into three equally large groups with group 3 having the highest expression).
+The mean methylation level of each group was smoothed with a tricube kernel of bandwidth 2000.
+    Data from \citet{argelaguet2019gastru}.}
+    \label{figure:correlation}
+\end{figure}
+
+To illustrate this effect with an example, we picked a gene where VMR methylation is a much better predictor of gene expression than promoter methylation (Ube2l6, highlighted in orange in Figure \ref{figure:correlation}A,B).
+Grouping all cells into four groups according to their level of Ube2l6 expression, it is clearly visible that the promoter of this gene is lowly methylated regardless of gene expression, while methylation at the nearest VMR is greatly decreased in cells with high Ube2l6 expression.
+
+\subsection{Identification of cellular sub-types}
+
+To assess the impact of our methods on dimensionality reduction and the ability to distinguish different (sub)types of cells, we used a data set from \citet{luo2017single}.
+This data set consists of 3377 single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal subtypes.
+We subjected this data set to the standard scBS workflow: defining a set of genomic intervals, quantifying methylation in these intervals, and subjecting the resulting matrix to PCA.
+We repeated this analysis several times, either using 100~kb tiles or VMRs, and either using methylation fractions or mean shrunken residuals as a measure of methylation.
+\todo{Is the following correct?} For both types of regions, the 100~kb tiles and the VMRs, we ranked the regions by the number of covering cells (i.e., cells with at least on read in the region) and kept only the top 100, 1000, or 10,000 regions.
+
+To assess performance of each combination of methods, we used a nearest neighbor count score that quantifies how well the different neuronal subtypes are separated in 15 dimensional PC space (see Methods).
+The results demonstrate that the 16 neuronal subtypes are more cleanly separated when using our proposed methods (Figure \ref{figure:score}A).
+Most of this gain in performance can be attributed to the use of VMRs over genomic tiles. The use of mean shrunken residuals over methylation fractions provides an additional small improvement.
+
+Since sc-BS-Seq protocols is still rather costly and labor-intensive, only few laboratories are currently able to obtain thousands of single cell methylomes.
+To simulate a smaller data set, we subsampled the frontal cortex data set to 500 cells and repeated our analysis.
+Here, our approach keeps its good score, while the simple approach falls behind (Figure \ref{figure:score}B).
+
+Of note, our results show that even few VMRs (here, the 1000 VMRs that have read coverage in the highest number of cells) already contain sufficient information to distinguish neuronal subtypes.
+In contrast, the neighbor count score for the top 1000 genomic tiles is much lower.
+To visualize the different degrees of separation, we further reduced the PC space from both analyses with UMAP, showing that some neuronal subtypes such as mL5-1 and mDL-3 are separated from the other subtypes when using 1000 VMRs (Figure \ref{figure:score}C).
+% why not just plot PC1 vs. PC2 here? % because people will want to see the UMAP. -S
+Again, the difference is even more pronounced in the sub-sampled analysis with only 500 cells (Figure \ref{figure:score}D).
+
+\begin{figure*}
+    \begin{center}
+    A\hspace{.4\textwidth}B\hspace*{.3\textwidth}~\\
+    \hspace{.3cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/complete_135k_12cm_log.pdf}
+    \hspace{1cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/cell500_20k_12cm_log.pdf}\\
+    C\hspace*{.7\textwidth}~\\
+    \hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_fulldataset.pdf}\\
+    D\hspace*{.7\textwidth}~\\
+    \hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_reduceddataset.pdf}
+    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of 16 neuronal subtypes}.
+\textbf{(A)} Nearest neighbor count score using four different method combinations and a variable number of genomic intervals (regions) as input to PCA.
+Exemplary UMAPs for encircled analyses are shown in (C).
+\textbf{(B)} Same benchmark as in (A), but for a smaller data set consisting of 500 randomly subsampled cells.
+Exemplary UMAPs for analyses are shown in (D).
+\textbf{(C)} Two exemplary UMAPs of the full data set and methylation fractions of 100~kb windows (left) or shrunken residuals of variable methylation intervals (right) \textbf{(D)} Two exemplary UMAPs of the smaller data set and methylation fraction of 100 kb windows (left) or shrunken residuals of variable methylation intervals (right).
+Data from \citet{luo2017single}.}
+    \label{figure:score}
+    \end{center}
+\end{figure*}
+
 \section{The \texttt{scbs} Python package}
@@
 The source code and the package are available at \href{https://github.com/LKremer/scbs}{https://github.com/LKremer/scbs}, where we also provide detailed documentation including a tutorial that demonstrates a complete \texttt{scbs} analysis on an example data set.
-
-\section{Benchmarks}
-To assess performance of our methods, we obtained two previously published scBS-Seq data sets and re-analysed them with \texttt{scbs}.
-
-\subsection{Correlation with gene expression}
-
-The first data set is a single-cell multi-omics data set generated by \citet{argelaguet2019gastru}.
-In their work, the authors obtained single-cell methylomes of mouse embryo cells across different stages of development (E4.5--E7.5) as well as single-cell transcriptomes from the same cells.
-In one of their analyses, the authors calculated the correlation between gene expression and promoter methylation to elucidate the implications of the epigenome on gene expression in development.
-The authors identified several genes where promoter methylation predicts gene expression, but obtained low correlations for the vast majority of genes.
-We reasoned that VMRs obtained with the previously described sliding window approach might be a better predictor of gene expression.
-
-In Figure \ref{figure:correlation}A, we compare correlations of gene expression and promoter methylation to correlations of gene expression and methylation of the closest VMR identified with \texttt{scbs scan}.
-This analysis revealed a number of genes whose expression does not correlate with promoter methylation, but correlates with VMR methylation.
-This effect is also observed when using our proposed measure of methylation, the shrunken mean of residuals (Figure \ref{figure:correlation}B).
-
-\begin{figure}
-    \begin{center}
-    A\hspace{.9\columnwidth}~\\
-    \includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/corr_methylation_fraction_4kbwindow.pdf} \\
-    B\hspace{.9\columnwidth}~\\
-    \includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/corr_shrunken_residual_4kbwindow.pdf} \\
-    C\hspace{.9\columnwidth}~\\
-    \includegraphics[width=.95\columnwidth]{part_leonie_git/leonie_plots/comparison_meth_RNA.pdf}
-    \end{center}
-    \caption{\small \textbf{Correlation of DNA methylation and gene expression}.
-    \textbf{(A)} Correlation of each gene's RNA expression values to the the methylation fractions that were measured at promoters (y-axis), or to methylation fractions of the nearest VMR (x-axis).
-Promoters are defined as intervals \textpm2~kb around the TSS.
-    \textbf{(B)} Same as in (A), but using mean shrunken residuals as a measure of methylation.
-    \textbf{(C)} Mean smoothed methylation near the gene Ube2l6 (ENSMUSG00000027078, marked in yellow in A \& B).
-Cells are assigned to 4 groups based on Ube2l6 expression (group 0: cells with no Ube2l6 expression, groups 1-3: cells that express Ube2l6, divided into three equally large groups with group 3 having the highest expression).
-The mean methylation level of each group was smoothed with a tricube kernel of bandwidth 2000.
-    Data from \citet{argelaguet2019gastru}.}
-    \label{figure:correlation}
-\end{figure}
-
-To illustrate this effect with an example, we picked a gene where VMR methylation is a much better predictor of gene expression than promoter methylation (Ube2l6, highlighted in orange in Figure \ref{figure:correlation}A,B).
-Grouping all cells into four groups according to their level of Ube2l6 expression, it is clearly visible that the promoter of this gene is lowly methylated regardless of gene expression, while methylation at the nearest VMR is greatly decreased in cells with high Ube2l6 expression.
-
-\subsection{Identification of cellular sub-types}
-
-To assess the impact of our methods on dimensionality reduction and the ability to distinguish different (sub)types of cells, we used a data set from \citet{luo2017single}.
-This data set consists of 3377 single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal subtypes.
-We subjected this data set to the standard scBS workflow: defining a set of genomic intervals, quantifying methylation in these intervals, and subjecting the resulting matrix to PCA.
-We repeated this analysis several times, either using 100~kb tiles or VMRs, and either using methylation fractions or mean shrunken residuals as a measure of methylation.
-\todo{Is the following correct?} For both types of regions, the 100~kb tiles and the VMRs, we ranked the regions by the number of covering cells (i.e., cells with at least on read in the region) and keep only the top 1000 regions.
-
-To assess performance of each combination of methods, we used a nearest neighbor count score that quantifies how well the different neuronal subtypes are separated in 15 dimensional PC space (see Methods).
-The results demonstrate that the 16 neuronal subtypes are more cleanly separated when using our proposed methods (Figure \ref{figure:score}A).
-Most of this gain in performance can be attributed to the use of VMRs over genomic tiles. The use of mean shrunken residuals over methylation fractions provides an additional small improvement.
-
-Since some scBS protocols are based on the use of 96-well-plates, and since scBS is costly and labor-intensive, only few laboratories are currently able to obtain thousands of single cell methylomes.
-To simulate a smaller data set, we subsampled the frontal cortex data set to 500 cells and repeated our analysis.
-Here, our approach keeps its good score, while the simple approach falls behind (Figure \ref{figure:score}B).
-
-Of note, our results show that even few VMRs (here, the 1000 VMRs that have read coverage in the highest number of cells) already contain sufficient information to distinguish neuronal subtypes.
-In contrast, the neighbor count score for the top 1000 genomic tiles is much lower.
-To visualize the different degrees of separation, we further reduced the PC space from both analyses with UMAP, showing that some neuronal subtypes such as mL5-1 and mDL-3 are separated from the other subtypes when using 1000 VMRs (Figure \ref{figure:score}C).
-% why not just plot PC1 vs. PC2 here? % because people will want to see the UMAP. -S
-Again, the difference is even more pronounced in the sub-sampled analysis with only 500 cells (Figure \ref{figure:score}D).
-
-\begin{figure*}
-    \begin{center}
-    A\hspace{.4\textwidth}B\hspace*{.3\textwidth}~\\
-    \hspace{.3cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/complete_135k_12cm_log.pdf}
-    \hspace{1cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/cell500_20k_12cm_log.pdf}\\
-    C\hspace*{.7\textwidth}~\\
-    \hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_fulldataset.pdf}\\
-    D\hspace*{.7\textwidth}~\\
-    \hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_reduceddataset.pdf}
-    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of 16 neuronal subtypes}.
-\textbf{(A)} Nearest neighbor count score using four different method combinations and a variable number of genomic intervals (regions) as input to PCA.
-Exemplary UMAPs for encircled analyses are shown in (C).
-\textbf{(B)} Same benchmark as in (A), but for a smaller data set consisting of 500 randomly subsampled cells.
-Exemplary UMAPs for analyses are shown in (D).
-\textbf{(C)} Two exemplary UMAPs of the full data set and methylation fractions of 100~kb windows (left) or shrunken residuals of variable methylation intervals (right) \textbf{(D)} Two exemplary UMAPs of the smaller data set and methylation fraction of 100 kb windows (left) or shrunken residuals of variable methylation intervals (right).
-Data from \citet{luo2017single}.}
-    \label{figure:score}
-    \end{center}
-\end{figure*}
 

### 2022-06-13 paper.tex
add discussion

@@
 
-%Here we propose a number of new methods for the analysis of single cell methylation data and provide them to the research community as a command line tool called \texttt{scbs}.
-%Instead of dividing the genome into coarse-grained tiles, we propose to discover variably-methylated regions (VMRs) in the data itself, which improves performance of common downstream applications.
-%We also think that VMRs, due to their smaller size, are easier to interpret than  and furthermore provides a list of candidate regions.
-%These regions may then be linked to individual genes, tested for celltype-specific methylation differences, or subjected to gene ontology or motif enrichment, depending on the research question at hand.
-%We also propose a number of methods specifically tailored to the sparse nature of scBS data.
-%First, we observe
+Here we propose a number of new methods for the analysis of single cell methylation data and provide them to the research community as a command line tool.
+Instead of dividing the genome into coarse-grained tiles, we propose to discover variably-methylated regions (VMRs) in the data itself, which improves performance of common downstream applications.
+VMRs help interpreting the results of scBS experiments, since they represent a list of candidate regions that are potentially biologically meaningful.
+Due to their smaller size, individual VMRs can also be related to nearby or intersecting genomic features such as gene bodies or known regulatory elements.
+Furthermore, VMRs can be tested for methylation differences between groups of cells, or subjected to gene ontology enrichment or motif enrichment, depending on the research questions at hand.
 
-%- propose a number of new methods for the analysis of scBS data
-%- developed methods to analyse scBS data that go beyond simple genome tiling
-% - instead we discover potential genomic regions of interest on the data itself, which improves performance of common downstream applications and aids interpretability.
-% - VMRs can also be used as input for e.g. motif enrichment, go enrichment etc
-%- devised a method to store genome-wide scBS data efficiently
-%- propose a new measure of DNA methylation in a genomic interval (~ relative methylation)
+We also propose a number of methods to address issues caused by the sparse nature of scBS data.
+First, we observe that incomplete read coverage of a genomic interval can lead to inaccurate methylation estimates.
+To remedy this problem, we propose to use relative methylation values instead of absolute methylation fractions, and to employ shrinkage to decrease the impact of intervals with very low sequencing coverage.
+Second, we suggest a method to handle missing values in PCA in an iterative manner.
+Both of these methods help remedy the influence of differences in sequencing coverage on dimensionality reduction and clustering.
 
-%- implement them in Python and release them to the community as a Python package with a CLI
-%-> no more custom scripts and manual data wrangling of 1000 text files required, instead every user can quickly go from raw data to count matrix
-%- lowers the entry barrier for scBS data analysis and makes this new and exciting data type available to a greater audience
-
-%whats missing?
-%-
+Finally, we provide the above-mentioned methods to the scientific community in the form of a command line tool.
+Our software, called \texttt{scbs}, allows the user to obtain a cell \times region matrix from raw methylation files using only few, simple shell commands
+This means that users no longer need to write custom data-wrangling scripts, which ensures reproducibility across and makes the analysis of single-cell methylomes faster and easier.
 

### 2022-06-13 paper.tex
addec conclusion

--- 2022-06-13 paper.tex
@@
 
+Here, we have proposed an improved strategy to preprocess single-cell bisulfite-sequencing data.
+On the one hand, we suggest a scoring scheme that is aware of a reads local context rather than just treating all reads in an interval alike, and on the other hand, we show a way to pinpoint minimal regions of high variability across cells, which we call variably-methylated regions (VMRs).
+By reanalyzing two published datasets we could show that these improvements to data preprocessing help to increase signal and descrease noise, resulting in a more informative intermediate-dimensional representation of the data.
+As examples of practical benefits, we demonstrate that our preprocessing allows for better distinction of cell sub-types, especially for smaller data sets with limited numbers of cells.
+In addition to this improvement of global signal-to-noise ratio (with regards to cell dissimilarity quantification), we also show that the determination of VMRs improve biological interpretability by pinpointing the regions that actually change and hence have a putative regulatory role.
+
+Furthermore, we have presented and published an open source software tool, called \texttt{scbs}, that provides an easy-to-use implementation of the described algorithms.
+The package also offers functionality for all other steps of data preprocessing taht are needed to get from the output of methylation callers such as Bismark and methylpy to a data matrix that is suitable as input for standard analysis tools that work in a reduced-dimensionality setting, thus allowing the user to use, e.g., well established tools from sc-RNA-Seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses etc.
+By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.
+In addition, by pinpointing exact regions of change, the VMR functionality provides useful input for genomics-style analyses.
+
+In conclusion, we have presented powerful improvements to sc-BS-Seq data preprocessing and a software tool that implements these and also completes a gap in existing tool chains.
+
 %Here we propose a number of new methods for the analysis of single cell methylation data and provide them to the research community as a command line tool called \texttt{scbs}.

### 2022-06-13 paper.tex

--- 2022-06-13 paper.tex
@@
 
+\subsection{Simon's draft}
+
+Here, we have proposed an improved strategy to preprocess single-cell bisulfite-sequencing data.
+On the one hand, we suggest a scoring scheme that is aware of a reads local context rather than just treating all reads in an interval alike, and on the other hand, we show a way to pinpoint minimal regions of high variability across cells, which we call variably-methylated regions (VMRs).
+By reanalyzing two published datasets we could show that these improvements to data preprocessing help to increase signal and descrease noise, resulting in a more informative intermediate-dimensional representation of the data.
+As examples of practical benefits, we demonstrate that our preprocessing allows for better distinction of cell sub-types, especially for smaller data sets with limited numbers of cells.
+In addition to this improvement of global signal-to-noise ratio (with regards to cell dissimilarity quantification), we also show that the determination of VMRs improve biological interpretability by pinpointing the regions that actually change and hence have a putative regulatory role.
+
+Furthermore, we have presented and published an open source software tool, called \texttt{scbs}, that provides an easy-to-use implementation of the described algorithms.
+The package also offers functionality for all other steps of data preprocessing taht are needed to get from the output of methylation callers such as Bismark and methylpy to a data matrix that is suitable as input for standard analysis tools that work in a reduced-dimensionality setting, thus allowing the user to use, e.g., well established tools from sc-RNA-Seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses etc.
+By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.
+In addition, by pinpointing exact regions of change, the VMR functionality provides useful input for genomics-style analyses.
+
+In conclusion, we have presented powerful improvements to sc-BS-Seq data preprocessing and a software tool that implements these and also completes a gap in existing tool chains.
+
+\subsection{Lukas's draft}
+
 Here we propose a number of new methods for the analysis of single cell methylation data and provide them to the research community as a command line tool.

### 2022-06-13 paper.tex
added imputation

@@
 Finally, we provide the above-mentioned methods to the scientific community in the form of a command line tool.
-Our software, called \texttt{scbs}, allows the user to obtain a cell \times region matrix from raw methylation files using only few, simple shell commands
+Our software, called \texttt{scbs}, allows the user to obtain a cell $\times$ region matrix from raw methylation files using only few, simple shell commands
 This means that users no longer need to write custom data-wrangling scripts, which ensures reproducibility across and makes the analysis of single-cell methylomes faster and easier.
@@
 \subsection{PCA with imputation}
-[...]
+
+Whenever a region is not covered by any read in a cell, the corresponding data entry in the input data matrix for the PCA will be missing.
+The standard approach to calculate PCA (commonly done using the IRLBA algorthm, \citep{Baglama2005}) is not suited to deal with missing data.
+We circumvent this issue by simply using the PCA itself to impute the missing value.
+
+Let us write $A$ for the matrix to which the PCA is to be applied.
+In contrast to the main text, and in order to follow statistics rather than bioinformatcis convention, we consider here the features (here: regions) to be represented by the matrix rows rather than the columns.
+The matrix has already been centered, i.e., $\sum_i a_{ij}=0$.
+To establish notation, we remind the reader that performing a PCA on A means finding the singular value decomposition (SVD), $A=UDR^\top$, with $D$ diagonal and $U$ and $R$ orthonormal.
+The PCA scores are contained in $X=UD$, the loadings in $R$.
+To reconstruct the input data $A$ from the PCA representation, one may use $A=XR^\top$, i.e., $a_{ij}=\sum_r x_{ir} r_{jr}$, where the equation is exact if $r$ runs over all principal components and approximate if it is truncated to the leading ones.
+
+Our iterative imputation strategy is now simply the following: we first replace all missing values in the row-centered input matrix $A$ with zeroes and perform the (truncated) PCA.
+Then, we use the PCA predictions for the missing values, i.e., the $a_{ij}=\sum_r x_{ir} r_{jr}$, as refined stand-ins for the missing values in $A$ and run the PCA once more.
+This can now be iterated utnil convergence, which is typically achieved after at most 4--5 rounds of PCA.
 

### 2022-06-13 paper.tex
added todo

@@
 The starting point in most scRNA-Seq analyses is a matrix of UMI counts (i.e.
-of counts of distinct RNA molecules), one row for each cell and one column for each gene.
+of counts of distinct RNA molecules), one row \todo{Which way round are our matrices?} for each cell and one column for each gene.
 A first goal is usually to assigning cell types or states to cells.
@@
 
-Let us write $A$ for the matrix to which the PCA is to be applied.
-In contrast to the main text, and in order to follow statistics rather than bioinformatcis convention, we consider here the features (here: regions) to be represented by the matrix rows rather than the columns.
+Let us write $A$ for the matrix to which the PCA is to be applied, with the features (here: regions) represented by the matrix rows.
 The matrix has already been centered, i.e., $\sum_i a_{ij}=0$.
@@
 Then, we use the PCA predictions for the missing values, i.e., the $a_{ij}=\sum_r x_{ir} r_{jr}$, as refined stand-ins for the missing values in $A$ and run the PCA once more.
-This can now be iterated utnil convergence, which is typically achieved after at most 4--5 rounds of PCA.
+This can now be iterated until convergence, which is typically achieved after at most 4--5 rounds of PCA.
 

### 2022-06-13 paper.tex
rewrote definition of nnc

@@
 To re-analyze scBS data sets by \citet{luo2017single} and \citet{argelaguet2019gastru}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb bins, or by acquiring VMRs with \texttt{scbs scan} using default options.
-\todo{Mention choice of top 1000 regions.} For all intervals (VMRs and tiles), the mean methylation fractions and mean shrunken residuals were obtained with \texttt{scbs matrix}.
+For all intervals (VMRs and tiles), the mean methylation fractions and mean shrunken residuals were obtained with \texttt{scbs matrix}.
 Intervals containing less than four measured CpGs, or that were measured in less than 10\% of cells, were discarded.
@@
 
-\todo{Improve wording} The nearest neighbor count score $nnc(k)$ is based on the $\Gamma$-score \citep{Kireeva_2014}.
-Each cell $c_j^i$ belongs to a cell type $j$.
-For each cell, $k$ nearest neighbors $nn_j^k$ were computed based on their Euclidean distance.
-Based on the cell type $j$ of the $k$ nearest neighbors of each cell, the nnc score $nnc(k)$ was computed with formula (\todo{ref}), in which $d(c_j^i, k_j^i)=1$ if at least 90\% of neighbor classes match the class of cell $c_j^i$ and $d(c_j^i, k_j^i)=0$ if less than 90\% of classes match, while $n$ are the total number of cells.
-\begin{equation}
-    \text{nnc}(k) = \frac{1}{n}d(c_j^i, k_j^i)
-\end{equation}
+Our nearest-neighbor-count score $\text{nnc}_k$, is based on the $\Gamma$-score \citep{Kireeva_2014} and is calculated as follows.
+For each cell $j$, we count how many of its $k$ nearest neighbors have been assigned to the same cluster as cell $j$.
+We denote this count $a^k_j$.
+We consider a cell as ``well-assigned'' if $a^k_j>\theta k$, where we have chosen $\theta=0.9$.
+In other words, a cell is ``well-assigned'' if at least 90\% of its $k$ nearest neighbors have been assigned the same class as the cell under consideration.
+
+The score $\text{nnc}_k$ is then simply the fraction of cells in the data set that are well-assigned in this sense.
 

### 2022-06-13 paper.tex
draft for abstract

@@
 \blfootnote{\hspace{-.5cm}\raggedright\textsuperscript{*} e-mail:  \texttt{l.kremer@dkfz-heidelberg.de} and\\ \hspace{1cm}\texttt{simon.anders@bioquant.uni-heidelberg.de}}
+
 \section{Abstract}
 
-[...]
+Single-cell-bisulfite sequencing (scBS) allows to asses DNA methylation with single-basepair and single-cell resolution.
+Analysing the big data sets thus obtained requires preprocessing to reduce data size, improve signal over noise, and provide interpretability.
+Currently, this is simply done by tiling teh genome into large windows and averaging in these.
+We show that this coarse-graining dilutes signal, and suggest improved approaches to determine informative regions to quantify for and to quantitate methylation in these windows.
+We demonstrate how our preprocessing approach helps to better distinguish cell types, reduces demands on cell number, and show that our method of finding variably-methylated regions finds biologically meaningful loci.
+Finally, we present a software tool, \texttt{scbs}, that implements this approach and offers further functionality requried to work with scBS data.
 

### 2022-06-13 paper.tex
merge discussions

--- 2022-06-13 paper.tex
@@
 
-\subsection{Simon's draft}
-
 Here, we have proposed an improved strategy to preprocess single-cell bisulfite-sequencing data.
-On the one hand, we suggest a scoring scheme that is aware of a reads local context rather than just treating all reads in an interval alike, and on the other hand, we show a way to pinpoint minimal regions of high variability across cells, which we call variably-methylated regions (VMRs).
-By reanalyzing two published datasets we could show that these improvements to data preprocessing help to increase signal and descrease noise, resulting in a more informative intermediate-dimensional representation of the data.
+Based on the observation that incomplete read coverage of a genomic interval can lead to inaccurate methylation estimates, we suggest a scoring scheme that is aware of a reads local context rather than just treating all reads in an interval alike
+Furthermore, we show a way to pinpoint minimal regions of high variability across cells, which we call variably-methylated regions (VMRs).
+By reanalyzing two published datasets we could show that these improvements to data preprocessing help to increase signal and decrease noise, resulting in a more informative intermediate-dimensional representation of the data.
 As examples of practical benefits, we demonstrate that our preprocessing allows for better distinction of cell sub-types, especially for smaller data sets with limited numbers of cells.
@@
 Furthermore, we have presented and published an open source software tool, called \texttt{scbs}, that provides an easy-to-use implementation of the described algorithms.
-The package also offers functionality for all other steps of data preprocessing taht are needed to get from the output of methylation callers such as Bismark and methylpy to a data matrix that is suitable as input for standard analysis tools that work in a reduced-dimensionality setting, thus allowing the user to use, e.g., well established tools from sc-RNA-Seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses etc.
+The package also offers functionality for all other steps of data preprocessing that are needed to get from the output of methylation callers such as Bismark and methylpy to a cell\texttimes region matrix that is suitable as input for standard analysis tools that work in a reduced-dimensionality setting, thus allowing the user to use, e.g., well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses etc.
 By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.
-In addition, by pinpointing exact regions of change, the VMR functionality provides useful input for genomics-style analyses.
 
-In conclusion, we have presented powerful improvements to sc-BS-Seq data preprocessing and a software tool that implements these and also completes a gap in existing tool chains.
+In addition, by pinpointing exact regions of interest, the VMR functionality provides useful input for genomics-style analyses.
+Depending on the research question at hand, individual VMRs can be related to nearby genomic features such as gene bodies or known regulatory elements, tested for methylation differences between groups of cells, or subjected to gene ontology enrichment or motif enrichment.
 
-\subsection{Lukas's draft}
-
-Here we propose a number of new methods for the analysis of single cell methylation data and provide them to the research community as a command line tool.
-Instead of dividing the genome into coarse-grained tiles, we propose to discover variably-methylated regions (VMRs) in the data itself, which improves performance of common downstream applications.
-VMRs help interpreting the results of scBS experiments, since they represent a list of candidate regions that are potentially biologically meaningful.
-Due to their smaller size, individual VMRs can also be related to nearby or intersecting genomic features such as gene bodies or known regulatory elements.
-Furthermore, VMRs can be tested for methylation differences between groups of cells, or subjected to gene ontology enrichment or motif enrichment, depending on the research questions at hand.
-
-We also propose a number of methods to address issues caused by the sparse nature of scBS data.
-First, we observe that incomplete read coverage of a genomic interval can lead to inaccurate methylation estimates.
-To remedy this problem, we propose to use relative methylation values instead of absolute methylation fractions, and to employ shrinkage to decrease the impact of intervals with very low sequencing coverage.
-Second, we suggest a method to handle missing values in PCA in an iterative manner.
-Both of these methods help remedy the influence of differences in sequencing coverage on dimensionality reduction and clustering.
-
-Finally, we provide the above-mentioned methods to the scientific community in the form of a command line tool.
-Our software, called \texttt{scbs}, allows the user to obtain a cell $\times$ region matrix from raw methylation files using only few, simple shell commands
-This means that users no longer need to write custom data-wrangling scripts, which ensures reproducibility across and makes the analysis of single-cell methylomes faster and easier.
+In conclusion, we have presented powerful improvements to scBS data preprocessing and a software tool that implements these and also completes a gap in existing tool chains.
 

### 2022-06-13 paper.tex
removed todos

--- 2022-06-13 paper.tex
@@
 Lukas P.~M.
-Kremer\textsuperscript{1,2,*}, Leonie Küchenhoff\textsuperscript{1,2}, Santiago Cerrizuela\textsuperscript{2}, \todo{...}, Ana Martin-Villalba\textsuperscript{2}, Simon Anders\textsuperscript{1,*}\\[1ex]
+Kremer\textsuperscript{1,2,*}, Leonie Küchenhoff\textsuperscript{1,2}, Santiago Cerrizuela\textsuperscript{2}, Ana Martin-Villalba\textsuperscript{2}, Simon Anders\textsuperscript{1,*}\\[1ex]
 {\footnotesize \textsuperscript{1} BioQuant Centre, University of Heidelberg, Heidelberg, Germany\\[-.5ex]
@@
 We repeated this analysis several times, either using 100~kb tiles or VMRs, and either using methylation fractions or mean shrunken residuals as a measure of methylation.
-\todo{Is the following correct?} For both types of regions, the 100~kb tiles and the VMRs, we ranked the regions by the number of covering cells (i.e., cells with at least on read in the region) and kept only the top 100, 1000, or 10,000 regions.
+For both types of regions, the 100~kb tiles and the VMRs, we ranked the regions by the number of covering cells (i.e., cells with at least on read in the region) and kept only the top 100, 1000, or 10,000 regions.
 

### 2022-06-13 paper.tex
try to always use the same word for 100kb windows

--- 2022-06-13 paper.tex
@@
 Analysing the big data sets thus obtained requires preprocessing to reduce data size, improve signal over noise, and provide interpretability.
-Currently, this is simply done by tiling teh genome into large windows and averaging in these.
+Currently, this is simply done by tiling the genome into large windows and averaging in these.
 We show that this coarse-graining dilutes signal, and suggest improved approaches to determine informative regions to quantify for and to quantitate methylation in these windows.
@@
 Besides providing better input for the distance calculations, the small size of identified VMRs also means that they can be readily compared with genomic annotation, which aids in their functional interpretation.
-While a large 100~kb bin will often contain multiple genes with distinct expression patterns, small VMRs located in known regulatory elements or gene bodies are more easily associated with individual genes and their associated function.
+While a large 100~kb tile will often contain multiple genes with distinct expression patterns, small VMRs located in known regulatory elements or gene bodies are more easily associated with individual genes and their associated function.
 Thus, VMRs provide a useful starting point for examining the data with methods from functional genomics.
@@
 We subjected this data set to the standard scBS workflow: defining a set of genomic intervals, quantifying methylation in these intervals, and subjecting the resulting matrix to PCA.
-We repeated this analysis several times, either using 100~kb tiles or VMRs, and either using methylation fractions or mean shrunken residuals as a measure of methylation.
-\todo{Is the following correct?} For both types of regions, the 100~kb tiles and the VMRs, we ranked the regions by the number of covering cells (i.e., cells with at least on read in the region) and kept only the top 100, 1000, or 10,000 regions.
+We repeated this analysis several times, either using 100~kb windows or VMRs, and either using methylation fractions or mean shrunken residuals as a measure of methylation.
+\todo{Is the following correct?} For both types of regions, the 100~kb windows and the VMRs, we ranked the regions by the number of covering cells (i.e., cells with at least on read in the region) and kept only the top 100, 1000, or 10,000 regions.
 
@@
 Of note, our results show that even few VMRs (here, the 1000 VMRs that have read coverage in the highest number of cells) already contain sufficient information to distinguish neuronal subtypes.
-In contrast, the neighbor count score for the top 1000 genomic tiles is much lower.
+In contrast, the neighbor count score for the top 1000 genomic 100~kb windows is much lower.
 To visualize the different degrees of separation, we further reduced the PC space from both analyses with UMAP, showing that some neuronal subtypes such as mL5-1 and mDL-3 are separated from the other subtypes when using 1000 VMRs (Figure \ref{figure:score}C).
@@
 Exemplary UMAPs for analyses are shown in (D).
-\textbf{(C)} Two exemplary UMAPs of the full data set and methylation fractions of 100~kb windows (left) or shrunken residuals of variable methylation intervals (right) \textbf{(D)} Two exemplary UMAPs of the smaller data set and methylation fraction of 100 kb windows (left) or shrunken residuals of variable methylation intervals (right).
+\textbf{(C)} Two exemplary UMAPs of the full data set and methylation fractions of 100~kb windows (left) or shrunken residuals of variable methylation intervals (right) \textbf{(D)} Two exemplary UMAPs of the smaller data set and methylation fraction of 100~kb windows (left) or shrunken residuals of variable methylation intervals (right).
 Data from \citet{luo2017single}.}
@@
 \subsection{Re-analysis of published scBS datasets}
-To re-analyze scBS data sets by \citet{luo2017single} and \citet{argelaguet2019gastru}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb bins, or by acquiring VMRs with \texttt{scbs scan} using default options.
-For all intervals (VMRs and tiles), the mean methylation fractions and mean shrunken residuals were obtained with \texttt{scbs matrix}.
-Intervals containing less than four measured CpGs, or that were measured in less than 10\% of cells, were discarded.
+To re-analyze scBS data sets by \citet{luo2017single} and \citet{argelaguet2019gastru}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb windows, or by acquiring VMRs with \texttt{scbs scan} using default options.
+For all regions (VMRs and 100~kb windows), the mean methylation fractions and mean shrunken residuals were obtained with \texttt{scbs matrix}.
+Regions containing less than four measured CpGs, or that were measured in less than 10\% of cells, were discarded.
 To select $n$ intervals for further analysis, the top $n$ intervals were picked based on the number of cells that the region was measured in.

### 2022-06-13 paper.tex
clarify caption

--- 2022-06-13 paper.tex
@@
     \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of 16 neuronal subtypes}.
-\textbf{(A)} Nearest neighbor count score using four different method combinations and a variable number of genomic intervals (regions) as input to PCA.
+\textbf{(A)} Nearest neighbor count score using four different method combinations and a variable number of genomic regions (VMRs or 100~kb windows) as input to PCA.
 Exemplary UMAPs for encircled analyses are shown in (C).
@@
 Exemplary UMAPs for analyses are shown in (D).
-\textbf{(C)} Two exemplary UMAPs of the full data set and methylation fractions of 100~kb windows (left) or shrunken residuals of variable methylation intervals (right) \textbf{(D)} Two exemplary UMAPs of the smaller data set and methylation fraction of 100~kb windows (left) or shrunken residuals of variable methylation intervals (right).
+\textbf{(C)} Two exemplary UMAPs of the full data set and methylation fractions of 100~kb windows (left) or mean shrunken residuals of VMRs (right) \textbf{(D)} Two exemplary UMAPs of the smaller data set and methylation fraction of 100~kb windows (left) or mean shrunken residuals of VMRs (right).
 Data from \citet{luo2017single}.}

### 2022-06-13 paper.tex
fmt abstract
moved figrues

@@
 
-\usepackage[a4paper,hmargin=1.5cm,vmargin=2.5cm,]{geometry}
+\usepackage[a4paper,hmargin=1.5cm,vmargin=2cm]{geometry}
 \setlength{\columnsep}{0.7cm}
@@
 }
+\renewcommand{\topfraction}{0.9}
 \renewcommand{\dbltopfraction}{0.9}
-\renewcommand{\textfraction}{0.15}
+\renewcommand{\textfraction}{0.1}
 
@@
 June 2022
-\vspace{6ex}
-}]
-\blfootnote{\hspace{-.5cm}\raggedright\textsuperscript{*} e-mail:  \texttt{l.kremer@dkfz-heidelberg.de} and\\ \hspace{1cm}\texttt{simon.anders@bioquant.uni-heidelberg.de}}
+\vspace{5.5ex}
 
-\section{Abstract}
-
-Single-cell-bisulfite sequencing (scBS) allows to asses DNA methylation with single-basepair and single-cell resolution.
+\begin{minipage}{0.7\textwidth}
+\small
+\textit{Abstract} -- Single-cell-bisulfite sequencing (scBS) allows to asses DNA methylation with single-basepair and single-cell resolution.
 Analysing the big data sets thus obtained requires preprocessing to reduce data size, improve signal over noise, and provide interpretability.
@@
 Finally, we present a software tool, \texttt{scbs}, that implements this approach and offers further functionality requried to work with scBS data.
+\end{minipage}
+\vspace{6.5ex}
+
+}]
 
@@
 Single-cell RNA sequencing (sc-RNA-seq) techniques have matured at great pace in recent years, with well developed analysis methodology, and methods to study epigenetics at single-cell resolution are rapidly catching up.
+\blfootnote{\hspace{-.5cm}\raggedright\textsuperscript{*} e-mail:  \texttt{l.kremer@dkfz-heidelberg.de} and\\ \hspace{1cm}\texttt{simon.anders@bioquant.uni-heidelberg.de}}
 
@@
 
-\begin{figure*}
+\begin{figure*}[t]
     \hspace{2.5cm}A\\[-7ex]
@@
 
+\begin{figure*}
+	\begin{center}
+		A\hspace{.4\textwidth}B\hspace*{.3\textwidth}~\\
+		\hspace{.3cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/complete_135k_12cm_log.pdf}
+		\hspace{1cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/cell500_20k_12cm_log.pdf}\\
+		C\hspace*{.7\textwidth}~\\
+		\hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_fulldataset.pdf}\\
+		D\hspace*{.7\textwidth}~\\
+		\hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_reduceddataset.pdf}
+		\caption{\small \textbf{Benchmark of our methods on single-cell methylomes of 16 neuronal subtypes}.
+\textbf{(A)} Nearest neighbor count score using four different method combinations and a variable number of genomic regions (VMRs or 100~kb windows) as input to PCA.
+Exemplary UMAPs for encircled analyses are shown in (C).
+\textbf{(B)} Same benchmark as in (A), but for a smaller data set consisting of 500 randomly subsampled cells.
+Exemplary UMAPs for analyses are shown in (D).
+\textbf{(C)} Two exemplary UMAPs of the full data set and methylation fractions of 100~kb windows (left) or mean shrunken residuals of VMRs (right) \textbf{(D)} Two exemplary UMAPs of the smaller data set and methylation fraction of 100~kb windows (left) or mean shrunken residuals of VMRs (right).
+Data from \citet{luo2017single}.}
+		\label{figure:score}
+	\end{center}
+\end{figure*}
+
 To assess the impact of our methods on dimensionality reduction and the ability to distinguish different (sub)types of cells, we used a data set from \citet{luo2017single}.
@@
 
-\begin{figure*}
-    \begin{center}
-    A\hspace{.4\textwidth}B\hspace*{.3\textwidth}~\\
-    \hspace{.3cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/complete_135k_12cm_log.pdf}
-    \hspace{1cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/cell500_20k_12cm_log.pdf}\\
-    C\hspace*{.7\textwidth}~\\
-    \hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_fulldataset.pdf}\\
-    D\hspace*{.7\textwidth}~\\
-    \hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_reduceddataset.pdf}
-    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of 16 neuronal subtypes}.
-\textbf{(A)} Nearest neighbor count score using four different method combinations and a variable number of genomic regions (VMRs or 100~kb windows) as input to PCA.
-Exemplary UMAPs for encircled analyses are shown in (C).
-\textbf{(B)} Same benchmark as in (A), but for a smaller data set consisting of 500 randomly subsampled cells.
-Exemplary UMAPs for analyses are shown in (D).
-\textbf{(C)} Two exemplary UMAPs of the full data set and methylation fractions of 100~kb windows (left) or mean shrunken residuals of VMRs (right) \textbf{(D)} Two exemplary UMAPs of the smaller data set and methylation fraction of 100~kb windows (left) or mean shrunken residuals of VMRs (right).
-Data from \citet{luo2017single}.}
-    \label{figure:score}
-    \end{center}
-\end{figure*}
-
 \section{The \texttt{scbs} Python package}
 
-\begin{figure*}
+\begin{figure*}[t]
 	\begin{center}

### 2022-06-13 paper.tex
added hrules
reduced title spacing
add #of region explanation

@@
 We repeated this analysis several times, either using 100~kb windows or VMRs, and either using methylation fractions or mean shrunken residuals as a measure of methylation. 
-\todo{Is the following correct?} For both types of regions, the 100~kb windows and the VMRs, we ranked the regions by the number of covering cells (i.e., cells with at least on read in the region) and kept only the top 100, 1000, or 10,000 regions.
+For both types of regions, we varied the number of input regions, ranging from 100 to 50,000 high-coverage regions that were selected for PCA.
+Specifically, we ranked regions by their number of coverging cells (i.e., cells with at least on read in the region) and selected those with the highest number.
 
@@
 In conclusion, we have presented powerful improvements to scBS data preprocessing and a software tool that implements these and also completes a gap in existing tool chains.
+
+\vspace{1.6ex}
+\noindent\hfil\rule{.6\columnwidth}{.2pt}\hfil
 
@@
 
+\vspace{1.6ex}
+\noindent\hfil\rule{.6\columnwidth}{.2pt}\hfil
+
 {\small \bibliography{scbs}}

### 2022-06-13 paper.tex
sc-BS-Seq -> scBS

@@
 \renewcommand{\textfraction}{0.1}
+
+\usepackage[compact]{titlesec}
 
@@
 
-Here, we discuss bisulfite sequencing (sc-BS-Seq), a method to study DNA methylation.
+Here, we discuss bisulfite sequencing, a method to study DNA methylation.
 In brief, DNA is treated with bisulfite, which converts unmethylated cytosines to uracils which are read as thymine in subsequent PCR, while methylated cytosines are protected from conversion.
@@
 
-In the present paper, we discuss strategies to analyze sc-BS-Seq data, suggest improvement to current approaches, show their value in a benchmark, and present software to perform the improved analysis methods.
+In the present paper, we discuss strategies to analyze scBS data, suggest improvement to current approaches, show their value in a benchmark, and present software to perform the improved analysis methods.
 
@@
 
-To demonstrate the value of our proposed improvements, we obtained two previously published sc-BS-Seq data sets and re-analysed them.
+To demonstrate the value of our proposed improvements, we obtained two previously published scBS data sets and re-analysed them.
 
@@
 
-Since sc-BS-Seq protocols is still rather costly and labor-intensive, only few laboratories are currently able to obtain thousands of single cell methylomes.
+Since scBS protocols is still rather costly and labor-intensive, only few laboratories are currently able to obtain thousands of single cell methylomes.
 To simulate a smaller data set, we subsampled the frontal cortex data set to 500 cells and repeated our analysis.
@@
 Methylated and unmethylated sites are coded as $+1$ and $-1$, respectively.
-Since the vast majority of values in this matrix are missing due to the sparsity of sc-BS-Seq data (and because rows for base-pairs not corresponding to a CpG site contain no data), we encode missing values as zero and then store the data in Compressed Sparse Row (CSR) format.
+Since the vast majority of values in this matrix are missing due to the sparsity of scBS data (and because rows for base-pairs not corresponding to a CpG site contain no data), we encode missing values as zero and then store the data in Compressed Sparse Row (CSR) format.
 This format does not explicitly store zeroes (here: missing values) and is optimized for row-wise access, which results in significant compression and allows fast access to the methylation status of genomic intervals.
@@
 
-\vspace{1.6ex}
+\vspace{1.4ex}
 \noindent\hfil\rule{.6\columnwidth}{.2pt}\hfil

### 2022-06-13 paper.tex
sc-RNA-Seq -> scRNA-seq

@@
 Sequencing-based assays with single-cell resolution have offered new means to understand the differences between the cells making up a sample.
-Single-cell RNA sequencing (sc-RNA-seq) techniques have matured at great pace in recent years, with well developed analysis methodology, and methods to study epigenetics at single-cell resolution are rapidly catching up.
+Single-cell RNA sequencing (scRNA-seq) techniques have matured at great pace in recent years, with well developed analysis methodology, and methods to study epigenetics at single-cell resolution are rapidly catching up.
 \blfootnote{\hspace{-.5cm}\raggedright\textsuperscript{*} e-mail:  \texttt{l.kremer@dkfz-heidelberg.de} and\\ \hspace{1cm}\texttt{simon.anders@bioquant.uni-heidelberg.de}}
@@
 After sequencing, these conversions allow to determine the methylation status of all cytosines covered by reads \citep{Frommer_1992}.
-Bisulfite sequencing can also be performed with single-cell resolution \citep{Smallwood_2014} and even in parallel with sc-RNA-Seq \citep{Clark2018}.
+Bisulfite sequencing can also be performed with single-cell resolution \citep{Smallwood_2014} and even in parallel with scRNA-seq \citep{Clark2018}.
 
@@
 This matrix is now subjected to PCA.
-After the PCA, one can proceed using approaches known from sc-RNA-Seq.
+After the PCA, one can proceed using approaches known from scRNA-seq.
 

### 2022-06-13 paper.tex
added a remark on where the cell types come from

--- 2022-06-13 paper.tex
@@
 We repeated this analysis several times, either using 100~kb windows or VMRs, and either using methylation fractions or mean shrunken residuals as a measure of methylation.
-For both types of regions, we varied the number of input regions, ranging from 100 to 50,000 high-coverage regions that were selected for PCA.
-Specifically, we ranked regions by their number of coverging cells (i.e., cells with at least on read in the region) and selected those with the highest number.
+\todo{Is the following correct?} For both types of regions, the 100~kb windows and the VMRs, we ranked the regions by the number of covering cells (i.e., cells with at least on read in the region) and kept only the top 100, 1000, or 10,000 regions.
 
-To assess performance of each combination of methods, we used a nearest neighbor count score that quantifies how well the different neuronal subtypes are separated in 15 dimensional PC space (see Methods).
+To assess performance of each combination of methods, we used a nearest neighbor count score (see Methods) that quantifies how well the different neuronal subtypes are separated in 15 dimensional PC space.
+We used the original assignment of cells to cell types from \citet{luo2017single}.
 The results demonstrate that the 16 neuronal subtypes are more cleanly separated when using our proposed methods (Figure \ref{figure:score}A).

### 2022-06-13 paper.tex

--- 2022-06-13 paper.tex
@@
 
-To assess performance of each combination of methods, we used a nearest neighbor count score that quantifies how well the different neuronal subtypes are separated in 15 dimensional PC space (see Methods).
+To assess performance of each combination of methods, we used a nearest neighbor count score (see Methods) that quantifies how well the different neuronal subtypes are separated in 15 dimensional PC space.
+We used the original assignment of cells to cell types from \citet{luo2017single}.
 The results demonstrate that the 16 neuronal subtypes are more cleanly separated when using our proposed methods (Figure \ref{figure:score}A).

### 2022-06-13 paper.tex
100kb window -> 100kb tile

--- 2022-06-13 paper.tex
@@
 Analysing the big data sets thus obtained requires preprocessing to reduce data size, improve signal over noise, and provide interpretability.
-Currently, this is simply done by tiling the genome into large windows and averaging in these.
-We show that this coarse-graining dilutes signal, and suggest improved approaches to determine informative regions to quantify for and to quantitate methylation in these windows.
+Currently, this is simply done by dividing the genome into large tiles and averaging in these.
+We show that this coarse-graining dilutes signal, and suggest improved approaches to determine informative regions to quantify for and to quantitate methylation in these tiles.
 We demonstrate how our preprocessing approach helps to better distinguish cell types, reduces demands on cell number, and show that our method of finding variably-methylated regions finds biologically meaningful loci.
@@
 The task is therefore how to construct a suitable matrix as input for the PCA.
-A simple and common approach, used for instance by \citet{luo2017single}, is to tile the genome into windows of, say, 100~kb size, and calculate for each cell the average methylation of the DNA within each window.
-To this end, we identify all CpG sites in the window for which we have coverage with at least one read and can hence call the CpG to be either unmethylated or methylated in the cell.
-Then we denote as average DNA methylation of this window in a given cell the proportion of observed CpG sites in the window that were found to be methylated (Figure \ref{fig:smoothres}A).
-This yields a matrix, with one row for each cell and one column for each window from the genome tiling, comprising numbers (methylation fractions) between 0 and 1.
+A simple and common approach, used for instance by \citet{luo2017single}, is to divide the genome into tiles of, say, 100~kb size, and calculate for each cell the average methylation of the DNA within each tile.
+To this end, we identify all CpG sites in the tile for which we have coverage with at least one read and can hence call the CpG to be either unmethylated or methylated in the cell.
+Then we denote as average DNA methylation of this tile in a given cell the proportion of observed CpG sites in the tile that were found to be methylated (Figure \ref{fig:smoothres}A).
+This yields a matrix, with one row for each cell and one column for each genomic tile, comprising numbers (methylation fractions) between 0 and 1.
 This matrix is now subjected to PCA.
@@
 The standard PCA algorithm cannot deal with missing data.
-Therefore, one would need to make intervals wide enough to ensure that every interval is covered by at least one read for most windows and discard any interval that is not covered by any single cell.
+Therefore, one would need to make intervals wide enough to ensure that every interval is covered by at least one read for most tiles and discard any interval that is not covered by any single cell.
 Hence, a naive analysis has to resort to a coarse tiling of the genome.
@@
 
-In the standard approach, one divides up (tiles) each chromosome into non-overlapping, equal-sized intervals, and quantitates the methylation of each such window.
+In the standard approach, one divides up (tiles) each chromosome into non-overlapping, equal-sized intervals, and quantitates the methylation of each such tiles.
 Such rigid placing of interval boundaries is unlikely to be optimal: for example, a VMR might be much smaller than a tile and its signal will hence be drowned out by the larger part of uninformative CpGs that are equal in all cells, when averaging over all the CpG sites in the tile.
@@
 		\caption{\small \textbf{Benchmark of our methods on single-cell methylomes of 16 neuronal subtypes}.
-\textbf{(A)} Nearest neighbor count score using four different method combinations and a variable number of genomic regions (VMRs or 100~kb windows) as input to PCA.
+\textbf{(A)} Nearest neighbor count score using four different method combinations and a variable number of genomic regions (VMRs or 100~kb tiles) as input to PCA.
 Exemplary UMAPs for encircled analyses are shown in (C).
@@
 Exemplary UMAPs for analyses are shown in (D).
-\textbf{(C)} Two exemplary UMAPs of the full data set and methylation fractions of 100~kb windows (left) or mean shrunken residuals of VMRs (right) \textbf{(D)} Two exemplary UMAPs of the smaller data set and methylation fraction of 100~kb windows (left) or mean shrunken residuals of VMRs (right).
+\textbf{(C)} Two exemplary UMAPs of the full data set and methylation fractions of 100~kb tiles (left) or mean shrunken residuals of VMRs (right) \textbf{(D)} Two exemplary UMAPs of the smaller data set and methylation fraction of 100~kb tiles (left) or mean shrunken residuals of VMRs (right).
 Data from \citet{luo2017single}.}
@@
 We subjected this data set to the standard scBS workflow: defining a set of genomic intervals, quantifying methylation in these intervals, and subjecting the resulting matrix to PCA.
-We repeated this analysis several times, either using 100~kb windows or VMRs, and either using methylation fractions or mean shrunken residuals as a measure of methylation. 
+We repeated this analysis several times, either using 100~kb tiles or VMRs, and either using methylation fractions or mean shrunken residuals as a measure of methylation. 
 For both types of regions, we varied the number of input regions, ranging from 100 to 50,000 high-coverage regions that were selected for PCA.
@@
 Of note, our results show that even few VMRs (here, the 1000 VMRs that have read coverage in the highest number of cells) already contain sufficient information to distinguish neuronal subtypes.
-In contrast, the neighbor count score for the top 1000 genomic 100~kb windows is much lower.
+In contrast, the neighbor count score for the top 1000 genomic 100~kb tiles is much lower.
 To visualize the different degrees of separation, we further reduced the PC space from both analyses with UMAP, showing that some neuronal subtypes such as mL5-1 and mDL-3 are separated from the other subtypes when using 1000 VMRs (Figure \ref{figure:score}C).
@@
 \subsection{Re-analysis of published scBS datasets}
-To re-analyze scBS data sets by \citet{luo2017single} and \citet{argelaguet2019gastru}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb windows, or by acquiring VMRs with \texttt{scbs scan} using default options.
-For all regions (VMRs and 100~kb windows), the mean methylation fractions and mean shrunken residuals were obtained with \texttt{scbs matrix}.
+To re-analyze scBS data sets by \citet{luo2017single} and \citet{argelaguet2019gastru}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using default options.
+For all regions (VMRs and 100~kb tiles), the mean methylation fractions and mean shrunken residuals were obtained with \texttt{scbs matrix}.
 Regions containing less than four measured CpGs, or that were measured in less than 10\% of cells, were discarded.

### 2022-06-13 paper.tex
club/widow pty
remove todo
fixed type

@@
 \renewcommand{\textfraction}{0.1}
+\clubpenalty=1000
+\widowpenalty=1000
+\displaywidowpenalty=1000
 
@@
 $$\overline{x}_i=\langle x_{ij} \rangle_{j\in C_i} = \frac{1}{|C_i|}\sum_{j\in C_i} x_{ij}$$ 
-for the average methylation at position $i$, where $\langle\cdot\rangle$ denotes averaging, and the average runs over all the cells $i\in C_j$, i.e.
+for the average methylation at position $i$, where $\langle\cdot\rangle$ denotes averaging, and the average runs over all the cells $j\in C_i$, i.e.
 over those cells for which methylation data is available for position $i$.

### 2022-06-13 paper.tex
minot

@@
 We consider a cell as ``well-assigned'' if $a^k_j>\theta k$, where we have chosen $\theta=0.9$.
-In other words, a cell is ``well-assigned'' if at least 90\% of its $k$ nearest neighbors have been assigned the same class as the cell under consideration.
+In other words, a cell is ``well-assigned'' if at least 90\% of its $k$ nearest neighbors have been assigned to the same class as the cell under consideration.
 

### 2022-06-13 paper.tex
edited abstract

@@
 \small
-\textit{Abstract} -- Single-cell-bisulfite sequencing (scBS) allows to asses DNA methylation with single-basepair and single-cell resolution.
+\textit{Abstract} -- Single-cell-bisulfite sequencing (scBS) allows for assessing DNA methylation with single-basepair and single-cell resolution.
 Analysing the big data sets thus obtained requires preprocessing to reduce data size, improve signal over noise, and provide interpretability.
-Currently, this is simply done by dividing the genome into large tiles and averaging in these.
-We show that this coarse-graining dilutes signal, and suggest improved approaches to determine informative regions to quantify for and to quantitate methylation in these tiles.
-We demonstrate how our preprocessing approach helps to better distinguish cell types, reduces demands on cell number, and show that our method of finding variably-methylated regions finds biologically meaningful loci.
-Finally, we present a software tool, \texttt{scbs}, that implements this approach and offers further functionality requried to work with scBS data.
+Typically, this is simply done by dividing the genome into large tiles and averaging in these.
+We show that this coarse-graining dilutes signal, and suggest improved approaches to determine more informative regions quantify methylation in, and a better quantitation method than simple averaging.
+We demonstrate how our approach enables better distinguishing of cell types and reduces demands on cell number.
+We also show that our method of finding variably-methylated regions finds biologically meaningful loci.
+Finally, we present a software tool, \texttt{scbs}, that implements this approach and offers further functionality required to work with scBS data.
 \end{minipage}
@@
 The starting point in most scRNA-Seq analyses is a matrix of UMI counts (i.e.
-of counts of distinct RNA molecules), one row \todo{Which way round are our matrices?} for each cell and one column for each gene.
+of counts of distinct RNA molecules), one row for each cell and one column for each gene.
 A first goal is usually to assigning cell types or states to cells.

### 2022-06-13 paper.tex
reduced spacing in refs
addressing Lukas's remakrs on Slack

@@
 \bibliographystyle{genome_research}
+\setlength{\bibsep}{0.3ex plus 0.2ex}
 %\setcitestyle{aysep={}} 
@@
 
-\subsection{Smoothing residuals} \label{residuals}
+\subsection{Read-position aware quantitation} \label{residuals}
 
@@
 We take the lengths of the red lines as signed values (``residual''), positive for lines extending upwards from the curve (methylated CpG) and negative for lines extending downwards (unmethylated CpG), and take the average over the residual for all the CpGs in the interval that are covered by reads from this cell.
-The value thus obtained is what we use to quantify this cell's (relative) methylation in the interval.
+In this average, we perform shrinkage towards zero via a pseudocount (to trade bias for variance, see Methods for detail) in order to dampen the signal in cells with low coverage of the interval.
+
+The average thus obtained is what we use to quantify this cell's (relative) methylation in the interval.
 For a genome tiled into such intervals, we thus obtain a matrix, one row per cell, one column per interval, that can be used for downstream analysis, e.g., as input for a PCA.
-
-The signal to noise ratio in a matrix thus obtained will be better than in a matrix obtained by simply averaging absolute methylation (0 or 1) over all the cells' covered CpG sites in a region.
+The signal to noise ratio in this matrix will be better than in a matrix obtained by simply averaging absolute methylation (0 or 1) over all the cells' covered CpG sites in a region.
 The reason for this is that we reduce the variation in situation as the one depicted in Figure \ref{fig:smoothres}A, where the methylation of the reads might differ strongly even though there is no actual evidence for a difference between the two cells.

### 2022-06-14 paper.tex
mention PyPI

@@
 
-\texttt{scbs} is free and open source, and the package can be installed via the Python package manager.
-The source code and the package are available at \href{https://github.com/LKremer/scbs}{https://github.com/LKremer/scbs}, where we also provide detailed documentation including a tutorial that demonstrates a complete \texttt{scbs} analysis on an example data set.
+\texttt{scbs} is free and open source.
+The package is available on the Python Package Index (PyPI) and can be installed with the Python package installer pip.
+The source code is hosted at \href{https://github.com/LKremer/scbs}{https://github.com/LKremer/scbs}, where we also provide detailed documentation including a tutorial that demonstrates a complete \texttt{scbs} analysis on an example data set.
 

### 2022-06-14 paper.tex
variably-methylated -> variably methylated

@@
 We demonstrate how our approach enables better distinguishing of cell types and reduces demands on cell number.
-We also show that our method of finding variably-methylated regions finds biologically meaningful loci.
+We also show that our method of finding variably methylated regions finds biologically meaningful loci.
 Finally, we present a software tool, \texttt{scbs}, that implements this approach and offers further functionality required to work with scBS data.
@@
 		To obtain a methylation matrix, similar to the count matrices used in scRNA-seq, the user must first decide in which genomic intervals methylation should be quantified.
-		The user may either provide genomic regions of \emph{a priori} interest, or they may choose to discover VMRs (variably-methylated regions) in the data with \texttt{scbs scan}.
+		The user may either provide genomic regions of \emph{a priori} interest, or they may choose to discover VMRs (variably methylated regions) in the data with \texttt{scbs scan}.
 		The resulting methylation matrix can then be used for downstream analysis such as cell clustering and dimensionality reduction.
@@
 Based on the observation that incomplete read coverage of a genomic interval can lead to inaccurate methylation estimates, we suggest a scoring scheme that is aware of a reads local context rather than just treating all reads in an interval alike
-Furthermore, we show a way to pinpoint minimal regions of high variability across cells, which we call variably-methylated regions (VMRs).
+Furthermore, we show a way to pinpoint minimal regions of high variability across cells, which we call variably methylated regions (VMRs).
 By reanalyzing two published datasets we could show that these improvements to data preprocessing help to increase signal and decrease noise, resulting in a more informative intermediate-dimensional representation of the data.

### 2022-06-14 paper.tex
added ack
rm hyphen
add funding

@@
 
+\subsection{Availability}
+
 \texttt{scbs} is free and open source.
@@
 
-Here, we have proposed an improved strategy to preprocess single-cell bisulfite-sequencing data.
+Here, we have proposed an improved strategy to preprocess single-cell bisulfite sequencing data.
 Based on the observation that incomplete read coverage of a genomic interval can lead to inaccurate methylation estimates, we suggest a scoring scheme that is aware of a reads local context rather than just treating all reads in an interval alike
@@
 In conclusion, we have presented powerful improvements to scBS data preprocessing and a software tool that implements these and also completes a gap in existing tool chains.
+
+\subsection{Acknowledgments}
+
+S.\ A.\ and L.\ P.\ M.\ K.\ acknowledge funding by the Klaus Tschira Foundation.
+L.\ P.\ M.\ K., S.\ C.\ and A.\ M.-V.\ acknowledge funding from the European Commission via ERC grant 771376.
 

### 2022-06-15 paper.tex
lots of minor changes, last pass

@@
 \setlength{\columnsep}{0.7cm}
-%\usepackage{microtype} % nice typesetting  % doesn't work for me :-( -S
+\usepackage{microtype} % nice typesetting  % doesn't work for me :-( -S
 \usepackage{amsmath}
@@
 \small
-\textit{Abstract} -- Single-cell-bisulfite sequencing (scBS) allows for assessing DNA methylation with single-basepair and single-cell resolution.
+\textit{Abstract} -- Single-cell bisulfite sequencing (scBS) allows for assessing DNA methylation at single-base pair and single-cell resolution.
 Analysing the big data sets thus obtained requires preprocessing to reduce data size, improve signal over noise, and provide interpretability.
 Typically, this is simply done by dividing the genome into large tiles and averaging in these.
-We show that this coarse-graining dilutes signal, and suggest improved approaches to determine more informative regions quantify methylation in, and a better quantitation method than simple averaging.
+We show that this coarse-graining dilutes signal, and suggest improved approaches to determine more informative regions to quantify methylation in, and a better quantitation method than simple averaging.
 We demonstrate how our approach enables better distinguishing of cell types and reduces demands on cell number.
@@
 After sequencing, these conversions allow to determine the methylation status of all cytosines covered by reads \citep{Frommer_1992}.
-Bisulfite sequencing can also be performed with single-cell resolution \citep{Smallwood_2014} and even in parallel with scRNA-seq \citep{Clark2018}.
+Bisulfite sequencing can also be performed at single-cell resolution \citep{Smallwood_2014} and even in parallel with scRNA-seq \citep{scMTseq,Clark2018}.
 
-In the present paper, we discuss strategies to analyze scBS data, suggest improvement to current approaches, show their value in a benchmark, and present software to perform the improved analysis methods.
+In the present paper, we discuss strategies to analyze scBS data, suggest improvements to current approaches, show their value in a benchmark, and present software to perform the improved analysis methods.
 
@@
     \end{center}
-    \caption{\small \textbf{Effect of read position:} \textbf{(A)} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified.
+    \caption{\small \textbf{Improved quantification of DNA methylation in genomic intervals:} \textbf{(A)} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified.
 Two cells cover differing parts of the interval with one read each.
@@
 
-The standard approach to analyse scBS-Seq data is based on methodology developed for the analysis of single-cell RNA-Seq data.
-Therefore, we start by briefly reviewing how single-cell \emph{RNA}-Seq data is commonly analysed, before we discuss scBS data analysis.
+The standard approach to analyze scBS-Seq data is based on methodology developed for the analysis of single-cell RNA-Seq data.
+Therefore, we start by briefly reviewing how single-cell \emph{RNA}-Seq data is commonly analyzed, before we discuss scBS data analysis.
 
@@
 of counts of distinct RNA molecules), one row for each cell and one column for each gene.
-A first goal is usually to assigning cell types or states to cells.
+A first goal is usually to assign cell types or states to cells.
 To this end, one needs to establish which cells are similar to each other, i.e., quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile.
@@
 
-This procedure is commonly adapted when working with single-cell DNA-methylation data, so that established methods used in scRNA-Seq can be used.
-The task is therefore how to construct a suitable matrix as input for the PCA.
+This procedure is commonly adapted when working with single-cell DNA methylation data, so that established methods used in scRNA-Seq can be used.
+The task is therefore how to construct a suitable matrix as input for PCA.
 A simple and common approach, used for instance by \citet{luo2017single}, is to divide the genome into tiles of, say, 100~kb size, and calculate for each cell the average methylation of the DNA within each tile.
@@
 This matrix is now subjected to PCA.
-After the PCA, one can proceed using approaches known from scRNA-seq.
+After PCA, one can proceed using approaches known from scRNA-seq.
 
@@
 We first discuss the task of quantifying the level of methylation in a given, fixed, genomic interval.
-In Figure \ref{fig:smoothres}A, such an interval is covered by a single read for each of the two cells shown.
+In Figure \ref{fig:smoothres}A, such an interval is covered by a single read for two of the three cells shown.
 The read from cell 2 shows much more methylation than the read from cell 1, and the standard analysis would therefore consider that cell as having stronger methylation in the interval.
@@
 
-Therefore, we propose to first obtain, for each CpG position, an average of the methylation across all cells and, then quantify for each individual cell its deviation from this average.
+Therefore, we propose to first obtain, for each CpG position, an average of the methylation across all cells, and then quantify for each individual cell its deviation from this average.
 In Figure \ref{fig:smoothres}B, the curved line depicts such an average over all cells, and the red vertical lines show an individual cell's deviation from the ensemble average.
@@
 The average thus obtained is what we use to quantify this cell's (relative) methylation in the interval.
-For a genome tiled into such intervals, we thus obtain a matrix, one row per cell, one column per interval, that can be used for downstream analysis, e.g., as input for a PCA.
-The signal to noise ratio in this matrix will be better than in a matrix obtained by simply averaging absolute methylation (0 or 1) over all the cells' covered CpG sites in a region.
-The reason for this is that we reduce the variation in situation as the one depicted in Figure \ref{fig:smoothres}A, where the methylation of the reads might differ strongly even though there is no actual evidence for a difference between the two cells.
+For a genome tiled into such intervals, we thus obtain a matrix, one row per cell, one column per interval, that can be used for downstream analysis, e.g., as input for PCA.
+The signal-to-noise ratio in this matrix will be better than in a matrix obtained by simply averaging absolute methylation (0 or 1) over all the cells' covered CpG sites in a region.
+The reason for this is that we reduce the variation in situations as the one depicted in Figure \ref{fig:smoothres}A, where the methylation of the reads might differ strongly even though there is no actual evidence for a difference between the two cells.
 
@@
 
-To demonstrate the value of our proposed improvements, we obtained two previously published scBS data sets and re-analysed them.
+To demonstrate the value of our proposed improvements, we obtained two previously published scBS data sets and re-analyzed them.
 
@@
 
-The first data set is a single-cell multi-omics data set generated by \citet{argelaguet2019gastru}.
+The first data set used for benchmarking is a single-cell multi-omics data set generated by \citet{argelaguet2019gastru}.
 In their work, the authors obtained single-cell methylomes of mouse embryo cells across different stages of development (E4.5--E7.5) as well as single-cell transcriptomes from the same cells.
@@
 
-In Figure \ref{figure:correlation}A, we compare correlations of gene expression and promoter methylation to correlations of gene expression and methylation of the closest VMR identified with \texttt{scbs scan}.
+In Figure \ref{figure:correlation}A, we compare correlations of gene expression and promoter methylation to correlations of gene expression and methylation of the closest VMR identified with our approach.
 This analysis revealed a number of genes whose expression does not correlate with promoter methylation, but correlates with VMR methylation.
@@
     \textbf{(B)} Same as in (A), but using mean shrunken residuals as a measure of methylation.
-    \textbf{(C)} Mean smoothed methylation near the gene Ube2l6 (ENSMUSG00000027078, marked in yellow in A \& B).
+    \textbf{(C)} Mean smoothed methylation near the gene Ube2l6 (ENSMUSG00000027078, marked in orange in A \& B).
 Cells are assigned to 4 groups based on Ube2l6 expression (group 0: cells with no Ube2l6 expression, groups 1-3: cells that express Ube2l6, divided into three equally large groups with group 3 having the highest expression).
@@
 To illustrate this effect with an example, we picked a gene where VMR methylation is a much better predictor of gene expression than promoter methylation (Ube2l6, highlighted in orange in Figure \ref{figure:correlation}A,B).
-Grouping all cells into four groups according to their level of Ube2l6 expression, it is clearly visible that the promoter of this gene is lowly methylated regardless of gene expression, while methylation at the nearest VMR is greatly decreased in cells with high Ube2l6 expression.
+Dividing all cells into four groups according to their level of Ube2l6 expression, it is clearly visible that the promoter of this gene is lowly methylated regardless of gene expression, while methylation at the nearest VMR is greatly decreased in cells with high Ube2l6 expression.
 
@@
 Exemplary UMAPs for analyses are shown in (D).
-\textbf{(C)} Two exemplary UMAPs of the full data set and methylation fractions of 100~kb tiles (left) or mean shrunken residuals of VMRs (right) \textbf{(D)} Two exemplary UMAPs of the smaller data set and methylation fraction of 100~kb tiles (left) or mean shrunken residuals of VMRs (right).
+\textbf{(C)} Two exemplary UMAPs of the full data set and methylation fractions of 100~kb tiles (left) or mean shrunken residuals of VMRs (right) \textbf{(D)} Two exemplary UMAPs of the smaller data set and methylation fractions of 100~kb tiles (left) or mean shrunken residuals of VMRs (right).
 Data from \citet{luo2017single}.}
@@
 We subjected this data set to the standard scBS workflow: defining a set of genomic intervals, quantifying methylation in these intervals, and subjecting the resulting matrix to PCA.
-We repeated this analysis several times, either using 100~kb tiles or VMRs, and either using methylation fractions or mean shrunken residuals as a measure of methylation. 
-For both types of regions, we varied the number of input regions, ranging from 100 to 50,000 high-coverage regions that were selected for PCA.
-Specifically, we ranked regions by their number of coverging cells (i.e., cells with at least on read in the region) and selected those with the highest number.
+We repeated this analysis several times, either using 100~kb tiles or VMRs, and either using methylation fractions or the shrunken mean of residuals as a measure of methylation. 
+For both types of regions, we varied the number of input regions, ranging from 100 to 50,000 regions with the highest sequencing coverage that were selected for PCA.
 
@@
 
-Since scBS protocols is still rather costly and labor-intensive, only few laboratories are currently able to obtain thousands of single cell methylomes.
+Since scBS protocols are still rather costly and labor-intensive, only few laboratories are currently able to obtain thousands of single cell methylomes.
 To simulate a smaller data set, we subsampled the frontal cortex data set to 500 cells and repeated our analysis.
@@
 
-\texttt{scbs prepare} also computes a number of summary statistics for each cell, including the mean genome-wide methylation level and the number of observed CpG sites, i.~e.
+\texttt{scbs prepare} also computes a number of summary statistics for each cell, including the mean genome-wide methylation level and the number of observed CpG sites, i.e.
 the number of CpG sites that have sequencing coverage.
@@
 
-After quality control, the user has access to the genome-wide VMR detection approach described above via \texttt{scbs scan}.
+After quality control, the user has access to the genome-wide VMR detection approach described earlier, using the \texttt{scbs scan} command.
 This produces a BED file that lists the genome coordinates of VMRs.
@@
 The command produces both the percentage of methylated CpG sites, as well as our proposed methylation measure (shrunken mean of the residuals) that is more robust to variation in sequencing coverage or stochastic differences in read position between cells.
-We note that both \texttt{matrix} and \texttt{profile} accept any valid BED-file as input, which means that the user can quantify and visualize methylation at any genomic feature of interest, such as promoters, enhancers or transcription factor binding sites, in single cells.
+We note that both \texttt{matrix} and \texttt{profile} accept any valid BED file as input, which means that the user can quantify and visualize methylation at any genomic feature of interest, such as promoters, enhancers or transcription factor binding sites, in single cells.
 The obtained methylation matrix can then serve as input for common single-cell analysis methods such as dimensionality reduction and cell clustering.
@@
 Here, we have proposed an improved strategy to preprocess single-cell bisulfite sequencing data.
-Based on the observation that incomplete read coverage of a genomic interval can lead to inaccurate methylation estimates, we suggest a scoring scheme that is aware of a reads local context rather than just treating all reads in an interval alike
+Based on the observation that incomplete read coverage of a genomic interval can lead to inaccurate methylation estimates, we suggest a scoring scheme that is aware of a reads local context rather than just treating all reads in an interval alike.
+
 Furthermore, we show a way to pinpoint minimal regions of high variability across cells, which we call variably methylated regions (VMRs).
-By reanalyzing two published datasets we could show that these improvements to data preprocessing help to increase signal and decrease noise, resulting in a more informative intermediate-dimensional representation of the data.
-As examples of practical benefits, we demonstrate that our preprocessing allows for better distinction of cell sub-types, especially for smaller data sets with limited numbers of cells.
+By re-analyzing two published data sets, we could show that these improvements to data preprocessing help to increase signal and decrease noise, resulting in a more informative intermediate-dimensional representation of the data.
+As examples of practical benefits, we demonstrate that our preprocessing allows for better distinction of cell sub-types, especially for smaller data sets with a limited number of cells.
 In addition to this improvement of global signal-to-noise ratio (with regards to cell dissimilarity quantification), we also show that the determination of VMRs improve biological interpretability by pinpointing the regions that actually change and hence have a putative regulatory role.
 
-Furthermore, we have presented and published an open source software tool, called \texttt{scbs}, that provides an easy-to-use implementation of the described algorithms.
+We also presented and published an open source software tool, called \texttt{scbs}, that provides an easy-to-use implementation of the described algorithms.
 The package also offers functionality for all other steps of data preprocessing that are needed to get from the output of methylation callers such as Bismark and methylpy to a cell\texttimes region matrix that is suitable as input for standard analysis tools that work in a reduced-dimensionality setting, thus allowing the user to use, e.g., well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses etc.
@@
 
-In conclusion, we have presented powerful improvements to scBS data preprocessing and a software tool that implements these and also completes a gap in existing tool chains.
+In conclusion, we have presented powerful improvements to scBS data preprocessing and a software tool that implements these and bridges a gap in existing tool chains.
 
@@
 The index $i$ runs over all CpG positions present in the genome, the index $j$ over all cells in the assay.
-We write $x_{ij}=0$ if position $i$ was found to be unmethylated in cell $j$ by bisulfite sequencing , $x_{ij}=1$ if it was methylated, and $x_{ij}=\text{NA}$ if position $i$ was not covered by reads from cell $j$ and the methylation status is therefore not available ("NA"),
+We write $x_{ij}=0$ if position $i$ was found to be unmethylated in cell $j$ by bisulfite sequencing, $x_{ij}=1$ if it was methylated, and $x_{ij}=\text{NA}$ if position $i$ was not covered by reads from cell $j$ and the methylation status is therefore not available ("NA"),
 
@@
 
-If multiple reads with the same cell barcode cover the positions, these will typically be PCR duplicates of each other and hence agree.
-Of course, the two alleles of a CpG may differ in their methylation status and while it is in principle possible that one obtains reads stemming the same position on both the paternal and the maternal chromosomes of the same cell, but this is so unlikely that we can ignore the case.
-Hence, whenever Bismarck reports multiple reads covering the same position in the same cell, we set $x_{ij}$ to 0 or 1 following the majority of reads and put $x_{ij}=\text{NA}$ in case of a tie.
+If multiple reads from the same cell cover a position, these will typically be PCR duplicates of each other and hence agree.
+Of course, the two alleles of a CpG may rarely differ in their methylation status.
+While it is in principle possible that one obtains discordant reads stemming from the same position on both the paternal and the maternal chromosomes of the same cell, this is so unlikely that we can ignore the case.
+Hence, whenever the methylation caller reports multiple reads covering the same position in the same cell, we set $x_{ij}$ to 0 or 1 whenever all reads agree.
+When there is disagreement, we put $x_{ij}=\text{NA}$ by default, or optionally follow the majority of reads whenever possible.
 
@@
 
-\subsection{Finding variably methylated regions (VMRs)}
+\subsection{Finding variably methylated regions}
 
@@
 
-Given a set, $\mathcal{V}=\{I^\text{v}_1,I^\text{v}_2,\dots\}$, of intervals corresponding to VMRs, we get a relative methylation fraction $r_{ij}$ for each VMR $I^\text{v}_i$ and each cell $j$ from Eq.\ ()\ref{avgres}).
+Given a set, $\mathcal{V}=\{I^\text{v}_1,I^\text{v}_2,\dots\}$, of intervals corresponding to VMRs, we get a relative methylation fraction $r_{ij}$ for each VMR $I^\text{v}_i$ and each cell $j$ from Eq.\ (\ref{avgres}).
 The matrix thus obtained can then be centered and used as input for a PCA.
@@
 For any two cells $j$, $j'$, we use the Euclidean distance $\|\mathbf{x}^\text{P}_j - \mathbf{x}^\text{P}_{j'}\|$ as the measure of dissimilarity of the two cells.
-Thus, the matrix of PC scores can be used as input to dimension reduction methods like t-SNE or UMAP, and to clustering methods like the Louvain or Leiden algorithm, which require such a matrix as input to the approximate mnearest neighbor (ANN) finding algorithm that is their first step.
+Thus, the matrix of PC scores can be used as input to dimension reduction methods like t-SNE or UMAP, and to clustering methods like the Louvain or Leiden algorithm, which require such a matrix as input to the approximate nearest neighbor (ANN) finding algorithm that is their first step.
 
@@
 
-Whenever a region is not covered by any read in a cell, the corresponding data entry in the input data matrix for the PCA will be missing.
-The standard approach to calculate PCA (commonly done using the IRLBA algorthm, \citep{Baglama2005}) is not suited to deal with missing data.
+Whenever a region is not covered by any read in a cell, the corresponding data entry in the input data matrix for PCA will be missing.
+The standard approach to calculate PCA (commonly done using the IRLBA algorithm, \citep{Baglama2005}) is not suited to deal with missing data.
 We circumvent this issue by simply using the PCA itself to impute the missing value.
@@
 Our iterative imputation strategy is now simply the following: we first replace all missing values in the row-centered input matrix $A$ with zeroes and perform the (truncated) PCA.
-Then, we use the PCA predictions for the missing values, i.e., the $a_{ij}=\sum_r x_{ir} r_{jr}$, as refined stand-ins for the missing values in $A$ and run the PCA once more.
+Then, we use the PCA predictions for the missing values, i.e., the $a_{ij}=\sum_r x_{ir} r_{jr}$, as refined stand-ins for the missing values in $A$ and run PCA once more.
 This can now be iterated until convergence, which is typically achieved after at most 4--5 rounds of PCA.
@@
 \subsection{Re-analysis of published scBS datasets}
-To re-analyze scBS data sets by \citet{luo2017single} and \citet{argelaguet2019gastru}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using default options.
+To re-analyze scBS data sets by \citet{luo2017single} and \citet{argelaguet2019gastru}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 2\%$).
 For all regions (VMRs and 100~kb tiles), the mean methylation fractions and mean shrunken residuals were obtained with \texttt{scbs matrix}.
@@
 To select $n$ intervals for further analysis, the top $n$ intervals were picked based on the number of cells that the region was measured in.
-For dimensionality reduction, missing values were imputed with an iterative PCA as described above. To visualize data, UMAP was performed on the final 15 principal components.
+For dimensionality reduction, missing values were imputed with the iterative PCA described in the previous section. To visualize data, UMAP was performed on the final 15 principal components.
 

### 2022-06-15 paper.tex
always use same term for the residuals

@@
 Promoters are defined as intervals \textpm2~kb around the TSS.
-    \textbf{(B)} Same as in (A), but using mean shrunken residuals as a measure of methylation.
+    \textbf{(B)} Same as in (A), but using the shrunken mean of residuals as a measure of methylation.
     \textbf{(C)} Mean smoothed methylation near the gene Ube2l6 (ENSMUSG00000027078, marked in orange in A \& B).
@@
 Exemplary UMAPs for analyses are shown in (D).
-\textbf{(C)} Two exemplary UMAPs of the full data set and methylation fractions of 100~kb tiles (left) or mean shrunken residuals of VMRs (right) \textbf{(D)} Two exemplary UMAPs of the smaller data set and methylation fractions of 100~kb tiles (left) or mean shrunken residuals of VMRs (right).
+\textbf{(C)} Two exemplary UMAPs of the full data set and methylation fractions of 100~kb tiles (left) or shrunken means of residuals of VMRs (right) \textbf{(D)} Two exemplary UMAPs of the smaller data set and methylation fractions of 100~kb tiles (left) or shrunken means of residuals of VMRs (right).
 Data from \citet{luo2017single}.}
@@
 We subjected this data set to the standard scBS workflow: defining a set of genomic intervals, quantifying methylation in these intervals, and subjecting the resulting matrix to PCA.
-We repeated this analysis several times, either using 100~kb tiles or VMRs, and either using methylation fractions or the shrunken mean of residuals as a measure of methylation. 
+We repeated this analysis several times, either using 100~kb tiles or VMRs, and either using methylation fractions or the shrunken means of residuals as a measure of methylation. 
 For both types of regions, we varied the number of input regions, ranging from 100 to 50,000 regions with the highest sequencing coverage that were selected for PCA.
@@
 The results demonstrate that the 16 neuronal subtypes are more cleanly separated when using our proposed methods (Figure \ref{figure:score}A).
-Most of this gain in performance can be attributed to the use of VMRs over genomic tiles. The use of mean shrunken residuals over methylation fractions provides an additional small improvement.
+Most of this gain in performance can be attributed to the use of VMRs over genomic tiles.
+The use of the shrunken means of residuals over methylation fractions provides an additional small improvement.
 
@@
 To re-analyze scBS data sets by \citet{luo2017single} and \citet{argelaguet2019gastru}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 2\%$).
-For all regions (VMRs and 100~kb tiles), the mean methylation fractions and mean shrunken residuals were obtained with \texttt{scbs matrix}.
+For all regions (VMRs and 100~kb tiles), the mean methylation fractions and the shrunken means of residuals were obtained with \texttt{scbs matrix}.
 Regions containing less than four measured CpGs, or that were measured in less than 10\% of cells, were discarded.

### 2022-06-15 paper.tex
add Ana's comments

@@
 
-While this simple procedure is straight-forward and produces satisfactory results, there is room for improvement.
+While this simple procedure is straight-forward and produces satisfactory results, the coarse resolution of large genomic tiles and the sparse read coverage of single-cell methylomes still pose significant challenges in the analysis of scBS data.
 Here, we propose three improvements that, as we show afterwards, considerably increase the information content of the processed data.
@@
 We also presented and published an open source software tool, called \texttt{scbs}, that provides an easy-to-use implementation of the described algorithms.
-The package also offers functionality for all other steps of data preprocessing that are needed to get from the output of methylation callers such as Bismark and methylpy to a cell\texttimes region matrix that is suitable as input for standard analysis tools that work in a reduced-dimensionality setting, thus allowing the user to use, e.g., well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses etc.
+The package also offers functionality for all other steps of data preprocessing that are needed to get from the output of methylation callers such as Bismark and methylpy to a cell\texttimes region matrix that is suitable as input for standard analysis tools that work in a reduced-dimensionality setting, thus allowing the user to use, e.g., well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc.
 By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.
@@
 The index $i$ runs over all CpG positions present in the genome, the index $j$ over all cells in the assay.
-We write $x_{ij}=0$ if position $i$ was found to be unmethylated in cell $j$ by bisulfite sequencing, $x_{ij}=1$ if it was methylated, and $x_{ij}=\text{NA}$ if position $i$ was not covered by reads from cell $j$ and the methylation status is therefore not available ("NA"),
+We write $x_{ij}=0$ if position $i$ was found to be unmethylated in cell $j$ by bisulfite sequencing, $x_{ij}=1$ if it was methylated, and $x_{ij}=\text{NA}$ if position $i$ was not covered by reads from cell $j$ and the methylation status is therefore not available (``NA''),
 
@@
 This is a shrunken average, with denominator $n+1$.
-This extra pseudocount has the effect of shrinking the value towards the "neutral" value 0, with the shrinkage becoming stronger if the data is "weak" because the number $|I\cap C_i|$ of positions covered by reads from cell $j$  is low.
+This extra pseudocount has the effect of shrinking the value towards the ``neutral'' value 0, with the shrinkage becoming stronger if the data is ``weak'' because the number $|I\cap C_i|$ of positions covered by reads from cell $j$  is low.
 In the extreme case of none of the reads from cell $j$ covering $I$, the sum becomes 0 and the denominator 1, i.e.

### 2022-06-15 paper.tex
add leonie to ack
add alexey to ack

@@
 
+\vspace{1.4ex}
+\noindent\hfil\rule{.6\columnwidth}{.2pt}\hfil
+
 \subsection{Acknowledgments}
@@
 S.\ A.\ and L.\ P.\ M.\ K.\ acknowledge funding by the Klaus Tschira Foundation.
-L.\ P.\ M.\ K., S.\ C.\ and A.\ M.-V.\ acknowledge funding from the European Commission via ERC grant 771376.
-
-\vspace{1.4ex}
-\noindent\hfil\rule{.6\columnwidth}{.2pt}\hfil
+L.\ P.\ M.\ K., L.\ K., S.\ C.\ and A.\ M.-V.\ acknowledge funding from the European Commission via ERC grant 771376.
+We thank Alexey Uvarovskii for code refactoring and helpful suggestions on the source code.
 

### 2022-06-15 paper.tex
implement Andre's suggestions

@@
 Therefore, one performs a principal component analysis (PCA), keeping only the top few (typically, 20 to 50) components.
-As Poisson noise is uncorrelated between genes, it will average out in the top principal components, as these are all linear combinations with weight on very many genes.
+As Poisson noise is uncorrelated between genes, it will average out in the top principal components, as these are all linear combinations with weight on a large number of genes.
 Therefore, Euclidean distances between these ``PCA space'' vectors provide a robust dissimilarity score.
@@
 Therefore, we propose to smoothen using a kernel smoother, i.e., by performing a kernel-weighted average over the CpG site's neighborhood.
-The kernel bandwidth (i.e., the size of the neighborhood to average over) is a tuning parameter; we got good results with 1000~bp and used this value for the examples presented here.
+The kernel bandwidth (i.e., the size of the neighborhood to average over) is a tuning parameter; for the examples presented here we used 1000~bp.
 
@@
 
-In the standard approach, one divides up (tiles) each chromosome into non-overlapping, equal-sized intervals, and quantitates the methylation of each such tiles.
+In the standard approach, one divides up (tiles) each chromosome into non-overlapping, equal-sized intervals, and quantitates the methylation of each tile.
 Such rigid placing of interval boundaries is unlikely to be optimal: for example, a VMR might be much smaller than a tile and its signal will hence be drowned out by the larger part of uninformative CpGs that are equal in all cells, when averaging over all the CpG sites in the tile.
@@
 Select, say, the top 2\% windows with the highest variances and mark them as VMRs.
-Wherever thus marked windows overlap or are divided by only a small gap, merge them into one larger VMR.
+Wherever thus marked windows overlap or are separated by only a small gap, merge them into one larger VMR.
 Then, calculate for each of these larger VMRs the methylation signal, as before, by averaging for each cell over the residuals of all contained CpG sites.
@@
 To illustrate this effect with an example, we picked a gene where VMR methylation is a much better predictor of gene expression than promoter methylation (Ube2l6, highlighted in orange in Figure \ref{figure:correlation}A,B).
-Dividing all cells into four groups according to their level of Ube2l6 expression, it is clearly visible that the promoter of this gene is lowly methylated regardless of gene expression, while methylation at the nearest VMR is greatly decreased in cells with high Ube2l6 expression.
+Dividing all cells into four groups according to their level of Ube2l6 expression, it is clearly visible that the promoter of this gene is lowly methylated regardless of gene expression, while methylation at the nearest VMR is greatly decreased in cells with high Ube2l6 expression (Figure \ref{figure:correlation}C).
 

### 2022-06-15 paper.tex
rm rare occurrences of british english

@@
 \centering
-\textbf{\Large Analysing single-cell bisulfite sequencing data with \texttt{scbs}}\\[1.5ex]
+\textbf{\Large Analyzing single-cell bisulfite sequencing data with \texttt{scbs}}\\[1.5ex]
 
@@
 \textit{Abstract} -- Single-cell bisulfite sequencing (scBS) allows for assessing DNA methylation at single-base pair and single-cell resolution.
-Analysing the big data sets thus obtained requires preprocessing to reduce data size, improve signal over noise, and provide interpretability.
+Analyzing the big data sets thus obtained requires preprocessing to reduce data size, improve signal over noise, and provide interpretability.
 Typically, this is simply done by dividing the genome into large tiles and averaging in these.
@@
 
-To illustrate this effect with an example, we picked a gene where VMR methylation is a much better predictor of gene expression than promoter methylation (Ube2l6, highlighted in orange in Figure \ref{figure:correlation}A,B).
+To illustrate this effect with an example, we picked a gene where VMR methylation is a much better predictor of gene expression than promoter methylation (Ube2l6, highlighted in orange in Figure \ref{figure:correlation}A-B).
 Dividing all cells into four groups according to their level of Ube2l6 expression, it is clearly visible that the promoter of this gene is lowly methylated regardless of gene expression, while methylation at the nearest VMR is greatly decreased in cells with high Ube2l6 expression (Figure \ref{figure:correlation}C).

### 2022-06-20 paper.tex
fmt

@@
 
-Lukas P.~M.
-Kremer\textsuperscript{1,2,*}, Leonie Küchenhoff\textsuperscript{1,2}, Santiago Cerrizuela\textsuperscript{2}, Ana Martin-Villalba\textsuperscript{2}, Simon Anders\textsuperscript{1,*}\\[1ex]
+Lukas P.~M.~Kremer\textsuperscript{1,2,*}, Leonie Küchenhoff\textsuperscript{1,2}, Santiago Cerrizuela\textsuperscript{2}, Ana Martin-Villalba\textsuperscript{2}, Simon Anders\textsuperscript{1,*}\\[1ex]
 {\footnotesize \textsuperscript{1} BioQuant Centre, University of Heidelberg, Heidelberg, Germany\\[-.5ex]

### 2023-03-09 paper.tex
rm Ensembl gene ID
inserted missing sum sign
add bookmarks pkg to fix PDF ToC
Added extra toc entries
moved hrule
start re-write

--- 2022-06-15 paper.tex
@@
 Lukas P.~M.
-Kremer\textsuperscript{1,2,*}, Leonie Küchenhoff\textsuperscript{1,2}, Santiago Cerrizuela\textsuperscript{2}, Ana Martin-Villalba\textsuperscript{2}, Simon Anders\textsuperscript{1,*}\\[1ex]
+Kremer\textsuperscript{1,2,*}, Martina Braun\textsuperscript{2}, Leonie Küchenhoff\textsuperscript{1,2}, Santiago Cerrizuela\textsuperscript{2},\\Ana Martin-Villalba\textsuperscript{2}, Simon Anders\textsuperscript{1,*}\\[1ex]
 {\footnotesize \textsuperscript{1} BioQuant Centre, University of Heidelberg, Heidelberg, Germany\\[-.5ex]
@@
 
-June 2022
+January 2023
 \vspace{5.5ex}
@@
 We demonstrate how our approach enables better distinguishing of cell types and reduces demands on cell number.
-We also show that our method of finding variably methylated regions finds biologically meaningful loci.
-Finally, we present a software tool, \texttt{scbs}, that implements this approach and offers further functionality required to work with scBS data.
+We also implement an approach to detect differentially methylated regions (DMRs) between groups of cells.
+Finally, we present a software tool, \texttt{scbs}, that implements these methods and offers further functionality required to work with scBS data.
 \end{minipage}
 \vspace{6.5ex}
-
 }]
@@
 Single-cell RNA sequencing (scRNA-seq) techniques have matured at great pace in recent years, with well developed analysis methodology, and methods to study epigenetics at single-cell resolution are rapidly catching up.
+
 \blfootnote{\hspace{-.5cm}\raggedright\textsuperscript{*} e-mail:  \texttt{l.kremer@dkfz-heidelberg.de} and\\ \hspace{1cm}\texttt{simon.anders@bioquant.uni-heidelberg.de}}
@@
 
-In the present paper, we discuss strategies to analyze scBS data, suggest improvements to current approaches, show their value in a benchmark, and present software to perform the improved analysis methods.
+In the present paper, we discuss strategies to analyze single-cell bisulfite-sequencing (scBS) data, suggest improvements to current approaches, show their value in a benchmark, and present software to perform the improved analysis methods.
 
@@
 \begin{figure*}[t]
-    \hspace{2.5cm}A\\[-7ex]
     \begin{center}
-        \includegraphics[width=0.7\textwidth]{figures/Fig_residuals_A.png}\\
+        \includegraphics[width=0.7\textwidth]{figures/Fig_residuals_AB.png}\\
     \end{center}
-    \hspace{2.5cm}B\\[-3ex]
-    \begin{center}
-        \includegraphics[width=0.7\textwidth]{figures/Fig_residuals_B.png}
-    \end{center}
-    \caption{\small \textbf{Improved quantification of DNA methylation in genomic intervals:} \textbf{(A)} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified.
+    \caption{\small \textbf{Improved quantification of DNA methylation in genomic intervals:}
+    \textbf{(A)} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified.
 Two cells cover differing parts of the interval with one read each.
@@
 The task is therefore how to construct a suitable matrix as input for PCA.
-A simple and common approach, used for instance by \citet{luo2017single}, is to divide the genome into tiles of, say, 100~kb size, and calculate for each cell the average methylation of the DNA within each tile.
+A simple and common approach, used for instance by \citet{luo2017single}, is to divide the genome into tiles of e.g.
+100~kb size, and calculate for each cell the average methylation of the DNA within each tile.
 To this end, we identify all CpG sites in the tile for which we have coverage with at least one read and can hence call the CpG to be either unmethylated or methylated in the cell.
@@
 This matrix is now subjected to PCA.
-After PCA, one can proceed using approaches known from scRNA-seq.
+After PCA, one can proceed with dimensionality reduction and clustering approaches known from scRNA-seq.
 
@@
 In Figure \ref{fig:smoothres}A, such an interval is covered by a single read for two of the three cells shown.
-The read from cell 2 shows much more methylation than the read from cell 1, and the standard analysis would therefore consider that cell as having stronger methylation in the interval.
+The read from cell 2 shows much more methylation than the read from cell 1, and the standard analysis would therefore consider that cell as having higher methylation in the interval.
 However, given that the two reads agree wherever they overlap, a more parsimonious interpretation would be that the cells do not show difference in methylation within the interval.
@@
 
-How should one obtain the ensemble average (the curved line in Fig.\ \ref{fig:smoothres}B)?
+How should one obtain the ensemble average (the curved line in Fig.~\ref{fig:smoothres}B)?
 A simple approach to get a value for a specific CpG would be to take all cells with read coverage for the CpG and use the fraction of these that show the CpG as methylated.
@@
 
+% remove or shorten this section cause it's not really novel, methods like this are already implemented in e.g. the "missMDA" R package
 \subsection{PCA with imputation}
@@
 
-\subsection{Finding variably methylated regions}
+\subsection{Finding variably methylated regions (VMRs)}
 
@@
     \includegraphics[width=\columnwidth]{figures/Fig_sliding.png}
-    \caption{\small \textbf{Finding variably methylated regions (VMR):} The chromosomes are divided up into overlapping windows (first five shown at the bottom), and for each window, the cells' methylation values are calculated as described and as depicted in Fig.\ \ref{fig:smoothres}B.
+    \caption{\small \textbf{Finding variably methylated regions (VMRs):} The chromosomes are divided up into overlapping windows (first five shown at the bottom), and for each window, the cells' methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
 Then, the variance of these values is calculated (each point represents one of the overlapping windows), a threshold (dashed line) is chosen such that a chosen quantile of windows have a variance exceeding the threshold.
@@
 In contrast, DNA methylation at certain genomic features such as enhancers is more dynamic \citep{argelaguet2019gastru}, and thus more variable across cells.
-Only the latter regions are of value for our goal of quantitating dissimilarity between cells. We call these the variably methylated regions (VMRs).
+Only the latter regions are of value for our goal of quantitating dissimilarity between cells.
+We call these the variably methylated regions (VMRs). % maybe mention the parallel to highly variable genes in scRNAseq?
 
 In the standard approach, one divides up (tiles) each chromosome into non-overlapping, equal-sized intervals, and quantitates the methylation of each tile.
-Such rigid placing of interval boundaries is unlikely to be optimal: for example, a VMR might be much smaller than a tile and its signal will hence be drowned out by the larger part of uninformative CpGs that are equal in all cells, when averaging over all the CpG sites in the tile.
+Such rigid placing of interval boundaries is unlikely to be optimal: for example, a VMR might be much smaller than a tile and its signal will hence be drowned out by the larger part of uninformative CpG sites that are equal in all cells, when averaging over all the CpG sites in the tile.
 
-Therefore, we propose the following approach (Fig.
-\ref{vmr}): Divide up the chromosome into many \emph{overlapping} windows, that start at regular multiples of a fixed, small, step size.
-Quantify the methylation of each cell in each window by averaging the cell's methylation residuals over all CpGs in the window, as described above and depicted in Figure\ \ref{fig:smoothres}B.
+Therefore, we propose the following approach (Fig.~\ref{vmr}): Divide up the chromosome into many \emph{overlapping} windows, that start at regular multiples of a fixed, small, step size.
+Quantify the methylation of each cell in each window by averaging the cell's methylation residuals over all CpGs in the window, as described above and depicted in Fig.~\ref{fig:smoothres}B.
 Then, calculate for each window the variance of these values over all cells.
@@
 
+\subsection{Finding differentially methylated regions (DMRs)}
+
+\begin{figure}
+    \includegraphics[width=\columnwidth]{figures/Fig_DMRdetection.png}
+    \caption{\small \textbf{Identification of differentially methylated regions (DMRs):}
+    The chromosome is divided into windows overlapping by a small and fixed size.
+    For each window, the cell’s methylation values are calculated as described and as depicted in Fig.~ \ref{fig:smoothres}B.
+    Then the t-statistic of these values is obtained as a measure of methylation difference (each point represents one of the windows).
+    Upper and lower thresholds (dashed lines) are chosen such that a defined quantile of windows have a t-statistic above or below the threshold.
+    Windows exceeding either threshold are merged if they overlap, yielding the ``differentially methylated regions``.}
+    \label{dmr}
+\end{figure}
+
+A common task in the analysis of bulk bisulfite-sequencing data is the detection of differentially methylated regions (DMRs) between conditions, tissues, or cell types \citep{Hebestreit2013, dmrseq}.
+As DNA methylation affects gene expression, DMRs can provide insights into the unique epigenetic and gene regulatory characteristics of cell types.
+However, to date no approach to detect DMRs in scBS data has been reported.
+To enable DMR detection in scBS data, we thus developed an approach that detects DMRs of variable size between two groups of cells, and controls the false discovery rate (Fig.~3):
+
+Similar to the previously described VMR identification approach (Fig.~\ref{vmr}), divide each chromosome into overlapping windows shifted by a small and fixed size (step size) and quantify the methylation of each cell in each window as described (Fig.~\ref{residuals}b).
+Next, instead of calculating the variance, obtain the t-statistic % Welch t-test
+as a measure of differential methylation between the two cell groups.
+Mark the windows with the most extreme t-statistics (e.g. 2\% upper and lower tails) as DMRs.
+Distinguishing between high and low t-statistics, merge any overlapping marked windows into larger DMRs.
+Re-calculate the t-statistic for each larger DMR.
+
+To assess statistical significance of DMRs, we repeat the same procedure on permutations of the scBS data.
+The DMR t-statistics obtained from permuted data are then used to estimate the false-discovery rate (FDR), yielding an adjusted p-value for each DMR.
+
+\subsection{???}  % guess we have to rewrite this closing paragraph a little bit, to include both DMRs and VMRs
+
 Besides providing better input for the distance calculations, the small size of identified VMRs also means that they can be readily compared with genomic annotation, which aids in their functional interpretation.
@@
 
-\section{Benchmarks}
+\section{Benchmarks} % re-name this to something more general? Application and benchmarks or something?
 
@@
     \begin{center}
-    A\hspace{.9\columnwidth}~\\
-    \includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/corr_methylation_fraction_4kbwindow.pdf} \\
-    B\hspace{.9\columnwidth}~\\
-    \includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/corr_shrunken_residual_4kbwindow.pdf} \\
-    C\hspace{.9\columnwidth}~\\
-    \includegraphics[width=.95\columnwidth]{part_leonie_git/leonie_plots/comparison_meth_RNA.pdf}
+    \includegraphics[width=.95\columnwidth]{figures/Fig_correlation.png}
     \end{center}
     \caption{\small \textbf{Correlation of DNA methylation and gene expression}.
-    \textbf{(A)} Correlation of each gene's RNA expression values to the the methylation fractions that were measured at promoters (y-axis), or to methylation fractions of the nearest VMR (x-axis).
+    \textbf{(A)} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red).
+    \textbf{(B)} Correlation of each gene's RNA expression values to methylation (mean shrunken residuals) at promoters (y-axis), or to methylation of the nearest VMR (x-axis).
     Promoters are defined as intervals \textpm2~kb around the TSS.
-    \textbf{(B)} Same as in (A), but using the shrunken mean of residuals as a measure of methylation.
-    \textbf{(C)} Mean smoothed methylation near the gene Ube2l6 (ENSMUSG00000027078, marked in orange in A \& B).
-Cells are assigned to 4 groups based on Ube2l6 expression (group 0: cells with no Ube2l6 expression, groups 1-3: cells that express Ube2l6, divided into three equally large groups with group 3 having the highest expression).
-The mean methylation level of each group was smoothed with a tricube kernel of bandwidth 2000.
-    Data from \citet{argelaguet2019gastru}.}
+    \textbf{(C)} Mean methylation near the gene Htra1 (ENSMUSG00000006205).
+    Cells are assigned to 4 groups based on Htra1 expression (group 0: cells with no Htra1 expression, groups 1-3: cells that express Htra1, divided into three equally large groups with group 3 having the highest expression).
+    Data from \citet{kremer_scnmt}.}
     \label{figure:correlation}
@@
 
-\begin{figure*}
+\begin{figure}
     \begin{center}
-        A\hspace{.4\textwidth}B\hspace*{.3\textwidth}~\\
-        \hspace{.3cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/complete_135k_12cm_log.pdf}
-        \hspace{1cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/cell500_20k_12cm_log.pdf}\\
-        C\hspace*{.7\textwidth}~\\
-        \hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_fulldataset.pdf}\\
-        D\hspace*{.7\textwidth}~\\
-        \hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_reduceddataset.pdf}
-        \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of 16 neuronal subtypes}.
-\textbf{(A)} Nearest neighbor count score using four different method combinations and a variable number of genomic regions (VMRs or 100~kb tiles) as input to PCA.
-Exemplary UMAPs for encircled analyses are shown in (C).
-\textbf{(B)} Same benchmark as in (A), but for a smaller data set consisting of 500 randomly subsampled cells.
-Exemplary UMAPs for analyses are shown in (D).
-\textbf{(C)} Two exemplary UMAPs of the full data set and methylation fractions of 100~kb tiles (left) or shrunken means of residuals of VMRs (right) \textbf{(D)} Two exemplary UMAPs of the smaller data set and methylation fractions of 100~kb tiles (left) or shrunken means of residuals of VMRs (right).
-Data from \citet{luo2017single}.}
+        \includegraphics[width=\columnwidth]{figures/Fig_benchmark.png}
+    \end{center}
+    \caption{\small \textbf{Benchmark of our methods on single-cell multi-omic data of cells of the murine forebrain}.
+    \textbf{(A)} Mean neighbor score obtained when analyzing data from \citet{kremer_scnmt} with different methods.
+    Either 100~kb genomic tiles or VMRs were subjected to PCA and UMAP.
+    Tiles or VMRs were pre-processed as proposed in this work (shrunken mean of residuals subjected to iterative PCA, labeled "pre-processing: ours") or using conventional methods (PCA on lightly imputed methylation fractions, based on \citep{luo2017single}).
+    The full 1566-cell data set was sub-sampled to simulate smaller data sets (x axis).
+    A higher neighbor score implies better separation of cell types inferred from single-cell transcriptomes of the same cells \textbf{(B)}.
+    \textbf{(C, D)} Exemplary UMAPs obtained when analyzing the data set with our proposed methods (C) or conventional methods based on genome tiling (D).}
     \label{figure:score}
-    \end{center}
-\end{figure*}
+\end{figure}
 
@@
 We repeated this analysis several times, either using 100~kb tiles or VMRs, and either using methylation fractions or the shrunken means of residuals as a measure of methylation.
+
 For both types of regions, we varied the number of input regions, ranging from 100 to 50,000 regions with the highest sequencing coverage that were selected for PCA.
@@
 To visualize the different degrees of separation, we further reduced the PC space from both analyses with UMAP, showing that some neuronal subtypes such as mL5-1 and mDL-3 are separated from the other subtypes when using 1000 VMRs (Figure \ref{figure:score}C).
-% why not just plot PC1 vs. PC2 here? % because people will want to see the UMAP. -S
 Again, the difference is even more pronounced in the sub-sampled analysis with only 500 cells (Figure \ref{figure:score}D).
+
+\subsection{Identification of biologically relevant DMRs}
+
+\begin{figure*}[t]
+    \begin{center}
+        \includegraphics[width=.8\textwidth]{part_martina/NbvsOlig_DMRs.png}
+    \end{center}
+    \caption{\small \textbf{identified neuroblast and oligodendrocyte DMRs have a putative regulatory role:} \textbf{(A)} Difference in methylation between oligodendrocytes and neuroblasts of DMRs plotted against the number of cells contained in a DMR.
+Colored by p-value.
+\textbf{(B)} Mean methylation of cells in the DMR covering MBP.
+Line indicates median.
+\textbf{(C)} Mean methylation of reads at individual sites along the DMR covering MBP.
+\textbf{(D)} GREAT \citet{mclean2010great} was used for GO-term analysis.
+The left side shows GO-terms for genes near the 3000 most significant neuroblast DMRs sorted by GO-enrichment p-value and t-statistic (green arrow) and the right side for oligodendrocyte DMRs (yellow arrow) respectively.
+    }
+    \label{dmr_oligVSnb}
+\end{figure*}
+
+As a demonstration and to assess the ability of the presented approach to detect relevant DMRs, the method was used to find regions differing in CpG methylation between vSVZ neuroblasts and oligodendrocytes.
+Figure 7A gives an overview of all identified DMRs and the relation of the adjusted p-value to the number of cells included in a DMR, and the difference in methylation between the cell types.
+Logically, if a DMR contains few cells, the methylation difference between cell types must be higher to obtain a low (and thus more significant) p-value.
+A low p-value is maintained, although the methylation difference decreases if the number of cells in the DMR increases.
+
+An oligodendrocyte DMR stretching across Mbp was chosen to assess the methylation patterns of an individual DMR (Figure 5B-C).
+Mbp is highly expressed in oligodendrocytes as it is essential for the myelination of neuronal axons.
+The DMR is highly methylated in neuroblasts and lowly methylated in oligodendrocytes (Figure 7B-C), including more substantial demethylation in oligodendrocytes at the position of Mbp (Figure 5C).
+Low methylation at this locus supports the expression of Mbp.
+
+Nevertheless, this is not the only indicator of the biological relevance of discovered DMRs.
+GO-term enrichment shows that neuroblast DMRs occur near genes involved in the regulation of neurogenesis, neuron differentiation, and the positive regulation of neural precursor cell proliferation, all of which are essential for neuroblasts (Figure 5D).
+On the other side, oligodendrocyte-defining DMRs are located near genes with implications for myelin sheath, myelination, and axon ensheathment and thus are linked to oligodendrocytes.
+Therefore, DMRs occur in the proximity of genes associated with oligodendrocyte or neuroblast function, respectively and are implied to regulate gene expression and contribute to fundamental cell type-specific function, which is supported by the repressive effect of CG methylation on gene expression.
 
@@
         The resulting methylation matrix can then be used for downstream analysis such as cell clustering and dimensionality reduction.
+        Identified groups of cells (e.~g.
+clusters) can then be scanned for DMRs with \texttt{scbs diff}.
     }
@@
 This produces a BED file that lists the genome coordinates of VMRs.
+Similarly, the user can identify DMRs after quality control with \texttt{scbs diff} using a text file listing the cells’ group assignment as input.
+The command produces a BED file listing the coordinates of the determined DMRs together with the t- statistic, the corresponding cell group, and the adjusted p-value.
 To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED file can be used as input for \texttt{scbs matrix}, which quantifies methylation at genomic intervals in all cells.
@@
 
-For later use: We write $C$ for the set of all cells in the assay (i.e., $C$ is the index set for the cell indices $j$). Moreover,
+For later use: We write $C$ for the set of all cells in the assay (i.e., $C$ is the index set for the cell indices $j$).
+Moreover,
 we define $C_i\subset C$ as the set of all those cells $j$ that have reads covering position $i$,
@@
 
-We then run a kernel smoother over these per-position averages to obtain the smoothed averages $\tilde x_i$. Specifically, we use
+We then run a kernel smoother over these per-position averages to obtain the smoothed averages $\tilde x_i$.
+Specifically, we use
 \[ \tilde x_i = \frac{\sum_{i'} \overline x_{i'}\, k_h(d_{ii'})}{\sum_{i'} k_h(d_{ii'})},\]
@@
 \begin{equation} 
-r_{Ij} = \frac{\scriptsize{1}}{\scriptsize{|I\cap C_i|+1}}\sum_{i\in I\cap C_i}\left( x_{ij} - \tilde x_i \right). \label{avgres}
+r_{Ij} = \frac{\scriptsize{1}}{\scriptsize{|I\cap C_i|+1}}\sum_{i\in I\cap C_i}\left( x_{ij} - \tilde x_i \right).
+\label{avgres}
 \end{equation}
@@
 To select $n$ intervals for further analysis, the top $n$ intervals were picked based on the number of cells that the region was measured in.
-For dimensionality reduction, missing values were imputed with the iterative PCA described in the previous section. To visualize data, UMAP was performed on the final 15 principal components.
+
+For dimensionality reduction, missing values were imputed with the iterative PCA described in the previous section.
+To visualize data, UMAP was performed on the final 15 principal components.
 

### 2023-03-09 paper.tex
fix crash and some warnings

@@
 \setlength{\columnsep}{0.7cm}
-\usepackage{microtype} % nice typesetting  % doesn't work for me :-( -S
+%\usepackage{microtype} % nice typesetting  % doesn't work for me :-( -S
 \usepackage{amsmath}
@@
 
+\phantomsection  % idk why we need this to silence hyperref's warning
 \subsection{Standard approach}
@@
 
+\phantomsection
 \subsection{Read-position aware quantitation} \label{residuals}
@@
 Windows with above-threshold variance are merged if they overlap, yielding the ``variably methylated regions''.}
-    \label{vmr}
+    \label{fig:vmr}
 \end{figure}
@@
 
-Therefore, we propose the following approach (Fig.~\ref{vmr}): Divide up the chromosome into many \emph{overlapping} windows, that start at regular multiples of a fixed, small, step size.
+Therefore, we propose the following approach (Fig.~\ref{fig:vmr}): Divide up the chromosome into many \emph{overlapping} windows, that start at regular multiples of a fixed, small, step size.
 Quantify the methylation of each cell in each window by averaging the cell's methylation residuals over all CpGs in the window, as described above and depicted in Fig.~\ref{fig:smoothres}B.
@@
     Windows exceeding either threshold are merged if they overlap, yielding the ``differentially methylated regions``.}
-    \label{dmr}
+    \label{fig:dmrscan}
 \end{figure}
@@
 However, to date no approach to detect DMRs in scBS data has been reported.
-To enable DMR detection in scBS data, we thus developed an approach that detects DMRs of variable size between two groups of cells, and controls the false discovery rate (Fig.~3):
+To enable DMR detection in scBS data, we thus developed an approach that detects DMRs of variable size between two groups of cells, and controls the false discovery rate (Fig.~\ref{fig:dmrscan}):
 
-Similar to the previously described VMR identification approach (Fig.~\ref{vmr}), divide each chromosome into overlapping windows shifted by a small and fixed size (step size) and quantify the methylation of each cell in each window as described (Fig.~\ref{residuals}b).
+Similar to the previously described VMR identification approach (Fig.~\ref{fig:vmr}), divide each chromosome into overlapping windows shifted by a small and fixed size (step size) and quantify the methylation of each cell in each window as described (Fig.~\ref{fig:smoothres}b).
 Next, instead of calculating the variance, obtain the t-statistic % Welch t-test
@@
 
+\phantomsection
 \subsection{Correlation with gene expression}
@@
 
-In Figure \ref{figure:correlation}A, we compare correlations of gene expression and promoter methylation to correlations of gene expression and methylation of the closest VMR identified with our approach.
+In Figure \ref{fig:correlation}A, we compare correlations of gene expression and promoter methylation to correlations of gene expression and methylation of the closest VMR identified with our approach.
 This analysis revealed a number of genes whose expression does not correlate with promoter methylation, but correlates with VMR methylation.
-This effect is also observed when using our proposed measure of methylation, the shrunken mean of residuals (Figure \ref{figure:correlation}B).
+This effect is also observed when using our proposed measure of methylation, the shrunken mean of residuals (Figure \ref{fig:correlation}B).
 
@@
     Data from \citet{kremer_scnmt}.}
-    \label{figure:correlation}
+    \label{fig:correlation}
 \end{figure}
 
-To illustrate this effect with an example, we picked a gene where VMR methylation is a much better predictor of gene expression than promoter methylation (Ube2l6, highlighted in orange in Figure \ref{figure:correlation}A-B).
-Dividing all cells into four groups according to their level of Ube2l6 expression, it is clearly visible that the promoter of this gene is lowly methylated regardless of gene expression, while methylation at the nearest VMR is greatly decreased in cells with high Ube2l6 expression (Figure \ref{figure:correlation}C).
+To illustrate this effect with an example, we picked a gene where VMR methylation is a much better predictor of gene expression than promoter methylation (Ube2l6, highlighted in orange in Figure \ref{fig:correlation}A-B).
+Dividing all cells into four groups according to their level of Ube2l6 expression, it is clearly visible that the promoter of this gene is lowly methylated regardless of gene expression, while methylation at the nearest VMR is greatly decreased in cells with high Ube2l6 expression (Figure \ref{fig:correlation}C).
 
@@
     \textbf{(C, D)} Exemplary UMAPs obtained when analyzing the data set with our proposed methods (C) or conventional methods based on genome tiling (D).}
-    \label{figure:score}
+    \label{fig:score}
 \end{figure}
@@
 We used the original assignment of cells to cell types from \citet{luo2017single}.
-The results demonstrate that the 16 neuronal subtypes are more cleanly separated when using our proposed methods (Figure \ref{figure:score}A).
+The results demonstrate that the 16 neuronal subtypes are more cleanly separated when using our proposed methods (Figure \ref{fig:score}A).
 Most of this gain in performance can be attributed to the use of VMRs over genomic tiles.
@@
 To simulate a smaller data set, we subsampled the frontal cortex data set to 500 cells and repeated our analysis.
-Here, our approach keeps its good score, while the simple approach falls behind (Figure \ref{figure:score}B).
+Here, our approach keeps its good score, while the simple approach falls behind (Figure \ref{fig:score}B).
 
@@
 In contrast, the neighbor count score for the top 1000 genomic 100~kb tiles is much lower.
-To visualize the different degrees of separation, we further reduced the PC space from both analyses with UMAP, showing that some neuronal subtypes such as mL5-1 and mDL-3 are separated from the other subtypes when using 1000 VMRs (Figure \ref{figure:score}C).
-Again, the difference is even more pronounced in the sub-sampled analysis with only 500 cells (Figure \ref{figure:score}D).
+To visualize the different degrees of separation, we further reduced the PC space from both analyses with UMAP, showing that some neuronal subtypes such as mL5-1 and mDL-3 are separated from the other subtypes when using 1000 VMRs (Figure \ref{fig:score}C).
+Again, the difference is even more pronounced in the sub-sampled analysis with only 500 cells (Figure \ref{fig:score}D).
 
@@
     \begin{center}
-        \includegraphics[width=.8\textwidth]{part_martina/NbvsOlig_DMRs.png}
+        \includegraphics[width=.8\textwidth]{figures/Fig_DMRs.png}
     \end{center}
@@
     }
-    \label{dmr_oligVSnb}
+    \label{fig:dmr}
 \end{figure*}
@@
 As a demonstration and to assess the ability of the presented approach to detect relevant DMRs, the method was used to find regions differing in CpG methylation between vSVZ neuroblasts and oligodendrocytes.
-Figure 7A gives an overview of all identified DMRs and the relation of the adjusted p-value to the number of cells included in a DMR, and the difference in methylation between the cell types.
+Fig.~\ref{fig:dmr} gives an overview of all identified DMRs and the relation of the adjusted p-value to the number of cells included in a DMR, and the difference in methylation between the cell types.
 Logically, if a DMR contains few cells, the methylation difference between cell types must be higher to obtain a low (and thus more significant) p-value.
@@
 
-An oligodendrocyte DMR stretching across Mbp was chosen to assess the methylation patterns of an individual DMR (Figure 5B-C).
+An oligodendrocyte DMR stretching across Mbp was chosen to assess the methylation patterns of an individual DMR (Fig.~\ref{fig:dmr}B-C).
 Mbp is highly expressed in oligodendrocytes as it is essential for the myelination of neuronal axons.
-The DMR is highly methylated in neuroblasts and lowly methylated in oligodendrocytes (Figure 7B-C), including more substantial demethylation in oligodendrocytes at the position of Mbp (Figure 5C).
+The DMR is highly methylated in neuroblasts and lowly methylated in oligodendrocytes (Fig.~\ref{fig:dmr}B-C), including more substantial demethylation in oligodendrocytes at the position of Mbp (Fig.~\ref{fig:dmr}C).
 Low methylation at this locus supports the expression of Mbp.
@@
 Nevertheless, this is not the only indicator of the biological relevance of discovered DMRs.
-GO-term enrichment shows that neuroblast DMRs occur near genes involved in the regulation of neurogenesis, neuron differentiation, and the positive regulation of neural precursor cell proliferation, all of which are essential for neuroblasts (Figure 5D).
+GO-term enrichment shows that neuroblast DMRs occur near genes involved in the regulation of neurogenesis, neuron differentiation, and the positive regulation of neural precursor cell proliferation, all of which are essential for neuroblasts (Fig.~\ref{fig:dmr}D).
 On the other side, oligodendrocyte-defining DMRs are located near genes with implications for myelin sheath, myelination, and axon ensheathment and thus are linked to oligodendrocytes.
@@
         The resulting methylation matrix can then be used for downstream analysis such as cell clustering and dimensionality reduction.
-        Identified groups of cells (e.~g.
-clusters) can then be scanned for DMRs with \texttt{scbs diff}.
+        Identified groups of cells (e.~g. clusters) can then be scanned for DMRs with \texttt{scbs diff}.
     }
-    \label{workflow}
+    \label{fig:workflow}
 \end{figure*}
 
-We have implemented the approach just described in a Python package with a command line interface, called \texttt{scbs}, which also offers a number of other functionalities for the analysis of scBS data (Figure \ref{workflow}).
+We have implemented the approach just described in a Python package with a command line interface, called \texttt{scbs}, which also offers a number of other functionalities for the analysis of scBS data (Figure \ref{fig:workflow}).
 The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark} or methylpy \citep{methylpy}.
@@
 
+\phantomsection
 \subsection{Acknowledgments}
@@
 
+\phantomsection
 \section{Methods}
 
+\phantomsection
 \subsection{Raw data}

### 2023-03-09 paper.tex
push minor stuff before lunch

@@
 
-The average thus obtained is what we use to quantify this cell's (relative) methylation in the interval.
+The average thus obtained, the shrunken mean of the residuals, is what we use to quantify this cell's (relative) methylation in the interval.
 For a genome tiled into such intervals, we thus obtain a matrix, one row per cell, one column per interval, that can be used for downstream analysis, e.g., as input for PCA.
@@
 % remove or shorten this section cause it's not really novel, methods like this are already implemented in e.g. the "missMDA" R package
-\subsection{PCA with imputation}
+%\subsection{PCA with imputation}
+%The standard PCA algorithm cannot deal with missing data.
+%Therefore, one would need to make intervals wide enough to ensure that every interval is covered by at least one read for most tiles and discard any interval that is not covered by any single cell.
+%Hence, a naive analysis has to resort to a coarse tiling of the genome.
+%To counter this, we propose a simple and straightforward way to deal with missing data in the input matrix to a PCA: In a first iteration, we replace each missing value in the centered input matrix with zero, then run the PCA.
+%Then, only these zeroes are replaced by the value predicted by the PCA and the PCA is rerun.
+%Convergence is typically achieved after two to five iterations.
+%For details, see Methods.
+%The possibility to perform PCA with missing data allows us to use shorter, and thus more intervals and hence provide the PCA with richer input data.
 
-The standard PCA algorithm cannot deal with missing data.
-Therefore, one would need to make intervals wide enough to ensure that every interval is covered by at least one read for most tiles and discard any interval that is not covered by any single cell.
-Hence, a naive analysis has to resort to a coarse tiling of the genome.
-To counter this, we propose a simple and straightforward way to deal with missing data in the input matrix to a PCA: In a first iteration, we replace each missing value in the centered input matrix with zero, then run the PCA.
-Then, only these zeroes are replaced by the value predicted by the PCA and the PCA is rerun.
-Convergence is typically achieved after two to five iterations.
-For details, see Methods.
-
-The possibility to perform PCA with missing data allows us to use shorter, and thus more intervals and hence provide the PCA with richer input data.
-
-\subsection{Finding variably methylated regions (VMRs)}
+\subsection{Finding variably methylated regions}
 
@@
 
-\subsection{Finding differentially methylated regions (DMRs)}
+\subsection{Finding differentially methylated regions}
 
@@
 
-Similar to the previously described VMR identification approach (Fig.~\ref{fig:vmr}), divide each chromosome into overlapping windows shifted by a small and fixed size (step size) and quantify the methylation of each cell in each window as described (Fig.~\ref{fig:smoothres}b).
-Next, instead of calculating the variance, obtain the t-statistic % Welch t-test
+Similar to the previously described VMR identification approach (Fig.~\ref{fig:vmr}), divide each chromosome into overlapping windows shifted by a small and fixed size (step size) and quantify the methylation of each cell in each window.
+Next, instead of the variance, obtain the t-statistic % Welch t-test
 as a measure of differential methylation between the two cell groups.
-Mark the windows with the most extreme t-statistics (e.g. 2\% upper and lower tails) as DMRs.
-Distinguishing between high and low t-statistics, merge any overlapping marked windows into larger DMRs.
-Re-calculate the t-statistic for each larger DMR.
+Identify the windows with the most extreme t-statistics, e.g. windows in the 2\% upper and lower tails.
+Merge any overlapping windows in the upper tail into larger DMRs.
+Repeat for windows in the lower tail, then re-calculate the t-statistic for each larger DMR.
 
-To assess statistical significance of DMRs, we repeat the same procedure on permutations of the scBS data.
+To assess statistical significance of DMRs, repeat the same procedure on permutations of the scBS data, i.e. the same data set with randomly shuffled cell labels.
 The DMR t-statistics obtained from permuted data are then used to estimate the false-discovery rate (FDR), yielding an adjusted p-value for each DMR.
 
-\subsection{???}  % guess we have to rewrite this closing paragraph a little bit, to include both DMRs and VMRs
+\subsection{Utility of DMRs and VMRs} % not sure if we need a subsection here
 
-Besides providing better input for the distance calculations, the small size of identified VMRs also means that they can be readily compared with genomic annotation, which aids in their functional interpretation.
-While a large 100~kb tile will often contain multiple genes with distinct expression patterns, small VMRs located in known regulatory elements or gene bodies are more easily associated with individual genes and their associated function.
+While the primary purpose of VMRs is to provide better input for the distance calculations, their small size also means that they can be readily compared with genomic annotation, which aids in their functional interpretation.
+Large 100~kb tiles will often contain multiple genes with distinct expression patterns, small VMRs located in known regulatory elements or gene bodies are more easily associated with individual genes and their associated function.
 Thus, VMRs provide a useful starting point for examining the data with methods from functional genomics.
+To enable a more targeted characterization of 
+The same principles may also be applied to DMRs, thus enabling in-depth characterization of epigenomic and gene regulatory differences between tissues, cell types or conditions.
 
-\section{Benchmarks} % re-name this to something more general? Application and benchmarks or something?
+\section{Application and benchmarks}
 
@@
 \phantomsection
-\subsection{Correlation with gene expression}
+\subsection{Correlation of VMR methylation with gene expression}
+
+The first data set used to test our methods is a single-cell multi-omics data set generated by \citet{kremer_scnmt}.
+This data set comprises the methylomes of 1566 cells isolated from mouse forebrains, as well as single-cell transcriptomes of the same cells.
 
@@
     \textbf{(A)} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red).
-    \textbf{(B)} Correlation of each gene's RNA expression values to methylation (mean shrunken residuals) at promoters (y-axis), or to methylation of the nearest VMR (x-axis).
+    \textbf{(B)} Correlation of each gene's RNA expression values to methylation (shrunken mean of residuals) at promoters (y-axis), or to methylation of the nearest VMR (x-axis).
     Promoters are defined as intervals \textpm2~kb around the TSS.

### 2023-03-09 paper.tex
write about benchmarking

@@
 Thus, VMRs provide a useful starting point for examining the data with methods from functional genomics.
-To enable a more targeted characterization of 
-The same principles may also be applied to DMRs, thus enabling in-depth characterization of epigenomic and gene regulatory differences between tissues, cell types or conditions.
+To enable a targeted characterization of epigenomic and gene regulatory differences between conditions or cell types, the same principles may also be applied to DMRs.
 
@@
 
-To demonstrate the value of our proposed improvements, we obtained two previously published scBS data sets and re-analyzed them.
+To demonstrate the value of our proposed improvements, we obtained two previously published scBS data sets and re-analyzed them using our methods.
 
 \phantomsection
-\subsection{Correlation of VMR methylation with gene expression}
+\subsection{Correlating VMR methylation with gene expression}
 
 The first data set used to test our methods is a single-cell multi-omics data set generated by \citet{kremer_scnmt}.
-This data set comprises the methylomes of 1566 cells isolated from mouse forebrains, as well as single-cell transcriptomes of the same cells.
-
-The first data set used for benchmarking is a single-cell multi-omics data set generated by \citet{argelaguet2019gastru}.
-In their work, the authors obtained single-cell methylomes of mouse embryo cells across different stages of development (E4.5--E7.5) as well as single-cell transcriptomes from the same cells.
-In one of their analyses, the authors calculated the correlation between gene expression and promoter methylation to elucidate the implications of the epigenome on gene expression in development.
-The authors identified several genes where promoter methylation predicts gene expression, but obtained low correlations for the vast majority of genes.
-We reasoned that VMRs obtained with the previously described sliding window approach might be a better predictor of gene expression.
-
-In Figure \ref{fig:correlation}A, we compare correlations of gene expression and promoter methylation to correlations of gene expression and methylation of the closest VMR identified with our approach.
-This analysis revealed a number of genes whose expression does not correlate with promoter methylation, but correlates with VMR methylation.
-This effect is also observed when using our proposed measure of methylation, the shrunken mean of residuals (Figure \ref{fig:correlation}B).
+This data set comprises the methylomes of 1566 cells isolated from mouse forebrains, as well as matched single-cell transcriptomes of the same cells.
+Among these cells are distinct cell types such as oligodendrocytes, oligodendrocyte precursor cells (OPCs) and endothelial cells, as well as cellular sub-states that are part of the continuous neural stem cell differentiation trajectory.
+To assess whether our VMR detection method captures genomic intervals that are biologically meaningful, we probed whether their methylation level correlates with the expression of nearby genes.
+Compared to promoter methylation, the obtained correlation coefficients are more extreme, indicating that VMR methylation often is a better predictor of gene expression than promoter methylation (Fig.~\ref{fig:correlation}A).
+Indeed, a gene-wise comparison revealed many genes whose expression is correlated with the nearest VMR, but not with promoter methylation (Fig.~\ref{fig:correlation}B).
+One such example gene, Htra1, is depicted in Fig.~\ref{fig:correlation}C.
+While the promoter of this gene is lowly methylated regardless of gene expression, a VMR located downstream of the promoter is lowly methylated in cells with high Htra1 expression.
 
@@
 
-To illustrate this effect with an example, we picked a gene where VMR methylation is a much better predictor of gene expression than promoter methylation (Ube2l6, highlighted in orange in Figure \ref{fig:correlation}A-B).
-Dividing all cells into four groups according to their level of Ube2l6 expression, it is clearly visible that the promoter of this gene is lowly methylated regardless of gene expression, while methylation at the nearest VMR is greatly decreased in cells with high Ube2l6 expression (Figure \ref{fig:correlation}C).
-
-\subsection{Identification of cellular sub-types}
+\subsection{Effects of our methods on cell type identification}
 
@@
     \caption{\small \textbf{Benchmark of our methods on single-cell multi-omic data of cells of the murine forebrain}.
-    \textbf{(A)} Mean neighbor score obtained when analyzing data from \citet{kremer_scnmt} with different methods.
+    \textbf{(A)} Mean neighbor score obtained when analyzing data from \citet{kremer_scnmt} with different combinations of methods.
     Either 100~kb genomic tiles or VMRs were subjected to PCA and UMAP.
@@
 
-To assess the impact of our methods on dimensionality reduction and the ability to distinguish different (sub)types of cells, we used a data set from \citet{luo2017single}.
-This data set consists of 3377 single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal subtypes.
-We subjected this data set to the standard scBS workflow: defining a set of genomic intervals, quantifying methylation in these intervals, and subjecting the resulting matrix to PCA.
-We repeated this analysis several times, either using 100~kb tiles or VMRs, and either using methylation fractions or the shrunken means of residuals as a measure of methylation.
+To assess the impact of our proposed methods, we next tested whether our methods improve the ability to distinguish cell types and cell states (Fig.~\ref{fig:score}).
+To this end, we obtained cluster labels based on the single-cell transcriptomes from \citet{kremer_scnmt} (Fig.~\ref{fig:score}B).
+We consider these cell labels as ground truth and next tested whether we are able to distinguish the same groups of cells based on their methylomes.
+To do this, we subjected the methylomes to the standard scBS workflow:
+defining a set of genomic intervals, quantifying methylation in these intervals, and subjecting the resulting matrix to PCA.
+To determine genomic features for distance calculation, we either tiled the genome into 100~kb bins as is commonly done, or we used our VMR detection approach.
+To quantify methylation in these intervals, we either averaged in these intervals, obtaining methylation percentages ("classic" pre-processing), or we calculated the shrunken mean of residuals.
+Visual inspection of the resulting UMAPs revealed that our proposed combination of methods results in more clearly separated cell types, compared to a UMAP obtained with default analysis methods (Fig.~\ref{fig:score}C,D).
+While all cells form a continuous point cloud when using default methods, our improvements led to a clear separation of oligodendrocytes, OPCs and endothelial cells.
+Furthermore, even cellular sub-types of cells in the continuous neural stem cell lineage were partially separated.
 
-For both types of regions, we varied the number of input regions, ranging from 100 to 50,000 regions with the highest sequencing coverage that were selected for PCA.
-
-To assess performance of each combination of methods, we used a nearest neighbor count score (see Methods) that quantifies how well the different neuronal subtypes are separated in 15 dimensional PC space.
-We used the original assignment of cells to cell types from \citet{luo2017single}.
-The results demonstrate that the 16 neuronal subtypes are more cleanly separated when using our proposed methods (Figure \ref{fig:score}A).
-Most of this gain in performance can be attributed to the use of VMRs over genomic tiles.
-The use of the shrunken means of residuals over methylation fractions provides an additional small improvement.
-
-Since scBS protocols are still rather costly and labor-intensive, only few laboratories are currently able to obtain thousands of single cell methylomes.
-To simulate a smaller data set, we subsampled the frontal cortex data set to 500 cells and repeated our analysis.
-Here, our approach keeps its good score, while the simple approach falls behind (Figure \ref{fig:score}B).
-
-Of note, our results show that even few VMRs (here, the 1000 VMRs that have read coverage in the highest number of cells) already contain sufficient information to distinguish neuronal subtypes.
-In contrast, the neighbor count score for the top 1000 genomic 100~kb tiles is much lower.
-To visualize the different degrees of separation, we further reduced the PC space from both analyses with UMAP, showing that some neuronal subtypes such as mL5-1 and mDL-3 are separated from the other subtypes when using 1000 VMRs (Figure \ref{fig:score}C).
-Again, the difference is even more pronounced in the sub-sampled analysis with only 500 cells (Figure \ref{fig:score}D).
+To quantify this performance gain in a more rigorious manner, we used a score that quantifies whether cells were placed in a neighborhood consisting of cells with the same cell type in 15-dimensional PC space ("neighbor score", see \ref{methods:score}).
+Since scBS protocols are costly and labor-intensive, only few laboratories are currently able to obtain thousands of single cell methylomes.
+To simulate smaller data sets, we thus sub-sampled the 1566-cell data set into smaller data sets, analyzed them with different method combinations, and calculated the mean neighbor score for each (Figure \ref{fig:score}A).
+The results demonstrate that the forebrain cell clusters are more cleanly separated when using our proposed methods.
+This performance gain was observed also in smaller data sets, although cell cluster separation generally becomes more difficult in those.
+Both the use of VMRs instead of genomic tiles, as well as the use of the mean shrunken residuals over methylation averages improved cell cluster separation.
 

### 2023-03-09 paper.tex

@@
 As we demonstrate below, a PCA performed on such a matrix provided a distance metric for the cells that contains more information on biological detail than one from a simpler analysis.
+Since not all genomic intervals have sequencing coverage in all cells, the resulting matrix contains missing values.
+To enable PCA in the presence of missing values, we propose to use a simple iterative approach (for details see Methods).
 
@@
 To assess the impact of our proposed methods, we next tested whether our methods improve the ability to distinguish cell types and cell states (Fig.~\ref{fig:score}).
-To this end, we obtained cluster labels based on the single-cell transcriptomes from \citet{kremer_scnmt} (Fig.~\ref{fig:score}B).
+To this end, we obtained cluster labels based on the single-cell transcriptomes from \citet{kremer_scnmt} (Fig.~\ref{fig:score}A).
 We consider these cell labels as ground truth and next tested whether we are able to distinguish the same groups of cells based on their methylomes.
@@
 To quantify methylation in these intervals, we either averaged in these intervals, obtaining methylation percentages ("classic" pre-processing), or we calculated the shrunken mean of residuals.
-Visual inspection of the resulting UMAPs revealed that our proposed combination of methods results in more clearly separated cell types, compared to a UMAP obtained with default analysis methods (Fig.~\ref{fig:score}C,D).
+Visual inspection of the resulting UMAPs revealed that our proposed combination of methods results in more clearly separated cell types, compared to a UMAP obtained with default analysis methods (Fig.~\ref{fig:score}B,C).
 While all cells form a continuous point cloud when using default methods, our improvements led to a clear separation of oligodendrocytes, OPCs and endothelial cells.
@@
 
-To quantify this performance gain in a more rigorious manner, we used a score that quantifies whether cells were placed in a neighborhood consisting of cells with the same cell type in 15-dimensional PC space ("neighbor score", see \ref{methods:score}).
+To quantify this performance gain in a more rigorous manner, we used a score that quantifies whether cells were placed in a neighborhood consisting of cells with the same cell type in 15-dimensional PC space ("neighbor score", for details see Methods).
 Since scBS protocols are costly and labor-intensive, only few laboratories are currently able to obtain thousands of single cell methylomes.
-To simulate smaller data sets, we thus sub-sampled the 1566-cell data set into smaller data sets, analyzed them with different method combinations, and calculated the mean neighbor score for each (Figure \ref{fig:score}A).
+To simulate smaller data sets, we sub-sampled the 1566-cell data set into smaller data sets, analyzed them with different method combinations, and calculated the mean neighbor score for each (Figure \ref{fig:score}D,E).
 The results demonstrate that the forebrain cell clusters are more cleanly separated when using our proposed methods.
@@
 Both the use of VMRs instead of genomic tiles, as well as the use of the mean shrunken residuals over methylation averages improved cell cluster separation.
+
+We repeated this benchmark on a second data set published by \citet{luo2017single}.
+This data set consists of 3377 single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal subtypes.
+The authors annotated cell types based on CH-methylation instead of CpG-methylation, and we used these labels as ground truth (\ref{fig:score_luo}A).
 
@@
 
-\subsection{Nearest neighbor count score}
+\subsection{Nearest neighbor count score} \label{methods:score}
 To assess performance of our methods, we processed mouse CpG methylation data from \citet{luo2017single} with several method combinations and scored each result.

### 2023-03-09 paper.tex
add supp fig

@@
 
-\begin{figure}
+\begin{figure*}[t]
     \begin{center}
-        \includegraphics[width=\columnwidth]{figures/Fig_benchmark.png}
+        \includegraphics[width=0.8\textwidth]{figures/Fig_benchmark.png}
     \end{center}
     \caption{\small \textbf{Benchmark of our methods on single-cell multi-omic data of cells of the murine forebrain}.
-    \textbf{(A)} Mean neighbor score obtained when analyzing data from \citet{kremer_scnmt} with different combinations of methods.
+    \textbf{(A)} Cell labels based on clustering of single-cell transcriptomes from \citet{kremer_scnmt}.
+    \textbf{(B, C)} Exemplary UMAPs obtained when analyzing the data set with our proposed methods (B) or conventional methods based on genome tiling (C).
+    \textbf{(D, E)} Mean neighbor score \textbf{(D)} obtained after analysing single-cell methylomes with different combinations of methods.
     Either 100~kb genomic tiles or VMRs were subjected to PCA and UMAP.
-    Tiles or VMRs were pre-processed as proposed in this work (shrunken mean of residuals subjected to iterative PCA, labeled "pre-processing: ours") or using conventional methods (PCA on lightly imputed methylation fractions, based on \citep{luo2017single}).
+    Tiles or VMRs were pre-processed as proposed in this work (shrunken mean of residuals subjected to iterative PCA, labeled "pre-processing: ours") or using conventional methods (PCA on lightly imputed methylation percentages, based on \citep{luo2017single}).
     The full 1566-cell data set was sub-sampled to simulate smaller data sets (x axis).
-    A higher neighbor score implies better separation of cell types inferred from single-cell transcriptomes of the same cells \textbf{(B)}.
-    \textbf{(C, D)} Exemplary UMAPs obtained when analyzing the data set with our proposed methods (C) or conventional methods based on genome tiling (D).}
+    A higher neighbor score implies better separation of cell types inferred from single-cell transcriptomes of the same cells \textbf{(A)}.
+    }
     \label{fig:score}
-\end{figure}
+\end{figure*}
 
@@
 The authors annotated cell types based on CH-methylation instead of CpG-methylation, and we used these labels as ground truth (\ref{fig:score_luo}A).
+
+\begin{figure}
+    \begin{center}
+        \includegraphics[width=0.9\columnwidth]{figures/SFig_benchmark.png}
+    \end{center}
+    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of neural subtypes of the murine frontal cortex}.
+    Our methods were benchmarked on single-cell methylomes from \citet{luo2017single} as described in Fig.~\ref{fig:score}.
+    }
+    \label{fig:score_luo}
+\end{figure}
 

### 2023-03-10 paper.tex
add note

@@
 This data set consists of 3377 single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal subtypes.
-The authors annotated cell types based on CH-methylation instead of CpG-methylation, and we used these labels as ground truth (\ref{fig:score_luo}A).
+The authors annotated cell types based on CH-methylation instead of CpG-methylation.
+Using these labels as ground truth we re-analyzed the full and sub-sampled data set with four methods combinations and found that our suggested methods consistently outperformed default methods (Fig.~\ref{fig:score_luo}).
 
-\begin{figure}
-    \begin{center}
-        \includegraphics[width=0.9\columnwidth]{figures/SFig_benchmark.png}
-    \end{center}
-    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of neural subtypes of the murine frontal cortex}.
-    Our methods were benchmarked on single-cell methylomes from \citet{luo2017single} as described in Fig.~\ref{fig:score}.
-    }
-    \label{fig:score_luo}
-\end{figure}
+\section{note to Simon: this is how far I got}   % todo rm
 
-\subsection{Identification of biologically relevant DMRs}
+\phantomsection
+\subsection{Detecting DMRs between oligodendrocytes and neuroblasts}
 
@@
 
-As a demonstration and to assess the ability of the presented approach to detect relevant DMRs, the method was used to find regions differing in CpG methylation between vSVZ neuroblasts and oligodendrocytes.
+Next, we used the single-cell multi-omics data set \citep{kremer_scnmt} to assess the ability of our DMR detection approach.
+We first selected two cell populations from the murine ventricular-subventricular zone, neuroblasts and oligodendrocytes, and then used the method depicted in Fig.~\ref{fig:dmrscan} to identify regions with differing methylation.
 Fig.~\ref{fig:dmr} gives an overview of all identified DMRs and the relation of the adjusted p-value to the number of cells included in a DMR, and the difference in methylation between the cell types.
@@
 In conclusion, we have presented powerful improvements to scBS data preprocessing and a software tool that implements these and bridges a gap in existing tool chains.
+
+\section{Supplementary Figures}
+
+\setcounter{figure}{0}
+%\renewcommand{\figurename}{Supplementary Figure}
+\renewcommand{\thefigure}{S\arabic{figure}}
+
+\begin{figure}
+    \begin{center}
+        \includegraphics[width=0.9\columnwidth]{figures/SFig_benchmark.png}
+    \end{center}
+    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of neural subtypes of the murine frontal cortex}.
+    Our methods were benchmarked on single-cell methylomes from \citet{luo2017single} as described in Fig.~\ref{fig:score}.
+    }
+    \label{fig:score_luo}
+\end{figure}
 

### 2023-03-14 paper.tex
Updates from Overleaf
minor stuff

@@
     \end{center}
-    \caption{\small \textbf{Improved quantification of DNA methylation in genomic intervals:}
+    \caption{\small \textbf{Improved quantification of DNA methylation in genomic intervals.}\\
     \textbf{(A)} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified.
@@
 \begin{figure}
-    \includegraphics[width=\columnwidth]{figures/Fig_sliding.png}
-    \caption{\small \textbf{Finding variably methylated regions (VMRs):} The chromosomes are divided up into overlapping windows (first five shown at the bottom), and for each window, the cells' methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
+    \begin{center}
+    \includegraphics[width=0.85\columnwidth]{figures/Fig_sliding.png}
+    \end{center}
+    \caption{\small \textbf{Finding variably methylated regions.}\\
+    The chromosomes are divided up into overlapping windows (first five shown at the bottom), and for each window, the cells' methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
     Then, the variance of these values is calculated (each point represents one of the overlapping windows), a threshold (dashed line) is chosen such that a chosen quantile of windows have a variance exceeding the threshold.
-    Windows with above-threshold variance are merged if they overlap, yielding the ``variably methylated regions''.}
+    Windows with above-threshold variance are merged if they overlap, yielding the ``variably methylated regions'' (VMRs).}
     \label{fig:vmr}
@@
     \includegraphics[width=\columnwidth]{figures/Fig_DMRdetection.png}
-    \caption{\small \textbf{Identification of differentially methylated regions (DMRs):}
+    \caption{\small \textbf{Identification of differentially methylated regions.}\\
     The chromosome is divided into windows overlapping by a small and fixed size.
-    For each window, the cell’s methylation values are calculated as described and as depicted in Fig.~ \ref{fig:smoothres}B.
+    For each window, the cell’s methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
     Then the t-statistic of these values is obtained as a measure of methylation difference (each point represents one of the windows).
     Upper and lower thresholds (dashed lines) are chosen such that a defined quantile of windows have a t-statistic above or below the threshold.
-    Windows exceeding either threshold are merged if they overlap, yielding the ``differentially methylated regions``.}
+    Windows exceeding either threshold are merged if they overlap, yielding the differentially methylated regions (DMRs).}
     \label{fig:dmrscan}
@@
     \end{center}
-    \caption{\small \textbf{Correlation of DNA methylation and gene expression}.
+    \caption{\small \textbf{Correlation of DNA methylation and gene expression}.\\
     \textbf{(A)} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red).
@@
 
-\subsection{Effects of our methods on cell type identification}
+\subsection{Our proposed methods improve identification of cell types}
 
@@
     \end{center}
-    \caption{\small \textbf{Benchmark of our methods on single-cell multi-omic data of cells of the murine forebrain}.
+    \caption{\small \textbf{Benchmark of our methods on single-cell multi-omic data of cells of the murine forebrain}.\\
     \textbf{(A)} Cell labels based on clustering of single-cell transcriptomes from \citet{kremer_scnmt}.
@@
     \begin{center}
-        \includegraphics[width=.8\textwidth]{figures/Fig_DMRs.png}
+        \includegraphics[width=.7\textwidth]{figures/Fig_DMRs.png}
     \end{center}
-    \caption{\small \textbf{identified neuroblast and oligodendrocyte DMRs have a putative regulatory role:} \textbf{(A)} Difference in methylation between oligodendrocytes and neuroblasts of DMRs plotted against the number of cells contained in a DMR.
-Colored by p-value.
-\textbf{(B)} Mean methylation of cells in the DMR covering MBP.
-Line indicates median.
-\textbf{(C)} Mean methylation of reads at individual sites along the DMR covering MBP.
-\textbf{(D)} GREAT \citet{mclean2010great} was used for GO-term analysis.
-The left side shows GO-terms for genes near the 3000 most significant neuroblast DMRs sorted by GO-enrichment p-value and t-statistic (green arrow) and the right side for oligodendrocyte DMRs (yellow arrow) respectively.
+    \caption{\small \textbf{Detection of differentially methylated regions between neural stem cells and oligodendrocytes.}\\
+    \textbf{(A)} DMRs detected between neural stem cells (NSCs) and oligodendrocytes (ODs) from \citep{kremer_scnmt} (top) and DMRs detected in the same data with randomly permuted cell labels (bottom, used to estimate FDR and determine adjusted p-values).
+    \textbf{(B)} Enrichment of GO terms associated with DMRs lowly methylated in ODs (top) or NSCs (bottom).
+    Depicted are the top 5 GO terms of the "Biological Process" (GO:BP) and "Cellular Component" (GO:CC) GO category, and their binomial test p-value and enrichment, as reported by GREAT \citep{mclean2010great}.
+    \textbf{(C)} Mean methylation of NSCs and ODs at an exemplary DMR.
+    Each point corresponds to a cell, black lines denote the median.
+    \textbf{(D)} Detailed view of the DMR (yellow dashed rectangle) from \textbf{C} in pseudo-bulk samples consisting of NSCs or ODs.
+    Vertical bars represent CpG sites.
     }
@@
     \end{center}
-    \caption{\small \textbf{Overview of the functionalities implemented in the \texttt{scbs} package.}
+    \caption{\small \textbf{Overview of the functionalities implemented in the \texttt{scbs} package.}\\
         \texttt{scbs prepare} parses methylation files produced by common bisulfite sequencing mappers and stores their contents in a compressed format optimised for efficient access to genomic intervals.
@@
 
-\begin{figure}
+\begin{figure}[h]
     \begin{center}
@@
     \end{center}
-    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of neural subtypes of the murine frontal cortex}.
+    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of neural subtypes of the murine frontal cortex}.\\
     Our methods were benchmarked on single-cell methylomes from \citet{luo2017single} as described in Fig.~\ref{fig:score}.

### 2023-03-14 paper.tex
describe dmr stuff

@@
     Either 100~kb genomic tiles or VMRs were subjected to PCA and UMAP.
-    Tiles or VMRs were pre-processed as proposed in this work (shrunken mean of residuals subjected to iterative PCA, labeled "pre-processing: ours") or using conventional methods (PCA on lightly imputed methylation percentages, based on \citep{luo2017single}).
+    Tiles or VMRs were pre-processed as proposed in this work (shrunken mean of residuals subjected to iterative PCA) or using conventional methods (PCA on lightly imputed methylation percentages, based on \citep{luo2017single}).
     The full 1566-cell data set was sub-sampled to simulate smaller data sets (x axis).
@@
 
-\section{note to Simon: this is how far I got}   % todo rm
-
 \phantomsection
-\subsection{Detecting DMRs between oligodendrocytes and neuroblasts}
+\subsection{Detecting DMRs between oligodendrocytes and neural stem cells}
 
@@
     \end{center}
-    \caption{\small \textbf{Detection of differentially methylated regions between neural stem cells and oligodendrocytes.}\\
-    \textbf{(A)} DMRs detected between neural stem cells (NSCs) and oligodendrocytes (ODs) from \citep{kremer_scnmt} (top) and DMRs detected in the same data with randomly permuted cell labels (bottom, used to estimate FDR and determine adjusted p-values).
+    \caption{\small \textbf{Detection of differentially methylated regions between oligodendrocytes and neural stem cells.}\\
+    \textbf{(A)} DMRs detected between 58 oligodendrocytes (ODs) and 130 neural stem cells (NSCs) from \citep{kremer_scnmt} (top) and DMRs detected in the same data with randomly permuted cell labels (bottom, used to estimate FDR and determine adjusted p-values).
     \textbf{(B)} Enrichment of GO terms associated with DMRs lowly methylated in ODs (top) or NSCs (bottom).
@@
 
-Next, we used the single-cell multi-omics data set \citep{kremer_scnmt} to assess the ability of our DMR detection approach.
-We first selected two cell populations from the murine ventricular-subventricular zone, neuroblasts and oligodendrocytes, and then used the method depicted in Fig.~\ref{fig:dmrscan} to identify regions with differing methylation.
-Fig.~\ref{fig:dmr} gives an overview of all identified DMRs and the relation of the adjusted p-value to the number of cells included in a DMR, and the difference in methylation between the cell types.
-Logically, if a DMR contains few cells, the methylation difference between cell types must be higher to obtain a low (and thus more significant) p-value.
-A low p-value is maintained, although the methylation difference decreases if the number of cells in the DMR increases.
+Next, we used the single-cell multi-omics data set \citep{kremer_scnmt} to evaluate our DMR detection approach.
+We first selected two cell populations from the healthy murine ventricular-subventricular zone: neural stem cells (NSCs, 130 cells) and oligodendrocytes (58 cells).
+Then we used the method depicted in Fig.~\ref{fig:dmrscan} to identify DMRs between these two cell types (Fig.~\ref{fig:dmr}).
+To estimate the false discovery rate and obtain adjusted p-values, we repeated the DMR detection approach on a permutation of the same data set, generated by shuffling cell labels.
+DMRs detected in this manner showed more modest methylation differences, and as a result many DMRs detected in the original data were assigned low adjusted p-values (Fig.~\ref{fig:dmr}A).
+Gene ontology (GO) enrichment with GREAT \citep{mclean2010great} revealed that DMRs lowly methylated in oligodendrocytes are located near genes involved in myelination, the main function of oligodendrocytes.
+Similarly, DMRs specifically demethylated in NSCs occur near genes involved in stem cell population maintenance.
+This demonstrates that our DMR detection approach is able to identify biologically meaningful DMRs, even in scBS data sets of modest cell number.
 
-An oligodendrocyte DMR stretching across Mbp was chosen to assess the methylation patterns of an individual DMR (Fig.~\ref{fig:dmr}B-C).
-Mbp is highly expressed in oligodendrocytes as it is essential for the myelination of neuronal axons.
-The DMR is highly methylated in neuroblasts and lowly methylated in oligodendrocytes (Fig.~\ref{fig:dmr}B-C), including more substantial demethylation in oligodendrocytes at the position of Mbp (Fig.~\ref{fig:dmr}C).
-Low methylation at this locus supports the expression of Mbp.
-
-Nevertheless, this is not the only indicator of the biological relevance of discovered DMRs.
-GO-term enrichment shows that neuroblast DMRs occur near genes involved in the regulation of neurogenesis, neuron differentiation, and the positive regulation of neural precursor cell proliferation, all of which are essential for neuroblasts (Fig.~\ref{fig:dmr}D).
-On the other side, oligodendrocyte-defining DMRs are located near genes with implications for myelin sheath, myelination, and axon ensheathment and thus are linked to oligodendrocytes.
-Therefore, DMRs occur in the proximity of genes associated with oligodendrocyte or neuroblast function, respectively and are implied to regulate gene expression and contribute to fundamental cell type-specific function, which is supported by the repressive effect of CG methylation on gene expression.
+Fig.~\ref{fig:dmr}C-D depicts an exemplary DMR, located at the gene body of myelin-basic protein (Mbp), the major component of myelin that is essential for myelination of neuronal axons \citep{mbp}.
+Our results suggest that oligodendrocyte-specific gene expression of Mbp is supported by low methylation at the detected DMR.
 
@@
 We have implemented the approach just described in a Python package with a command line interface, called \texttt{scbs}, which also offers a number of other functionalities for the analysis of scBS data (Figure \ref{fig:workflow}).
-The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark} or methylpy \citep{methylpy}.
+The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark}, methylpy \citep{methylpy} or biscuit \citep{biscuit}.
 For each cell, these tools produce a text file that lists the methylation status of all CpG sites.
@@
 This produces a BED file that lists the genome coordinates of VMRs.
-Similarly, the user can identify DMRs after quality control with \texttt{scbs diff} using a text file listing the cells’ group assignment as input.
-The command produces a BED file listing the coordinates of the determined DMRs together with the t- statistic, the corresponding cell group, and the adjusted p-value.
 To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED file can be used as input for \texttt{scbs matrix}, which quantifies methylation at genomic intervals in all cells.
@@
 The obtained methylation matrix can then serve as input for common single-cell analysis methods such as dimensionality reduction and cell clustering.
+
+Finally, after annotation and exploration of the data set, the user may specify two groups of cells for DMR detection with \texttt{scbs diff}.
+This command produces a BED file listing the genome coordinates, adjusted p-values, as well as several other metrics associated with DMRs.
+These DMRs may then be associated with nearby genes, or used for GO enrichment with tools such as GREAT \citep{mclean2010great} to enable functional interpretation.
 
@@
 The source code is hosted at \href{https://github.com/LKremer/scbs}{https://github.com/LKremer/scbs}, where we also provide detailed documentation including a tutorial that demonstrates a complete \texttt{scbs} analysis on an example data set.
+
+\section{note to Simon: this is how far I got}   % todo rm
 

### 2023-03-14 paper.tex
update discussion

@@
 
-\section{note to Simon: this is how far I got}   % todo rm
-
 \section{Discussion and Conclusion}
@@
 By re-analyzing two published data sets, we could show that these improvements to data preprocessing help to increase signal and decrease noise, resulting in a more informative intermediate-dimensional representation of the data.
-As examples of practical benefits, we demonstrate that our preprocessing allows for better distinction of cell sub-types, especially for smaller data sets with a limited number of cells.
-In addition to this improvement of global signal-to-noise ratio (with regards to cell dissimilarity quantification), we also show that the determination of VMRs improve biological interpretability by pinpointing the regions that actually change and hence have a putative regulatory role.
+As examples of practical benefits, we demonstrate that our preprocessing allows for better distinction of cell sub-types, especially for challenging data sets comprising cellular sub-states and lineage transitions.
+
+To furthermore aid interpretation of scBS data, we developed and implemented the first single-cell DMR detection algorithm.
+FDR estimation via permutation allows us to report statistical significance of each DMR.
+% maybe here discuss pseudoreplication issues?
+Applying our approach to NSCs and oligodendrocytes demonstrated that the obtained DMRs locate near meaningful loci associated with cell-type specific functions.
+Thus, our method aids biological interpretability by pinpointing the regions that differ between groups of cells and hence have a putative regulatory role.
 
@@
 
-In addition, by pinpointing exact regions of interest, the VMR functionality provides useful input for genomics-style analyses.
-Depending on the research question at hand, individual VMRs can be related to nearby genomic features such as gene bodies or known regulatory elements, tested for methylation differences between groups of cells, or subjected to gene ontology enrichment or motif enrichment.
+In addition, by pinpointing exact regions of interest, the VMR and DMR functionalities provide useful input for genomics-style analyses.
+Depending on the research question at hand, individual DMRs can be related to nearby genomic features such as gene bodies or known regulatory elements, tested for methylation differences between groups of cells, or subjected to gene ontology enrichment or motif enrichment.
 
@@
 \phantomsection
-\section{Methods}
+\section{Methods (note to Simon: didnt go through this yet)}  % todo rm
 

### 2023-03-15 paper.tex

@@
 This performance gain was observed also in smaller data sets, although cell cluster separation generally becomes more difficult in those.
-Both the use of VMRs instead of genomic tiles, as well as the use of the mean shrunken residuals over methylation averages improved cell cluster separation.
+Both the use of VMRs instead of genomic tiles, as well as the use of the shrunken mean of the residuals over methylation averages improved cell cluster separation.
 
@@
 
+\begin{figure}[h]
+    \begin{center}
+        \includegraphics[width=0.9\columnwidth]{figures/SFig_benchmark.png}
+    \end{center}
+    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of neural subtypes of the murine frontal cortex}.\\
+    Our methods were benchmarked on single-cell methylomes from \citet{luo2017single} as described in Fig.~\ref{fig:score}.
+    }
+    \label{fig:score_luo}
+\end{figure}
+
 \vspace{1.4ex}
@@
 Whenever a region is not covered by any read in a cell, the corresponding data entry in the input data matrix for PCA will be missing.
-The standard approach to calculate PCA (commonly done using the IRLBA algorithm, \citep{Baglama2005}) is not suited to deal with missing data.
+The standard approach to calculate PCA, commonly done using the IRLBA algorithm \citep{Baglama2005}, is not suited to deal with missing data.
 We circumvent this issue by simply using the PCA itself to impute the missing value.
@@
 Then, we use the PCA predictions for the missing values, i.e., the $a_{ij}=\sum_r x_{ir} r_{jr}$, as refined stand-ins for the missing values in $A$ and run PCA once more.
-This can now be iterated until convergence, which is typically achieved after at most 4--5 rounds of PCA.
+This can now be iterated until convergence.
 
 \subsection{Re-analysis of published scBS datasets}
-To re-analyze scBS data sets by \citet{luo2017single} and \citet{argelaguet2019gastru}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 2\%$).
-For all regions (VMRs and 100~kb tiles), the mean methylation fractions and the shrunken means of residuals were obtained with \texttt{scbs matrix}.
+To re-analyze scBS data sets by \citet{kremer_scnmt} and \citet{luo2017single}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 2\%$).
+For all regions (VMRs and 100~kb tiles), the mean methylation fractions and the shrunken means of the residuals were obtained with \texttt{scbs matrix}.
+
 Regions containing less than four measured CpGs, or that were measured in less than 10\% of cells, were discarded.

### 2023-03-15 paper.tex
minor stuff

@@
     \begin{center}
-        \includegraphics[width=0.9\columnwidth]{figures/SFig_benchmark.png}
+        \includegraphics[width=\columnwidth]{figures/SFig_parameter_sweep.png}
     \end{center}
-    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of neural subtypes of the murine frontal cortex}.\\
-    Our methods were benchmarked on single-cell methylomes from \citet{luo2017single} as described in Fig.~\ref{fig:score}.
+    \caption{\small \textbf{Effect of VMR detection parameters on separation of cell types}.\\
+    \texttt{(A - B)} Neighbor score obtained after analyzing single-cell methylomes from \citet{kremer_scnmt} (\textbf{A}) or \citettt{luo2017single} (\textbf{B}) with our proposed methods.
+    VMRs were detected with \citettt{scbs scan} using various sliding window bandwidths (\citettt{--bandwidth} parameter) or variance thresholds (\citettt{--var-threshold} parameter).
     }
-    \label{fig:score_luo}
+    \label{fig:sweep}
 \end{figure}

### 2023-03-15 paper.tex
work on methods

--- 2023-03-15 paper.tex
@@
 
+Lastly, we assessed whether our proposed workflow requires fine-tuning of parameters (Fig.~\ref{fig:sweep}).
+To this end, we re-analyzed both data sets, as well as sub-samples of the data with different VMR detection parameters (the width of the sliding window in bp, \texttt{--bandwidth}, default 2000 and the variance threshold above which windows are merged to VMRs, \texttt{--var-threshold}, default 0.02).
+This parameter sweep showed that our workflow is robust to various parameter selections and depends mostly on the overall cell number.
+In the more challenging data set (Fig.~\ref{fig:sweep}A), cell types were less cleanly separated when very large bandwidths, or very strict variance thresholds were selected.
+However, in our experience small bandwidths or lenient thresholds resulted in a much higher number of VMRs and thus longer computing times.
+Our default parameter combination provided good results and fast compute times in both data sets.
+
 \phantomsection
@@
 DMRs detected in this manner showed more modest methylation differences, and as a result many DMRs detected in the original data were assigned low adjusted p-values (Fig.~\ref{fig:dmr}A).
+
 Gene ontology (GO) enrichment with GREAT \citep{mclean2010great} revealed that DMRs lowly methylated in oligodendrocytes are located near genes involved in myelination, the main function of oligodendrocytes.
@@
 This demonstrates that our DMR detection approach is able to identify biologically meaningful DMRs, even in scBS data sets of modest cell number.
-
 Fig.~\ref{fig:dmr}C-D depicts an exemplary DMR, located at the gene body of myelin-basic protein (Mbp), the major component of myelin that is essential for myelination of neuronal axons \citep{mbp}.
@@
 In conclusion, we have presented powerful improvements to scBS data preprocessing and a software tool that implements these and bridges a gap in existing tool chains.
-
-\section{Supplementary Figures}
-
-\setcounter{figure}{0}
-%\renewcommand{\figurename}{Supplementary Figure}
-\renewcommand{\thefigure}{S\arabic{figure}}
-
-\begin{figure}[h]
-    \begin{center}
-        \includegraphics[width=0.9\columnwidth]{figures/SFig_benchmark.png}
-    \end{center}
-    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of neural subtypes of the murine frontal cortex}.\\
-    Our methods were benchmarked on single-cell methylomes from \citet{luo2017single} as described in Fig.~\ref{fig:score}.
-    }
-    \label{fig:score_luo}
-\end{figure}
-
-\begin{figure}[h]
-    \begin{center}
-        \includegraphics[width=0.9\columnwidth]{figures/SFig_benchmark.png}
-    \end{center}
-    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of neural subtypes of the murine frontal cortex}.\\
-    Our methods were benchmarked on single-cell methylomes from \citet{luo2017single} as described in Fig.~\ref{fig:score}.
-    }
-    \label{fig:score_luo}
-\end{figure}
 
@@
 \subsection{Re-analysis of published scBS datasets}
-To re-analyze scBS data sets by \citet{kremer_scnmt} and \citet{luo2017single}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 2\%$).
-For all regions (VMRs and 100~kb tiles), the mean methylation fractions and the shrunken means of the residuals were obtained with \texttt{scbs matrix}.
+To re-analyze scBS data sets by \citet{kremer_scnmt} and \citet{luo2017single}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 0.02$).
+VMRs with sequencing coverage in less than 
+% WIP
+For all intervals (VMRs and 100~kb tiles), the mean methylation fractions and the shrunken means of the residuals were obtained with \texttt{scbs matrix}.
 
-Regions containing less than four measured CpGs, or that were measured in less than 10\% of cells, were discarded.
+Regions observed in less than four measured CpGs, or that were measured in less than 10\% of cells, were discarded.
 To select $n$ intervals for further analysis, the top $n$ intervals were picked based on the number of cells that the region was measured in.
@@
 
+\section{Supplementary Figures}
+
+\setcounter{figure}{0}
+%\renewcommand{\figurename}{Supplementary Figure}
+\renewcommand{\thefigure}{S\arabic{figure}}
+
+\begin{figure*}[h]
+    \begin{center}
+        \includegraphics[width=0.8\columnwidth]{figures/SFig_benchmark.png}
+    \end{center}
+    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of neural subtypes of the murine frontal cortex}.\\
+    Our methods were benchmarked on single-cell methylomes from \citet{luo2017single} as described in Fig.~\ref{fig:score}.
+    }
+    \label{fig:score_luo}
+\end{figure*}
+
+\begin{figure*}[h]
+    \begin{center}
+        \includegraphics[width=.6\textwidth]{figures/SFig_parameter_sweep.png}
+    \end{center}
+    \caption{\small \textbf{Effect of VMR detection parameters on separation of cell types}.\\
+    \textbf{(A-B)} Mean neighbor scores obtained after analyzing single-cell methylomes from \citet{kremer_scnmt} (\textbf{A}) or \citet{luo2017single} (\textbf{B}) with our proposed methods.
+    VMRs were detected with \texttt{scbs scan} using various sliding window bandwidths or variance thresholds, and on sub-samples of the full data sets.
+    }
+    \label{fig:sweep}
+\end{figure*}
+
 \end{document}

### 2023-03-15 paper.tex
Updates from Overleaf

--- 2023-03-15 paper.tex
@@
     \end{center}
-    \caption{\small \textbf{Improved quantification of DNA methylation in genomic intervals.}\\
+    \caption{\small \textbf{Improved quantification of DNA methylation in genomic intervals:}
     \textbf{(A)} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified.
@@
 \begin{figure}
-    \begin{center}
-    \includegraphics[width=0.85\columnwidth]{figures/Fig_sliding.png}
-    \end{center}
-    \caption{\small \textbf{Finding variably methylated regions.}\\
-    The chromosomes are divided up into overlapping windows (first five shown at the bottom), and for each window, the cells' methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
+    \includegraphics[width=\columnwidth]{figures/Fig_sliding.png}
+    \caption{\small \textbf{Finding variably methylated regions (VMRs):} The chromosomes are divided up into overlapping windows (first five shown at the bottom), and for each window, the cells' methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
     Then, the variance of these values is calculated (each point represents one of the overlapping windows), a threshold (dashed line) is chosen such that a chosen quantile of windows have a variance exceeding the threshold.
-    Windows with above-threshold variance are merged if they overlap, yielding the ``variably methylated regions'' (VMRs).}
+    Windows with above-threshold variance are merged if they overlap, yielding the ``variably methylated regions''.}
     \label{fig:vmr}
@@
     \includegraphics[width=\columnwidth]{figures/Fig_DMRdetection.png}
-    \caption{\small \textbf{Identification of differentially methylated regions.}\\
+    \caption{\small \textbf{Identification of differentially methylated regions (DMRs):}
     The chromosome is divided into windows overlapping by a small and fixed size.
-    For each window, the cell’s methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
+    For each window, the cell’s methylation values are calculated as described and as depicted in Fig.~ \ref{fig:smoothres}B.
     Then the t-statistic of these values is obtained as a measure of methylation difference (each point represents one of the windows).
     Upper and lower thresholds (dashed lines) are chosen such that a defined quantile of windows have a t-statistic above or below the threshold.
-    Windows exceeding either threshold are merged if they overlap, yielding the differentially methylated regions (DMRs).}
+    Windows exceeding either threshold are merged if they overlap, yielding the ``differentially methylated regions``.}
     \label{fig:dmrscan}
@@
     \end{center}
-    \caption{\small \textbf{Correlation of DNA methylation and gene expression}.\\
+    \caption{\small \textbf{Correlation of DNA methylation and gene expression}.
     \textbf{(A)} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red).
@@
 
-\subsection{Our proposed methods improve identification of cell types}
+\subsection{Effects of our methods on cell type identification}
 
@@
     \end{center}
-    \caption{\small \textbf{Benchmark of our methods on single-cell multi-omic data of cells of the murine forebrain}.\\
+    \caption{\small \textbf{Benchmark of our methods on single-cell multi-omic data of cells of the murine forebrain}.
     \textbf{(A)} Cell labels based on clustering of single-cell transcriptomes from \citet{kremer_scnmt}.
@@
     Either 100~kb genomic tiles or VMRs were subjected to PCA and UMAP.
-    Tiles or VMRs were pre-processed as proposed in this work (shrunken mean of residuals subjected to iterative PCA) or using conventional methods (PCA on lightly imputed methylation percentages, based on \citep{luo2017single}).
+    Tiles or VMRs were pre-processed as proposed in this work (shrunken mean of residuals subjected to iterative PCA, labeled "pre-processing: ours") or using conventional methods (PCA on lightly imputed methylation percentages, based on \citet{luo2017single}).
     The full 1566-cell data set was sub-sampled to simulate smaller data sets (x axis).
-    A higher neighbor score implies better separation of cell types inferred from single-cell transcriptomes of the same cells \textbf{(A)}.
+    A higher neighbor score implies better separation of cell types inferred from single-cell transcriptomes of the same cells (A).
     }
@@
 This performance gain was observed also in smaller data sets, although cell cluster separation generally becomes more difficult in those.
-Both the use of VMRs instead of genomic tiles, as well as the use of the shrunken mean of the residuals over methylation averages improved cell cluster separation.
+Both the use of VMRs instead of genomic tiles, as well as the use of the mean shrunken residuals over methylation averages improved cell cluster separation.
 
@@
 
+\section{note to Simon: this is how far I got}   % todo rm
+
 \phantomsection
-\subsection{Detecting DMRs between oligodendrocytes and neural stem cells}
+\subsection{Detecting DMRs between oligodendrocytes and neuroblasts}
 
@@
     \begin{center}
-        \includegraphics[width=.7\textwidth]{figures/Fig_DMRs.png}
+        \includegraphics[width=.8\textwidth]{figures/Fig_DMRs.png}
     \end{center}
-    \caption{\small \textbf{Detection of differentially methylated regions between oligodendrocytes and neural stem cells.}\\
-    \textbf{(A)} DMRs detected between 58 oligodendrocytes (ODs) and 130 neural stem cells (NSCs) from \citep{kremer_scnmt} (top) and DMRs detected in the same data with randomly permuted cell labels (bottom, used to estimate FDR and determine adjusted p-values).
-    \textbf{(B)} Enrichment of GO terms associated with DMRs lowly methylated in ODs (top) or NSCs (bottom).
-    Depicted are the top 5 GO terms of the "Biological Process" (GO:BP) and "Cellular Component" (GO:CC) GO category, and their binomial test p-value and enrichment, as reported by GREAT \citep{mclean2010great}.
-    \textbf{(C)} Mean methylation of NSCs and ODs at an exemplary DMR.
-    Each point corresponds to a cell, black lines denote the median.
-    \textbf{(D)} Detailed view of the DMR (yellow dashed rectangle) from \textbf{C} in pseudo-bulk samples consisting of NSCs or ODs.
-    Vertical bars represent CpG sites.
+    \caption{\small \textbf{identified neuroblast and oligodendrocyte DMRs have a putative regulatory role:} \textbf{(A)} Difference in methylation between oligodendrocytes and neuroblasts of DMRs plotted against the number of cells contained in a DMR.
+Colored by p-value.
+\textbf{(B)} Mean methylation of cells in the DMR covering MBP.
+Line indicates median.
+\textbf{(C)} Mean methylation of reads at individual sites along the DMR covering MBP.
+\textbf{(D)} GREAT \citet{mclean2010great} was used for GO-term analysis.
+The left side shows GO-terms for genes near the 3000 most significant neuroblast DMRs sorted by GO-enrichment p-value and t-statistic (green arrow) and the right side for oligodendrocyte DMRs (yellow arrow) respectively.
     }
@@
 
-Next, we used the single-cell multi-omics data set \citep{kremer_scnmt} to evaluate our DMR detection approach.
-We first selected two cell populations from the healthy murine ventricular-subventricular zone: neural stem cells (NSCs, 130 cells) and oligodendrocytes (58 cells).
-Then we used the method depicted in Fig.~\ref{fig:dmrscan} to identify DMRs between these two cell types (Fig.~\ref{fig:dmr}).
-To estimate the false discovery rate and obtain adjusted p-values, we repeated the DMR detection approach on a permutation of the same data set, generated by shuffling cell labels.
-DMRs detected in this manner showed more modest methylation differences, and as a result many DMRs detected in the original data were assigned low adjusted p-values (Fig.~\ref{fig:dmr}A).
-Gene ontology (GO) enrichment with GREAT \citep{mclean2010great} revealed that DMRs lowly methylated in oligodendrocytes are located near genes involved in myelination, the main function of oligodendrocytes.
-Similarly, DMRs specifically demethylated in NSCs occur near genes involved in stem cell population maintenance.
-This demonstrates that our DMR detection approach is able to identify biologically meaningful DMRs, even in scBS data sets of modest cell number.
+Next, we used the single-cell multi-omics data set \citep{kremer_scnmt} to assess the ability of our DMR detection approach.
+We first selected two cell populations from the murine ventricular-subventricular zone, neuroblasts and oligodendrocytes, and then used the method depicted in Fig.~\ref{fig:dmrscan} to identify regions with differing methylation.
+Fig.~\ref{fig:dmr} gives an overview of all identified DMRs and the relation of the adjusted p-value to the number of cells included in a DMR, and the difference in methylation between the cell types.
+Logically, if a DMR contains few cells, the methylation difference between cell types must be higher to obtain a low (and thus more significant) p-value.
+A low p-value is maintained, although the methylation difference decreases if the number of cells in the DMR increases.
 
-Fig.~\ref{fig:dmr}C-D depicts an exemplary DMR, located at the gene body of myelin-basic protein (Mbp), the major component of myelin that is essential for myelination of neuronal axons \citep{mbp}.
-Our results suggest that oligodendrocyte-specific gene expression of Mbp is supported by low methylation at the detected DMR.
+An oligodendrocyte DMR stretching across Mbp was chosen to assess the methylation patterns of an individual DMR (Fig.~\ref{fig:dmr}B-C).
+Mbp is highly expressed in oligodendrocytes as it is essential for the myelination of neuronal axons.
+The DMR is highly methylated in neuroblasts and lowly methylated in oligodendrocytes (Fig.~\ref{fig:dmr}B-C), including more substantial demethylation in oligodendrocytes at the position of Mbp (Fig.~\ref{fig:dmr}C).
+Low methylation at this locus supports the expression of Mbp.
+
+Nevertheless, this is not the only indicator of the biological relevance of discovered DMRs.
+GO-term enrichment shows that neuroblast DMRs occur near genes involved in the regulation of neurogenesis, neuron differentiation, and the positive regulation of neural precursor cell proliferation, all of which are essential for neuroblasts (Fig.~\ref{fig:dmr}D).
+On the other side, oligodendrocyte-defining DMRs are located near genes with implications for myelin sheath, myelination, and axon ensheathment and thus are linked to oligodendrocytes.
+Therefore, DMRs occur in the proximity of genes associated with oligodendrocyte or neuroblast function, respectively and are implied to regulate gene expression and contribute to fundamental cell type-specific function, which is supported by the repressive effect of CG methylation on gene expression.
 
@@
     \end{center}
-    \caption{\small \textbf{Overview of the functionalities implemented in the \texttt{scbs} package.}\\
+    \caption{\small \textbf{Overview of the functionalities implemented in the \texttt{scbs} package.}
         \texttt{scbs prepare} parses methylation files produced by common bisulfite sequencing mappers and stores their contents in a compressed format optimised for efficient access to genomic intervals.
@@
 We have implemented the approach just described in a Python package with a command line interface, called \texttt{scbs}, which also offers a number of other functionalities for the analysis of scBS data (Figure \ref{fig:workflow}).
-The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark}, methylpy \citep{methylpy} or biscuit \citep{biscuit}.
+The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark} or methylpy \citep{methylpy}.
 For each cell, these tools produce a text file that lists the methylation status of all CpG sites.
@@
 The obtained methylation matrix can then serve as input for common single-cell analysis methods such as dimensionality reduction and cell clustering.
-
-Finally, after annotation and exploration of the data set, the user may specify two groups of cells for DMR detection with \texttt{scbs diff}.
-This command produces a BED file listing the genome coordinates, adjusted p-values, as well as several other metrics associated with DMRs.
-These DMRs may then be associated with nearby genes, or used for GO enrichment with tools such as GREAT \citep{mclean2010great} to enable functional interpretation.
 
@@
 By re-analyzing two published data sets, we could show that these improvements to data preprocessing help to increase signal and decrease noise, resulting in a more informative intermediate-dimensional representation of the data.
-As examples of practical benefits, we demonstrate that our preprocessing allows for better distinction of cell sub-types, especially for challenging data sets comprising cellular sub-states and lineage transitions.
-
-To furthermore aid interpretation of scBS data, we developed and implemented the first single-cell DMR detection algorithm.
-FDR estimation via permutation allows us to report statistical significance of each DMR.
-% maybe here discuss pseudoreplication issues?
-Applying our approach to NSCs and oligodendrocytes demonstrated that the obtained DMRs locate near meaningful loci associated with cell-type specific functions.
-Thus, our method aids biological interpretability by pinpointing the regions that differ between groups of cells and hence have a putative regulatory role.
+As examples of practical benefits, we demonstrate that our preprocessing allows for better distinction of cell sub-types, especially for smaller data sets with a limited number of cells.
+In addition to this improvement of global signal-to-noise ratio (with regards to cell dissimilarity quantification), we also show that the determination of VMRs improve biological interpretability by pinpointing the regions that actually change and hence have a putative regulatory role.
 
@@
 
-In addition, by pinpointing exact regions of interest, the VMR and DMR functionalities provide useful input for genomics-style analyses.
-Depending on the research question at hand, individual DMRs can be related to nearby genomic features such as gene bodies or known regulatory elements, tested for methylation differences between groups of cells, or subjected to gene ontology enrichment or motif enrichment.
+In addition, by pinpointing exact regions of interest, the VMR functionality provides useful input for genomics-style analyses.
+Depending on the research question at hand, individual VMRs can be related to nearby genomic features such as gene bodies or known regulatory elements, tested for methylation differences between groups of cells, or subjected to gene ontology enrichment or motif enrichment.
 
@@
 
-\begin{figure}[h]
+\begin{figure}
     \begin{center}
@@
     \end{center}
-    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of neural subtypes of the murine frontal cortex}.\\
-    Our methods were benchmarked on single-cell methylomes from \citet{luo2017single} as described in Fig.~\ref{fig:score}.
-    }
-    \label{fig:score_luo}
-\end{figure}
-
-\begin{figure}[h]
-    \begin{center}
-        \includegraphics[width=0.9\columnwidth]{figures/SFig_benchmark.png}
-    \end{center}
-    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of neural subtypes of the murine frontal cortex}.\\
+    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of neural subtypes of the murine frontal cortex}.
     Our methods were benchmarked on single-cell methylomes from \citet{luo2017single} as described in Fig.~\ref{fig:score}.
@@
 S.\ A.\ and L.\ P.\ M.\ K.\ acknowledge funding by the Klaus Tschira Foundation.
-L.\ P.\ M.\ K., L.\ K., S.\ C.\ and A.\ M.-V.\ acknowledge funding from the European Commission via ERC grant 771376.
+L.\ P.\ M.\ K., M.\ B., L.\ K., S.\ C.\ and A.\ M.-V.\ acknowledge funding from the European Commission via ERC grant 771376.
 We thank Alexey Uvarovskii for code refactoring and helpful suggestions on the source code.
@@
 \phantomsection
-\section{Methods (note to Simon: didnt go through this yet)}  % todo rm
+\section{Methods}
 
@@
 Whenever a region is not covered by any read in a cell, the corresponding data entry in the input data matrix for PCA will be missing.
-The standard approach to calculate PCA, commonly done using the IRLBA algorithm \citep{Baglama2005}, is not suited to deal with missing data.
+The standard approach to calculate PCA (commonly done using the IRLBA algorithm, \citep{Baglama2005}) is not suited to deal with missing data.
 We circumvent this issue by simply using the PCA itself to impute the missing value.
@@
 Then, we use the PCA predictions for the missing values, i.e., the $a_{ij}=\sum_r x_{ir} r_{jr}$, as refined stand-ins for the missing values in $A$ and run PCA once more.
-This can now be iterated until convergence.
+This can now be iterated until convergence, which is typically achieved after at most 4--5 rounds of PCA.
 
 \subsection{Re-analysis of published scBS datasets}
-To re-analyze scBS data sets by \citet{kremer_scnmt} and \citet{luo2017single}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 2\%$).
-For all regions (VMRs and 100~kb tiles), the mean methylation fractions and the shrunken means of the residuals were obtained with \texttt{scbs matrix}.
-
+To re-analyze scBS data sets by \citet{luo2017single} and \citet{argelaguet2019gastru}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 2\%$).
+For all regions (VMRs and 100~kb tiles), the mean methylation fractions and the shrunken means of residuals were obtained with \texttt{scbs matrix}.
 Regions containing less than four measured CpGs, or that were measured in less than 10\% of cells, were discarded.

### 2023-03-16 paper.tex
minor stuff

--- 2023-03-15 paper.tex
@@
 
-The standard approach to analyze scBS-Seq data is based on methodology developed for the analysis of single-cell RNA-Seq data.
-Therefore, we start by briefly reviewing how single-cell \emph{RNA}-Seq data is commonly analyzed, before we discuss scBS data analysis.
+The standard approach to analyze scBS-seq data is based on methodology developed for the analysis of scRNA-seq data.
+Therefore, we start by briefly reviewing how sc\emph{RNA}-seq data is commonly analyzed, before we discuss scBS data analysis.
 
-The starting point in most scRNA-Seq analyses is a matrix of UMI counts (i.e.
+The starting point in most scRNA-seq analyses is a matrix of UMI counts (i.e.
 of counts of distinct RNA molecules), one row for each cell and one column for each gene.
@@
 A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm.
-In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-4}$) to the normalized fractions before taking the logarithm.
+\textbf{In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-4}$) to the normalized fractions before taking the logarithm.}  % is this even correct?
 Now, one could use Euclidean distances of these vectors of logarithmized fractions as dissimilarity score.
@@
 
-This procedure is commonly adapted when working with single-cell DNA methylation data, so that established methods used in scRNA-Seq can be used.
+This procedure is commonly adapted when working with single-cell DNA methylation data, so that established methods used in scRNA-seq can be used.
 The task is therefore how to construct a suitable matrix as input for PCA.
@@
     \begin{center}
-    \includegraphics[width=0.85\columnwidth]{figures/Fig_sliding.png}
+    \includegraphics[width=\columnwidth]{figures/Fig_sliding.png}
     \end{center}
@@
 
-\begin{figure}
-    \includegraphics[width=\columnwidth]{figures/Fig_DMRdetection.png}
+\begin{figure*}
+    \begin{center}
+    \includegraphics[width=.65\textwidth]{figures/Fig_DMRdetection.png}
+    \end{center}
     \caption{\small \textbf{Identification of differentially methylated regions.}\\
@@
     \label{fig:dmrscan}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}
+\begin{figure*}
     \begin{center}
-    \includegraphics[width=.95\columnwidth]{figures/Fig_correlation.png}
+    \includegraphics[width=.65\textwidth]{figures/Fig_correlation.png}
     \end{center}
@@
     \label{fig:correlation}
-\end{figure}
+\end{figure*}
 
@@
     \begin{center}
-        \includegraphics[width=0.8\textwidth]{figures/Fig_benchmark.png}
+        \includegraphics[width=0.7\textwidth]{figures/Fig_benchmark.png}
     \end{center}

### 2023-03-16 paper.tex
write methods

@@
 \subsection{Re-analysis of published scBS datasets}
-To re-analyze scBS data sets by \citet{kremer_scnmt} and \citet{luo2017single}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 0.02$).
-VMRs with sequencing coverage in less than 
-% WIP
-For all intervals (VMRs and 100~kb tiles), the mean methylation fractions and the shrunken means of the residuals were obtained with \texttt{scbs matrix}.
+To re-analyze scBS data from \citet{kremer_scnmt}, single-cell CpG methylation reports were first stored with \texttt{scbs prepare} and then smoothed with \texttt{scbs smooth} using the default bandwidth of 1000~bp.
+The mm10 mouse genome was then annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 0.02$, minimum cell requirement $=6$).
+Methylation fractions, commonly represented as percentages, and the shrunken means of residuals were quantified with \texttt{scbs matrix} for both VMRs and tiles.
+The same analysis was repeated for scBS data from \citet{luo2017single}, and for smaller data sets generated by randomly sub-sampling cells separately from both data sets.
+For each data set, we thus obtained four methylation matrices:
+either 100~kb tiles or VMRs, quantified either by simple averaging or with the shrunken mean of the residuals.
+All four matrices were then subjected to PCA, followed by UMAP with parameters min\_dist $=0.2$, init $=$ spca, and n\_neighbors $=\frac{\sqrt{n}}{1.5}$ where $n$ is the total cell number, to adapt to data sets of different cell numbers.
 
-Regions observed in less than four measured CpGs, or that were measured in less than 10\% of cells, were discarded.
-To select $n$ intervals for further analysis, the top $n$ intervals were picked based on the number of cells that the region was measured in.
+\subsection{Mean neighbor score} \label{methods:score}
+To assess performance of our methods, we employed a score
 
-For dimensionality reduction, missing values were imputed with the iterative PCA described in the previous section.
-To visualize data, UMAP was performed on the final 15 principal components.
-
-\subsection{Nearest neighbor count score} \label{methods:score}
-To assess performance of our methods, we processed mouse CpG methylation data from \citet{luo2017single} with several method combinations and scored each result.
+we processed mouse CpG methylation data from \citet{luo2017single} with several method combinations and scored each result.
 The score evaluates the ability of each method combination to separate cell types in 15-dimensional PCA space, based on their CpG methylation.
@@
 
+% WIP
+For all intervals (VMRs and 100~kb tiles), the mean methylation fractions and the shrunken means of the residuals were obtained with \texttt{scbs matrix}.
+
+Regions observed in less than four measured CpGs, or that were measured in less than 10\% of cells, were discarded.
+To select $n$ intervals for further analysis, the top $n$ intervals were picked based on the number of cells that the region was measured in.
+
+For dimensionality reduction, missing values were imputed with the iterative PCA described in the previous section.
+To visualize data, UMAP was performed on the final 15 principal components.
+
 \vspace{1.4ex}
@@
 
+\newpage
 \section{Supplementary Figures}
@@
 
-\begin{figure*}[h]
+\begin{figure*}
     \begin{center}

### 2023-03-16 paper.tex
work on methods

@@
         The resulting methylation matrix can then be used for downstream analysis such as cell clustering and dimensionality reduction.
-        Identified groups of cells (e.~g. clusters) can then be scanned for DMRs with \texttt{scbs diff}.
+        Identified groups of cells (e.g. clusters) can then be scanned for DMRs with \texttt{scbs diff}.
     }
@@
 
+\subsection{Finding differentially methylated regions}
+
 \subsection{Calculating cell to cell distances}
@@
 For each data set, we thus obtained four methylation matrices:
-either 100~kb tiles or VMRs, quantified either by simple averaging or with the shrunken mean of the residuals.
-All four matrices were then subjected to PCA, followed by UMAP with parameters min\_dist $=0.2$, init $=$ spca, and n\_neighbors $=\frac{\sqrt{n}}{1.5}$ where $n$ is the total cell number, to adapt to data sets of different cell numbers.
+either 100~kb tiles or VMRs, quantified either by simple averaging (methylation fractions) or with the shrunken mean of the residuals.
+All four matrices were then subjected to PCA.
+In the case of methylation fractions, we closely followed methods described in \citet{luo2017single}:
+We selected tiles spanning $\ge20$ CpG sites and with sequencing coverage in at least 70\% of cells.
+We then imputed missing values by imputing the mean for each tile, centered the values, and performed PCA.
+In the case of VMRs, we used iterative PCA to handle missing values, as described earlier.
+For visualization, the first 15 PCs were subjected to UMAP with parameters min\_dist $=0.2$, init $=$ "spca", and n\_neighbors $=\frac{\sqrt{n}}{1.5}$ where $n$ is the total cell number, to adapt to data sets of different size.
 
 \subsection{Mean neighbor score} \label{methods:score}
-To assess performance of our methods, we employed a score
-
-we processed mouse CpG methylation data from \citet{luo2017single} with several method combinations and scored each result.
-The score evaluates the ability of each method combination to separate cell types in 15-dimensional PCA space, based on their CpG methylation.
-Each score varies between 0 and 1, where higher scores reflect better separation.
-
-Our nearest-neighbor-count score $\text{nnc}_k$, is based on the $\Gamma$-score \citep{Kireeva_2014} and is calculated as follows.
-For each cell $j$, we count how many of its $k$ nearest neighbors have been assigned to the same cluster as cell $j$.
-We denote this count $a^k_j$.
-We consider a cell as ``well-assigned'' if $a^k_j>\theta k$, where we have chosen $\theta=0.9$.
-In other words, a cell is ``well-assigned'' if at least 90\% of its $k$ nearest neighbors have been assigned to the same class as the cell under consideration.
-
-The score $\text{nnc}_k$ is then simply the fraction of cells in the data set that are well-assigned in this sense.
-
-% WIP
-For all intervals (VMRs and 100~kb tiles), the mean methylation fractions and the shrunken means of the residuals were obtained with \texttt{scbs matrix}.
-
-Regions observed in less than four measured CpGs, or that were measured in less than 10\% of cells, were discarded.
-To select $n$ intervals for further analysis, the top $n$ intervals were picked based on the number of cells that the region was measured in.
-
-For dimensionality reduction, missing values were imputed with the iterative PCA described in the previous section.
-To visualize data, UMAP was performed on the final 15 principal components.
+To assess performance of our methods, we employed a score that quantifies how well cell types (or cell states) are separated in 15-dimensional PCA space.
+For data from \citet{luo2017single}, we used cell type labels that the authors manually curated based on CH-methylation.
+For the multi-omic data set, we repeated the single-cell transcriptomics analysis described in \citet{kremer_scnmt} with two adjustments:
+We did not filter off-target cells such as endothelial cells, and we annotated cell types using Leiden clustering with the Seurat \citep{seurat} function \texttt{FindClusters(resolution = 0.1)}.
+The score, based on the $\Gamma$-score \citep{Kireeva_2014}, varies between 0 and 1, where higher scores reflect better separation of cell types.
+For each cell $j$, we count how many of its $k$ nearest neighbors have been assigned to the same cell type as cell $j$.
+We denote this count $a^k_j$, where we have chosen $k=\frac{\sqrt{n}}{1.5}$.
+The overall score used to evaluate a given combination of methods is then simply the mean of all cell-wise scores.
 

### 2023-03-16 paper.tex
Updates from Overleaf

@@
 We demonstrate how our approach enables better distinguishing of cell types and reduces demands on cell number.
-We also implement an approach to detect differentially methylated regions (DMRs) between groups of cells.
+We also implement an approach to detect differentially methylated regions (DMRs) between groups of cells, and demonstrate its ability to detect biologically meaningful regions near 
 Finally, we present a software tool, \texttt{scbs}, that implements these methods and offers further functionality required to work with scBS data.
@@
 A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm.
-\textbf{In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-4}$) to the normalized fractions before taking the logarithm.}  % is this even correct?
+In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-4}$) to the normalized fractions before taking the logarithm \todo{is this even correct?}.
 Now, one could use Euclidean distances of these vectors of logarithmized fractions as dissimilarity score.
@@
     Promoters are defined as intervals \textpm2~kb around the TSS.
-    \textbf{(C)} Mean methylation near the gene Htra1 (ENSMUSG00000006205).
+    \textbf{(C)} \todo{plot this as discussed, as in Fig.~6D} Mean methylation near the gene Htra1 (ENSMUSG00000006205).
     Cells are assigned to 4 groups based on Htra1 expression (group 0: cells with no Htra1 expression, groups 1-3: cells that express Htra1, divided into three equally large groups with group 3 having the highest expression).
@@
 over those cells for which methylation data is available for position $i$.
-
-%\todo{NOTE: Here, we could already do shrinkage and use $\frac{1}{|C_i|+1}\sum_i x_{ij}$ instead. But we don't do that, right?}
 
@@
 \subsection{Re-analysis of published scBS datasets}
-To re-analyze scBS data from \citet{kremer_scnmt}, single-cell CpG methylation reports were first stored with \texttt{scbs prepare} and then smoothed with \texttt{scbs smooth} using the default bandwidth of 1000~bp.
+To re-analyze scBS data from \citet{kremer_scnmt}, single-cell CpG methylation reports from all conditions were first stored with \texttt{scbs prepare} and then smoothed with \texttt{scbs smooth} using the default bandwidth of 1000~bp.
 The mm10 mouse genome was then annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 0.02$, minimum cell requirement $=6$).
 Methylation fractions, commonly represented as percentages, and the shrunken means of residuals were quantified with \texttt{scbs matrix} for both VMRs and tiles.
-The same analysis was repeated for scBS data from \citet{luo2017single}, and for smaller data sets generated by randomly sub-sampling cells separately from both data sets.
+The same analysis was repeated for mouse scBS data from \citet{luo2017single}, and for smaller data sets generated by randomly sub-sampling cells separately from both data sets.
 For each data set, we thus obtained four methylation matrices:
@@
 The overall score used to evaluate a given combination of methods is then simply the mean of all cell-wise scores.
+
+\subsection{Correlation of DNA methylation and gene expression}
+To assess correlation between gene expression and the methylation status of VMRs or promoters, we
+first detected VMRs with \texttt{scbs scan --bandwidth 1000 --var-treshold 0.05}.
+We then quantified DNA methylation at VMRs and promoters with \citettt{scbs matrix}.
+We defined promoters as $\pm2000$~bp regions centered on a genes transcription start site (TSS).
+When multiple TSSs were annotated, we chose the TSS of the "principal" isoform \citep{appris}.
+Log-normalized expression values reported by \citet{kremer_scnmt} were then correlated with methylation of the genes promoter, or with methylation of the VMR closest to the gene body.
+When multiple VMRs intersected the gene body, we chose the VMR with the highest methylation variance.
+As a measure of methylation we used the shrunken mean of the residuals.
+We omitted lowly expressed genes (with scRNA-seq counts in $\lt5\%$ of cells) and promoters and VMRs with low scBS coverage (in $\lt10$ cells).
 

### 2023-03-16 paper.tex
final touches

@@
 We demonstrate how our approach enables better distinguishing of cell types and reduces demands on cell number.
-We also implement an approach to detect differentially methylated regions (DMRs) between groups of cells, and demonstrate its ability to detect biologically meaningful regions near 
+We also implement an approach to detect differentially methylated regions (DMRs) between groups of cells, and demonstrate its ability to detect biologically meaningful regions near genes involved in core functions of specific cell types.
 Finally, we present a software tool, \texttt{scbs}, that implements these methods and offers further functionality required to work with scBS data.
@@
 A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm.
-In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-4}$) to the normalized fractions before taking the logarithm \todo{is this even correct?}.
+In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-4}$) \todo{is this even correct?} to the normalized fractions before taking the logarithm.
 Now, one could use Euclidean distances of these vectors of logarithmized fractions as dissimilarity score.
@@
     \caption{\small \textbf{Correlation of DNA methylation and gene expression}.\\
-    \textbf{(A)} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red).
+    \textbf{(A)} \todo{maybe remove panel A or B, and put the remaining panel left of C?} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red).
     \textbf{(B)} Correlation of each gene's RNA expression values to methylation (shrunken mean of residuals) at promoters (y-axis), or to methylation of the nearest VMR (x-axis).
     Promoters are defined as intervals \textpm2~kb around the TSS.
-    \textbf{(C)} \todo{plot this as discussed, as in Fig.~6D} Mean methylation near the gene Htra1 (ENSMUSG00000006205).
+    \textbf{(C)} Mean methylation near the gene Htra1 (ENSMUSG00000006205).
     Cells are assigned to 4 groups based on Htra1 expression (group 0: cells with no Htra1 expression, groups 1-3: cells that express Htra1, divided into three equally large groups with group 3 having the highest expression).
@@
 For each cell $j$, we count how many of its $k$ nearest neighbors have been assigned to the same cell type as cell $j$.
-We denote this count $a^k_j$, where we have chosen $k=\frac{\sqrt{n}}{1.5}$.
+We denote this count $a^k_j$, where we have chosen $k=\frac{\sqrt{n}}{1.5}$, rounded to the nearest integer.
 The overall score used to evaluate a given combination of methods is then simply the mean of all cell-wise scores.
@@
 first detected VMRs with \texttt{scbs scan --bandwidth 1000 --var-treshold 0.05}.
-We then quantified DNA methylation at VMRs and promoters with \citettt{scbs matrix}.
+We then quantified DNA methylation at VMRs and promoters with \texttt{scbs matrix}.
 We defined promoters as $\pm2000$~bp regions centered on a genes transcription start site (TSS).
 When multiple TSSs were annotated, we chose the TSS of the "principal" isoform \citep{appris}.
-Log-normalized expression values reported by \citet{kremer_scnmt} were then correlated with methylation of the genes promoter, or with methylation of the VMR closest to the gene body.
+Log-normalized expression values reported by \citet{kremer_scnmt} were then correlated with methylation of the gene's promoter, or with methylation of the VMR closest to the gene body.
 When multiple VMRs intersected the gene body, we chose the VMR with the highest methylation variance.
 As a measure of methylation we used the shrunken mean of the residuals.
-We omitted lowly expressed genes (with scRNA-seq counts in $\lt5\%$ of cells) and promoters and VMRs with low scBS coverage (in $\lt10$ cells).
+We omitted lowly expressed genes (with scRNA-seq counts in $<5\%$ of cells) and promoters and VMRs with low scBS coverage (in $<10$ cells).
 

### 2023-03-16 paper.tex
Updates from Overleaf

--- 2023-03-16 paper.tex
@@
 We demonstrate how our approach enables better distinguishing of cell types and reduces demands on cell number.
-We also implement an approach to detect differentially methylated regions (DMRs) between groups of cells.
+We also implement an approach to detect differentially methylated regions (DMRs) in scBS data.
+DMRs detected between oligodendrocytes and neural stem cells occur near genes related to myelination and stem cell maintenance respectively.
+These represent core functions of these cell types, which highlights our ability to detect biologically meaningful DMRs.
 Finally, we present a software tool, \texttt{scbs}, that implements these methods and offers further functionality required to work with scBS data.
@@
 
-The standard approach to analyze scBS-seq data is based on methodology developed for the analysis of scRNA-seq data.
+The standard approach to analyze scBS-Seq data is based on methodology developed for the analysis of scRNA-seq data.
 Therefore, we start by briefly reviewing how sc\emph{RNA}-seq data is commonly analyzed, before we discuss scBS data analysis.
 
-The starting point in most scRNA-seq analyses is a matrix of UMI counts (i.e.
-of counts of distinct RNA molecules), one row for each cell and one column for each gene.
+The starting point in most scRNA-seq analyses is a matrix of UMI counts (i.e. of counts of distinct RNA molecules), one row for each cell and one column for each gene.
 A first goal is usually to assign cell types or states to cells.
@@
 \textbf{In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-4}$) to the normalized fractions before taking the logarithm.}  % is this even correct?
+-L
 Now, one could use Euclidean distances of these vectors of logarithmized fractions as dissimilarity score.
@@
 Then, we will demonstrate the improvement using benchmarks.
+Finally, we describe an approach to detect differentially methylated regions (DMRs) in scBS data, and showcase it on real data.
 
@@
     \begin{center}
-    \includegraphics[width=\columnwidth]{figures/Fig_sliding.png}
+    \includegraphics[width=0.85\columnwidth]{figures/Fig_sliding.png}
     \end{center}
@@
 
-\begin{figure*}
-    \begin{center}
-    \includegraphics[width=.65\textwidth]{figures/Fig_DMRdetection.png}
-    \end{center}
+\begin{figure}
+    \includegraphics[width=\columnwidth]{figures/Fig_DMRdetection.png}
     \caption{\small \textbf{Identification of differentially methylated regions.}\\
@@
     \label{fig:dmrscan}
-\end{figure*}
+\end{figure}
 
@@
 
-\subsection{Utility of DMRs and VMRs} % not sure if we need a subsection here
-
-While the primary purpose of VMRs is to provide better input for the distance calculations, their small size also means that they can be readily compared with genomic annotation, which aids in their functional interpretation.
-Large 100~kb tiles will often contain multiple genes with distinct expression patterns, small VMRs located in known regulatory elements or gene bodies are more easily associated with individual genes and their associated function.
-Thus, VMRs provide a useful starting point for examining the data with methods from functional genomics.
-To enable a targeted characterization of epigenomic and gene regulatory differences between conditions or cell types, the same principles may also be applied to DMRs.
+While the primary purpose of VMRs is to provide better input for PCA and distance calculations, DMR detection facilitates the discovery of epigenetic differences between conditions or cell types.
 
@@
 
-\begin{figure*}
+\begin{figure}
     \begin{center}
-    \includegraphics[width=.65\textwidth]{figures/Fig_correlation.png}
+    \includegraphics[width=.95\columnwidth]{figures/Fig_correlation.png}
     \end{center}
@@
     \label{fig:correlation}
-\end{figure*}
+\end{figure}
 
@@
     \begin{center}
-        \includegraphics[width=0.7\textwidth]{figures/Fig_benchmark.png}
+        \includegraphics[width=0.8\textwidth]{figures/Fig_benchmark.png}
     \end{center}
@@
         The resulting methylation matrix can then be used for downstream analysis such as cell clustering and dimensionality reduction.
-        Identified groups of cells (e.g. clusters) can then be scanned for DMRs with \texttt{scbs diff}.
+        Identified groups of cells (e.~g. clusters) can then be scanned for DMRs with \texttt{scbs diff}.
     }
@@
 We have implemented the approach just described in a Python package with a command line interface, called \texttt{scbs}, which also offers a number of other functionalities for the analysis of scBS data (Figure \ref{fig:workflow}).
-The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark}, methylpy \citep{methylpy} or biscuit \citep{biscuit}.
+The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark}, methylpy \citep{methylpy} or BISCUIT.
 For each cell, these tools produce a text file that lists the methylation status of all CpG sites.
@@
 
-\subsection{Finding differentially methylated regions}
-
 \subsection{Calculating cell to cell distances}
@@
 \subsection{Re-analysis of published scBS datasets}
-To re-analyze scBS data from \citet{kremer_scnmt}, single-cell CpG methylation reports were first stored with \texttt{scbs prepare} and then smoothed with \texttt{scbs smooth} using the default bandwidth of 1000~bp.
-The mm10 mouse genome was then annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 0.02$, minimum cell requirement $=6$).
-Methylation fractions, commonly represented as percentages, and the shrunken means of residuals were quantified with \texttt{scbs matrix} for both VMRs and tiles.
-The same analysis was repeated for scBS data from \citet{luo2017single}, and for smaller data sets generated by randomly sub-sampling cells separately from both data sets.
-For each data set, we thus obtained four methylation matrices:
-either 100~kb tiles or VMRs, quantified either by simple averaging (methylation fractions) or with the shrunken mean of the residuals.
-All four matrices were then subjected to PCA.
-In the case of methylation fractions, we closely followed methods described in \citet{luo2017single}:
-We selected tiles spanning $\ge20$ CpG sites and with sequencing coverage in at least 70\% of cells.
-We then imputed missing values by imputing the mean for each tile, centered the values, and performed PCA.
-In the case of VMRs, we used iterative PCA to handle missing values, as described earlier.
-For visualization, the first 15 PCs were subjected to UMAP with parameters min\_dist $=0.2$, init $=$ "spca", and n\_neighbors $=\frac{\sqrt{n}}{1.5}$ where $n$ is the total cell number, to adapt to data sets of different size.
+To re-analyze scBS data sets by \citet{kremer_scnmt} and \citet{luo2017single}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 0.02$).
+VMRs with sequencing coverage in less than 
+% WIP
+For all intervals (VMRs and 100~kb tiles), the mean methylation fractions and the shrunken means of the residuals were obtained with \texttt{scbs matrix}.
 
-\subsection{Mean neighbor score} \label{methods:score}
-To assess performance of our methods, we employed a score that quantifies how well cell types (or cell states) are separated in 15-dimensional PCA space.
-For data from \citet{luo2017single}, we used cell type labels that the authors manually curated based on CH-methylation.
-For the multi-omic data set, we repeated the single-cell transcriptomics analysis described in \citet{kremer_scnmt} with two adjustments:
-We did not filter off-target cells such as endothelial cells, and we annotated cell types using Leiden clustering with the Seurat \citep{seurat} function \texttt{FindClusters(resolution = 0.1)}.
-The score, based on the $\Gamma$-score \citep{Kireeva_2014}, varies between 0 and 1, where higher scores reflect better separation of cell types.
-For each cell $j$, we count how many of its $k$ nearest neighbors have been assigned to the same cell type as cell $j$.
-We denote this count $a^k_j$, where we have chosen $k=\frac{\sqrt{n}}{1.5}$.
-The overall score used to evaluate a given combination of methods is then simply the mean of all cell-wise scores.
+Regions observed in less than four measured CpGs, or that were measured in less than 10\% of cells, were discarded.
+To select $n$ intervals for further analysis, the top $n$ intervals were picked based on the number of cells that the region was measured in.
+
+For dimensionality reduction, missing values were imputed with the iterative PCA described in the previous section.
+To visualize data, UMAP was performed on the final 15 principal components.
+
+\subsection{Nearest neighbor count score} \label{methods:score}
+To assess performance of our methods, we processed mouse CpG methylation data from \citet{luo2017single} with several method combinations and scored each result.
+The score evaluates the ability of each method combination to separate cell types in 15-dimensional PCA space, based on their CpG methylation.
+Each score varies between 0 and 1, where higher scores reflect better separation.
+
+Our nearest-neighbor-count score $\text{nnc}_k$, is based on the $\Gamma$-score \citep{Kireeva_2014} and is calculated as follows.
+For each cell $j$, we count how many of its $k$ nearest neighbors have been assigned to the same cluster as cell $j$.
+We denote this count $a^k_j$.
+We consider a cell as ``well-assigned'' if $a^k_j>\theta k$, where we have chosen $\theta=0.9$.
+In other words, a cell is ``well-assigned'' if at least 90\% of its $k$ nearest neighbors have been assigned to the same class as the cell under consideration.
+
+The score $\text{nnc}_k$ is then simply the fraction of cells in the data set that are well-assigned in this sense.
 
@@
 
-\newpage
 \section{Supplementary Figures}
@@
 
-\begin{figure*}
+\begin{figure*}[h]
     \begin{center}

### 2023-03-16 paper.tex
...

--- 2023-03-16 paper.tex
@@
 
+\todo{...}
+
 \subsection{Calculating cell to cell distances}
@@
 In the case of VMRs, we used iterative PCA to handle missing values, as described earlier.
-For visualization, the first 15 PCs were subjected to UMAP with parameters min\_dist $=0.2$, init $=$ "spca", and n\_neighbors $=\frac{\sqrt{n}}{1.5}$ where $n$ is the total cell number, to adapt to data sets of different size.
+For visualization, the first 15 PCs were subjected to UMAP with parameters min\_dist $=0.2$, init $=$ "spca".
+To flexibly adapt to data sets of different sizes, we set n\_neighbors  $=\frac{\sqrt{n}}{1.5}$ (rounded to the nearest integer), where $n$ is the total cell number.
 
@@
 For each cell $j$, we count how many of its $k$ nearest neighbors have been assigned to the same cell type as cell $j$.
-We denote this count $a^k_j$, where we have chosen $k=\frac{\sqrt{n}}{1.5}$, rounded to the nearest integer.
+We denote this count $a^k_j$, where we have chosen $k=\frac{\sqrt{n}}{1.5}$, rounded.
 The overall score used to evaluate a given combination of methods is then simply the mean of all cell-wise scores.

### 2023-04-06 paper.tex
apply Ana's changes

@@
 \small
-\textit{Abstract} -- Single-cell bisulfite sequencing (scBS) allows for assessing DNA methylation at single-base pair and single-cell resolution.
-Analyzing the big data sets thus obtained requires preprocessing to reduce data size, improve signal over noise, and provide interpretability.
-Typically, this is simply done by dividing the genome into large tiles and averaging in these.
-We show that this coarse-graining dilutes signal, and suggest improved approaches to determine more informative regions to quantify methylation in, and a better quantitation method than simple averaging.
-We demonstrate how our approach enables better distinguishing of cell types and reduces demands on cell number.
-We also implement an approach to detect differentially methylated regions (DMRs) between groups of cells, and demonstrate its ability to detect biologically meaningful regions near genes involved in core functions of specific cell types.
-Finally, we present a software tool, \texttt{scbs}, that implements these methods and offers further functionality required to work with scBS data.
+\textit{Abstract} -- Single-cell bisulfite sequencing (scBS) is a technique that enables the assessment of DNA methylation at single-base pair and single-cell resolution.
+However, the analysis of the large datasets obtained from scBS requires preprocessing to reduce data size, improve signal-to-noise ratio, and provide interpretability.
+Typically, this is achieved by dividing the genome into large tiles and averaging the methylation signals within each tile.
+Here, we demonstrate that this coarse-graining approach can lead to signal dilution, and propose improved strategies to identify more informative regions for methylation quantification, and a more accurate quantitation method than simple averaging.
+Our approach enables better discrimination of cell types and reduces the need for large numbers of cells.
+We also present an approach to detect differentially methylated regions (DMRs) between groups of cells, and demonstrate its ability to identify biologically meaningful regions that are associated with genes involved in the core functions of specific cell types.
+To facilitate the analysis of scBS data, we have developed a software tool called "\textttt{scbs}" that implements these methods and provides additional functionality.
+In summary, our work provides a more robust and accurate approach to analyze scBS data, which has the potential to enhance our understanding of DNA methylation dynamics in single cells and improve the identification of epigenetic changes associated with different cell states or diseases.
 \end{minipage}

### 2023-04-06 paper.tex
apply Ana's changes

@@
 We also present an approach to detect differentially methylated regions (DMRs) between groups of cells, and demonstrate its ability to identify biologically meaningful regions that are associated with genes involved in the core functions of specific cell types.
-To facilitate the analysis of scBS data, we have developed a software tool called "\textttt{scbs}" that implements these methods and provides additional functionality.
+To facilitate the analysis of scBS data, we have developed a software tool called ``\texttt{scbs}'' that implements these methods and provides additional functionality.
 In summary, our work provides a more robust and accurate approach to analyze scBS data, which has the potential to enhance our understanding of DNA methylation dynamics in single cells and improve the identification of epigenetic changes associated with different cell states or diseases.

### 2023-04-19 paper.tex
some edits to first two pages

@@
 
-\setcounter{secnumdepth}{0}
-
 \twocolumn[{%
@@
 }]
+\setcounter{secnumdepth}{0}
 
@@
 
-\blfootnote{\hspace{-.5cm}\raggedright\textsuperscript{*} e-mail:  \texttt{l.kremer@dkfz-heidelberg.de} and\\ \hspace{1cm}\texttt{simon.anders@bioquant.uni-heidelberg.de}}
-
 Here, we discuss bisulfite sequencing, a method to study DNA methylation.
@@
 
-In the present paper, we discuss strategies to analyze single-cell bisulfite-sequencing (scBS) data, suggest improvements to current approaches, show their value in a benchmark, and present software to perform the improved analysis methods.
+In the present paper, we discuss strategies to analyze single-cell bisulfite-sequencing (scBS) data.
+We suggest improvements to current approaches, and demonstrate their value in benchmarks, using two real-world datasets.
+Furthermore, we discuss how to perform comparative analyses.
+Finally, we present ``scbs'', a comprehensive software toolkit to perform scBS data analysis, using the described methods.
+\blfootnote{\hspace{-.5cm}\raggedright\textsuperscript{*} e-mail:  \texttt{l.kremer@dkfz-heidelberg.de} and\\ \hspace{1cm}\texttt{simon.anders@bioquant.uni-heidelberg.de}}
 
@@
     \end{center}
-    \caption{\small \textbf{Improved quantification of DNA methylation in genomic intervals.}\\
+    \caption{\small \textbf{Improved quantification of DNA methylation in a given genomic interval.}\\
     \textbf{(A)} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified.
@@
 A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm.
-In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-4}$) \todo{is this even correct?} to the normalized fractions before taking the logarithm.
+In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-4}$) to the normalized fractions before taking the logarithm.
 Now, one could use Euclidean distances of these vectors of logarithmized fractions as dissimilarity score.
@@
 
-This procedure is commonly adapted when working with single-cell DNA methylation data, so that established methods used in scRNA-seq can be used.
-The task is therefore how to construct a suitable matrix as input for PCA.
+This procedure is commonly adapted when working with single-cell DNA methylation data, because once on gets to the PCA step, one can then continue with the established methods just mentioned.
+The task is therefore to construct from the methylation data a matrix suitable as input for PCA.
 A simple and common approach, used for instance by \citet{luo2017single}, is to divide the genome into tiles of e.g.
 100~kb size, and calculate for each cell the average methylation of the DNA within each tile.
-To this end, we identify all CpG sites in the tile for which we have coverage with at least one read and can hence call the CpG to be either unmethylated or methylated in the cell.
-Then we denote as average DNA methylation of this tile in a given cell the proportion of observed CpG sites in the tile that were found to be methylated (Figure \ref{fig:smoothres}A).
-This yields a matrix, with one row for each cell and one column for each genomic tile, comprising numbers (methylation fractions) between 0 and 1.
+To this end, one identifies in the tile all CpG sites that are covered by at least one read and averages their methylation state, i.e., one denotes the as average DNA methylation of the tile in a given cell the proportion of the observed CpG sites in the tile that were found to be methylated (Figure \ref{fig:smoothres}A).
+This yields a matrix, with one row for each cell and one column for each genomic tile, comprising numbers (``methylation fractions'') between 0 and 1.
 This matrix is now subjected to PCA.
@@
 
-While this simple procedure is straight-forward and produces satisfactory results, the coarse resolution of large genomic tiles and the sparse read coverage of single-cell methylomes still pose significant challenges in the analysis of scBS data.
-Here, we propose three improvements that, as we show afterwards, considerably increase the information content of the processed data.
-We will first explain the proposed improvement in a qualitative manner, while mathematical details are given later in the Methods section.
-Then, we will demonstrate the improvement using benchmarks.
-Finally, we describe an approach to detect differentially methylated regions (DMRs) in scBS data, and showcase it on real data.
+While this simple procedure is straight-forward and produces useable results, it is not optimal.
+In this paper, we discuss weaknesses of the simple approach and suggest several refinements to overcome them.
+Using benchmarks and application to real data, we show that our improvements substantially increase the information content of the processed data.
+In the main text, we explain the proposed methods and their motivation in a qualitative manner, while mathematical details are given later in the Methods section.
+Then, we will demonstrate the value of our methods using benchmarks and application to real data.
+Finally, we describe and demonstrate an approach to detect differentially methylated regions (DMRs) in scBS data.
+We also describe our software toolkit, scbs, that facilitates all these analyses.
 
@@
 We first discuss the task of quantifying the level of methylation in a given, fixed, genomic interval.
-In Figure \ref{fig:smoothres}A, such an interval is covered by a single read for two of the three cells shown.
-The read from cell 2 shows much more methylation than the read from cell 1, and the standard analysis would therefore consider that cell as having higher methylation in the interval.
+Typically, read coverage per cell is sparse in scBS data.
+In the example of Figure \ref{fig:smoothres}A, teh depicted interval is covered by a single read for two of the three cells shown and no read in the third.
+The read from cell 2 shows much more methylation than the read from cell 1, and the standard analysis would therefore consider cell 1 to have higher methylation in the interval than cell 2.
 However, given that the two reads agree wherever they overlap, a more parsimonious interpretation would be that the cells do not show difference in methylation within the interval.
@@
 In Figure \ref{fig:smoothres}B, the curved line depicts such an average over all cells, and the red vertical lines show an individual cell's deviation from the ensemble average.
-We take the lengths of the red lines as signed values (``residual''), positive for lines extending upwards from the curve (methylated CpG) and negative for lines extending downwards (unmethylated CpG), and take the average over the residual for all the CpGs in the interval that are covered by reads from this cell.
+We take the lengths of the red lines as signed values (``residuals''), positive for lines extending upwards from the curve (methylated CpG) and negative for lines extending downwards (unmethylated CpG).
+For each cell, we then take the average over the residuals for all the CpGs in the interval that are covered by reads from this cell.
 In this average, we perform shrinkage towards zero via a pseudocount (to trade bias for variance, see Methods for detail) in order to dampen the signal in cells with low coverage of the interval.
 
-The average thus obtained, the shrunken mean of the residuals, is what we use to quantify this cell's (relative) methylation in the interval.
+The average thus obtained, the shrunken mean of the residuals, is what we use to quantify the cell's (relative) methylation in the interval.
 For a genome tiled into such intervals, we thus obtain a matrix, one row per cell, one column per interval, that can be used for downstream analysis, e.g., as input for PCA.
@@
 
+A remaining issue is how to deal with the case that a cell has no reads at all within a given interval.
+Here, it is justified to simply put zero into the matrix element, because a shrunken residual average of zero indicates that there is no evidence of the cell deviating from the mean.
+We slightly refine this by using an iterative imputation within the PCA (see Methdos for details).
+
+\todo{Make sure this is shown in the methods.}
 % remove or shorten this section cause it's not really novel, methods like this are already implemented in e.g. the "missMDA" R package

### 2023-04-19 paper.tex
changed Figure 1 (residuals): introduced blue lines

@@
     \caption{\small \textbf{Improved quantification of DNA methylation in a given genomic interval.}\\
-    \textbf{(A)} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified.
+    \textbf{(A)} Depicted is a genomic interval (between vertical blue lines) along a chromosome, for which DNA methylation is to be quantified.
     Two cells cover differing parts of the interval with one read each.
@@
 Typically, some regions in a chromosome will have very similar methylation status in all cells, while other regions show variability in methylation across cells.
-For instance, it is long known that CpG-rich promoters of housekeeping genes are unmethylated, and that a large proportion of the remaining genome is highly methylated regardless of cell type \citep{bird1986cpg}.
-In contrast, DNA methylation at certain genomic features such as enhancers is more dynamic \citep{argelaguet2019gastru}, and thus more variable across cells.
+For instance, it is known  since long that CpG-rich promoters of housekeeping genes are unmethylated, and that a large proportion of the remaining genome is highly methylated regardless of cell type \citep{bird1986cpg}.
+In contrast, DNA methylation at certain genomic features such as enhancers is more dynamic (e.g., \citet{argelaguet2019gastru}), and thus more variable across cells.
 Only the latter regions are of value for our goal of quantitating dissimilarity between cells.
@@
 In the standard approach, one divides up (tiles) each chromosome into non-overlapping, equal-sized intervals, and quantitates the methylation of each tile.
-Such rigid placing of interval boundaries is unlikely to be optimal: for example, a VMR might be much smaller than a tile and its signal will hence be drowned out by the larger part of uninformative CpG sites that are equal in all cells, when averaging over all the CpG sites in the tile.
+Such rigid placing of interval boundaries is unlikely to be optimal: for example, a VMR might be much smaller than a tile and the signal from its CpG sites will hence be drowned out by the larger number of uninformative CpG sites that are equal in all cells, when averaging over all the CpG sites in the tile.
 
-Therefore, we propose the following approach (Fig.~\ref{fig:vmr}): Divide up the chromosome into many \emph{overlapping} windows, that start at regular multiples of a fixed, small, step size.
+Therefore, we propose the following approach (Fig.~\ref{fig:vmr}): Divide up the chromosome into many \emph{overlapping} windows that start at regular multiples of a fixed, small, step size.
 Quantify the methylation of each cell in each window by averaging the cell's methylation residuals over all CpGs in the window, as described above and depicted in Fig.~\ref{fig:smoothres}B.

### 2023-04-19 paper.tex
made suppl figs non-floating

@@
 \usepackage{graphicx}
+\usepackage{caption}
 
@@
 
-S.\ A.\ and L.\ P.\ M.\ K.\ acknowledge funding by the Klaus Tschira Foundation.
-L.\ P.\ M.\ K., L.\ K., S.\ C.\ and A.\ M.-V.\ acknowledge funding from the European Commission via ERC grant 771376.
+SA and LPMK acknowledge funding by the Klaus Tschira Foundation (project 00.022.2019).
+LPMK, LK, SC, and AMV acknowledge funding from the European Commission via ERC grant 771376.
 We thank Alexey Uvarovskii for code refactoring and helpful suggestions on the source code.
+\bigskip
 
@@
 
+\onecolumn 
 \newpage
 \section{Supplementary Figures}
+\bigskip
 
@@
 
-\begin{figure*}
+\noindent%
+\begin{minipage}{\linewidth}
 	\begin{center}
-        \includegraphics[width=0.8\columnwidth]{figures/SFig_benchmark.png}
+	    \includegraphics[width=0.6\textwidth]{figures/SFig_benchmark.png}
 	\end{center}
-    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of neural subtypes of the murine frontal cortex}.\\
+	\captionof{figure}{\small \textbf{Benchmark of our methods on single-cell methylomes of neural subtypes of the murine frontal cortex}.\\
 	Our methods were benchmarked on single-cell methylomes from \citet{luo2017single} as described in Fig.~\ref{fig:score}. 
@@
 	\label{fig:score_luo}
-\end{figure*}
+\end{minipage}
+\bigskip
 
-\begin{figure*}
+\noindent%
+\begin{minipage}{\linewidth}
     \begin{center}
@@
     \end{center}
-    \caption{\small \textbf{Effect of VMR detection parameters on separation of cell types}.\\
+    \captionof{figure}{\small \textbf{Effect of VMR detection parameters on separation of cell types}.\\
     \textbf{(A-B)} Mean neighbor scores obtained after analyzing single-cell methylomes from \citet{kremer_scnmt} (\textbf{A}) or \citet{luo2017single} (\textbf{B}) with our proposed methods.
@@
     \label{fig:sweep}
-\end{figure*}
+\end{minipage}
 

### 2023-04-19 paper.tex
chaned bibtex style to inlcude DOI links

@@
 \usepackage{natbib}
-\bibliographystyle{genome_research}
+\bibliographystyle{simon}
 \setlength{\bibsep}{0.3ex plus 0.2ex}
@@
 
-January 2023
+April 2023
 \vspace{5.5ex}
@@
 \begin{minipage}{0.7\textwidth}
-\small
-\textit{Abstract} -- Single-cell bisulfite sequencing (scBS) is a technique that enables the assessment of DNA methylation at single-base pair and single-cell resolution.
-However, the analysis of the large datasets obtained from scBS requires preprocessing to reduce data size, improve signal-to-noise ratio, and provide interpretability.
+\small \setlength{\parindent}{1em}
+\noindent \textit{Abstract} -- Single-cell bisulfite sequencing (scBS) is a technique that enables the assessment of DNA methylation at single-base pair and single-cell resolution.
+The analysis of the large datasets obtained from scBS requires preprocessing to reduce data size, improve signal-to-noise ratio, and provide interpretability.
 Typically, this is achieved by dividing the genome into large tiles and averaging the methylation signals within each tile.
+
 Here, we demonstrate that this coarse-graining approach can lead to signal dilution, and propose improved strategies to identify more informative regions for methylation quantification, and a more accurate quantitation method than simple averaging.
@@
 We also present an approach to detect differentially methylated regions (DMRs) between groups of cells, and demonstrate its ability to identify biologically meaningful regions that are associated with genes involved in the core functions of specific cell types.
+
 To facilitate the analysis of scBS data, we have developed a software tool called ``\texttt{scbs}'' that implements these methods and provides additional functionality.
-In summary, our work provides a more robust and accurate approach to analyze scBS data, which has the potential to enhance our understanding of DNA methylation dynamics in single cells and improve the identification of epigenetic changes associated with different cell states or diseases.
+
+%In summary, our work provides a more robust and accurate approach to analyze scBS data, which has the potential to enhance our understanding of DNA methylation dynamics in single cells and improve the identification of epigenetic changes associated with different cell states or diseases.
 \end{minipage}
@@
 
-Therefore, we propose to first obtain, for each CpG position, an average of the methylation across all cells, and then quantify for each individual cell its deviation from this average.
+Therefore, we propose to first obtain, for each CpG position, an average of the methylation across all cells, and then quantify each cell's deviation from this average.
 In Figure \ref{fig:smoothres}B, the curved line depicts such an average over all cells, and the red vertical lines show an individual cell's deviation from the ensemble average.
@@
 Here, it is justified to simply put zero into the matrix element, because a shrunken residual average of zero indicates that there is no evidence of the cell deviating from the mean.
-We slightly refine this by using an iterative imputation within the PCA (see Methdos for details).
-
-\todo{Make sure this is shown in the methods.}
-% remove or shorten this section cause it's not really novel, methods like this are already implemented in e.g. the "missMDA" R package
-%\subsection{PCA with imputation}
-%The standard PCA algorithm cannot deal with missing data.
-%Therefore, one would need to make intervals wide enough to ensure that every interval is covered by at least one read for most tiles and discard any interval that is not covered by any single cell.
-%Hence, a naive analysis has to resort to a coarse tiling of the genome.
-%To counter this, we propose a simple and straightforward way to deal with missing data in the input matrix to a PCA: In a first iteration, we replace each missing value in the centered input matrix with zero, then run the PCA.
-%Then, only these zeroes are replaced by the value predicted by the PCA and the PCA is rerun.
-%Convergence is typically achieved after two to five iterations.
-%For details, see Methods.
-%The possibility to perform PCA with missing data allows us to use shorter, and thus more intervals and hence provide the PCA with richer input data.
+We slightly refine this by using an iterative imputation within the PCA (see Methods for details).
 
@@
 
+\begin{figure*}
+	\begin{center}
+		\includegraphics[width=.65\textwidth]{figures/Fig_correlation.png}
+	\end{center}
+	\caption{\small \textbf{Correlation of DNA methylation and gene expression}.\\
+		\textbf{(A)} \todo{maybe remove panel A or B, and put the remaining panel left of C?} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red).
+		\textbf{(B)} Correlation of each gene's RNA expression values to methylation (shrunken mean of residuals) at promoters (y-axis), or to methylation of the nearest VMR (x-axis).
+		Promoters are defined as intervals \textpm2~kb around the TSS.
+		\textbf{(C)} Mean methylation near the gene Htra1 (ENSMUSG00000006205).
+		Cells are assigned to 4 groups based on Htra1 expression (group 0: cells with no Htra1 expression, groups 1-3: cells that express Htra1, divided into three equally large groups with group 3 having the highest expression).
+		Data from \citet{kremer_scnmt}.}
+	\label{fig:correlation}
+\end{figure*}
+
 A common task in the analysis of bulk bisulfite-sequencing data is the detection of differentially methylated regions (DMRs) between conditions, tissues, or cell types \citep{Hebestreit2013, dmrseq}.
@@
 
-\begin{figure*}
-    \begin{center}
-    \includegraphics[width=.65\textwidth]{figures/Fig_correlation.png}
-    \end{center}
-    \caption{\small \textbf{Correlation of DNA methylation and gene expression}.\\
-    \textbf{(A)} \todo{maybe remove panel A or B, and put the remaining panel left of C?} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red).
-    \textbf{(B)} Correlation of each gene's RNA expression values to methylation (shrunken mean of residuals) at promoters (y-axis), or to methylation of the nearest VMR (x-axis).
-    Promoters are defined as intervals \textpm2~kb around the TSS.
-    \textbf{(C)} Mean methylation near the gene Htra1 (ENSMUSG00000006205).
-    Cells are assigned to 4 groups based on Htra1 expression (group 0: cells with no Htra1 expression, groups 1-3: cells that express Htra1, divided into three equally large groups with group 3 having the highest expression).
-    Data from \citet{kremer_scnmt}.}
-    \label{fig:correlation}
-\end{figure*}
-
-\subsection{Our proposed methods improve identification of cell types}
-
 \begin{figure*}[t]
@@
 \end{figure*}
+
+\subsection{Our proposed methods improve identification of cell types}
 
@@
 
+We note that similar approaches have also been used elsewhere, e.g., in the missMDA method of \citet{missMDA}.
+
 \subsection{Re-analysis of published scBS datasets}
@@
 \section{Supplementary Figures}
-\bigskip
+\vspace{6ex}
 
@@
 	\begin{center}
-	    \includegraphics[width=0.6\textwidth]{figures/SFig_benchmark.png}
+	    \includegraphics[width=0.4\textwidth]{figures/SFig_benchmark.png}
 	\end{center}
@@
 \end{minipage}
-\bigskip
+\vspace{6ex}
 

### 2023-04-19 paper.tex
moved mention of iterated PCA

@@
 
-A remaining issue is how to deal with the case that a cell has no reads at all within a given interval.
+A minor remaining issue is how to deal with the case that a cell has no reads at all within a given interval.
 Here, it is justified to simply put zero into the matrix element, because a shrunken residual average of zero indicates that there is no evidence of the cell deviating from the mean.
 We slightly refine this by using an iterative imputation within the PCA (see Methods for details).
+
+Taken together, this shrunken mean of residuals quantitation reduces variance in comparison to simple averaging of raw methylation calls. We will show further below how this improves results.
 

### 2023-04-19 paper.tex
a few more edits

@@
 
-\section{Improved processing of scBS data}
+\section{Improved preprocessing of scBS data}
 
@@
 
+\begin{figure*}
+	\begin{center}
+		\includegraphics[width=.65\textwidth]{figures/Fig_correlation.png}
+	\end{center}
+	\caption{\small \textbf{Correlation of DNA methylation and gene expression}.\\
+		\textbf{(A)} \todo{maybe remove panel A or B, and put the remaining panel left of C?} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red): for most promotors, correlation is weak.
+		\textbf{(B)} Correlation of each gene's RNA expression values to methylation (shrunken mean of residuals) at promoters (y-axis), or to methylation of the nearest VMR (x-axis).
+\todo{Yes, I would remove B.}
+		Promoters are defined as intervals \textpm2~kb around the TSS.
+		\textbf{(C)} Mean methylation near the gene Htra1 (ENSMUSG00000006205).
+		Cells are assigned to 4 groups based on Htra1 expression (group 0: cells with no Htra1 expression, groups 1-3: cells that express Htra1, divided into three equally large groups with group 3 having the highest expression).
+		Data from \citet{kremer_scnmt}.}
+	\label{fig:correlation}
+\end{figure*}
+
 Typically, some regions in a chromosome will have very similar methylation status in all cells, while other regions show variability in methylation across cells.
@@
 As we demonstrate below, a PCA performed on such a matrix provided a distance metric for the cells that contains more information on biological detail than one from a simpler analysis.
-Since not all genomic intervals have sequencing coverage in all cells, the resulting matrix contains missing values.
-To enable PCA in the presence of missing values, we propose to use a simple iterative approach (for details see Methods).
 
-\subsection{Finding differentially methylated regions}
+\begin{figure*}
+	\begin{center}
+		\includegraphics[width=0.7\textwidth]{figures/Fig_benchmark.png}
+	\end{center}
+	\caption{\small \textbf{Benchmark of our methods on single-cell multi-omic data of cells of the murine forebrain}.\\
+		\textbf{(A)} Cell labels based on clustering of single-cell transcriptomes from \citet{kremer_scnmt}.
+		\textbf{(B, C)} Exemplary UMAPs obtained when analyzing the data set with our proposed methods (B) or conventional methods based on genome tiling (C).
+		\textbf{(D)} Illustration of the neighbor score. Note that, actually,
+		distances are measured in 15-dimensional PC space, not in the 2-dimensional
+		UMAP space.
+		\textbf{(E)} Mean neighbor score obtained after analysing single-cell methylomes with different combinations of methods.
+		Either 100~kb genomic tiles or VMRs were subjected to PCA and UMAP.
+		Tiles or VMRs were pre-processed as proposed in this work (shrunken mean of residuals subjected to iterative PCA) or using conventional methods (PCA on methylation percentages, lightly imputed, as done in \citet{luo2017single}).
+		The full 1566-cell data set was sub-sampled to simulate smaller data sets (x axis).
+		A higher neighbor score implies better separation of cell types inferred from single-cell transcriptomes of the same cells \textbf{(A)}.
+	}
+	\label{fig:score}
+\end{figure*}
+
+\subsection{Application and benchmarks}
+
+To demonstrate the value of our proposed improvements, we employ our own data \citep{kremer_scnmt} as well as a previously published data set, namely the one by  \citet{luo2017single} already mentioned in the introduction.
+
+\phantomsection
+\subsubsection{Correlating VMR methylation with gene expression}
+
+Our data set \citep{kremer_scnmt} comprises the methylomes of 1566 cells isolated from mouse forebrains, as well as matched single-cell transcriptomes of the same cells.
+Among these cells are distinct cell types such as oligodendrocytes, oligodendrocyte precursor cells (OPCs) and endothelial cells, as well as cellular sub-states that are part of the continuous neural stem cell differentiation trajectory.
+To assess whether our VMR detection method captures genomic intervals that are biologically meaningful, we probed whether their methylation level correlates with the expression of nearby genes.
+
+We first note that gene expression is more strongly correlated with the methylation of nearby VMRs than with the methylation of their promotors (Fig.~\ref{fig:correlation}A), indicating that VMR methylation is often a better predictor of gene expression than promoter methylation.
+Indeed, a gene-wise comparison revealed many genes whose expression is correlated with the nearest VMR, but not with promoter methylation (Fig.~\ref{fig:correlation}B).
+One such example gene, Htra1, is depicted in Fig.~\ref{fig:correlation}C.
+While the promoter of this gene is lowly methylated regardless of gene expression, a VMR located downstream of the promoter is lowly methylated in cells with high Htra1 expression.
+
+\subsubsection{Improved identification of cell types}
+
+We next tested whether our methods improve the ability to distinguish cell types and cell states (Fig.~\ref{fig:score}).
+To this end, we obtained cell type/state labels based on the single-cell transcriptomes from \citet{kremer_scnmt} (Fig.~\ref{fig:score}A).
+We consider these cell labels as ground truth and tested whether we are able to distinguish the same groups of cells based on their methylomes.
+To do this, we subjected the methylomes to four variants of the workflow described above: we defining genomic intervals in two ways and quantified methylation in these intervals in two ways, then always subjected the resulting matrix to PCA.
+To define the intervals, we either simply tiled the genome into 100~kb bins as is commonly done, or we used our VMR detection approach.
+To quantify methylation in these intervals, we either averaged in these intervals, obtaining methylation percentages (``classic'' pre-processing), or we calculated the shrunken mean of residuals.
+
+Visual inspection of the resulting UMAPs revealed that our proposed combination of methods results in more clearly separated cell types, compared to a UMAP obtained with default analysis methods (Fig.~\ref{fig:score}B,C).
+While all cells form a continuous point cloud when using default methods, our improvements led to a clear separation of oligodendrocytes, OPCs and endothelial cells.
+Furthermore, even cellular sub-types of cells in the continuous neural stem cell lineage were partially separated.
 
@@
 
-\begin{figure*}
-	\begin{center}
-		\includegraphics[width=.65\textwidth]{figures/Fig_correlation.png}
-	\end{center}
-	\caption{\small \textbf{Correlation of DNA methylation and gene expression}.\\
-		\textbf{(A)} \todo{maybe remove panel A or B, and put the remaining panel left of C?} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red).
-		\textbf{(B)} Correlation of each gene's RNA expression values to methylation (shrunken mean of residuals) at promoters (y-axis), or to methylation of the nearest VMR (x-axis).
-		Promoters are defined as intervals \textpm2~kb around the TSS.
-		\textbf{(C)} Mean methylation near the gene Htra1 (ENSMUSG00000006205).
-		Cells are assigned to 4 groups based on Htra1 expression (group 0: cells with no Htra1 expression, groups 1-3: cells that express Htra1, divided into three equally large groups with group 3 having the highest expression).
-		Data from \citet{kremer_scnmt}.}
-	\label{fig:correlation}
-\end{figure*}
-
-A common task in the analysis of bulk bisulfite-sequencing data is the detection of differentially methylated regions (DMRs) between conditions, tissues, or cell types \citep{Hebestreit2013, dmrseq}.
-As DNA methylation affects gene expression, DMRs can provide insights into the unique epigenetic and gene regulatory characteristics of cell types.
-However, to date no approach to detect DMRs in scBS data has been reported.
-To enable DMR detection in scBS data, we thus developed an approach that detects DMRs of variable size between two groups of cells, and controls the false discovery rate (Fig.~\ref{fig:dmrscan}):
-
-Similar to the previously described VMR identification approach (Fig.~\ref{fig:vmr}), divide each chromosome into overlapping windows shifted by a small and fixed size (step size) and quantify the methylation of each cell in each window.
-Next, instead of the variance, obtain the t-statistic % Welch t-test
-as a measure of differential methylation between the two cell groups.
-Identify the windows with the most extreme t-statistics, e.g. windows in the 2\% upper and lower tails.
-Merge any overlapping windows in the upper tail into larger DMRs.
-Repeat for windows in the lower tail, then re-calculate the t-statistic for each larger DMR.
-
-To assess statistical significance of DMRs, repeat the same procedure on permutations of the scBS data, i.e. the same data set with randomly shuffled cell labels.
-The DMR t-statistics obtained from permuted data are then used to estimate the false-discovery rate (FDR), yielding an adjusted p-value for each DMR.
-
-While the primary purpose of VMRs is to provide better input for PCA and distance calculations, DMR detection facilitates the discovery of epigenetic differences between conditions or cell types.
-
-\section{Application and benchmarks}
-
-To demonstrate the value of our proposed improvements, we obtained two previously published scBS data sets and re-analyzed them using our methods.
-
-\phantomsection
-\subsection{Correlating VMR methylation with gene expression}
-
-The first data set used to test our methods is a single-cell multi-omics data set generated by \citet{kremer_scnmt}.
-This data set comprises the methylomes of 1566 cells isolated from mouse forebrains, as well as matched single-cell transcriptomes of the same cells.
-Among these cells are distinct cell types such as oligodendrocytes, oligodendrocyte precursor cells (OPCs) and endothelial cells, as well as cellular sub-states that are part of the continuous neural stem cell differentiation trajectory.
-To assess whether our VMR detection method captures genomic intervals that are biologically meaningful, we probed whether their methylation level correlates with the expression of nearby genes.
-Compared to promoter methylation, the obtained correlation coefficients are more extreme, indicating that VMR methylation often is a better predictor of gene expression than promoter methylation (Fig.~\ref{fig:correlation}A).
-Indeed, a gene-wise comparison revealed many genes whose expression is correlated with the nearest VMR, but not with promoter methylation (Fig.~\ref{fig:correlation}B).
-One such example gene, Htra1, is depicted in Fig.~\ref{fig:correlation}C.
-While the promoter of this gene is lowly methylated regardless of gene expression, a VMR located downstream of the promoter is lowly methylated in cells with high Htra1 expression.
-
-\begin{figure*}[t]
-    \begin{center}
-        \includegraphics[width=0.7\textwidth]{figures/Fig_benchmark.png}
-    \end{center}
-    \caption{\small \textbf{Benchmark of our methods on single-cell multi-omic data of cells of the murine forebrain}.\\
-    \textbf{(A)} Cell labels based on clustering of single-cell transcriptomes from \citet{kremer_scnmt}.
-    \textbf{(B, C)} Exemplary UMAPs obtained when analyzing the data set with our proposed methods (B) or conventional methods based on genome tiling (C).
-    \textbf{(D, E)} Mean neighbor score \textbf{(D)} obtained after analysing single-cell methylomes with different combinations of methods.
-    Either 100~kb genomic tiles or VMRs were subjected to PCA and UMAP.
-    Tiles or VMRs were pre-processed as proposed in this work (shrunken mean of residuals subjected to iterative PCA) or using conventional methods (PCA on lightly imputed methylation percentages, based on \citep{luo2017single}).
-    The full 1566-cell data set was sub-sampled to simulate smaller data sets (x axis).
-    A higher neighbor score implies better separation of cell types inferred from single-cell transcriptomes of the same cells \textbf{(A)}.
-    }
-    \label{fig:score}
-\end{figure*}
-
-\subsection{Our proposed methods improve identification of cell types}
-
-To assess the impact of our proposed methods, we next tested whether our methods improve the ability to distinguish cell types and cell states (Fig.~\ref{fig:score}).
-To this end, we obtained cluster labels based on the single-cell transcriptomes from \citet{kremer_scnmt} (Fig.~\ref{fig:score}A).
-We consider these cell labels as ground truth and next tested whether we are able to distinguish the same groups of cells based on their methylomes.
-To do this, we subjected the methylomes to the standard scBS workflow:
-defining a set of genomic intervals, quantifying methylation in these intervals, and subjecting the resulting matrix to PCA.
-To determine genomic features for distance calculation, we either tiled the genome into 100~kb bins as is commonly done, or we used our VMR detection approach.
-To quantify methylation in these intervals, we either averaged in these intervals, obtaining methylation percentages ("classic" pre-processing), or we calculated the shrunken mean of residuals.
-Visual inspection of the resulting UMAPs revealed that our proposed combination of methods results in more clearly separated cell types, compared to a UMAP obtained with default analysis methods (Fig.~\ref{fig:score}B,C).
-While all cells form a continuous point cloud when using default methods, our improvements led to a clear separation of oligodendrocytes, OPCs and endothelial cells.
-Furthermore, even cellular sub-types of cells in the continuous neural stem cell lineage were partially separated.
-
-To quantify this performance gain in a more rigorous manner, we used a score that quantifies whether cells were placed in a neighborhood consisting of cells with the same cell type in 15-dimensional PC space ("neighbor score", for details see Methods).
-Since scBS protocols are costly and labor-intensive, only few laboratories are currently able to obtain thousands of single cell methylomes.
-To simulate smaller data sets, we sub-sampled the 1566-cell data set into smaller data sets, analyzed them with different method combinations, and calculated the mean neighbor score for each (Figure \ref{fig:score}D,E).
+To quantify this performance gain in a more rigorous manner, we used a score that quantifies whether cells were placed, in 15-dimensional PC space, in a neighborhood comprising cells of the same cell type ("neighbor score", see Figure \ref{fig:score}D, and Methods for details).
+We then ask how this score depends on the number of available cells.
+This is relevant as scBS protocols are costly and labor-intensive, and only few laboratories are currently able to obtain thousands of single cell methylomes.
+To simulate smaller data sets, we sub-sampled the 1566-cell data set into smaller data sets, analyzed them again with the four workflow variants, and calculated the mean neighbor score for each (Figure \ref{fig:score}E).
 The results demonstrate that the forebrain cell clusters are more cleanly separated when using our proposed methods.
@@
 
-We repeated this benchmark on a second data set published by \citet{luo2017single}.
-This data set consists of 3377 single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal subtypes.
+We repeated this benchmark on the data set published by \citet{luo2017single}, which consists of 3377 single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal subtypes.
 The authors annotated cell types based on CH-methylation instead of CpG-methylation.
-Using these labels as ground truth we re-analyzed the full and sub-sampled data set with four methods combinations and found that our suggested methods consistently outperformed default methods (Fig.~\ref{fig:score_luo}).
+Using these labels as ground truth we re-analyzed the full and sub-sampled CpG methylation data with the four methods combinations and found that our suggested methods consistently outperformed default methods (Fig.~\ref{fig:score_luo}).
+
+\subsubsection{Robustness to parameter changes}
 
 Lastly, we assessed whether our proposed workflow requires fine-tuning of parameters (Fig.~\ref{fig:sweep}).
-To this end, we re-analyzed both data sets, as well as sub-samples of the data with different VMR detection parameters (the width of the sliding window in bp, \texttt{--bandwidth}, default 2000 and the variance threshold above which windows are merged to VMRs, \texttt{--var-threshold}, default 0.02).
-This parameter sweep showed that our workflow is robust to various parameter selections and depends mostly on the overall cell number.
-In the more challenging data set (Fig.~\ref{fig:sweep}A), cell types were less cleanly separated when very large bandwidths, or very strict variance thresholds were selected.
-However, in our experience small bandwidths or lenient thresholds resulted in a much higher number of VMRs and thus longer computing times.
-Our default parameter combination provided good results and fast compute times in both data sets.
+To this end, we re-analyzed both data sets, as well as sub-samples of the data with different VMR detection parameters, namely the width of the sliding window in bp, (set with option \texttt{--bandwidth} in our ``scbs'' software, default 2000 )and the variance threshold above which windows are merged to VMRs (\texttt{--var-threshold}, default 0.02).
+This parameter sweep showed that our workflow gives good results over a wide range of parameter values.
+For the data of \citet{luo2017single}, results are nearly independent of the parameters (Fig.~\ref{fig:sweep}B).
+In the more challenging data set of \citet{kremer_scnmt} (Fig.~\ref{fig:sweep}A), cell types were less cleanly separated when very large bandwidths, or very strict variance thresholds were selected.
+However, very small bandwidths or very lenient thresholds resulted in a much higher number of VMRs and thus long computing times.
+Overall, however, our default parameter combination provided good results and fast compute times in both data sets.
 
-\phantomsection
-\subsection{Detecting DMRs between oligodendrocytes and neural stem cells}
-
-\begin{figure*}[t]
+\begin{figure*}
 	\begin{center}
@@
 
-Next, we used the single-cell multi-omics data set \citep{kremer_scnmt} to evaluate our DMR detection approach.
+\section{Finding differentially methylated regions}
+
+A common task in the analysis of bulk bisulfite-sequencing data is the detection of differentially methylated regions (DMRs) between conditions, tissues, or cell types \citep{Hebestreit2013, dmrseq}.
+As DNA methylation affects gene expression, DMRs can provide insights into the unique epigenetic and gene regulatory characteristics of cell types.
+However, to date no approach to detect DMRs in scBS data has been reported.
+To enable DMR detection in scBS data, we thus developed an approach that detects DMRs of variable size between two groups of cells, and controls the false discovery rate (Fig.~\ref{fig:dmrscan}):
+
+Similar to the previously described approach for VMR identification (Fig.~\ref{fig:vmr}), we divide each chromosome into overlapping windows shifted by a small and fixed size (step size) and quantify the methylation of each cell in each window.
+Next, instead of the variance, we obtain the t-statistic % Welch t-test
+as a measure of differential methylation between the two cell groups.
+We identify the windows with the most extreme t-statistics, e.g.
+windows in the 2\% upper and lower tails.
+We the merge any overlapping windows in the upper tail into larger DMRs, and do the same for windows in the lower tail, and then we re-calculate the t-statistic for each larger DMR.
+
+To assess statistical significance of DMRs, we repeat the same procedure on permutations of the scBS data, i.e.
+the same data set with randomly shuffled cell labels.
+The DMR t-statistics obtained from permuted data are then used to estimate the false-discovery rate (FDR), yielding an adjusted p-value for each DMR.
+
+While the primary purpose of VMRs is to provide better input for PCA and distance calculations, DMR detection facilitates the discovery of epigenetic differences between conditions or cell types, as we demonstrate next.
+
+\phantomsection
+\subsection{Detecting DMRs between oligodendrocytes and neural stem cells}
+
+We used the single-cell multi-omics data set \citep{kremer_scnmt} to evaluate our DMR detection approach.
 We first selected two cell populations from the healthy murine ventricular-subventricular zone: neural stem cells (NSCs, 130 cells) and oligodendrocytes (58 cells).
-Then we used the method depicted in Fig.~\ref{fig:dmrscan} to identify DMRs between these two cell types (Fig.~\ref{fig:dmr}).
-To estimate the false discovery rate and obtain adjusted p-values, we repeated the DMR detection approach on a permutation of the same data set, generated by shuffling cell labels.
-DMRs detected in this manner showed more modest methylation differences, and as a result many DMRs detected in the original data were assigned low adjusted p-values (Fig.~\ref{fig:dmr}A).
+Then we used the method just described to identify DMRs between these two cell types (Fig.~\ref{fig:dmr}A).
+Repeating this after permuting the cell-type labels, in order to obtain a null distribution fo the t statistics, yielded
+DMRs with much weaker methylation differences. Consequently, we could assign to many of the DMRs detected in the unpermuted data a low adjusted p-values (colours in Fig.~\ref{fig:dmr}A).
 
@@
 This demonstrates that our DMR detection approach is able to identify biologically meaningful DMRs, even in scBS data sets of modest cell number.
+
 Fig.~\ref{fig:dmr}C-D depicts an exemplary DMR, located at the gene body of myelin-basic protein (Mbp), the major component of myelin that is essential for myelination of neuronal axons \citep{mbp}.
@@
 
-\section{The \texttt{scbs} Python package}
+\section{The \texttt{scbs} software toolkit}
 
@@
 We have implemented the approach just described in a Python package with a command line interface, called \texttt{scbs}, which also offers a number of other functionalities for the analysis of scBS data (Figure \ref{fig:workflow}).
-The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark}, methylpy \citep{methylpy} or BISCUIT.
+The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark}, methylpy \citep{methylpy} or BISCUIT \citep{biscuit}.
 For each cell, these tools produce a text file that lists the methylation status of all CpG sites.
@@
 To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED file can be used as input for \texttt{scbs matrix}, which quantifies methylation at genomic intervals in all cells.
-The command produces both the percentage of methylated CpG sites, as well as our proposed methylation measure (shrunken mean of the residuals) that is more robust to variation in sequencing coverage or stochastic differences in read position between cells.
-We note that both \texttt{matrix} and \texttt{profile} accept any valid BED file as input, which means that the user can quantify and visualize methylation at any genomic feature of interest, such as promoters, enhancers or transcription factor binding sites, in single cells.
-The obtained methylation matrix can then serve as input for common single-cell analysis methods such as dimensionality reduction and cell clustering.
+The command produces both, the simple percentage of methylated CpG sites, as well as our proposed methylation measure, the shrunken mean of the residuals, which is more robust to variation in sequencing coverage or stochastic differences in read position between cells.
+We note that both \texttt{matrix} and \texttt{profile} accept any valid BED file as input, which means that the user can quantify and visualize methylation at any set of genomic features of interest, such as promoters, enhancers or transcription factor binding sites, obtaining one profile plot per cell.
+
+The methylation matrix produces by the \texttt{matrix} function can then serve as input for standard single-cell analysis tools like Seurat \citep{seurat} or scanpy \citep{scanpy}, in order to perform, e.g., dimensionality reduction and cell clustering.
+For this, the methylation matrix should in lieu of the log-normalized expression matrix usually expected by these tools.
 
@@
 \texttt{scbs} is free and open source.
-The package is available on the Python Package Index (PyPI) and can be installed with the Python package installer pip.
+The package is available on the Python Package Index (PyPI) and can be installed with the Python package installer \texttt{pip}.
 The source code is hosted at \href{https://github.com/LKremer/scbs}{https://github.com/LKremer/scbs}, where we also provide detailed documentation including a tutorial that demonstrates a complete \texttt{scbs} analysis on an example data set.
@@
 Here, we have proposed an improved strategy to preprocess single-cell bisulfite sequencing data.
-Based on the observation that incomplete read coverage of a genomic interval can lead to inaccurate methylation estimates, we suggest a scoring scheme that is aware of a reads local context rather than just treating all reads in an interval alike.
+Based on the observation that incomplete read coverage of a genomic interval can lead to inaccurate methylation estimates, we suggest a scoring scheme that is aware of a read's local context rather than just treating all reads in an interval alike.
 
@@
 By re-analyzing two published data sets, we could show that these improvements to data preprocessing help to increase signal and decrease noise, resulting in a more informative intermediate-dimensional representation of the data.
-As examples of practical benefits, we demonstrate that our preprocessing allows for better distinction of cell sub-types, especially for challenging data sets comprising cellular sub-states and lineage transitions.
+As examples of practical benefits, we demonstrate that our preprocessing allows for better distinction of cell sub-types, especially for challenging data sets comprising cellular sub-states and lineage transitions or for data sets with small cell number.
 
@@
 We also presented and published an open source software tool, called \texttt{scbs}, that provides an easy-to-use implementation of the described algorithms.
-The package also offers functionality for all other steps of data preprocessing that are needed to get from the output of methylation callers such as Bismark and methylpy to a cell\texttimes region matrix that is suitable as input for standard analysis tools that work in a reduced-dimensionality setting, thus allowing the user to use, e.g., well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc.
+The package also offers functionality for all other steps of data preprocessing that are needed to get from the output of methylation callers such as Bismark and methylpy to a cell\texttimes region matrix that is suitable as input for standard analysis tools that work in a reduced-dimensionality setting.
+This allows the user to use well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc.
 By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.

### 2023-04-19 paper.tex
finished editing, appended Methods, changed callouts etc

@@
 \usepackage{caption}
+\usepackage{balance}
+\usepackage{palatino}
+\usepackage[varqu]{inconsolata}
 
 \usepackage[utf8]{inputenc}
-\usepackage{hyperref}
-\urlstyle{same} % no tt font for URLs
 \usepackage{natbib}
@@
 \usepackage[dvipsnames]{xcolor}
+\usepackage[compact]{titlesec}
+\usepackage[hypertexnames=false]{hyperref}
 \hypersetup{
@@
 }
+\urlstyle{same} % no tt font for URLs
 \renewcommand{\topfraction}{0.9}
@@
 \displaywidowpenalty=1000
-
-\usepackage[compact]{titlesec}
 
@@
 
-\phantomsection  % idk why we need this to silence hyperref's warning
+%\phantomsection  % idk why we need this to silence hyperref's warning
 \subsection{Standard approach}
@@
 
-\phantomsection
+%\phantomsection
 \subsection{Read-position aware quantitation} \label{residuals}
@@
 
-\begin{figure*}
-	\begin{center}
-		\includegraphics[width=.65\textwidth]{figures/Fig_correlation.png}
-	\end{center}
-	\caption{\small \textbf{Correlation of DNA methylation and gene expression}.\\
-		\textbf{(A)} \todo{maybe remove panel A or B, and put the remaining panel left of C?} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red): for most promotors, correlation is weak.
-		\textbf{(B)} Correlation of each gene's RNA expression values to methylation (shrunken mean of residuals) at promoters (y-axis), or to methylation of the nearest VMR (x-axis).
-\todo{Yes, I would remove B.}
-		Promoters are defined as intervals \textpm2~kb around the TSS.
-		\textbf{(C)} Mean methylation near the gene Htra1 (ENSMUSG00000006205).
-		Cells are assigned to 4 groups based on Htra1 expression (group 0: cells with no Htra1 expression, groups 1-3: cells that express Htra1, divided into three equally large groups with group 3 having the highest expression).
-		Data from \citet{kremer_scnmt}.}
-	\label{fig:correlation}
-\end{figure*}
-
 Typically, some regions in a chromosome will have very similar methylation status in all cells, while other regions show variability in methylation across cells.
@@
 	\begin{center}
-		\includegraphics[width=0.7\textwidth]{figures/Fig_benchmark.png}
+		\includegraphics[width=\textwidth]{figures/Fig_correlation.png}
+	\end{center}
+	\caption{\small \textbf{Correlation of DNA methylation and gene expression}.\\
+		\textbf{(A)} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red): for most promotors, correlation is weak.
+Promoters are defined as intervals \textpm2~kb around the TSS.
+		\textbf{(B)} Mean methylation near the gene Htra1.
+		Cells are assigned to 4 groups based on Htra1 expression (group 0: cells with no Htra1 expression, groups 1-3: cells that express Htra1, divided into three equally large groups with group 3 having the highest expression).
+		Data from \citet{kremer_scnmt}.}
+	\label{fig:correlation}
+\end{figure*}
+
+\begin{figure*}
+	\begin{center}
+		\includegraphics[width=0.8\textwidth]{figures/Fig_benchmark.png}
 	\end{center}
@@
 		Either 100~kb genomic tiles or VMRs were subjected to PCA and UMAP.
-		Tiles or VMRs were pre-processed as proposed in this work (shrunken mean of residuals subjected to iterative PCA) or using conventional methods (PCA on methylation percentages, lightly imputed, as done in \citet{luo2017single}).
+		Tiles or VMRs were pre-processed as proposed in this work (shrunken mean of residuals subjected to iterative PCA) or using conventional methods (PCA on methylation percentages, lightly imputed as done in \citet{luo2017single}; see Methods for details).
 		The full 1566-cell data set was sub-sampled to simulate smaller data sets (x axis).
@@
 
-\phantomsection
+%\phantomsection
 \subsubsection{Correlating VMR methylation with gene expression}
@@
 Furthermore, even cellular sub-types of cells in the continuous neural stem cell lineage were partially separated.
-
-\begin{figure*}
-	\begin{center}
-		\includegraphics[width=.65\textwidth]{figures/Fig_DMRdetection.png}
-	\end{center}
-	\caption{\small \textbf{Identification of differentially methylated regions.}\\
-		The chromosome is divided into windows overlapping by a small and fixed size.
-		For each window, the cell’s methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
-		Then the t-statistic of these values is obtained as a measure of methylation difference (each point represents one of the windows).
-		Upper and lower thresholds (dashed lines) are chosen such that a defined quantile of windows have a t-statistic above or below the threshold.
-		Windows exceeding either threshold are merged if they overlap, yielding the differentially methylated regions (DMRs).}
-	\label{fig:dmrscan}
-\end{figure*}
 
@@
 	\begin{center}
-		\includegraphics[width=.7\textwidth]{figures/Fig_DMRs.png}
+		\includegraphics[width=.75\textwidth]{figures/Fig_DMRdetection.png}
 	\end{center}
-	\caption{\small \textbf{Detection of differentially methylated regions between oligodendrocytes and neural stem cells.}\\
-		\textbf{(A)} DMRs detected between 58 oligodendrocytes (ODs) and 130 neural stem cells (NSCs) from \citep{kremer_scnmt} (top) and DMRs detected in the same data with randomly permuted cell labels (bottom, used to estimate FDR and determine adjusted p-values).
-		\textbf{(B)} Enrichment of GO terms associated with DMRs lowly methylated in ODs (top) or NSCs (bottom).
-		Depicted are the top 5 GO terms of the "Biological Process" (GO:BP) and "Cellular Component" (GO:CC) GO category, and their binomial test p-value and enrichment, as reported by GREAT \citep{mclean2010great}.
-		\textbf{(C)} Mean methylation of NSCs and ODs at an exemplary DMR.
-		Each point corresponds to a cell, black lines denote the median.
-		\textbf{(D)} Detailed view of the DMR (yellow dashed rectangle) from \textbf{C} in pseudo-bulk samples consisting of NSCs or ODs.
-		Vertical bars represent CpG sites.
-	}
-	\label{fig:dmr}
+	\caption{\small \textbf{Identification of differentially methylated regions.}\\
+		The chromosome is divided into windows overlapping by a small and fixed size.
+		For each window, the cell’s methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
+		Then the t-statistic of these values is obtained as a measure of methylation difference (each point represents one of the windows).
+		Upper and lower thresholds (dashed lines) are chosen such that a defined quantile of windows have a t-statistic above or below the threshold.
+		Windows exceeding either threshold are merged if they overlap, yielding the differentially methylated regions (DMRs).}
+	\label{fig:dmrscan}
 \end{figure*}
@@
 
-\phantomsection
+\begin{figure*}
+	\begin{center}
+		\includegraphics[width=.7\textwidth]{figures/Fig_DMRs.png}
+	\end{center}
+	\caption{\small \textbf{Detection of differentially methylated regions between oligodendrocytes and neural stem cells.}\\
+		\textbf{(A)} DMRs detected between 58 oligodendrocytes (ODs) and 130 neural stem cells (NSCs) from \citep{kremer_scnmt} (top) and DMRs detected in the same data with randomly permuted cell labels (bottom, used to estimate FDR and determine adjusted p-values).
+		\textbf{(B)} Enrichment of GO terms associated with DMRs lowly methylated in ODs (top) or NSCs (bottom).
+		Depicted are the top 5 GO terms of the "Biological Process" (GO:BP) and "Cellular Component" (GO:CC) GO category, and their binomial test p-value and enrichment, as reported by GREAT \citep{mclean2010great}.
+		\textbf{(C)} Mean methylation of NSCs and ODs at an exemplary DMR.
+		Each point corresponds to a cell, black lines denote the median.
+		\textbf{(D)} Detailed view of the DMR (yellow dashed rectangle) from \textbf{C} in pseudo-bulk samples consisting of NSCs or ODs.
+		Vertical bars represent CpG sites.
+	}
+	\label{fig:dmr}
+\end{figure*}
+
+%\phantomsection
 \subsection{Detecting DMRs between oligodendrocytes and neural stem cells}
@@
 
-The methylation matrix produces by the \texttt{matrix} function can then serve as input for standard single-cell analysis tools like Seurat \citep{seurat} or scanpy \citep{scanpy}, in order to perform, e.g., dimensionality reduction and cell clustering.
+The methylation matrix produces by the \texttt{matrix} function can then serve as input for standard single-cell analysis tools like Seurat \citep{seurat} or scanpy \citep{Wolf_2018}, in order to perform, e.g., dimensionality reduction and cell clustering.
 For this, the methylation matrix should in lieu of the log-normalized expression matrix usually expected by these tools.
@@
 
-\phantomsection
-\subsection{Acknowledgments}
-
-SA and LPMK acknowledge funding by the Klaus Tschira Foundation (project 00.022.2019).
-LPMK, LK, SC, and AMV acknowledge funding from the European Commission via ERC grant 771376.
-We thank Alexey Uvarovskii for code refactoring and helpful suggestions on the source code.
-\bigskip
-
-\phantomsection
+%\phantomsection
 \section{Methods}
 
-\phantomsection
 \subsection{Raw data}
 
-We write $x_{ij}$ for the methylation status of CpG $i$ in cell $j$.
+Let us write $x_{ij}$ for the methylation status of CpG $i$ in cell $j$.
 The index $i$ runs over all CpG positions present in the genome, the index $j$ over all cells in the assay.
@@
 Coding NA (the most common value) as 0 leverages SciPy's sparse matrix storage.
-In all the follows here, any mention of $x$ will, however, always mean the encoding as $x_{ij}\in\{0,1,\text{NA}\}$.
+In all that follows here, any mention of $x$ will, however, always mean the encoding as $x_{ij}\in\{0,1,\text{NA}\}$.
 
@@
 
-$$ v_I = \frac{\scriptsize{1}}{\scriptsize{|C_I|}}\left( r_{Ij} - \langle r_{Ij'}\rangle_{j'\in C_I} \right)^2,$$
+\begin{equation} 
+v_I = \frac{\scriptsize{1}}{\scriptsize{|C_I|}}\sum_{j\in C_I}\left( r_{Ij} - \langle r_{Ij'}\rangle_{j'\in C_I} \right)^2, \label{var}
+\end{equation}
 
-where the average runs only over the set $C_I=\bigcup_{i\in I}C_i$ of those cells $j$ which have reads covering interval I.
+where the average runs only over the set $C_I=\bigcup_{i\in I}C_i$ of those cells $j$ which have reads covering interval I. \todo{No Bessel correction?}
 
@@
 
-\todo{...}
+When searching for DMRs, we compare two groups of cells, whose index sets we denote with $C_\text{A}$ and $C_\text{B}$.
+For a given interval $I$, we calculate the mean each of the mean shrunken residuals $r_{ij}$ (see Eq.~(\ref{avgres})) over the cells $j$ in each of the two groups,
+$$ \mu^g_I = \langle r_{Ij} \rangle_{j\in g},\qquad g=\text{A},\text{B} $$
+
+We also calculate a variance as in Eq.(\ref{var}):
+$$ v^g_I = \frac{1}{|C_\text{g}|+1}  \sum_{j\in g} \left( r_{Ij} - \mu_I^g \right)^2, 
+\qquad g=\text{A},\text{B}$$
+\todo{Is it really not pooled? Is there a Bessel correction?}
+
+From this, we calculate Welch's t statistic as usual:
+$$ t_I = \frac{\mu^\text{B}_I - \mu^\text{A}_I}{\sqrt{v^\text{B}_I - v^\text{A}_I}}.$$
+
+In order to find candidate DMRs, we again define overlapping and stepwise shifted intervals $I_1, I_2, \dots$ as for the VMRs and calulate a t statistics $t_1, t_2, \dots$ for these.
+As before, we take the top 2-percentile of these values, fuse intervals that overlap and call the regions thus obtained candidate DMRs.
+We repeat the procedure for the bottom 2-percentile to get the candidate DMRs for the other sign.
+
+Next, we need to check these candidate DMRs for statistical significance.
+We first remind the readers here that, as this is a within-sample analysis, and cells, not samples, are the statistical unit.
+Therefore, a call as significant implies that the same DMR is likely to be called again if we repated the analysis with another set of cells taken from the same biological sample, not that it would generalize to further samples.
+This fact, although often overlooked, is common to all within-sample analyses in the single-cell fields, e.g.
+also to the differential expression tests performed in scRNASeq analyses to find marker genes that differentiate clusters.
+
+It may seem that we could use the standard procedure for the Welch t-test here, i.e.
+use the Welch-Satterthwaite formula to get an approximate degree of freedom and then calculate the tail probability of the corresponding t distribution.
+However, this is unlikely to hold for two reasons: First, the Welch-Sattersthwaite degrees of freedom are only based on the number of cells per group and do not account for the fact that the read coverage might vary from cell to cell.
+Second, the fusing of the DMRs obtained in the scanning step to obtain fused candidate DMRs would invalidate subsequent p-value based adjustment for multiple testing.
+
+Therefore, we have instead implemented a permutation procedure, which works as follows: We randomly shuffle the assignment of the cells in $C_\text{A}\cup C_\text{A}$ to either of the two groups and then rerun the whole procedure, i.e., the scanning step, the DMR fusing and the calculation of t values from the (potentially fused) candidate DMRs.
+This needs to be done for a sufficiently large number of permutation.
+Running through the whole genome for each permutation would be too computationally expensive.
+Instead, we go through the genome only once \todo{correct?}, but reshuffle the group labels every \todo{???} kb.
+
+All the t values obtained from this permutation procedure are taken together to get an empirical null distibution.
+Then, we can use this null distribution to control false discovery rate (FDR) by applying the Benjamini-Hochberg procedure in its p-value-free form: Let us write $T$ for the set of all t values obtained from the ``real'' assignment of cells to group labels and $T_0$ from the set of all t values obtained from the shuffled assignments, i.e., the empirical null distribution.
+To adjust a specific t value $t_i\in T$, we calculate
+$$ p^\text{adj}_i = \frac{ \left|\left\{t_j\in T_0\mid|t_j| > |t_i| \right\}\right| \big/ |T_0|}
+{ \left|\left\{t_j\in T\mid|t_j| > |t_i| \right\}\right| \big/ |T|}.$$
+In words: we calculate which fraction of the null t values is greater than t by absolute value, and which fraction of the real t values is.
+The ratio gives us the false discovery rate we should expect if we used the t value as threshold.
+\todo{Check if this is what we do.}
 
@@
 
-We note that similar approaches have also been used elsewhere, e.g., in the missMDA method of \citet{missMDA}.
+We note that similar approaches have also been used elsewhere \citep{josse2012}.
 
-\subsection{Re-analysis of published scBS datasets}
-To re-analyze scBS data from \citet{kremer_scnmt}, single-cell CpG methylation reports from all conditions were first stored with \texttt{scbs prepare} and then smoothed with \texttt{scbs smooth} using the default bandwidth of 1000~bp.
+\subsection{Analysis of scBS datasets for benchmarks}
+To analyze scBS data from \citet{kremer_scnmt}, single-cell CpG methylation reports from all conditions were first stored with \texttt{scbs prepare} and then smoothed with \texttt{scbs smooth} using the default bandwidth of 1000~bp.
 The mm10 mouse genome was then annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 0.02$, minimum cell requirement $=6$).
@@
 
-{\small \bibliography{scbs}}
+\phantomsection\addcontentsline{toc}{section}{Back Matter}
+\subsection{Acknowledgments}
+
+SA and LPMK acknowledge funding by the Klaus Tschira Foundation (project 00.022.2019).
+LPMK, LK, SC, and AMV acknowledge funding from the European Commission via ERC grant 771376.
+We thank Alexey Uvarovskii for code refactoring and helpful suggestions on the source code.
+
+{\phantomsection\addcontentsline{toc}{subsection}{References}
+\small \balance \bibliography{scbs} \balance}
 

### 2023-04-20 paper.tex
Updates from Overleaf

--- 2023-04-19 paper.tex
@@
 \usepackage{graphicx}
-\usepackage{caption}
 
@@
 \usepackage{natbib}
-\bibliographystyle{simon}
+\bibliographystyle{genome_research}
 \setlength{\bibsep}{0.3ex plus 0.2ex}
@@
 
+\setcounter{secnumdepth}{0}
+
 \twocolumn[{%
@@
 
-April 2023
+January 2023
 \vspace{5.5ex}
@@
 \begin{minipage}{0.7\textwidth}
-\small \setlength{\parindent}{1em}
-\noindent \textit{Abstract} -- Single-cell bisulfite sequencing (scBS) is a technique that enables the assessment of DNA methylation at single-base pair and single-cell resolution.
-The analysis of the large datasets obtained from scBS requires preprocessing to reduce data size, improve signal-to-noise ratio, and provide interpretability.
-Typically, this is achieved by dividing the genome into large tiles and averaging the methylation signals within each tile.
-
-Here, we demonstrate that this coarse-graining approach can lead to signal dilution, and propose improved strategies to identify more informative regions for methylation quantification, and a more accurate quantitation method than simple averaging.
-Our approach enables better discrimination of cell types and reduces the need for large numbers of cells.
-We also present an approach to detect differentially methylated regions (DMRs) between groups of cells, and demonstrate its ability to identify biologically meaningful regions that are associated with genes involved in the core functions of specific cell types.
-
-To facilitate the analysis of scBS data, we have developed a software tool called ``\texttt{scbs}'' that implements these methods and provides additional functionality.
-
-%In summary, our work provides a more robust and accurate approach to analyze scBS data, which has the potential to enhance our understanding of DNA methylation dynamics in single cells and improve the identification of epigenetic changes associated with different cell states or diseases.
+\small
+\textit{Abstract} -- Single-cell bisulfite sequencing (scBS) allows for assessing DNA methylation at single-base pair and single-cell resolution.
+Analyzing the big data sets thus obtained requires preprocessing to reduce data size, improve signal over noise, and provide interpretability.
+Typically, this is simply done by dividing the genome into large tiles and averaging in these.
+We show that this coarse-graining dilutes signal, and suggest improved approaches to determine more informative regions to quantify methylation in, and a better quantitation method than simple averaging.
+We demonstrate how our approach enables better distinguishing of cell types and reduces demands on cell number.
+We also implement an approach to detect differentially methylated regions (DMRs) between groups of cells, and demonstrate its ability to detect biologically meaningful regions near genes involved in core functions of specific cell types.
+Finally, we present a software tool, \texttt{scbs}, that implements these methods and offers further functionality required to work with scBS data.
 \end{minipage}
@@
 }]
-\setcounter{secnumdepth}{0}
 
@@
 
+\blfootnote{\hspace{-.5cm}\raggedright\textsuperscript{*} e-mail:  \texttt{l.kremer@dkfz-heidelberg.de} and\\ \hspace{1cm}\texttt{simon.anders@bioquant.uni-heidelberg.de}}
+
 Here, we discuss bisulfite sequencing, a method to study DNA methylation.
@@
 
-In the present paper, we discuss strategies to analyze single-cell bisulfite-sequencing (scBS) data.
-We suggest improvements to current approaches, and demonstrate their value in benchmarks, using two real-world datasets.
-Furthermore, we discuss how to perform comparative analyses.
-Finally, we present ``scbs'', a comprehensive software toolkit to perform scBS data analysis, using the described methods.
-\blfootnote{\hspace{-.5cm}\raggedright\textsuperscript{*} e-mail:  \texttt{l.kremer@dkfz-heidelberg.de} and\\ \hspace{1cm}\texttt{simon.anders@bioquant.uni-heidelberg.de}}
+In the present paper, we discuss strategies to analyze single-cell bisulfite-sequencing (scBS) data, suggest improvements to current approaches, show their value in a benchmark, and present software to perform the improved analysis methods.
 
@@
     \end{center}
-    \caption{\small \textbf{Improved quantification of DNA methylation in a given genomic interval.}\\
-    \textbf{(A)} Depicted is a genomic interval (between vertical blue lines) along a chromosome, for which DNA methylation is to be quantified.
+    \caption{\small \textbf{Improved quantification of DNA methylation in genomic intervals.}\\
+    \textbf{(A)} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified.
     Two cells cover differing parts of the interval with one read each.
@@
 A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm.
-In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-4}$) to the normalized fractions before taking the logarithm.
+In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-4}$) \todo{is this even correct?} to the normalized fractions before taking the logarithm.
 Now, one could use Euclidean distances of these vectors of logarithmized fractions as dissimilarity score.
@@
 
-This procedure is commonly adapted when working with single-cell DNA methylation data, because once on gets to the PCA step, one can then continue with the established methods just mentioned.
-The task is therefore to construct from the methylation data a matrix suitable as input for PCA.
+This procedure is commonly adapted when working with single-cell DNA methylation data, so that established methods used in scRNA-seq can be used.
+The task is therefore how to construct a suitable matrix as input for PCA.
 A simple and common approach, used for instance by \citet{luo2017single}, is to divide the genome into tiles of e.g.
 100~kb size, and calculate for each cell the average methylation of the DNA within each tile.
-To this end, one identifies in the tile all CpG sites that are covered by at least one read and averages their methylation state, i.e., one denotes the as average DNA methylation of the tile in a given cell the proportion of the observed CpG sites in the tile that were found to be methylated (Figure \ref{fig:smoothres}A).
-This yields a matrix, with one row for each cell and one column for each genomic tile, comprising numbers (``methylation fractions'') between 0 and 1.
+To this end, we identify all CpG sites in the tile for which we have coverage with at least one read and can hence call the CpG to be either unmethylated or methylated in the cell.
+Then we denote as average DNA methylation of this tile in a given cell the proportion of observed CpG sites in the tile that were found to be methylated (Figure \ref{fig:smoothres}A).
+This yields a matrix, with one row for each cell and one column for each genomic tile, comprising numbers (methylation fractions) between 0 and 1.
 This matrix is now subjected to PCA.
@@
 
-While this simple procedure is straight-forward and produces useable results, it is not optimal.
-In this paper, we discuss weaknesses of the simple approach and suggest several refinements to overcome them.
-Using benchmarks and application to real data, we show that our improvements substantially increase the information content of the processed data.
-In the main text, we explain the proposed methods and their motivation in a qualitative manner, while mathematical details are given later in the Methods section.
-Then, we will demonstrate the value of our methods using benchmarks and application to real data.
-Finally, we describe and demonstrate an approach to detect differentially methylated regions (DMRs) in scBS data.
-We also describe our software toolkit, scbs, that facilitates all these analyses.
+While this simple procedure is straight-forward and produces satisfactory results, the coarse resolution of large genomic tiles and the sparse read coverage of single-cell methylomes still pose significant challenges in the analysis of scBS data.
+Here, we propose three improvements that, as we show afterwards, considerably increase the information content of the processed data.
+We will first explain the proposed improvement in a qualitative manner, while mathematical details are given later in the Methods section.
+Then, we will demonstrate the improvement using benchmarks.
+Finally, we describe an approach to detect differentially methylated regions (DMRs) in scBS data, and showcase it on real data.
 
@@
 We first discuss the task of quantifying the level of methylation in a given, fixed, genomic interval.
-Typically, read coverage per cell is sparse in scBS data.
-In the example of Figure \ref{fig:smoothres}A, teh depicted interval is covered by a single read for two of the three cells shown and no read in the third.
-The read from cell 2 shows much more methylation than the read from cell 1, and the standard analysis would therefore consider cell 1 to have higher methylation in the interval than cell 2.
+In Figure \ref{fig:smoothres}A, such an interval is covered by a single read for two of the three cells shown.
+The read from cell 2 shows much more methylation than the read from cell 1, and the standard analysis would therefore consider that cell as having higher methylation in the interval.
 However, given that the two reads agree wherever they overlap, a more parsimonious interpretation would be that the cells do not show difference in methylation within the interval.
@@
 
-Therefore, we propose to first obtain, for each CpG position, an average of the methylation across all cells, and then quantify each cell's deviation from this average.
+Therefore, we propose to first obtain, for each CpG position, an average of the methylation across all cells, and then quantify for each individual cell its deviation from this average.
 In Figure \ref{fig:smoothres}B, the curved line depicts such an average over all cells, and the red vertical lines show an individual cell's deviation from the ensemble average.
-We take the lengths of the red lines as signed values (``residuals''), positive for lines extending upwards from the curve (methylated CpG) and negative for lines extending downwards (unmethylated CpG).
-For each cell, we then take the average over the residuals for all the CpGs in the interval that are covered by reads from this cell.
+We take the lengths of the red lines as signed values (``residual''), positive for lines extending upwards from the curve (methylated CpG) and negative for lines extending downwards (unmethylated CpG), and take the average over the residual for all the CpGs in the interval that are covered by reads from this cell.
 In this average, we perform shrinkage towards zero via a pseudocount (to trade bias for variance, see Methods for detail) in order to dampen the signal in cells with low coverage of the interval.
 
-The average thus obtained, the shrunken mean of the residuals, is what we use to quantify the cell's (relative) methylation in the interval.
+The average thus obtained, the shrunken mean of the residuals, is what we use to quantify this cell's (relative) methylation in the interval.
 For a genome tiled into such intervals, we thus obtain a matrix, one row per cell, one column per interval, that can be used for downstream analysis, e.g., as input for PCA.
@@
 
-A minor remaining issue is how to deal with the case that a cell has no reads at all within a given interval.
-Here, it is justified to simply put zero into the matrix element, because a shrunken residual average of zero indicates that there is no evidence of the cell deviating from the mean.
-We slightly refine this by using an iterative imputation within the PCA (see Methods for details).
-
-Taken together, this shrunken mean of residuals quantitation reduces variance in comparison to simple averaging of raw methylation calls. We will show further below how this improves results.
+% remove or shorten this section cause it's not really novel, methods like this are already implemented in e.g. the "missMDA" R package
+%\subsection{PCA with imputation}
+%The standard PCA algorithm cannot deal with missing data.
+%Therefore, one would need to make intervals wide enough to ensure that every interval is covered by at least one read for most tiles and discard any interval that is not covered by any single cell.
+%Hence, a naive analysis has to resort to a coarse tiling of the genome.
+%To counter this, we propose a simple and straightforward way to deal with missing data in the input matrix to a PCA: In a first iteration, we replace each missing value in the centered input matrix with zero, then run the PCA.
+%Then, only these zeroes are replaced by the value predicted by the PCA and the PCA is rerun.
+%Convergence is typically achieved after two to five iterations.
+%For details, see Methods.
+%The possibility to perform PCA with missing data allows us to use shorter, and thus more intervals and hence provide the PCA with richer input data.
 
@@
 Typically, some regions in a chromosome will have very similar methylation status in all cells, while other regions show variability in methylation across cells.
-For instance, it is known  since long that CpG-rich promoters of housekeeping genes are unmethylated, and that a large proportion of the remaining genome is highly methylated regardless of cell type \citep{bird1986cpg}.
-In contrast, DNA methylation at certain genomic features such as enhancers is more dynamic (e.g., \citet{argelaguet2019gastru}), and thus more variable across cells.
+For instance, it is long known that CpG-rich promoters of housekeeping genes are unmethylated, and that a large proportion of the remaining genome is highly methylated regardless of cell type \citep{bird1986cpg}.
+In contrast, DNA methylation at certain genomic features such as enhancers is more dynamic \citep{argelaguet2019gastru}, and thus more variable across cells.
 Only the latter regions are of value for our goal of quantitating dissimilarity between cells.
@@
 In the standard approach, one divides up (tiles) each chromosome into non-overlapping, equal-sized intervals, and quantitates the methylation of each tile.
-Such rigid placing of interval boundaries is unlikely to be optimal: for example, a VMR might be much smaller than a tile and the signal from its CpG sites will hence be drowned out by the larger number of uninformative CpG sites that are equal in all cells, when averaging over all the CpG sites in the tile.
+Such rigid placing of interval boundaries is unlikely to be optimal: for example, a VMR might be much smaller than a tile and its signal will hence be drowned out by the larger part of uninformative CpG sites that are equal in all cells, when averaging over all the CpG sites in the tile.
 
-Therefore, we propose the following approach (Fig.~\ref{fig:vmr}): Divide up the chromosome into many \emph{overlapping} windows that start at regular multiples of a fixed, small, step size.
+Therefore, we propose the following approach (Fig.~\ref{fig:vmr}): Divide up the chromosome into many \emph{overlapping} windows, that start at regular multiples of a fixed, small, step size.
 Quantify the methylation of each cell in each window by averaging the cell's methylation residuals over all CpGs in the window, as described above and depicted in Fig.~\ref{fig:smoothres}B.
@@
     \label{fig:dmrscan}
-\end{figure*}
-
-\begin{figure*}
-	\begin{center}
-		\includegraphics[width=.65\textwidth]{figures/Fig_correlation.png}
-	\end{center}
-	\caption{\small \textbf{Correlation of DNA methylation and gene expression}.\\
-		\textbf{(A)} \todo{maybe remove panel A or B, and put the remaining panel left of C?} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red).
-		\textbf{(B)} Correlation of each gene's RNA expression values to methylation (shrunken mean of residuals) at promoters (y-axis), or to methylation of the nearest VMR (x-axis).
-		Promoters are defined as intervals \textpm2~kb around the TSS.
-		\textbf{(C)} Mean methylation near the gene Htra1 (ENSMUSG00000006205).
-		Cells are assigned to 4 groups based on Htra1 expression (group 0: cells with no Htra1 expression, groups 1-3: cells that express Htra1, divided into three equally large groups with group 3 having the highest expression).
-		Data from \citet{kremer_scnmt}.}
-	\label{fig:correlation}
 \end{figure*}
@@
 
+\begin{figure*}
+    \begin{center}
+    \includegraphics[width=.65\textwidth]{figures/Fig_correlation.png}
+    \end{center}
+    \caption{\small \textbf{Correlation of DNA methylation and gene expression}.\\
+    \textbf{(A)} \todo{maybe remove panel A or B, and put the remaining panel left of C?} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red).
+    \textbf{(B)} Correlation of each gene's RNA expression values to methylation (shrunken mean of residuals) at promoters (y-axis), or to methylation of the nearest VMR (x-axis).
+    Promoters are defined as intervals \textpm2~kb around the TSS.
+    \textbf{(C)} Mean methylation near the gene Htra1 (ENSMUSG00000006205).
+    Cells are assigned to 4 groups based on Htra1 expression (group 0: cells with no Htra1 expression, groups 1-3: cells that express Htra1, divided into three equally large groups with group 3 having the highest expression).
+    Data from \citet{kremer_scnmt}.}
+    \label{fig:correlation}
+\end{figure*}
+
+\subsection{Our proposed methods improve identification of cell types}
+
 \begin{figure*}[t]
@@
 \end{figure*}
-
-\subsection{Our proposed methods improve identification of cell types}
 
@@
 
-SA and LPMK acknowledge funding by the Klaus Tschira Foundation (project 00.022.2019).
-LPMK, LK, SC, and AMV acknowledge funding from the European Commission via ERC grant 771376.
+S.\ A.\ and L.\ P.\ M.\ K.\ acknowledge funding by the Klaus Tschira Foundation.
+L.\ P.\ M.\ K., L.\ K., S.\ C.\ and A.\ M.-V.\ acknowledge funding from the European Commission via ERC grant 771376.
 We thank Alexey Uvarovskii for code refactoring and helpful suggestions on the source code.
-\bigskip
 
@@
 
-We note that similar approaches have also been used elsewhere, e.g., in the missMDA method of \citet{missMDA}.
-
 \subsection{Re-analysis of published scBS datasets}
@@
 
-\onecolumn 
-\newpage
+\begin{samepage}
 \section{Supplementary Figures}
-\vspace{6ex}
 
@@
 
-\noindent%
-\begin{minipage}{\linewidth}
+\begin{figure*}
     \begin{center}
-	    \includegraphics[width=0.4\textwidth]{figures/SFig_benchmark.png}
+        \includegraphics[width=0.8\columnwidth]{figures/SFig_benchmark.png}
     \end{center}
-	\captionof{figure}{\small \textbf{Benchmark of our methods on single-cell methylomes of neural subtypes of the murine frontal cortex}.\\
+    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of neural subtypes of the murine frontal cortex}.\\
     Our methods were benchmarked on single-cell methylomes from \citet{luo2017single} as described in Fig.~\ref{fig:score}.
@@
     \label{fig:score_luo}
-\end{minipage}
-\vspace{6ex}
+\end{figure*}
 
-\noindent%
-\begin{minipage}{\linewidth}
+\begin{figure*}
     \begin{center}
@@
     \end{center}
-    \captionof{figure}{\small \textbf{Effect of VMR detection parameters on separation of cell types}.\\
+    \caption{\small \textbf{Effect of VMR detection parameters on separation of cell types}.\\
     \textbf{(A-B)} Mean neighbor scores obtained after analyzing single-cell methylomes from \citet{kremer_scnmt} (\textbf{A}) or \citet{luo2017single} (\textbf{B}) with our proposed methods.
@@
     \label{fig:sweep}
-\end{minipage}
+\end{figure*}
+
+\end{samepage}
 

### 2023-04-20 paper.tex
typos etc

--- 2023-04-19 paper.tex
@@
 
-The starting point in most scRNA-seq analyses is a matrix of UMI counts (i.e.
-of counts of distinct RNA molecules), one row for each cell and one column for each gene.
+The starting point in most scRNA-seq analyses is a matrix of UMI counts (i.e. of counts of distinct RNA molecules), one row for each cell and one column for each gene.
 A first goal is usually to assign cell types or states to cells.
@@
 100~kb size, and calculate for each cell the average methylation of the DNA within each tile.
-To this end, one identifies in the tile all CpG sites that are covered by at least one read and averages their methylation state, i.e., one denotes the as average DNA methylation of the tile in a given cell the proportion of the observed CpG sites in the tile that were found to be methylated (Figure \ref{fig:smoothres}A).
+To this end, one identifies in the tile all CpG sites that are covered by at least one read and averages their methylation state, i.e., one denotes as average DNA methylation of the tile in a given cell the proportion of the observed CpG sites in the tile that were found to be methylated (Figure \ref{fig:smoothres}A).
 This yields a matrix, with one row for each cell and one column for each genomic tile, comprising numbers (``methylation fractions'') between 0 and 1.
@@
 
-%\phantomsection
 \subsection{Read-position aware quantitation} \label{residuals}
@@
 Typically, read coverage per cell is sparse in scBS data.
-In the example of Figure \ref{fig:smoothres}A, teh depicted interval is covered by a single read for two of the three cells shown and no read in the third.
+In the example of Figure \ref{fig:smoothres}A, the depicted interval is covered by a single read for two of the three cells shown and no read in the third.
 The read from cell 2 shows much more methylation than the read from cell 1, and the standard analysis would therefore consider cell 1 to have higher methylation in the interval than cell 2.
@@
 
+\subsection{Application and benchmarks}
+
+To demonstrate the value of our proposed improvements, we employ our own data \citep{kremer_scnmt} as well as a previously published data set, namely the one by  \citet{luo2017single} already mentioned in the introduction.
+
+%\phantomsection
+\subsubsection{Correlating VMR methylation with gene expression}
+
+Our data set \citep{kremer_scnmt} comprises the methylomes of 1566 cells isolated from mouse forebrains, as well as matched single-cell transcriptomes of the same cells.
+Among these cells are distinct cell types such as oligodendrocytes, oligodendrocyte precursor cells (OPCs) and endothelial cells, as well as cellular sub-states that are part of the continuous neural stem cell differentiation trajectory.
+To assess whether our VMR detection method captures genomic intervals that are biologically meaningful, we probed whether their methylation level correlates with the expression of nearby genes.
+
+We first note that gene expression is more strongly correlated with the methylation of nearby VMRs than with the methylation of their promotors (Fig.~\ref{fig:correlation}A), indicating that VMR methylation is often a better predictor of gene expression than promoter methylation.
+Indeed, a gene-wise comparison revealed many genes whose expression is correlated with the nearest VMR, but not with promoter methylation.
+One such example gene, \textit{Htra1}, is depicted in Fig.~\ref{fig:correlation}C.
+While the promoter of this gene is lowly methylated regardless of gene expression, a VMR located downstream of the promoter is lowly methylated in cells with high \textit{Htra1} expression.
+
+\subsubsection{Improved identification of cell types}
+
+We next tested whether our methods improve the ability to distinguish cell types and cell states (Fig.~\ref{fig:score}).
+To this end, we obtained cell type/state labels based on the single-cell transcriptomes from \citet{kremer_scnmt} (Fig.~\ref{fig:score}A).
+We consider these cell labels as ground truth and tested whether we are able to distinguish the same groups of cells based on their methylomes.
+To do this, we subjected the methylomes to four variants of the workflow described above: we defined genomic intervals in two ways and quantified methylation in these intervals in two ways, then always subjected the resulting matrix to PCA.
+To define the intervals, we either simply tiled the genome into 100~kb bins as is commonly done, or we used our VMR detection approach.
+To quantify methylation in these intervals, we either averaged in these intervals, obtaining methylation percentages (``classic'' pre-processing), or we calculated the shrunken mean of residuals.
+
+Visual inspection of the resulting UMAPs revealed that our proposed combination of methods results in more clearly separated cell types, compared to a UMAP obtained with default analysis methods (Fig.~\ref{fig:score}B,C).
+While all cells form a continuous point cloud when using default methods, our improvements led to a clear separation of oligodendrocytes, OPCs and endothelial cells.
+Furthermore, even cellular sub-states of cells in the continuous neural stem cell lineage were partially separated.
+
+To quantify this performance gain in a more rigorous manner, we used a score that quantifies whether cells were placed, in 15-dimensional PC space, in a neighborhood comprising cells of the same cell type ("neighbor score", see Figure \ref{fig:score}D, and Methods for details).
+We then ask how this score depends on the number of available cells.
+This is relevant as scBS protocols are costly and labor-intensive, and only few laboratories are currently able to obtain thousands of single cell methylomes.
+To simulate smaller data sets, we sub-sampled the 1566-cell data set into smaller data sets, analyzed them again with the four workflow variants, and calculated the mean neighbor score for each (Figure \ref{fig:score}E).
+The results demonstrate that the forebrain cell clusters are more cleanly separated when using our proposed methods.
+This performance gain was observed also in smaller data sets, although cell cluster separation generally becomes more difficult in those.
+Both the use of VMRs instead of genomic tiles, as well as the use of the shrunken mean of the residuals over methylation averages improved cell cluster separation.
+
+We repeated this benchmark on the data set published by \citet{luo2017single}, which consists of 3377 single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal subtypes.
+The authors annotated cell types based on CH-methylation instead of CpG-methylation.
+Using these labels as ground truth we re-analyzed the full and sub-sampled CpG methylation data with the four methods combinations and found that our suggested methods consistently outperformed default methods (Fig.~\ref{fig:score_luo}).
+
+\subsubsection{Robustness to parameter changes}
+
+Lastly, we assessed whether our proposed workflow requires fine-tuning of parameters (Fig.~\ref{fig:sweep}).
+To this end, we re-analyzed both data sets, as well as sub-samples of the data with different VMR detection parameters, namely the width of the sliding window in bp, (set with option \texttt{--bandwidth} in our ``scbs'' software, default 2000) and the variance threshold above which windows are merged to VMRs (\texttt{--var-threshold}, default 0.02).
+This parameter sweep showed that our workflow gives good results over a wide range of parameter values.
+For the data of \citet{luo2017single}, results are nearly independent of the parameters (Fig.~\ref{fig:sweep}B).
+In the more challenging data set of \citet{kremer_scnmt} (Fig.~\ref{fig:sweep}A), cell types were less cleanly separated when very large bandwidths, or very strict variance thresholds were selected.
+However, very small bandwidths or very lenient thresholds resulted in a much higher number of VMRs and thus long computing times.
+Overall, however, our default parameter combination provided good results and fast compute times in both data sets.
+
 \begin{figure*}
@@
 Promoters are defined as intervals \textpm2~kb around the TSS.
-		\textbf{(B)} Mean methylation near the gene Htra1.
-		Cells are assigned to 4 groups based on Htra1 expression (group 0: cells with no Htra1 expression, groups 1-3: cells that express Htra1, divided into three equally large groups with group 3 having the highest expression).
+		\textbf{(B)} Mean methylation near the gene \textit{Htra1}.
+		Cells are assigned to 4 groups based on \textit{Htra1} expression (group 0: cells with no \textit{Htra1} expression, groups 1-3: cells that express \textit{Htra1}, divided into three equally large groups with group 3 having the highest expression).
 		Data from \citet{kremer_scnmt}.}
@@
 
-\subsection{Application and benchmarks}
+\section{Finding differentially methylated regions}
 
-To demonstrate the value of our proposed improvements, we employ our own data \citep{kremer_scnmt} as well as a previously published data set, namely the one by  \citet{luo2017single} already mentioned in the introduction.
-
-%\phantomsection
-\subsubsection{Correlating VMR methylation with gene expression}
-
-Our data set \citep{kremer_scnmt} comprises the methylomes of 1566 cells isolated from mouse forebrains, as well as matched single-cell transcriptomes of the same cells.
-Among these cells are distinct cell types such as oligodendrocytes, oligodendrocyte precursor cells (OPCs) and endothelial cells, as well as cellular sub-states that are part of the continuous neural stem cell differentiation trajectory.
-To assess whether our VMR detection method captures genomic intervals that are biologically meaningful, we probed whether their methylation level correlates with the expression of nearby genes.
-
-We first note that gene expression is more strongly correlated with the methylation of nearby VMRs than with the methylation of their promotors (Fig.~\ref{fig:correlation}A), indicating that VMR methylation is often a better predictor of gene expression than promoter methylation.
-Indeed, a gene-wise comparison revealed many genes whose expression is correlated with the nearest VMR, but not with promoter methylation (Fig.~\ref{fig:correlation}B).
-One such example gene, Htra1, is depicted in Fig.~\ref{fig:correlation}C.
-While the promoter of this gene is lowly methylated regardless of gene expression, a VMR located downstream of the promoter is lowly methylated in cells with high Htra1 expression.
-
-\subsubsection{Improved identification of cell types}
-
-We next tested whether our methods improve the ability to distinguish cell types and cell states (Fig.~\ref{fig:score}).
-To this end, we obtained cell type/state labels based on the single-cell transcriptomes from \citet{kremer_scnmt} (Fig.~\ref{fig:score}A).
-We consider these cell labels as ground truth and tested whether we are able to distinguish the same groups of cells based on their methylomes.
-To do this, we subjected the methylomes to four variants of the workflow described above: we defining genomic intervals in two ways and quantified methylation in these intervals in two ways, then always subjected the resulting matrix to PCA.
-To define the intervals, we either simply tiled the genome into 100~kb bins as is commonly done, or we used our VMR detection approach.
-To quantify methylation in these intervals, we either averaged in these intervals, obtaining methylation percentages (``classic'' pre-processing), or we calculated the shrunken mean of residuals.
-
-Visual inspection of the resulting UMAPs revealed that our proposed combination of methods results in more clearly separated cell types, compared to a UMAP obtained with default analysis methods (Fig.~\ref{fig:score}B,C).
-While all cells form a continuous point cloud when using default methods, our improvements led to a clear separation of oligodendrocytes, OPCs and endothelial cells.
-Furthermore, even cellular sub-types of cells in the continuous neural stem cell lineage were partially separated.
-
-To quantify this performance gain in a more rigorous manner, we used a score that quantifies whether cells were placed, in 15-dimensional PC space, in a neighborhood comprising cells of the same cell type ("neighbor score", see Figure \ref{fig:score}D, and Methods for details).
-We then ask how this score depends on the number of available cells.
-This is relevant as scBS protocols are costly and labor-intensive, and only few laboratories are currently able to obtain thousands of single cell methylomes.
-To simulate smaller data sets, we sub-sampled the 1566-cell data set into smaller data sets, analyzed them again with the four workflow variants, and calculated the mean neighbor score for each (Figure \ref{fig:score}E).
-The results demonstrate that the forebrain cell clusters are more cleanly separated when using our proposed methods.
-This performance gain was observed also in smaller data sets, although cell cluster separation generally becomes more difficult in those.
-Both the use of VMRs instead of genomic tiles, as well as the use of the shrunken mean of the residuals over methylation averages improved cell cluster separation.
-
-We repeated this benchmark on the data set published by \citet{luo2017single}, which consists of 3377 single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal subtypes.
-The authors annotated cell types based on CH-methylation instead of CpG-methylation.
-Using these labels as ground truth we re-analyzed the full and sub-sampled CpG methylation data with the four methods combinations and found that our suggested methods consistently outperformed default methods (Fig.~\ref{fig:score_luo}).
-
-\subsubsection{Robustness to parameter changes}
-
-Lastly, we assessed whether our proposed workflow requires fine-tuning of parameters (Fig.~\ref{fig:sweep}).
-To this end, we re-analyzed both data sets, as well as sub-samples of the data with different VMR detection parameters, namely the width of the sliding window in bp, (set with option \texttt{--bandwidth} in our ``scbs'' software, default 2000 )and the variance threshold above which windows are merged to VMRs (\texttt{--var-threshold}, default 0.02).
-This parameter sweep showed that our workflow gives good results over a wide range of parameter values.
-For the data of \citet{luo2017single}, results are nearly independent of the parameters (Fig.~\ref{fig:sweep}B).
-In the more challenging data set of \citet{kremer_scnmt} (Fig.~\ref{fig:sweep}A), cell types were less cleanly separated when very large bandwidths, or very strict variance thresholds were selected.
-However, very small bandwidths or very lenient thresholds resulted in a much higher number of VMRs and thus long computing times.
-Overall, however, our default parameter combination provided good results and fast compute times in both data sets.
+A common task in the analysis of bulk bisulfite-sequencing data is the detection of differentially methylated regions (DMRs) between conditions, tissues, or cell types \citep{Hebestreit2013, dmrseq}.
+As DNA methylation affects gene expression, DMRs can provide insights into the unique epigenetic and gene regulatory characteristics of cell types.
+However, to date no approach to detect DMRs in scBS data has been reported.
+To enable DMR detection in scBS data, we thus developed an approach that detects DMRs of variable size between two groups of cells, and controls the false discovery rate (Fig.~\ref{fig:dmrscan}):
 
@@
 		For each window, the cell’s methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
-		Then the t-statistic of these values is obtained as a measure of methylation difference (each point represents one of the windows).
-		Upper and lower thresholds (dashed lines) are chosen such that a defined quantile of windows have a t-statistic above or below the threshold.
+		Then the t statistic of these values is obtained as a measure of methylation difference (each point represents one of the windows).
+		Upper and lower thresholds (dashed lines) are chosen such that a defined quantile of windows have a t statistic above or below the threshold.
 		Windows exceeding either threshold are merged if they overlap, yielding the differentially methylated regions (DMRs).}
@@
 
-\section{Finding differentially methylated regions}
-
-A common task in the analysis of bulk bisulfite-sequencing data is the detection of differentially methylated regions (DMRs) between conditions, tissues, or cell types \citep{Hebestreit2013, dmrseq}.
-As DNA methylation affects gene expression, DMRs can provide insights into the unique epigenetic and gene regulatory characteristics of cell types.
-However, to date no approach to detect DMRs in scBS data has been reported.
-To enable DMR detection in scBS data, we thus developed an approach that detects DMRs of variable size between two groups of cells, and controls the false discovery rate (Fig.~\ref{fig:dmrscan}):
-
 Similar to the previously described approach for VMR identification (Fig.~\ref{fig:vmr}), we divide each chromosome into overlapping windows shifted by a small and fixed size (step size) and quantify the methylation of each cell in each window.
-Next, instead of the variance, we obtain the t-statistic % Welch t-test
+Next, instead of the variance, we obtain the t statistic % Welch t-test
 as a measure of differential methylation between the two cell groups.
-We identify the windows with the most extreme t-statistics, e.g.
+We identify the windows with the most extreme t statistics, e.g.
 windows in the 2\% upper and lower tails.
-We the merge any overlapping windows in the upper tail into larger DMRs, and do the same for windows in the lower tail, and then we re-calculate the t-statistic for each larger DMR.
+We then merge any overlapping windows in the upper tail into larger DMRs, and do the same for windows in the lower tail, and then we re-calculate the t statistic for each larger DMR.
 
@@
 the same data set with randomly shuffled cell labels.
-The DMR t-statistics obtained from permuted data are then used to estimate the false-discovery rate (FDR), yielding an adjusted p-value for each DMR.
+The DMR t statistics obtained from permuted data are then used to estimate the false-discovery rate (FDR), yielding an adjusted p-value for each DMR.
 
@@
 
-Fig.~\ref{fig:dmr}C-D depicts an exemplary DMR, located at the gene body of myelin-basic protein (Mbp), the major component of myelin that is essential for myelination of neuronal axons \citep{mbp}.
-Our results suggest that oligodendrocyte-specific gene expression of Mbp is supported by low methylation at the detected DMR.
+Fig.~\ref{fig:dmr}C-D depicts an exemplary DMR, located at the gene body of myelin-basic protein (\textit{Mbp}), the major component of myelin that is essential for myelination of neuronal axons \citep{mbp}.
+Our results suggest that oligodendrocyte-specific gene expression of \textit{Mbp} is supported by low methylation at the detected DMR.
 
@@
 The methylation matrix produces by the \texttt{matrix} function can then serve as input for standard single-cell analysis tools like Seurat \citep{seurat} or scanpy \citep{Wolf_2018}, in order to perform, e.g., dimensionality reduction and cell clustering.
-For this, the methylation matrix should in lieu of the log-normalized expression matrix usually expected by these tools.
+For this, the methylation matrix should be provided in lieu of the log-normalized expression matrix usually expected by these tools.
 
@@
 
-where the average runs only over the set $C_I=\bigcup_{i\in I}C_i$ of those cells $j$ which have reads covering interval I. \todo{No Bessel correction?}
+where the average runs only over the set $C_I=\bigcup_{i\in I}C_i$ of those cells $j$ which have reads covering interval I.
 
@@
 We also calculate a variance as in Eq.(\ref{var}):
-$$ v^g_I = \frac{1}{|C_\text{g}|+1}  \sum_{j\in g} \left( r_{Ij} - \mu_I^g \right)^2, 
+$$ v^g_I = \frac{1}{|C_\text{g}|-1}  \sum_{j\in g} \left( r_{Ij} - \mu_I^g \right)^2, 
 \qquad g=\text{A},\text{B}$$
-\todo{Is it really not pooled? Is there a Bessel correction?}
 
 From this, we calculate Welch's t statistic as usual:
-$$ t_I = \frac{\mu^\text{B}_I - \mu^\text{A}_I}{\sqrt{v^\text{B}_I - v^\text{A}_I}}.$$
+$$ t_I = \frac{\mu^\text{A}_I - \mu^\text{B}_I}{\sqrt{v^\text{A}_I/|C_A| + v^\text{B}_I/|C_B|}}.$$
 
-In order to find candidate DMRs, we again define overlapping and stepwise shifted intervals $I_1, I_2, \dots$ as for the VMRs and calulate a t statistics $t_1, t_2, \dots$ for these.
+In order to find candidate DMRs, we again define overlapping and stepwise shifted intervals $I_1, I_2, \dots$ as for the VMRs and calulate t statistics $t_1, t_2, \dots$ for these.
 As before, we take the top 2-percentile of these values, fuse intervals that overlap and call the regions thus obtained candidate DMRs.
@@
 We first remind the readers here that, as this is a within-sample analysis, and cells, not samples, are the statistical unit.
-Therefore, a call as significant implies that the same DMR is likely to be called again if we repated the analysis with another set of cells taken from the same biological sample, not that it would generalize to further samples.
+Therefore, a call as significant implies that the same DMR is likely to be called again if we repeated the analysis with another set of cells taken from the same biological sample, not that it would generalize to further samples.
 This fact, although often overlooked, is common to all within-sample analyses in the single-cell fields, e.g.
-also to the differential expression tests performed in scRNASeq analyses to find marker genes that differentiate clusters.
+also to the differential expression tests performed in scRNA-seq analyses to find marker genes that differentiate clusters.
 
@@
 use the Welch-Satterthwaite formula to get an approximate degree of freedom and then calculate the tail probability of the corresponding t distribution.
-However, this is unlikely to hold for two reasons: First, the Welch-Sattersthwaite degrees of freedom are only based on the number of cells per group and do not account for the fact that the read coverage might vary from cell to cell.
+However, this is unlikely to hold for two reasons: First, the Welch-Satterthwaite degrees of freedom are only based on the number of cells per group and do not account for the fact that the read coverage might vary from cell to cell.
 Second, the fusing of the DMRs obtained in the scanning step to obtain fused candidate DMRs would invalidate subsequent p-value based adjustment for multiple testing.
@@
 Therefore, we have instead implemented a permutation procedure, which works as follows: We randomly shuffle the assignment of the cells in $C_\text{A}\cup C_\text{A}$ to either of the two groups and then rerun the whole procedure, i.e., the scanning step, the DMR fusing and the calculation of t values from the (potentially fused) candidate DMRs.
-This needs to be done for a sufficiently large number of permutation.
+This needs to be done for a sufficiently large number of permutations.
 Running through the whole genome for each permutation would be too computationally expensive.
-Instead, we go through the genome only once \todo{correct?}, but reshuffle the group labels every \todo{???} kb.
+Instead, we go through the genome only once, but reshuffle the group labels every 2 megabasepairs.
+% didn't abbreviate cause one of our example genes is "Mbp"...
 
@@
 { \left|\left\{t_j\in T\mid|t_j| > |t_i| \right\}\right| \big/ |T|}.$$
+
+\todo{In the code we do, for a given t-stat threshold: adjPval = fdisc / (fdisc + tdisc). This seems slightly different than the formula here.}
+
 In words: we calculate which fraction of the null t values is greater than t by absolute value, and which fraction of the real t values is.
 The ratio gives us the false discovery rate we should expect if we used the t value as threshold.
-\todo{Check if this is what we do.}
 
@@
 \setcounter{figure}{0}
-%\renewcommand{\figurename}{Supplementary Figure}
 \renewcommand{\thefigure}{S\arabic{figure}}

### 2023-04-20 paper.tex
minor changes

--- 2023-04-19 paper.tex
@@
 	\caption{\small \textbf{Detection of differentially methylated regions between oligodendrocytes and neural stem cells.}\\
-		\textbf{(A)} DMRs detected between 58 oligodendrocytes (ODs) and 130 neural stem cells (NSCs) from \citep{kremer_scnmt} (top) and DMRs detected in the same data with randomly permuted cell labels (bottom, used to estimate FDR and determine adjusted p-values).
+		\textbf{(A)} DMRs detected between 58 oligodendrocytes (ODs) and 130 neural stem cells (NSCs) from \citet{kremer_scnmt} (top) and DMRs detected in the same data with randomly permuted cell labels (bottom, used to estimate FDR and determine adjusted p-values).
 		\textbf{(B)} Enrichment of GO terms associated with DMRs lowly methylated in ODs (top) or NSCs (bottom).
@@
 		Each point corresponds to a cell, black lines denote the median.
-		\textbf{(D)} Detailed view of the DMR (yellow dashed rectangle) from \textbf{C} in pseudo-bulk samples consisting of NSCs or ODs.
+		\textbf{(D)} Detailed view of the DMR (yellow dashed rectangle) from C in pseudo-bulk samples consisting of NSCs or ODs.
 		Vertical bars represent CpG sites.

### 2023-04-21 paper.tex
two minor changes

@@
 April 2023
-\vspace{5.5ex}
+\vspace{4.5ex}
 
@@
 From this, we calculate Welch's t statistic as usual:
-$$ t_I = \frac{\mu^\text{A}_I - \mu^\text{B}_I}{\sqrt{v^\text{A}_I/|C_A| + v^\text{B}_I/|C_B|}}.$$
+$$ t_I = \frac{\mu^\text{A}_I - \mu^\text{B}_I}{\sqrt{\frac{v^\text{A}_I}{|C_A|} + \frac{v^\text{B}_I}{|C_B|}}}.$$
 

### 2023-04-24 paper.tex
minor
add GREAT options to methods

--- 2023-04-20 paper.tex
@@
 	\caption{\small \textbf{Detection of differentially methylated regions between oligodendrocytes and neural stem cells.}\\
-		\textbf{(A)} DMRs detected between 58 oligodendrocytes (ODs) and 130 neural stem cells (NSCs) from \citet{kremer_scnmt} (top) and DMRs detected in the same data with randomly permuted cell labels (bottom, used to estimate FDR and determine adjusted p-values).
+		\textbf{(A)} DMRs detected between 58 oligodendrocytes (ODs) and 130 neural stem cells (NSCs) from \citep{kremer_scnmt} (top) and DMRs detected in the same data with randomly permuted cell labels (bottom, used to estimate FDR and determine adjusted p-values).
 		\textbf{(B)} Enrichment of GO terms associated with DMRs lowly methylated in ODs (top) or NSCs (bottom).
@@
 		Each point corresponds to a cell, black lines denote the median.
-		\textbf{(D)} Detailed view of the DMR (yellow dashed rectangle) from C in pseudo-bulk samples consisting of NSCs or ODs.
+		\textbf{(D)} Detailed view of the DMR (yellow dashed rectangle) from \textbf{C} in pseudo-bulk samples consisting of NSCs or ODs.
 		Vertical bars represent CpG sites.
@@
 
-\todo{In the code we do, for a given t-stat threshold: adjPval = fdisc / (fdisc + tdisc). This seems slightly different than the formula here.}
-
 In words: we calculate which fraction of the null t values is greater than t by absolute value, and which fraction of the real t values is.
@@
 To flexibly adapt to data sets of different sizes, we set n\_neighbors  $=\frac{\sqrt{n}}{1.5}$ (rounded to the nearest integer), where $n$ is the total cell number.
+
+To test our DMR detection approach, we selected oligodendrocytes and NSCs from healthy wild type mice of \citet{kremer_scnmt} and ran \texttt{scbs diff} with default parameters.
+For GO enrichment analysis, DMRs with an adjusted p-value $<0.01$ were uploaded to GREAT \citep{mclean2010great} with association rules basal plus extension, 0~bp upstream, 20~kb downstream, 1~Mbp max extension, curated regulatory domains included.
 

### 2023-04-24 paper.tex

--- 2023-04-21 paper.tex
@@
 \twocolumn[{%
-\centering
+\centering\phantomsection\addcontentsline{toc}{section}{Title}
 \textbf{\Large Analyzing single-cell bisulfite sequencing data with \texttt{scbs}}\\[1.5ex]
@@
 \small \setlength{\parindent}{1em}
+\phantomsection\addcontentsline{toc}{subsection}{Abstract}
 \noindent \textit{Abstract} -- Single-cell bisulfite sequencing (scBS) is a technique that enables the assessment of DNA methylation at single-base pair and single-cell resolution.

### 2023-04-24 paper.tex
minor

@@
 To test our DMR detection approach, we selected oligodendrocytes and NSCs from healthy wild type mice of \citet{kremer_scnmt} and ran \texttt{scbs diff} with default parameters.
-For GO enrichment analysis, DMRs with an adjusted p-value $<0.01$ were uploaded to GREAT \citep{mclean2010great} with association rules basal plus extension, 0~bp upstream, 20~kb downstream, 1~Mbp max extension, curated regulatory domains included.
+For GO enrichment analysis, DMRs with $p^\text{adj}<0.01$ were uploaded to GREAT \citep{mclean2010great} with the association rule "basal plus extension, 0~bp upstream, 20~kb downstream, 1~Mbp max extension, curated regulatory domains included".
 
@@
 \subsection{Correlation of DNA methylation and gene expression}
-To assess correlation between gene expression and the methylation status of VMRs or promoters, we
-first detected VMRs with \texttt{scbs scan --bandwidth 1000 --var-treshold 0.05}.
+To assess correlation between gene expression and the methylation status of VMRs or promoters, we first detected VMRs with \texttt{scbs scan --bandwidth 1000 --var-treshold 0.05}.
 We then quantified DNA methylation at VMRs and promoters with \texttt{scbs matrix}.

### 2023-04-24 paper.tex
address Leonie's comments & fix some typos

@@
 \noindent \textit{Abstract} -- Single-cell bisulfite sequencing (scBS) is a technique that enables the assessment of DNA methylation at single-base pair and single-cell resolution.
-The analysis of the large datasets obtained from scBS requires preprocessing to reduce data size, improve signal-to-noise ratio, and provide interpretability.
+The analysis of large datasets obtained from scBS requires preprocessing to reduce data size, improve signal-to-noise ratio, and provide interpretability.
 Typically, this is achieved by dividing the genome into large tiles and averaging the methylation signals within each tile.
 
-Here, we demonstrate that this coarse-graining approach can lead to signal dilution, and propose improved strategies to identify more informative regions for methylation quantification, and a more accurate quantitation method than simple averaging.
+Here, we demonstrate that this coarse-graining approach can lead to signal dilution.
+As an alternative, we propose improved strategies to identify more informative regions for methylation quantification, and a more accurate quantitation method than simple averaging.
 Our approach enables better discrimination of cell types and reduces the need for large numbers of cells.
@@
 
-The standard approach to analyze scBS-seq data is based on methodology developed for the analysis of scRNA-seq data.
+The standard approach to analyze scBS data is based on methodology developed for the analysis of scRNA-seq data.
 Therefore, we start by briefly reviewing how sc\emph{RNA}-seq data is commonly analyzed, before we discuss scBS data analysis.
@@
 To this end, one needs to establish which cells are similar to each other, i.e., quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile.
-A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm.
+A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows:
+One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm.
 In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-4}$) to the normalized fractions before taking the logarithm.
@@
 
-This procedure is commonly adapted when working with single-cell DNA methylation data, because once on gets to the PCA step, one can then continue with the established methods just mentioned.
-The task is therefore to construct from the methylation data a matrix suitable as input for PCA.
-A simple and common approach, used for instance by \citet{luo2017single}, is to divide the genome into tiles of e.g.
+This procedure is commonly adapted when working with single-cell DNA methylation data, because once one gets to the PCA step, one can then continue with the established methods just mentioned.
+However, constructing a matrix suitable as input for PCA from methylation data is requires deviation from the standard scRNA-seq workflow due to considerable differences in data structure.
+First, while scRNA-seq quantifies RNA abundance of genes or transcripts, scBS is genome-wide and thus lacks a natural choice for features in which methylation is to be quantified.
+Second, instead of counts, scBS generates binary data that informs us whether certain cytosines in a given cell are methylated or not.
+A simple and common approach to construct a methylation matrix suitable for PCA, used for instance by \citet{luo2017single}, is to divide the genome into tiles of e.g.
 100~kb size, and calculate for each cell the average methylation of the DNA within each tile.
@@
 
-While this simple procedure is straight-forward and produces useable results, it is not optimal.
+While this simple procedure is straight-forward and produces usable results, it is not optimal.
 In this paper, we discuss weaknesses of the simple approach and suggest several refinements to overcome them.
@@
 Select, say, the top 2\% windows with the highest variances and mark them as VMRs.
-Wherever thus marked windows overlap or are separated by only a small gap, merge them into one larger VMR.
+Wherever thus marked windows overlap, merge them into one larger VMR.
 Then, calculate for each of these larger VMRs the methylation signal, as before, by averaging for each cell over the residuals of all contained CpG sites.
@@
 
-Our data set \citep{kremer_scnmt} comprises the methylomes of 1566 cells isolated from mouse forebrains, as well as matched single-cell transcriptomes of the same cells.
+Our data set \citep{kremer_scnmt} comprises the single-cell methylomes of 1566 cells isolated from mouse forebrains, as well as matched single-cell transcriptomes of the same cells.
 Among these cells are distinct cell types such as oligodendrocytes, oligodendrocyte precursor cells (OPCs) and endothelial cells, as well as cellular sub-states that are part of the continuous neural stem cell differentiation trajectory.
@@
 
-We first note that gene expression is more strongly correlated with the methylation of nearby VMRs than with the methylation of their promotors (Fig.~\ref{fig:correlation}A), indicating that VMR methylation is often a better predictor of gene expression than promoter methylation.
+We first note that gene expression is more strongly correlated with the methylation of nearby VMRs than with the methylation of their promoters (Fig.~\ref{fig:correlation}A), indicating that VMR methylation is often a better predictor of gene expression than promoter methylation.
 Indeed, a gene-wise comparison revealed many genes whose expression is correlated with the nearest VMR, but not with promoter methylation.
@@
 
-We repeated this benchmark on the data set published by \citet{luo2017single}, which consists of 3377 single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal subtypes.
+We repeated this benchmark on the data set published by \citet{luo2017single}, which consists of 3377 single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal sub-types.
 The authors annotated cell types based on CH-methylation instead of CpG-methylation.
@@
 	\caption{\small \textbf{Correlation of DNA methylation and gene expression}.\\
-		\textbf{(A)} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red): for most promotors, correlation is weak.
+		\textbf{(A)} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red): for most promoters, correlation is weak.
 Promoters are defined as intervals \textpm2~kb around the TSS.
@@
 		UMAP space.
-		\textbf{(E)} Mean neighbor score obtained after analysing single-cell methylomes with different combinations of methods.
+		\textbf{(E)} Mean neighbor score obtained after analyzing single-cell methylomes with different combinations of methods.
 		Either 100~kb genomic tiles or VMRs were subjected to PCA and UMAP.
@@
 		For each window, the cell’s methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
-		Then the t statistic of these values is obtained as a measure of methylation difference (each point represents one of the windows).
+		Then the t statistic of these values is obtained as a measure of methylation difference between two groups of cells (each point represents one of the windows).
 		Upper and lower thresholds (dashed lines) are chosen such that a defined quantile of windows have a t statistic above or below the threshold.
@@
 Then we used the method just described to identify DMRs between these two cell types (Fig.~\ref{fig:dmr}A).
-Repeating this after permuting the cell-type labels, in order to obtain a null distribution fo the t statistics, yielded
-DMRs with much weaker methylation differences. Consequently, we could assign to many of the DMRs detected in the unpermuted data a low adjusted p-values (colours in Fig.~\ref{fig:dmr}A).
+Repeating this after permuting the cell-type labels, in order to obtain a null distribution of the t statistics, yielded DMRs with much weaker methylation differences.
+Consequently, we could assign to many of the DMRs detected in the unpermuted data a low adjusted p-values (colors in Fig.~\ref{fig:dmr}A).
 
@@
     \caption{\small \textbf{Overview of the functionalities implemented in the \texttt{scbs} package.}\\
-        \texttt{scbs prepare} parses methylation files produced by common bisulfite sequencing mappers and stores their contents in a compressed format optimised for efficient access to genomic intervals.
+        \texttt{scbs prepare} parses methylation files produced by common bisulfite sequencing mappers and stores their contents in a compressed format optimized for efficient access to genomic intervals.
         \texttt{scbs} also produces cell-wise summary statistics and quality plots (here: average methylation around the transcription start site) that are used to detect low-quality cells.
@@
 
-The methylation matrix produces by the \texttt{matrix} function can then serve as input for standard single-cell analysis tools like Seurat \citep{seurat} or scanpy \citep{Wolf_2018}, in order to perform, e.g., dimensionality reduction and cell clustering.
+The methylation matrix produces by the \texttt{matrix} function can then serve as input for standard single-cell analysis tools like Seurat \citep{seurat} or ScanPy \citep{Wolf_2018}, in order to perform, e.g., dimensionality reduction and cell clustering.
 For this, the methylation matrix should be provided in lieu of the log-normalized expression matrix usually expected by these tools.
@@
 This allows the user to use well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc.
-By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.
+By offering these functions, \texttt{scbs} \todo{bridges a gap in the chain of existing tools} that so far hindered practitioners in their data analysis.
 
@@
 
-In conclusion, we have presented powerful improvements to scBS data preprocessing and a software tool that implements these and bridges a gap in existing tool chains.
+In conclusion, we have presented powerful improvements to scBS data preprocessing and a software tool that implements these and \todo{bridges a gap in existing tool chains}.
 
@@
 
-For later use: We write $C$ for the set of all cells in the assay (i.e., $C$ is the index set for the cell indices $j$).
-Moreover,
-we define $C_i\subset C$ as the set of all those cells $j$ that have reads covering position $i$,
+For later use: We define $C$ as the set of all cells in the assay (i.e., $C$ is the index set for the cell indices $j$).
+Moreover, we define $C_i\subset C$ as the set of all those cells $j$ that have reads covering position $i$,
 $$ C_i=\{j\in C: x_{ij}\neq\text{NA}\}.$$
@@
 Running through the whole genome for each permutation would be too computationally expensive.
-Instead, we go through the genome only once, but reshuffle the group labels every 2 megabasepairs.
-% didn't abbreviate cause one of our example genes is "Mbp"...
+Instead, we go through the genome only once, but reshuffle the group labels every 2~Mbp.
 
-All the t values obtained from this permutation procedure are taken together to get an empirical null distibution.
+All the t values obtained from this permutation procedure are taken together to get an empirical null distribution.
 Then, we can use this null distribution to control false discovery rate (FDR) by applying the Benjamini-Hochberg procedure in its p-value-free form: Let us write $T$ for the set of all t values obtained from the ``real'' assignment of cells to group labels and $T_0$ from the set of all t values obtained from the shuffled assignments, i.e., the empirical null distribution.
@@
 \subsection{Correlation of DNA methylation and gene expression}
-To assess correlation between gene expression and the methylation status of VMRs or promoters, we first detected VMRs with \texttt{scbs scan --bandwidth 1000 --var-treshold 0.05}.
+To assess correlation between gene expression and the methylation status of VMRs or promoters, we first detected VMRs with \texttt{scbs scan --bandwidth 1000 --var-threshold 0.05}.
 We then quantified DNA methylation at VMRs and promoters with \texttt{scbs matrix}.
@@
 SA and LPMK acknowledge funding by the Klaus Tschira Foundation (project 00.022.2019).
-LPMK, LK, SC, and AMV acknowledge funding from the European Commission via ERC grant 771376.
+LPMK, MB, LK, SC, and AMV acknowledge funding from the European Commission via ERC grant 771376.
 We thank Alexey Uvarovskii for code refactoring and helpful suggestions on the source code.
@@
 	\end{center}
-	\captionof{figure}{\small \textbf{Benchmark of our methods on single-cell methylomes of neural subtypes of the murine frontal cortex}.\\
+	\captionof{figure}{\small \textbf{Benchmark of our methods on single-cell methylomes of neural sub-types of the murine frontal cortex}.\\
 	Our methods were benchmarked on single-cell methylomes from \citet{luo2017single} as described in Fig.~\ref{fig:score}. 

### 2023-04-25 paper.tex
rm claim that our matrix works with seurat & rm a repetition

@@
 Lukas P.~M.
-Kremer\textsuperscript{1,2,*}, Martina Braun\textsuperscript{2}, Leonie Küchenhoff\textsuperscript{1,2}, Santiago Cerrizuela\textsuperscript{2},\\Ana Martin-Villalba\textsuperscript{2}, Simon Anders\textsuperscript{1,*}\\[1ex]
+Kremer\textsuperscript{1,2,*}, Martina M.
+Braun\textsuperscript{2}, Leonie Küchenhoff\textsuperscript{1,2}, Santiago Cerrizuela\textsuperscript{2},\\Ana Martin-Villalba\textsuperscript{2}, Simon Anders\textsuperscript{1,*}\\[1ex]
 {\footnotesize \textsuperscript{1} BioQuant Centre, University of Heidelberg, Heidelberg, Germany\\[-.5ex] 
@@
 Furthermore, we discuss how to perform comparative analyses.
-Finally, we present ``scbs'', a comprehensive software toolkit to perform scBS data analysis, using the described methods.
+Finally, we present ``\texttt{scbs}'', a comprehensive software toolkit to perform scBS data analysis, using the described methods.
 \blfootnote{\hspace{-.5cm}\raggedright\textsuperscript{*} e-mail:  \texttt{l.kremer@dkfz-heidelberg.de} and\\ \hspace{1cm}\texttt{simon.anders@bioquant.uni-heidelberg.de}}
@@
 as a measure of differential methylation between the two cell groups.
-We identify the windows with the most extreme t statistics, e.g.
-windows in the 2\% upper and lower tails.
+We identify the windows with the most extreme t statistics, e.g. windows in the 2\% upper and lower tails.
 We then merge any overlapping windows in the upper tail into larger DMRs, and do the same for windows in the lower tail, and then we re-calculate the t statistic for each larger DMR.
 
-To assess statistical significance of DMRs, we repeat the same procedure on permutations of the scBS data, i.e.
-the same data set with randomly shuffled cell labels.
+To assess statistical significance of DMRs, we repeat the same procedure on permutations of the scBS data, i.e. the same data set with randomly shuffled cell labels.
 The DMR t statistics obtained from permuted data are then used to estimate the false-discovery rate (FDR), yielding an adjusted p-value for each DMR.
@@
 We note that both \texttt{matrix} and \texttt{profile} accept any valid BED file as input, which means that the user can quantify and visualize methylation at any set of genomic features of interest, such as promoters, enhancers or transcription factor binding sites, obtaining one profile plot per cell.
-
-The methylation matrix produces by the \texttt{matrix} function can then serve as input for standard single-cell analysis tools like Seurat \citep{seurat} or ScanPy \citep{Wolf_2018}, in order to perform, e.g., dimensionality reduction and cell clustering.
+% This is not really true cause you still need to deal with missing values and all the default transformations don't make sense when you're not dealing with count data: "The methylation matrix produced by the \texttt{matrix} function can then serve as input for standard single-cell analysis tools like Seurat \citep{seurat} or ScanPy \citep{Wolf_2018}, in order to perform, e.g., dimensionality reduction and cell clustering.
 For this, the methylation matrix should be provided in lieu of the log-normalized expression matrix usually expected by these tools.
+The methylation matrix produced by the \texttt{matrix} function can then serve as input for established methods used for the exploration of single-cell data, such as dimensionality reduction and cell clustering.
 
@@
 FDR estimation via permutation allows us to report statistical significance of each DMR.
-% maybe here discuss pseudoreplication issues?
 Applying our approach to NSCs and oligodendrocytes demonstrated that the obtained DMRs locate near meaningful loci associated with cell-type specific functions.
@@
 This allows the user to use well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc.
-By offering these functions, \texttt{scbs} \todo{bridges a gap in the chain of existing tools} that so far hindered practitioners in their data analysis.
+By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.
 
@@
 
-In conclusion, we have presented powerful improvements to scBS data preprocessing and a software tool that implements these and \todo{bridges a gap in existing tool chains}.
+In conclusion, we have presented powerful improvements to scBS data preprocessing and a software tool that implements these.
 

### 2023-04-25 paper.tex
minor stuff

@@
 In brief, DNA is treated with bisulfite, which converts unmethylated cytosines to uracils which are read as thymine in subsequent PCR, while methylated cytosines are protected from conversion.
-After sequencing, these conversions allow to determine the methylation status of all cytosines covered by reads \citep{Frommer_1992}.
+After sequencing, these conversions allow for the determination of the methylation status of all cytosines covered by reads \citep{Frommer_1992}.
 Bisulfite sequencing can also be performed at single-cell resolution \citep{Smallwood_2014} and even in parallel with scRNA-seq \citep{scMTseq,Clark2018}.
@@
 This procedure is commonly adapted when working with single-cell DNA methylation data, because once one gets to the PCA step, one can then continue with the established methods just mentioned.
-However, constructing a matrix suitable as input for PCA from methylation data is requires deviation from the standard scRNA-seq workflow due to considerable differences in data structure.
+However, constructing a matrix suitable as input for PCA from methylation data requires deviation from the standard scRNA-seq workflow due to considerable differences in data structure.
 First, while scRNA-seq quantifies RNA abundance of genes or transcripts, scBS is genome-wide and thus lacks a natural choice for features in which methylation is to be quantified.
 Second, instead of counts, scBS generates binary data that informs us whether certain cytosines in a given cell are methylated or not.
-A simple and common approach to construct a methylation matrix suitable for PCA, used for instance by \citet{luo2017single}, is to divide the genome into tiles of e.g.
-100~kb size, and calculate for each cell the average methylation of the DNA within each tile.
+A simple and common approach to construct a methylation matrix suitable for PCA, used for instance by \citet{luo2017single}, is to divide the genome into tiles of e.g., 100~kb size, and calculate for each cell the average methylation of the DNA within each tile.
 To this end, one identifies in the tile all CpG sites that are covered by at least one read and averages their methylation state, i.e., one denotes as average DNA methylation of the tile in a given cell the proportion of the observed CpG sites in the tile that were found to be methylated (Figure \ref{fig:smoothres}A).
@@
 Finally, we describe and demonstrate an approach to detect differentially methylated regions (DMRs) in scBS data.
-We also describe our software toolkit, scbs, that facilitates all these analyses.
+We also describe our software toolkit, \texttt{scbs}, that facilitates all these analyses.
 
@@
 In the standard approach, one divides up (tiles) each chromosome into non-overlapping, equal-sized intervals, and quantitates the methylation of each tile.
-Such rigid placing of interval boundaries is unlikely to be optimal: for example, a VMR might be much smaller than a tile and the signal from its CpG sites will hence be drowned out by the larger number of uninformative CpG sites that are equal in all cells, when averaging over all the CpG sites in the tile.
+Such rigid placement of interval boundaries is unlikely to be optimal: for example, a VMR might be much smaller than a tile and the signal from its CpG sites will hence be drowned out by the larger number of uninformative CpG sites that are equal in all cells, when averaging over all the CpG sites in the tile.
 
@@
 Quantify the methylation of each cell in each window by averaging the cell's methylation residuals over all CpGs in the window, as described above and depicted in Fig.~\ref{fig:smoothres}B.
-Then, calculate for each window the variance of these values over all cells.
+Next, calculate for each window the variance of these values over all cells.
 Select, say, the top 2\% windows with the highest variances and mark them as VMRs.
 Wherever thus marked windows overlap, merge them into one larger VMR.
-Then, calculate for each of these larger VMRs the methylation signal, as before, by averaging for each cell over the residuals of all contained CpG sites.
+Then, calculate for each of these merged VMRs the methylation signal, as before, by averaging for each cell over the residuals of all contained CpG sites.
 
@@
 Lastly, we assessed whether our proposed workflow requires fine-tuning of parameters (Fig.~\ref{fig:sweep}).
-To this end, we re-analyzed both data sets, as well as sub-samples of the data with different VMR detection parameters, namely the width of the sliding window in bp, (set with option \texttt{--bandwidth} in our ``scbs'' software, default 2000) and the variance threshold above which windows are merged to VMRs (\texttt{--var-threshold}, default 0.02).
+To this end, we re-analyzed both data sets, as well as sub-samples of the data with different VMR detection parameters, namely the width of the sliding window in bp, (set with option \texttt{--bandwidth} in our \texttt{scbs} software, default 2000) and the variance threshold above which windows are merged to VMRs (\texttt{--var-threshold}, default 0.02).
 This parameter sweep showed that our workflow gives good results over a wide range of parameter values.
@@
 In addition, by pinpointing exact regions of interest, the VMR and DMR functionalities provide useful input for genomics-style analyses.
-Depending on the research question at hand, individual DMRs can be related to nearby genomic features such as gene bodies or known regulatory elements, tested for methylation differences between groups of cells, or subjected to gene ontology enrichment or motif enrichment.
+Depending on the research question at hand, individual DMRs may be related to nearby genomic features such as gene bodies or known regulatory elements, tested for methylation differences between groups of cells, or subjected to gene ontology and motif enrichment.
 
@@
 The index $i$ runs over all CpG positions present in the genome, the index $j$ over all cells in the assay.
-We write $x_{ij}=0$ if position $i$ was found to be unmethylated in cell $j$ by bisulfite sequencing, $x_{ij}=1$ if it was methylated, and $x_{ij}=\text{NA}$ if position $i$ was not covered by reads from cell $j$ and the methylation status is therefore not available (``NA''),
+We write $x_{ij}=0$ if position $i$ was found to be unmethylated in cell $j$ by bisulfite sequencing, $x_{ij}=1$ if it was methylated, and $x_{ij}=\text{NA}$ if position $i$ was not covered by reads from cell $j$ and the methylation status is therefore not available (``NA'').
 

### 2023-04-26 paper.tex
implement Santi's comments

@@
 {\footnotesize \textsuperscript{1} BioQuant Centre, University of Heidelberg, Heidelberg, Germany\\[-.5ex] 
-\textsuperscript{2} Molecular Neurobiology Dept., German Cancer Research Center (DKFZ), Heidelberg, Germany}
+\textsuperscript{2} Division of Molecular Neurobiology, German Cancer Research Center (DKFZ), Heidelberg, Germany}
 \vspace{1.5ex}
@@
 In the example of Figure \ref{fig:smoothres}A, the depicted interval is covered by a single read for two of the three cells shown and no read in the third.
-The read from cell 2 shows much more methylation than the read from cell 1, and the standard analysis would therefore consider cell 1 to have higher methylation in the interval than cell 2.
+The read from cell 2 shows much more methylation than the read from cell 1, and the standard analysis would therefore consider cell 2 to have higher methylation in the interval than cell 1.
 However, given that the two reads agree wherever they overlap, a more parsimonious interpretation would be that the cells do not show difference in methylation within the interval.
@@
 SA and LPMK acknowledge funding by the Klaus Tschira Foundation (project 00.022.2019).
-LPMK, MB, LK, SC, and AMV acknowledge funding from the European Commission via ERC grant 771376.
+LPMK, MMB, LK, SC, and AMV acknowledge funding from the European Commission via ERC grant 771376.
 We thank Alexey Uvarovskii for code refactoring and helpful suggestions on the source code.

### 2023-04-26 paper.tex
one sentence per line

@@
 
-Taken together, this shrunken mean of residuals quantitation reduces variance in comparison to simple averaging of raw methylation calls. We will show further below how this improves results.
+Taken together, this shrunken mean of residuals quantitation reduces variance in comparison to simple averaging of raw methylation calls.
+We will show further below how this improves results.
 
@@
 For instance, it is known  since long that CpG-rich promoters of housekeeping genes are unmethylated, and that a large proportion of the remaining genome is highly methylated regardless of cell type \citep{bird1986cpg}.
-In contrast, DNA methylation at certain genomic features such as enhancers is more dynamic (e.g., \citet{argelaguet2019gastru}), and thus more variable across cells.
+In contrast, DNA methylation at certain genomic features such as enhancers is more dynamic \citep[e.g.,][]{argelaguet2019gastru}, and thus more variable across cells.
 Only the latter regions are of value for our goal of quantitating dissimilarity between cells.
@@
 
-Therefore, we propose the following approach (Fig.~\ref{fig:vmr}): Divide up the chromosome into many \emph{overlapping} windows that start at regular multiples of a fixed, small, step size.
+Therefore, we propose the following approach (Fig.~\ref{fig:vmr}):
+Divide up the chromosome into many \emph{overlapping} windows that start at regular multiples of a fixed, small, step size.
 Quantify the methylation of each cell in each window by averaging the cell's methylation residuals over all CpGs in the window, as described above and depicted in Fig.~\ref{fig:smoothres}B.
@@
 		\textbf{(B, C)} Exemplary UMAPs obtained when analyzing the data set with our proposed methods (B) or conventional methods based on genome tiling (C).
-		\textbf{(D)} Illustration of the neighbor score. Note that, actually,
-		distances are measured in 15-dimensional PC space, not in the 2-dimensional
-		UMAP space.
+		\textbf{(D)} Illustration of the neighbor score.
+        Note that, actually, distances are measured in 15-dimensional PC space, not in the 2-dimensional UMAP space.
 		\textbf{(E)} Mean neighbor score obtained after analyzing single-cell methylomes with different combinations of methods.
 		Either 100~kb genomic tiles or VMRs were subjected to PCA and UMAP.
-		Tiles or VMRs were pre-processed as proposed in this work (shrunken mean of residuals subjected to iterative PCA) or using conventional methods (PCA on methylation percentages, lightly imputed as done in \citet{luo2017single}; see Methods for details).
+		Tiles or VMRs were pre-processed as proposed in this work (shrunken mean of residuals subjected to iterative PCA) or using conventional methods (PCA on
+        methylation percentages, lightly imputed as done in \citet{luo2017single}; see Methods for details).
 		The full 1566-cell data set was sub-sampled to simulate smaller data sets (x axis).
@@
 
-The function \texttt{scbs prepare} reads a set of methylation files (e.g.
-produced by Bismark) and produces a file that stores the matrix $x$ in a space-efficient format, as follows: $x$ is represented as a SciPy sparse matrix \citep{SciPy}, encoding the actual values 0, 1, and NA as -1, 1, and 0, respectively.
+The function \texttt{scbs prepare} reads a set of methylation files (e.g. produced by Bismark) and produces a file that stores the matrix $x$ in a space-efficient format, as follows:
+$x$ is represented as a SciPy sparse matrix \citep{SciPy}, encoding the actual values 0, 1, and NA as -1, 1, and 0, respectively.
 Coding NA (the most common value) as 0 leverages SciPy's sparse matrix storage.
@@
 We take the union of all these intervals, split the union into connected components, and call each component a VMR.
-Putting that last step in other words: We take all the intervals with variance in the top 2-percentile, fuse intervals that overlap and call the regions thus obtained the VMRs.
+Putting that last step in other words:
+We take all the intervals with variance in the top 2-percentile, fuse intervals that overlap and call the regions thus obtained the VMRs.
 
@@
 use the Welch-Satterthwaite formula to get an approximate degree of freedom and then calculate the tail probability of the corresponding t distribution.
-However, this is unlikely to hold for two reasons: First, the Welch-Satterthwaite degrees of freedom are only based on the number of cells per group and do not account for the fact that the read coverage might vary from cell to cell.
+However, this is unlikely to hold for two reasons:
+First, the Welch-Satterthwaite degrees of freedom are only based on the number of cells per group and do not account for the fact that the read coverage might vary from cell to cell.
 Second, the fusing of the DMRs obtained in the scanning step to obtain fused candidate DMRs would invalidate subsequent p-value based adjustment for multiple testing.
 
-Therefore, we have instead implemented a permutation procedure, which works as follows: We randomly shuffle the assignment of the cells in $C_\text{A}\cup C_\text{A}$ to either of the two groups and then rerun the whole procedure, i.e., the scanning step, the DMR fusing and the calculation of t values from the (potentially fused) candidate DMRs.
+Therefore, we have instead implemented a permutation procedure, which works as follows:
+We randomly shuffle the assignment of the cells in $C_\text{A}\cup C_\text{A}$ to either of the two groups and then rerun the whole procedure, i.e., the scanning step, the DMR fusing and the calculation of t values from the (potentially fused) candidate DMRs.
 This needs to be done for a sufficiently large number of permutations.
@@
 All the t values obtained from this permutation procedure are taken together to get an empirical null distribution.
-Then, we can use this null distribution to control false discovery rate (FDR) by applying the Benjamini-Hochberg procedure in its p-value-free form: Let us write $T$ for the set of all t values obtained from the ``real'' assignment of cells to group labels and $T_0$ from the set of all t values obtained from the shuffled assignments, i.e., the empirical null distribution.
+Then, we can use this null distribution to control false discovery rate (FDR) by applying the Benjamini-Hochberg procedure in its p-value-free form:
+Let us write $T$ for the set of all t values obtained from the ``real'' assignment of cells to group labels and $T_0$ from the set of all t values obtained from the shuffled assignments, i.e., the empirical null distribution.
 To adjust a specific t value $t_i\in T$, we calculate

### 2023-04-26 paper.tex
replaced $$ by \[

@@
 Moreover, we define $C_i\subset C$ as the set of all those cells $j$ that have reads covering position $i$,
-$$ C_i=\{j\in C: x_{ij}\neq\text{NA}\}.$$
+\[ C_i=\{j\in C: x_{ij}\neq\text{NA}\}.\]
 Conversely, we define $G_j$ as the set of all the CpG positions $i$ covered by reads from cell $j$ 
-$$ G_j=\{i: x_{ij}\neq\text{NA}\}.$$
+\[ G_j=\{i: x_{ij}\neq\text{NA}\}.\]
 
@@
 For each CpG position $i$, we write 
-$$\overline{x}_i=\langle x_{ij} \rangle_{j\in C_i} = \frac{1}{|C_i|}\sum_{j\in C_i} x_{ij}$$ 
+\[\overline{x}_i=\langle x_{ij} \rangle_{j\in C_i} = \frac{1}{|C_i|}\sum_{j\in C_i} x_{ij}\] 
 for the average methylation at position $i$, where $\langle\cdot\rangle$ denotes averaging, and the average runs over all the cells $j\in C_i$, i.e.
@@
 
-$$ m_{Ij} = \left< x_{ij} \right>_{i\in I\cap G_j}.$$
+\[ m_{Ij} = \left< x_{ij} \right>_{i\in I\cap G_j}.\]
 
@@
 
-$$ z_{Ij} = m_{Ij} - \langle m_{Ij'}\rangle_{j'\in C}.$$
+\[ z_{Ij} = m_{Ij} - \langle m_{Ij'}\rangle_{j'\in C}.\]
 
 As an alternative, we suggest to consider the residuals of the individual CpG methylation values $x_{ij}$ from the smoothed average $\tilde x_i$,
-$$ r_{ij} = x_{ij} - \tilde x_i, $$
+\[ r_{ij} = x_{ij} - \tilde x_i, \]
 and averaging over these, obtaining
@@
 For a given interval $I$, we calculate the mean each of the mean shrunken residuals $r_{ij}$ (see Eq.~(\ref{avgres})) over the cells $j$ in each of the two groups,
-$$ \mu^g_I = \langle r_{Ij} \rangle_{j\in g},\qquad g=\text{A},\text{B} $$
+\[ \mu^g_I = \langle r_{Ij} \rangle_{j\in g},\qquad g=\text{A},\text{B}. \]
 
 We also calculate a variance as in Eq.(\ref{var}):
-$$ v^g_I = \frac{1}{|C_\text{g}|-1}  \sum_{j\in g} \left( r_{Ij} - \mu_I^g \right)^2, 
-\qquad g=\text{A},\text{B}$$
+\[ v^g_I = \frac{1}{|C_\text{g}|-1}  \sum_{j\in g} \left( r_{Ij} - \mu_I^g \right)^2, 
+\qquad g=\text{A},\text{B}.\]
 
 From this, we calculate Welch's t statistic as usual:
-$$ t_I = \frac{\mu^\text{A}_I - \mu^\text{B}_I}{\sqrt{\frac{v^\text{A}_I}{|C_A|} + \frac{v^\text{B}_I}{|C_B|}}}.$$
+\[ t_I = \frac{\mu^\text{A}_I - \mu^\text{B}_I}{\sqrt{\frac{v^\text{A}_I}{|C_A|} + \frac{v^\text{B}_I}{|C_B|}}}.\]
 
@@
 To adjust a specific t value $t_i\in T$, we calculate
-$$ p^\text{adj}_i = \frac{ \left|\left\{t_j\in T_0\mid|t_j| > |t_i| \right\}\right| \big/ |T_0|}
-{ \left|\left\{t_j\in T\mid|t_j| > |t_i| \right\}\right| \big/ |T|}.$$
+\[ p^\text{adj}_i = \frac{ \left|\left\{t_j\in T_0\mid|t_j| > |t_i| \right\}\right| \big/ |T_0|}
+{ \left|\left\{t_j\in T\mid|t_j| > |t_i| \right\}\right| \big/ |T|}.\]
 

### 2023-11-15 paper.tex
cover letter as sent to NBT
add data accessibility section
update benchmark fig and text

@@
 \newcommand{\todo}[1]{[\textcolor{orange}{#1}]}
+\newcommand{\new}[1]{\textcolor{blue}{#1}}
 
@@
 
-Lukas P.~M.
-Kremer\textsuperscript{1,2,*}, Martina M.
-Braun\textsuperscript{2}, Leonie Küchenhoff\textsuperscript{1,2}, Santiago Cerrizuela\textsuperscript{2},\\Ana Martin-Villalba\textsuperscript{2}, Simon Anders\textsuperscript{1,*}\\[1ex]
+Lukas P.~M. Kremer\textsuperscript{1,2,*},
+Martina M. Braun\textsuperscript{2},
+Svetlana Ovchinnikova\textsuperscript{1},
+Leonie Küchenhoff\textsuperscript{1,2},\\
+Santiago Cerrizuela\textsuperscript{2},
+Ana Martin-Villalba\textsuperscript{2},
+Simon Anders\textsuperscript{1,*}\\[1ex]
 {\footnotesize \textsuperscript{1} BioQuant Centre, University of Heidelberg, Heidelberg, Germany\\[-.5ex] 
@@
 
-April 2023
+May 2023
 \vspace{4.5ex}
@@
 
-To demonstrate the value of our proposed improvements, we employ our own data \citep{kremer_scnmt} as well as a previously published data set, namely the one by  \citet{luo2017single} already mentioned in the introduction.
+To demonstrate the value of our proposed improvements, we benchmarked various combinations of analysis methods on \new{\todo{N} diverse single-cell methylome data sets.}
 
@@
 We consider these cell labels as ground truth and tested whether we are able to distinguish the same groups of cells based on their methylomes.
-To do this, we subjected the methylomes to four variants of the workflow described above: we defined genomic intervals in two ways and quantified methylation in these intervals in two ways, then always subjected the resulting matrix to PCA.
-To define the intervals, we either simply tiled the genome into 100~kb bins as is commonly done, or we used our VMR detection approach.
-To quantify methylation in these intervals, we either averaged in these intervals, obtaining methylation percentages (``classic'' pre-processing), or we calculated the shrunken mean of residuals.
-
+\new{
+To do this, we subjected the methylomes to various combinations of analysis methods including our own proposed methods and others that are commonly used.
+Specifically, we selected four different sets of genomic intervals at which DNA methylation is to be quantified:
+either VMRs detected with our approach, 100~kb genomic tiles, promoter regions (TSS\textpm2000~bp), or candidate cis-regulatory elements \citep[ENCODE cCREs,][]{encode2020expanded}.
+To quantify methylation at these features, we either averaged in these intervals, obtaining methylation percentages, or we calculated the shrunken mean of residuals.
+Finally, the resulting methylation matrices were subjected to iterative PCA (see Methods).
+}
 Visual inspection of the resulting UMAPs revealed that our proposed combination of methods results in more clearly separated cell types, compared to a UMAP obtained with default analysis methods (Fig.~\ref{fig:score}B,C).
@@
 
-To quantify this performance gain in a more rigorous manner, we used a score that quantifies whether cells were placed, in 15-dimensional PC space, in a neighborhood comprising cells of the same cell type ("neighbor score", see Figure \ref{fig:score}D, and Methods for details).
+To quantify this performance gain in a more rigorous manner, we used a score that quantifies whether cells were placed, in 15-dimensional PC space, in a neighborhood comprising cells of the same cell type ("neighbor score", see Fig.~\ref{fig:score}E, and Methods for details).
 We then ask how this score depends on the number of available cells.
 This is relevant as scBS protocols are costly and labor-intensive, and only few laboratories are currently able to obtain thousands of single cell methylomes.
-To simulate smaller data sets, we sub-sampled the 1566-cell data set into smaller data sets, analyzed them again with the four workflow variants, and calculated the mean neighbor score for each (Figure \ref{fig:score}E).
-The results demonstrate that the forebrain cell clusters are more cleanly separated when using our proposed methods.
-This performance gain was observed also in smaller data sets, although cell cluster separation generally becomes more difficult in those.
-Both the use of VMRs instead of genomic tiles, as well as the use of the shrunken mean of the residuals over methylation averages improved cell cluster separation.
+To simulate smaller data sets, we sub-sampled the 1566-cell data set into smaller data sets, analyzed them again with all combinations of analysis methods, and calculated the mean neighbor score for each (Figure \ref{fig:score}D).
+\new{
+The results demonstrate that quantifying methylation at VMRs or regulatory elements leads to a cleaner separation of cell states than using promoter regions or 100~kb tiles.
+Using the shrunken mean of residuals as a measure of DNA methylation furthermore improved results.
+This effect was most noticeable in promoters and genomic tiles, presumably because an individual promoter region or tile might span genomic regions with varying levels of DNA methylation as depicted in Fig.~\ref{fig:smoothres}B, which our method accounts for.
+Although cell cluster separation generally becomes more difficult in those, performance gains were observed also in smaller data sets.
+}
 
-We repeated this benchmark on the data set published by \citet{luo2017single}, which consists of 3377 single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal sub-types.
-The authors annotated cell types based on CH-methylation instead of CpG-methylation.
-Using these labels as ground truth we re-analyzed the full and sub-sampled CpG methylation data with the four methods combinations and found that our suggested methods consistently outperformed default methods (Fig.~\ref{fig:score_luo}).
+\new{
+We repeated this benchmark on \todo{N} published single-cell methylome data sets.
+These include neuronal sub-types of the murine cortex \citep[][using cell type labels derived from CH-methylation in genomic tiles, instead of CpG-methylation, as ground truth]{luo2017single}, cells isolated from mouse embryos during the onset of gastrulation \citep[][using RNA-derived cell clusters, or alternatively embryonic stage as ground truth]{argelaguet2019gastru} and human colorectal cancer cells \citep[][using \todo{tumor region(?)} as ground truth]{bian2018single}.
+Again, we subjected each data set to all possible combinations of genomic feature selection and methylation quantification.
+We furthermore included three additional approaches to perform dimensionality reduction in our benchmarks, including PCA with two different pre-processing strategies, as well as MOFA+, a dimensionality reduction technique designed for multi-modal single-cell data that can also process methylation data \citep{argelaguet2020mofa}.
+}
+
+\new{
+These extensive benchmarks revealed that our proposed combination of methods, i.e.~using the shrunken means of residuals of VMRs for dimensionality reduction, is able to distinguish diverse cellular properties such as cell type, \todo{colorectal cancer sampling region(?)}, embryonic stage, and germ layer (Fig.~\ref{fig:score_luo}).
+In addition, the use of ENCODE regulatory elements as input features also yielded good results.
+However, due to their large number (339\,815 in mouse, 1\,063\,878 in human) using regulatory regions as features resulted in greater run times and increased RAM usage, even when accounting for the additional step of VMR detection (Fig.~\ref{fig:score}F).
+}
 
@@
 	\begin{center}
-		\includegraphics[width=0.8\textwidth]{figures/Fig_benchmark.png}
+		\includegraphics[width=0.9\textwidth]{figures/Fig_benchmark.png}
 	\end{center}
@@
 
-\subsection{Availability}
+\subsection{Data Availability}
+
+Single-cell multi-omic data of cells of the murine forebrain \citep{kremer_scnmt} is available at the NCBI Gene Expression Omnibus (GEO) under the accession \texttt{GSE210806}.
+
+\subsection{Code Availability}
 
@@
 We thank Alexey Uvarovskii for code refactoring and helpful suggestions on the source code.
+\bigskip
 

### 2023-11-16 paper.tex
discuss benchmark

@@
 To do this, we subjected the methylomes to various combinations of analysis methods including our own proposed methods and others that are commonly used.
-Specifically, we selected four different sets of genomic intervals at which DNA methylation is to be quantified:
+Specifically, we selected four different sets of genomic intervals at which CpG methylation is to be quantified:
 either VMRs detected with our approach, 100~kb genomic tiles, promoter regions (TSS\textpm2000~bp), or candidate cis-regulatory elements \citep[ENCODE cCREs,][]{encode2020expanded}.
 To quantify methylation at these features, we either averaged in these intervals, obtaining methylation percentages, or we calculated the shrunken mean of residuals.
-Finally, the resulting methylation matrices were subjected to iterative PCA (see Methods).
+Finally, the resulting methylation matrices were subjected to iterative PCA (see Methods) and subsequent UMAP for visualization.
 }
+
 Visual inspection of the resulting UMAPs revealed that our proposed combination of methods results in more clearly separated cell types, compared to a UMAP obtained with default analysis methods (Fig.~\ref{fig:score}B,C).
@@
 \new{
-The results demonstrate that quantifying methylation at VMRs or regulatory elements leads to a cleaner separation of cell states than using promoter regions or 100~kb tiles.
+The results confirmed that quantifying methylation at VMRs leads to a cleaner separation of cell states than using promoter regions or 100~kb tiles.
 Using the shrunken mean of residuals as a measure of DNA methylation furthermore improved results.
@@
 \new{
-We repeated this benchmark on \todo{N} published single-cell methylome data sets.
+Overall, VMR quantification yielded results similar to those obtained when quantifying ENCODE regulatory regions.
+This demonstrates that our VMR detection approach identifies genomic regions in scBS data that are as informative as state-of-the-art epigenomic annotations.
+An added benefit of VMR detection is that this approach is available even in the absence of such annotations, for instance when studying species other than human or mouse.
+Furthermore, using VMRs over regulatory regions resulted in decreased RAM requirements as well as a shorter runtime, even when accounting for the additional step of VMR detection (Fig.~\ref{fig:score}F).
+This is likely due to the large number of annotated regulatory features (339\,815 in mouse, 1\,063\,878 in human) as opposed to the 63\,421 detected VMRs.
+Filtering regulatory regions by sequencing coverage reduced resource requirements, but at the cost of a reduced ability to distinguish cell types (Fig.~\ref{fig:todo}F).
+}
+
+\new{
+We repeated this benchmark on an additional \todo{3(?)} published single-cell methylome data sets.
 These include neuronal sub-types of the murine cortex \citep[][using cell type labels derived from CH-methylation in genomic tiles, instead of CpG-methylation, as ground truth]{luo2017single}, cells isolated from mouse embryos during the onset of gastrulation \citep[][using RNA-derived cell clusters, or alternatively embryonic stage as ground truth]{argelaguet2019gastru} and human colorectal cancer cells \citep[][using \todo{tumor region(?)} as ground truth]{bian2018single}.
@@
 We furthermore included three additional approaches to perform dimensionality reduction in our benchmarks, including PCA with two different pre-processing strategies, as well as MOFA+, a dimensionality reduction technique designed for multi-modal single-cell data that can also process methylation data \citep{argelaguet2020mofa}.
-}
-
-\new{
-These extensive benchmarks revealed that our proposed combination of methods, i.e.~using the shrunken means of residuals of VMRs for dimensionality reduction, is able to distinguish diverse cellular properties such as cell type, \todo{colorectal cancer sampling region(?)}, embryonic stage, and germ layer (Fig.~\ref{fig:score_luo}).
-In addition, the use of ENCODE regulatory elements as input features also yielded good results.
-However, due to their large number (339\,815 in mouse, 1\,063\,878 in human) using regulatory regions as features resulted in greater run times and increased RAM usage, even when accounting for the additional step of VMR detection (Fig.~\ref{fig:score}F).
+These extensive benchmarks confirmed that our proposed combination of methods, i.e.~using the shrunken means of residuals of VMRs for dimensionality reduction, is able to distinguish diverse cellular properties such as cell type, \todo{colorectal cancer sampling region(?)}, embryonic stage, and germ layer (Fig.~\ref{fig:todo}).
 }
@@
 Lastly, we assessed whether our proposed workflow requires fine-tuning of parameters (Fig.~\ref{fig:sweep}).
-To this end, we re-analyzed both data sets, as well as sub-samples of the data with different VMR detection parameters, namely the width of the sliding window in bp, (set with option \texttt{--bandwidth} in our \texttt{scbs} software, default 2000) and the variance threshold above which windows are merged to VMRs (\texttt{--var-threshold}, default 0.02).
+To this end, we re-analyzed \new{two} data sets, as well as sub-samples of the data with different VMR detection parameters namely the width of the sliding window in bp, (set with option \texttt{--bandwidth} in our \texttt{scbs} software, default 2000), the variance threshold above which windows are merged to VMRs (\texttt{--var-threshold}, default 0.02) \new{and the step size of the sliding window (\texttt{--stepsize}, default 100bp).}
 This parameter sweep showed that our workflow gives good results over a wide range of parameter values.
 For the data of \citet{luo2017single}, results are nearly independent of the parameters (Fig.~\ref{fig:sweep}B).
-In the more challenging data set of \citet{kremer_scnmt} (Fig.~\ref{fig:sweep}A), cell types were less cleanly separated when very large bandwidths, or very strict variance thresholds were selected.
+In the more challenging data set of \citet{kremer_scnmt} (Fig.~\ref{fig:sweep}A), cell types were less cleanly separated when very large bandwidths, very strict variance thresholds, \new{or a very large step size was} selected.
 However, very small bandwidths or very lenient thresholds resulted in a much higher number of VMRs and thus long computing times.
-Overall, however, our default parameter combination provided good results and fast compute times in both data sets.
+Overall, our default parameter combination´ provided good results and fast compute times in both data sets.
 

### 2023-11-17 paper.tex
minor stuff

@@
 		\textbf{(A)} Cell labels based on clustering of single-cell transcriptomes from \citet{kremer_scnmt}.
-		\textbf{(B, C)} Exemplary UMAPs obtained when analyzing the data set with our proposed methods (B) or conventional methods based on genome tiling (C).
-		\textbf{(D)} Illustration of the neighbor score.
-        Note that, actually, distances are measured in 15-dimensional PC space, not in the 2-dimensional UMAP space.
-		\textbf{(E)} Mean neighbor score obtained after analyzing single-cell methylomes with different combinations of methods.
-		Either 100~kb genomic tiles or VMRs were subjected to PCA and UMAP.
-		Tiles or VMRs were pre-processed as proposed in this work (shrunken mean of residuals subjected to iterative PCA) or using conventional methods (PCA on
-        methylation percentages, lightly imputed as done in \citet{luo2017single}; see Methods for details).
+		\textbf{(B, C)} Exemplary UMAPs obtained when analyzing the data set with \new{conventional methods based on genome tiling (B) or our proposed methods (C)}.
+		\new{
+		\textbf{(D, E)} Mean degree of cell type separation (neighbor score, E, computed in 15-dimensional PC space) obtained when analyzing single-cell methylomes with different combinations of methods.
+		Either VMRs, ENCODE regulatory elements, 100~kb genomic tiles or promoter regions (TSS\textpm2kb) were subjected to iterative PCA and UMAP.
+		CpG methylation in these intervals was either quantified by averaging (dotted lines) or using the shrunken mean of the residuals as proposed in this work (solid lines).
+	    }
 		The full 1566-cell data set was sub-sampled to simulate smaller data sets (x axis).
-		A higher neighbor score implies better separation of cell types inferred from single-cell transcriptomes of the same cells \textbf{(A)}.
+		A higher neighbor score implies better separation of cell types inferred from single-cell transcriptomes of the same cells (A).
+		\new{
+		\textbf{(F)} Resource and time requirements for dimensionality reduction of VMRs, regulatory elements and genomic tiles.
+		VMR runtime includes VMR detection (light blue).
+		}
 	}
@@
 We also presented and published an open source software tool, called \texttt{scbs}, that provides an easy-to-use implementation of the described algorithms.
-The package also offers functionality for all other steps of data preprocessing that are needed to get from the output of methylation callers such as Bismark and methylpy to a cell\texttimes region matrix that is suitable as input for standard analysis tools that work in a reduced-dimensionality setting.
+\todo{Rewrite this part to discuss other tools: The package also offers functionality for all other steps of data preprocessing that are needed to get from the output of methylation callers such as Bismark and methylpy to a cell\texttimes region matrix that is suitable as input for standard analysis tools that work in a reduced-dimensionality setting.
 This allows the user to use well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc.
-By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.
-
+By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.}
+\new{
+Our benchmarks demonstrate that selecting informative genomic features is a crucial aspect of scBS data analysis.
+Unlike other tools for scBS data analysis \citep{}, which rely on the user to manually specify which genomic intervals should be quantified, \texttt{scbs} implements an approach to discover these intervals in the data itself.
+Furthermore, \todo{DMR detection and residuals}
+}
 In addition, by pinpointing exact regions of interest, the VMR and DMR functionalities provide useful input for genomics-style analyses.
@@
 To analyze scBS data from \citet{kremer_scnmt}, single-cell CpG methylation reports from all conditions were first stored with \texttt{scbs prepare} and then smoothed with \texttt{scbs smooth} using the default bandwidth of 1000~bp.
-The mm10 mouse genome was then annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 0.02$, minimum cell requirement $=6$).
+The mm10 mouse genome was then annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, \new{stepsize $=100$}, variance threshold $= 0.02$, minimum cell requirement $=6$).
 Methylation fractions, commonly represented as percentages, and the shrunken means of residuals were quantified with \texttt{scbs matrix} for both VMRs and tiles.
@@
 SA and LPMK acknowledge funding by the Klaus Tschira Foundation (project 00.022.2019).
-LPMK, MMB, LK, SC, and AMV acknowledge funding from the European Commission via ERC grant 771376.
+LPMK, MMB, LK, SC, and AMV acknowledge funding from the European Commission via ERC grant 771376 \todo{add synergy ERC?}.
 We thank Alexey Uvarovskii for code refactoring and helpful suggestions on the source code.

### 2023-11-17 paper.tex
add fig

@@
 This is likely due to the large number of annotated regulatory features (339\,815 in mouse, 1\,063\,878 in human) as opposed to the 63\,421 detected VMRs.
-Filtering regulatory regions by sequencing coverage reduced resource requirements, but at the cost of a reduced ability to distinguish cell types (Fig.~\ref{fig:todo}F).
+Filtering regulatory regions by sequencing coverage reduced resource requirements, but at the cost of a reduced ability to distinguish cell types (Fig.~\ref{fig:todo}).
 }
@@
 We furthermore included three additional approaches to perform dimensionality reduction in our benchmarks, including PCA with two different pre-processing strategies, as well as MOFA+, a dimensionality reduction technique designed for multi-modal single-cell data that can also process methylation data \citep{argelaguet2020mofa}.
-These extensive benchmarks confirmed that our proposed combination of methods, i.e.~using the shrunken means of residuals of VMRs for dimensionality reduction, is able to distinguish diverse cellular properties such as cell type, \todo{colorectal cancer sampling region(?)}, embryonic stage, and germ layer (Fig.~\ref{fig:todo}).
+These extensive benchmarks confirmed that our proposed combination of methods, i.e.~using the shrunken means of residuals of VMRs for dimensionality reduction, is able to distinguish diverse cellular properties such as cell type, \todo{colorectal cancer sampling region(?)}, embryonic stage, and germ layer (Fig.~\ref{fig:resource_usage}).
 }
@@
 This parameter sweep showed that our workflow gives good results over a wide range of parameter values.
-For the data of \citet{luo2017single}, results are nearly independent of the parameters (Fig.~\ref{fig:sweep}B).
+For the \new{CpG-methylation} data of \citet{luo2017single}, results are nearly independent of the parameters (Fig.~\ref{fig:sweep}B).
+\new{To assess whether our proposed methods are also suitable for the analysis of CH-methylation, we furthermore performed VMR detection with default options on the mCH-portion of this data set.}
 In the more challenging data set of \citet{kremer_scnmt} (Fig.~\ref{fig:sweep}A), cell types were less cleanly separated when very large bandwidths, very strict variance thresholds, \new{or a very large step size was} selected.
@@
 Unlike other tools for scBS data analysis \citep{}, which rely on the user to manually specify which genomic intervals should be quantified, \texttt{scbs} implements an approach to discover these intervals in the data itself.
-Furthermore, \todo{DMR detection and residuals}
+Furthermore, \todo{other tools don't offer DMR detection and our residuals}.
+\todo{scbs is compatible with other tools that are up- or downstream of its capabilities, as we've shown here for MOFA+.}
 }
@@
 To analyze scBS data from \citet{kremer_scnmt}, single-cell CpG methylation reports from all conditions were first stored with \texttt{scbs prepare} and then smoothed with \texttt{scbs smooth} using the default bandwidth of 1000~bp.
-The mm10 mouse genome was then annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, \new{stepsize $=100$}, variance threshold $= 0.02$, minimum cell requirement $=6$).
+The mm10 mouse genome was then annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 0.02$, minimum cell requirement $=6$).
 Methylation fractions, commonly represented as percentages, and the shrunken means of residuals were quantified with \texttt{scbs matrix} for both VMRs and tiles.
@@
 	\end{center}
-	\captionof{figure}{\small \textbf{Benchmark of our methods on single-cell methylomes of neural sub-types of the murine frontal cortex}.\\
+	\captionof{figure}{\small \textbf{Benchmark of our methods on single-cell methylomes of neural sub-types of the murine frontal cortex.}\\
 	Our methods were benchmarked on single-cell methylomes from \citet{luo2017single} as described in Fig.~\ref{fig:score}.
@@
     \end{center}
-    \captionof{figure}{\small \textbf{Effect of VMR detection parameters on separation of cell types}.\\
+    \captionof{figure}{\small \textbf{Effect of VMR detection parameters on separation of cell types.}\\
     \textbf{(A-B)} Mean neighbor scores obtained after analyzing single-cell methylomes from \citet{kremer_scnmt} (\textbf{A}) or \citet{luo2017single} (\textbf{B}) with our proposed methods.
@@
 
+\noindent%
+\begin{minipage}{\linewidth}
+	\begin{center}
+		\includegraphics[width=.8\textwidth]{figures/SFig_benchmark_resources.png}
+	\end{center}
+	\captionof{figure}{\small \new{\textbf{VMRs are as informative as regulatory regions, but more resource-efficient.}\\
+	\textbf{(A-B)} Mean neighbor score obtained when quantifying CpG methylation at VMRs, all ENCODE regulatory regions, or a subset of regulatory regions.
+	The subset of high-quality (HQ) regulatory regions was selected in such a way that it matches the number of VMRs and contains those regulatory regions with the highest coverage.
+	\textbf{(C-D)} Runtime (top) and RAM usage (bottom) of the two dimensionality reduction techniques shown in A and B.
+	}}
+	\label{fig:resource_usage}
+\end{minipage}
+
 \end{document}

### 2023-11-18 paper.tex
update Supp Figs and add captions

@@
 Martina M. Braun\textsuperscript{2},
-Svetlana Ovchinnikova\textsuperscript{1},
+\new{Svetlana Ovchinnikova\textsuperscript{1},}  % feel free to adjust the author order, Simon.
 Leonie Küchenhoff\textsuperscript{1,2},\\
@@
 For the \new{CpG-methylation} data of \citet{luo2017single}, results are nearly independent of the parameters (Fig.~\ref{fig:sweep}B).
-\new{To assess whether our proposed methods are also suitable for the analysis of CH-methylation, we furthermore performed VMR detection with default options on the mCH-portion of this data set.}
+\new{To assess whether our proposed methods are also suitable for the analysis of CH-methylation (mCH), we furthermore performed VMR detection with default options on the mCH-portion of this data set (Fig.~\ref{fig:sweep}C).
+We obtained results that are qualitatively similar to those reported in \citep{luo2017single}, who quantified mCH in 100~kb tiles, suggesting that our methods also function on mCH data.}
+
 In the more challenging data set of \citet{kremer_scnmt} (Fig.~\ref{fig:sweep}A), cell types were less cleanly separated when very large bandwidths, very strict variance thresholds, \new{or a very large step size was} selected.
 However, very small bandwidths or very lenient thresholds resulted in a much higher number of VMRs and thus long computing times.
-Overall, our default parameter combination´ provided good results and fast compute times in both data sets.
+Overall, our default parameter combination provided good results and fast compute times in both data sets.
 
@@
 \section{Supplementary Figures}
-\vspace{6ex}
 
@@
 	\begin{center}
-	    \includegraphics[width=0.4\textwidth]{figures/SFig_benchmark.png}
+	    \includegraphics[width=\textwidth]{figures/SFig_benchmark.png}
 	\end{center}
-	\captionof{figure}{\small \textbf{Benchmark of our methods on single-cell methylomes of neural sub-types of the murine frontal cortex.}\\
-	Our methods were benchmarked on single-cell methylomes from \citet{luo2017single} as described in Fig.~\ref{fig:score}.
-	}
+	\captionof{figure}{\small \new{ \textbf{Benchmark of several combinations of analysis methods on four single-cell methylation data sets.}\\
+	\textbf{(A-E)} Benchmarking the ability to separate groups of cells based on CpG methylation data, using different analysis approaches.
+	Methylation matrices were obtained either by averaging CpG methylation in genomic intervals (dotted lines, hollow points) or using the shrunken mean of residuals (solid lines and points).
+	Each analysis was performed for the full data sets and sub-samples of each.
+	Four sets of genomic intervals (VMRs, ENCODE regulatory elements, 100~kb tiles or promoters) were quantified and separately subjected to four dimensionality reduction techniques (see Methods for details):
+	iterative PCA as proposed by us, mean-imputed PCA on intervals with high sequencing coverage \citep{luo2017single}, mean-imputed PCA on all intervals, and MOFA+ \citep{argelaguet2020mofa} on all intervals.
+	For each combination of methods, we quantified the ability to separate ground-truth cell labels reported by the authors in 15-dimensional PCA space (neighbor score).
+	\textbf{(A)} Separation of neural cell types/states annotated based on scRNA-seq of the same cells \citep{kremer_scnmt}.
+	\textbf{(B)} Separation of neuron sub-types, annotated by averaging CH-methylation (mCH) in 100~kb genomic tiles \citep{luo2017single}.
+	\textbf{(C)} Separation of human colorectal cancer sampling regions (normal adjacent tissue, primary tumor, lymph node metastasis, liver metastasis before and after treatment, omental metastasis) \citep{bian2018single}.
+	\textbf{(D-E)} Separation of three properties of mouse embryo cells:
+	The developmental stage (E4.5, E5.5, E6.5, E7.5), the germ layer or extra-embryonic tissue of origin, or the cell lineage.
+	Germ layer and lineage (\textbf{E}, only E7.5-cells) were annotated by \citet{argelaguet2019gastru} based on scRNA-seq of the same cells.
+	}}
 	\label{fig:score_luo}
 \end{minipage}
-\vspace{6ex}
 
@@
     \begin{center}
-        \includegraphics[width=.6\textwidth]{figures/SFig_parameter_sweep.png}
+        \includegraphics[width=.85\textwidth]{figures/SFig_parameter_sweep.png}
     \end{center}
-    \captionof{figure}{\small \textbf{Effect of VMR detection parameters on separation of cell types.}\\
+    \captionof{figure}{\small \textbf{Effect of VMR detection parameters \new{and methylation context} on separation of cell types.}\\
+	\new{
     \textbf{(A-B)} Mean neighbor scores obtained after analyzing single-cell methylomes from \citet{kremer_scnmt} (\textbf{A}) or \citet{luo2017single} (\textbf{B}) with our proposed methods.
-    VMRs were detected with \texttt{scbs scan} using various sliding window bandwidths or variance thresholds, and on sub-samples of the full data sets.
-    }
+    VMRs were detected with \texttt{scbs} \texttt{scan} using various sliding window bandwidths, variance thresholds, or sliding window step sizes (\textbf{A}, bottom right), and on sub-samples of the full data sets.
+    \textbf{(C)} UMAPs obtained when analysing data from \citep{luo2017single} with three different strategies for producing a methylation matrix:
+    top: averaging CH-methylation (mCH) in 100~kb genomic tiles \citep[as done in][to obtain the depicted cell type labels]{luo2017single},
+    center: using the shrunken mean of residuals to quantify mCH in mCH-VMRs,
+    bottom: shrunken mean of residuals in CpG-VMRs.
+    }}
     \label{fig:sweep}
@@
 	\begin{center}
-		\includegraphics[width=.8\textwidth]{figures/SFig_benchmark_resources.png}
+		\includegraphics[width=.85\textwidth]{figures/SFig_benchmark_resources.png}
 	\end{center}
 	\captionof{figure}{\small \new{\textbf{VMRs are as informative as regulatory regions, but more resource-efficient.}\\
-	\textbf{(A-B)} Mean neighbor score obtained when quantifying CpG methylation at VMRs, all ENCODE regulatory regions, or a subset of regulatory regions.
+	\textbf{(A-B)} Mean neighbor score obtained when quantifying CpG methylation at VMRs, all ENCODE regulatory regions, or a subset of regulatory regions for the data sets of \citep{kremer_scnmt} and \citet{luo2017single}.
 	The subset of high-quality (HQ) regulatory regions was selected in such a way that it matches the number of VMRs and contains those regulatory regions with the highest coverage.

### 2023-12-19 response.md
Lukas's draft, copied over from scbs-figures repo

# scbs benchmarks

correlation_fig = Figure 3: Correlation of DNA methylation and gene expression.

benchmark_fig = Fig 4 and Figs. S1 - S2

dmr_fig = Figure 6: Detection of differentially methylated regions between oligodendrocytes and neural stem cells.

## Revisions to do list:
- [ ] compare scbs with other software:
  - [ ] LIGER
  - [ ] scAI
  - [ ] EpiScanpy
  - [x] MOFA+
  - [ ] DeepCpG
  - [ ] epiclonal
  - [ ] MELISSA
  - [ ] scMET
- [ ] benchmark scbs on other data sets
  - [ ] Liu: https://doi.org/10.1038/s41586-020-03182-8 *[currently running scbs prepare]*
  - [ ] Chatterton: https://doi.org/10.1186/s13578-022-00938-9 (only raw data available)
  - [x] scNMT gastrulation data set?
- [x] demonstrate DMR / VMR detection in conditions such as age, cancer
- [x] use regulatory elements as features
  - [x] TSS ±2 kb
  - [x] Encode features as done in episcanpy
- [x] try on CH-methylation data (Luo et al) *ongoing*
- [x] disentangle the effect of iterative PCA from the effect of the use of mean shrunken residuals
  - [x] do a VMR PCA with imputed zeroes instead of iterative PCA
- [x] try different slding window step sizes and show that it doesn't matter so much

minor stuff
- [ ] determine cell number and seq coverage that is required for a sensible analysis
- [x] explore of smoothing in fixed intervals of a certain CpG number makes sense (as opposed to 2kb windows)
- [x] during variable window detection, two ends of the window would cancel each other out during variance calculation. Explore whether this matters or happens at all.

## Reviewers' Comments:

### Reviewer #1:
> Remarks to the Author:
> Kremer and colleagues have proposed an innovative dimensionality
> reduction technique, specifically tailored for the analysis of
> single-cell bisulfite sequencing data. This novel approach is predicated
> on the understanding that fixed-size window tiling can result in the
> dilution of cell type-specific signals. As an alternative, the authors
> have suggested a process whereby variably methylated regions (VMRs) are
> identified first, followed by the execution of dimensionality reduction
> on these VMRs. The study compellingly demonstrates that this novel
> approach can yield improved discrimination of cell types.

> Although the limitations of a fixed size window average have long been
> suspected, no previous studies have explored this issue with the depth
> and rigor seen in the current work. The authors compellingly
> demonstrated the superiority of the VMR-based method over the
> conventional 100kb bin average.

> However, a few points for consideration arise:

> There are several existing methods that handle single-cell methylome
> dimensionality reduction and imputation, such as LIGER, scAI, EpiScanpy
> MOFA/MOFA+, which are general-purpose tools that support DNA methylation
> analysis, as well as others like DeepCpG, epiclomal, MELISSA and scMET
> that are specifically focused on DNA methylation (see PMID 35718270 for
> a recent review). The present manuscript, however, only appears to
> compare the new method with the simplest fixed-size window approach,
> without discussing these other, more recently developed methods.

- we expanded the discussion of the paper to discuss these tools and how their functionality compares to scbs
- e.g. none of them is designed find VMRs or DMRs
- furthermore, we demonstrate that MOFA+ is fully compatible with scbs, i.e. it can be used for dimensionality reduction of methylation matrices (%meth or residuals) and, when using VMRs, performs similar to iterative PCA while using less RAM at the cost of a longer runtime.

> It's not entirely clear how this method would be applicable to datasets
> where the primary source of biological variation isn't cell type but
> rather other biological factors such as normal vs. cancer, age, or
> different cell cycle stages. These factors could potentially lead to
> VMRs being present at various genomic scales. Are parameter adjustments
> necessary under these conditions?

- Lukas: I'm re-visiting the Argelaguet scNMT-seq gastrulation data set for this question. We can use it as another benchmarking data set to distinguish cell types, but I will also check if we can distinguish embryonic developmental stages and not just cell type.

> While the direct detection of VMRs from data is certainly a desirable
> feature, it's worth noting that this approach could be computationally
> demanding. A comparison with a less complex alternative that aggregates
> methylation using known biological annotations of regulatory elements
> (for example, from ENCODE, as implemented in EpiScanpy) could provide
> additional insight into the relative advantages and disadvantages of the
> proposed method.

- we added the latest annotation of ENCODE candidate cis-Regulatory Elements (cCREs) in our benchmarks
- these regions are indeed suitable to distinguish cell types, as they outperformed both 100kb tiles and promoter regions in out benchmarks and performed similarly to VMRs  
![image](https://github.com/LKremer/scbs-figures/assets/15158974/f43be414-65f3-46ef-be93-bf0a5a83471d)  
![image](https://github.com/LKremer/scbs-figures/assets/15158974/1759a17e-2b90-484b-9d19-3b21069104cd)  
- however, due to the large number of annotated cCREs (~340k in mouse) the resulting methylation matrices occupied lots of RAM even in data sets of modest size and took longer to process than the ~60k detected VMRs  
![image](https://github.com/LKremer/scbs-figures/assets/15158974/70c678a3-b2fd-4d38-b05c-57bbd2011f03)  
![image](https://github.com/LKremer/scbs-figures/assets/15158974/8caaa5b2-5eaa-459d-a380-501f1b08cb99)  
- of course, it might not be necessary to use all cCREs to distinguish cell types. In fact, many dimensionality reduction tools such as MOFA+ recommend strict feature selection, i.e. the exclusion of genomic features with low seuqencing coverage. Thus, we obtained the ~60k cCREs with the greatest sequencing coverage and repeated the benchmarks on this reduced data set. As expected, this reduced runtime and RAM usage, but cell type distinction was compromised compared to the full set of cCREs or VMRs.  
![image](https://github.com/LKremer/scbs-figures/assets/15158974/25a82dd2-d579-41fa-8b42-39ca038c148d)
![image](https://github.com/LKremer/scbs-figures/assets/15158974/9ea24835-689a-4cd5-977a-5908bc7df535)  
- thus, VMR detection might more computationally efficient, even when accounting for the additional VMR detection step
- its also kind of cool that our ~60k VMRs perform equally well to the ~340k cCREs which were identified by ENCODE, hundreds of sequencing data set, while we can just find equally informative regions in 1500 methylomes
- furthermore, VMR detection offers a data-driven approach for sc-methylome analysis that does not require any prior knowledge on the genomic regions that might be informative. While cCREs seem to be suitable to distinguish cell types, not all biological factors of interest might be detectible at cCREs. Furthermore, cCREs are only available for mouse and human.

> The method's effectiveness when applied to non-CG methylation should be
> explored, given that this form of methylation is recognized for its
> greater discriminatory power among neurons.

> Philosophically, how does the method address the balance of information
> and sparsity? Smaller VMRs could indeed provide higher discriminatory
> power, but they may also be more affected by data sparsity, making
> imputation necessary for cross-cell comparison.

> It seems the method still relies on choosing a small window size and a
> step size for VMR detection. How would this parameter be chosen and how
> does this affect the cell clustering?

- To address this valid question, we performed an additional parameter sweep of the step size parameter. This benchmark revealed that step sizes between 5bp and 200bp perform very similarly, while performance dropped when using a step size of 500bp or 1000bp.  
![image](https://github.com/LKremer/scbs-figures/assets/15158974/e7e25ab3-ad31-4e21-8bd9-02b74be4ff15)  
- As a greater step size leads to faster runtime of the VMR detection algorithm, we thus increased the default step size from 10bp to 100bp. We thank the reviewer for their valuable suggestion.

> The methodology was benchmarked against the authors' mouse brain data.
> It would be valuable to extend this comparison to other datasets, such
> as those produced by the research groups led by Joe Ecker and Wolf Reik.

- it's challenging to find a data set with ground truth cell type labels, i.e. cell type labels derived not from methylation data but from transcriptomic data, so currently we are a bit limited in our possibilities
- we include the gastrulation scNMT data set from Wolf Reiks lab

> In reference to Figure 4B and C, the utilization of neighbor scores to
> evaluate separation seems ambiguous. Lower neighbor scores can still
> result in clear separability, making the choice of these scores for
> analysis somewhat counterintuitive. Furthermore, the choice to perform
> this neighbor score analysis on PC space rather than the UMAP as
> depicted in Figures B and C could use further clarification.

- how can low neighbor scores result in clear separability?
- UMAP is just for visualization, and of course reducing the whole complexity of the data set to 2D loses lots of detail. So its better to look at the underlying PC space.

> Has the team attempted to use tSNE with varying perplexity parameters?
> This could be another beneficial comparison to further evaluate the
> effectiveness of their proposed method.

- why though

### Reviewer #2:
> Remarks to the Author:
> Kremer et al have presented a paper describing a novel approach to the
> quanitation, normalisation and analysis of single cell bisulphite data
> which aims to address issues of technical bias and signal dilution in
> some of the existing approaches. They provide a description of their
> method, a python software package to implment the processing steps
> (supplemented by example R code for some parts of the described
> analysis), and an example analysis of data from both their group and and
> an external dataset.

> The paper is well written, with a clear description of the problems with
> current approaches, and justifications for their new methods. The method
> itself is clearly and concisely described. I was able to install their
> software from the PyPi repository and could follow through all the steps
> in their analysis using the example code on their site.

> In general this method is an interesting and well thought out approach
> which offers improvements over existing methods, however there are some
> places where additional clarification or illustration would be useful.

> The initial step in the analysis turns the measured methylation values
> into residuals to a globally calculated methylation value across all
> cells. The initial description of this in the paper talks about
> subtracting the mean methylation for each position, but if I'm reading
> it correctly it's actually the smoothed running value which is used.

> [1] It would be useful to have some comment on the density of data which
> would be necessary for this approach. The datasets used by the authors
> feature thousands of cells, but many scBS experiments are much smaller
> than that, and in those a large proportion of all CpG positions will be
> measured in only 1 or 2 cells, giving little opportunity for calculating
> sensible global values against which to normalise.

To explore how our methods fare on smaller data sets, we performed our benchmarks on sub-sampled data sets, the smallest of which comprise only 100 cells.
Importantly, these sub-samples were processed completely independently, which means that the smoothed averages against which we later normalize were computed separately for data sets of various cell numbers.
These benchmarks show that cell types are indeed more difficult to assign for such small data sets, as indicated by the lower neighbor scores.
However, this is also true when using the average methylation % instead of the shrunken mean of the residuals as a measure of DNA methylation, suggesting that our method does not worsen performance, and in contrast it even increased performance when using 100kb tiles or promoter regions as features.

> [2] In a few places in the paper (kernel smoothing and detection of
> variable and differentially methlylated regions) the methods use small,
> fixed size windows, as the basis for analysis - often 1000bp. This size
> of window can be problematic in BS-Seq data given the uneven
> distribution of CpGs across the genome, and the potential for this to
> introduce bias into the results. Would smoothing in windows of fixed
> numbers of CpGs make more sense?

This is an excellent suggestion.
We imagine that using a fixed number of CpG sites instead of a fixed number of base pairs might indeed be more sensible, as this approach would prevent the evaluation of near-empty genomic windows, i.e. windows in genomic regions with little CpG density that might only contain one or two CpG sites.
Furthermore, the proposed approach would create smaller windows in CpG-dense areas, which would lead to a more fine-grained evaluation of CpG islands.
To put this suggestion to the test, we created a fork of the scbs package at https://github.com/LKremer/scbs/tree/adaptive_bandwidth where the sliding window used for VMR detection always comprises a fixed number of CpG sites.
Testing this approach on our own multi-omics data set, using 9 CpG sites per window, produced very promising results:
We found that VMRs detected with this approach led to a slightly better neighbor score (i.e. better cell type separation), as well as decreased runtime of the VMR detection procedure.
For now, we consider this implementation experimental as its currently lacking unittests as well as extensive benchmarks on multiple data sets.
Furthermore, there are some implementation details that we have not resolved yet:
Especially in smaller data sets with low coverage and/or few cells, there might be substantial gaps with zero coverage between individual CpG sites.
Does it nonetheless make sense to include adjacent CpG sites into one window, or should there be a maximum distance at which windows are broken into smaller pieces?
For now, we thus decided to stick to the "classic" sliding window approach, but if future benchmarks confirm the superiority of this approach we will update the scbs package accordingly.

> [3] When selecting the variably methylated regions the method takes the
> mean of the residuals in a given region. Does this mean that if adjacent
> regions changed in opposite directions that this would be averaged away
> and the region discarded as uninteresting?

VMRs coordinates are determined by merging overlapping genomic windows above a variance threshold (2% by default).
Although the variance is then re-calculated for the entire VMR, this value is not used for a secondary filtering step but only reported in the output file of `scbs scan`.
The rationale behind this is that we also want to report regions where the variance of individual windows briefly drops below the variance threshold - such as the one depicted in Fig. 2:  
![Fig_sliding](https://github.com/LKremer/scbs-figures/assets/15158974/e8f68d28-806e-4a4a-be6f-5a4d05fde243)  
In this example, two of the 2kb windows in the center of the VMR are just barely below the variance threshold, but we nonetheless report the entire region because the 2kb windows above the threshold are 2kb wide and are thus able to "bridge the gap".
Similarly, in the scenario described by the reviewer, a VMR with two opposite trends would not be discarded as long as the individual 2kb windows are above the 2% variance threshold.
However, our data indicate that such regions are rather rare.
A closer look at the variances of all 2kb windows of the largest chromosome, as well as the variances of reported VMRs, shows that the majority of VMRs are above the 2% variance threshold (red line).  
![image](https://github.com/LKremer/scbs-figures/assets/15158974/9bf7845a-e9cd-4887-8b32-3176ee8baeaa)  
As expected due to the two scenarios described above, the region-wide variance of some VMRs is slightly below the 2% window variance threshold.
However, these variances are still well above genome-wide background levels (blue line: average of all genomic windows), suggesting that cancellation of two opposing signals as described by the reviewer is not common.

> The authors illustrate the effectiveness of their method by showing a
> PCA of a set of cells separated by their VMR process, compared to a more
> conventional separation by simple methylation calculation. In the VMR
> PCA there are two potentially influential steps - the conversion of
> methylation to residuals, and the imputation of missing values by
> iterative PCA. It would be useful to know which of these was having the
> greatest influence, since it's possible that it is the imputation which
> is causing the clearer separation through the reinforcement of the
> initial signal, rather than that the initial values are clearer.

It is true that our previous benchmarks did not allow us to distinguish the effects of the choice of dimensionality reduction technique (e.g. iterative PCA) from the choice of DNA methylation measure (average methylation % or shrunken means of the residuals).
To disentangle the influence of these analysis options, we refined our benchmarks by testing all possible combinations of the following choices:
- genomic features at which DNA methylation is quantified (100kb tiles, VMRs, promoters or ENCODE candidate cis-regulatory elements (cCREs))
- the measure used to quantify DNA methylation at these features (average methylation % or the shrunken mean of the residuals)
- dimensionality reduction methods (iterative PCA as proposed by us, lightly-imputed PCA on high-coverage features as proposed by Luo et al., or PCA with the missing values set to the column-mean (i.e. zero when using residuals) as suggested by this reviewer in comment [4])  
![image](https://github.com/LKremer/scbs-figures/assets/15158974/f43be414-65f3-46ef-be93-bf0a5a83471d)  
![image](https://github.com/LKremer/scbs-figures/assets/15158974/1759a17e-2b90-484b-9d19-3b21069104cd)  
Based on these new results, we can now draw more detailed conclusions on the impact various choices of methods:
1. Using the shrunken means of the residuals over average methylation percentages provides the greatest benefit when quantifying promoters or 100kb tiles.
   The likely reason for this is that these regions are more heterogeneously methylated than VMRs or cCREs:
   100kb tiles are very large and thus each tile will comprise stretches of DNA with high and low methylation.
   Similarly, many promoter regions (here: TSS±2kb) will comprise a lowly-methylated center region near the TSS, as well as more highly methylated promoter flanks.
   As the shrunken mean of the residuals is designed to account for this heterogeneity within a region (Fig. 1), it makes sense that we see the greatest benefit of their use when quantifying promoters or large tiles.
2. The use of iterative PCA over other PCA-variations constantly improved the ability to distinguish cell types.
   The main reason for this performance gain is that iterative PCA alleviates differences between low-quality and high-quality cells, as is clearly visible in these plots:
   [add plot here]
3. Overall, it is clearly visible that the most important choice is the set of genomic features to be quantified. Our results indicate that both VMRs and ENCODE cCREs are suitable for this task and offer good ability to distinguish cell types, although using cCREs requires more RAM and runtime.

> [4] What would a PCA of the VMR values with missing values still set to
> 0 look like? This would be a more direct comparison to the global
> calculation and would demonstrate which step is more important in the
> improved separation they show.

This comment was addressed and illustrated in our response to the previous comment, where we show benchmarking results and exemplary plots of zero-imputed PCAs.

> [5] In the DMR detection they show a comparison of 130 cells and 58
> cells, but am I right in thinking that these cells were split off only
> after calculating the normalised VMR values from a much larger
> population of cells? Does the approach still work as effectively if
> starting from the 188 cells in the DMR analysis, or do you require the
> context of the larger dataset to get accurate VMR values to put into the
> DMR detection?

Yes, the 130 vs 58 cells used for DMR detection were split off from a larger data set (comprising 540 cells) and compared against each other, as is commonly done in scRNA-seq where users might select two cell clusters for differential gene expression testing.
To clarify, the DMR detection procedure does not use any VMR coordinates that were detected on the full data set, as DMR coordinates are newly discovered on the selected sub-data set.
However, as the reviewer rightly points out, the DMR detection uses the shrunken mean of the residuals as a measure of methylation, which in turn uses the smoothed average genomic methylation values for normalization.
These smoothed averages were indeed calculated on the full data set and not on the reduced set of 130 + 58 cells.
To assess whether this would affect DMR detection, we thus repeated the DMR analysis depicted in Fig. 6 on the 130 + 58 cells only, using smoothed genomic averages calculated on these cells only.
The obtained oligodendrocyte- and NSC-DMR sets are largely consistent with those obtained using smoothed averages obtained from the full data set.
For comparison, here is Fig. 6 from the manuscript (full data set) and Fig. 6 based on the reduced data set only:

All 540 cells (current manuscript version of the figure):  
![Fig_DMRs](https://github.com/LKremer/scbs-figures/assets/15158974/51287393-fba8-4e2f-a82f-6359a57ed985)  
130 + 58 cells only:  
![Fig_DMR_showcase](https://github.com/LKremer/scbs-figures/assets/15158974/18ac2686-be83-47df-ba9d-579c1670867b)  
While the DMR lists obtained using both approaches are not 100% identical, the results are qualitatively similar and yield similar largely identical top 5 enriched GO terms.

> [6] In the DMR calculation the false discovery rate is estimated by
> shuffling cell labels and repeating the analysis - which seems to be a
> reasonable idea for this type of data. However, the FDR estimation is
> done only after the initially calculated windows are filtered for the
> top 2% then merged if they are adjacent. The values tested are those
> from the recalculated t-statistic after merging. I would assume that the
> approach shown heavily favours regions which are much larger than the
> initial window size, and that the major contribution to significance is
> adjacency in the initially filtered top 2% of regions. Does this method
> work if the FDR values are calculated from the original window data,
> before merging adjacent windows? If not, then what size of DMR is
> realistically detectable with this method?

L: Ok, here we have to discuss a bit and explain that we want to be able to detect DMRs of flexible size.
Calculating FDR for individual windows would work, but once we merge them the FDR of the entire region is unclear.  
Maybe its also worth demonstrating that DMR detection works even if we don't merge windows, so its not the window merging thats driving significance.
Without merging, we get some nice DMRs but all of them have the same width, of course, which doesn't make so much sense:
![image](https://github.com/LKremer/scbs-figures/assets/15158974/fd2301ca-00d5-4bff-a3fb-a70099d367bd)

### Reviewer #3:
> Remarks to the Author:
> Kremer et al describe ‘scbs’ – a computational toolset for the analysis
> of single-cell methylation data. In the field there is certainly a need
> for a suite of tools to handle single-cell methylation data, and the
> manuscript details several compelling strategies and frameworks for
> doing so. The most significant aspect is the identification of variable
> regions that are then used for subsequent dimensionality reduction,
> clustering and visualization; followed by tools for DMR calling. The
> authors apply ‘scbs’ to their own, published, scNMT-seq dataset on brain
> as well as some applications to a dataset produced by the Ecker Lab,
> also brain. While there are positives to the work, there are key
> limitations in the assessment of the tool that need to be performed in
> order to properly evaluate the tool and its broader utility. Lastly, the
> name itself is a term that is already used to define an experimental
> approach” “single-cell bisulfite sequencing” – whereas this reports on a
> computational tool for that type of data (or enzymatic-converted data,
> which is not bisulfite). I strongly suggest changing the name of the
> tool to something informative – e.g. SCMtools for “single-cell methylome
> tools” – anything that indicates that it is a computational tool and
> used for analysis of single-cell methylome data.

> Major Comments:

> The authors perform almost all of their evaluation solely on their own
> dataset with very little analysis of the Luo et al 2017 dataset.
> Furthermore, their only comparison of methodology is by using 100 kbp
> tiling windows for CpG methylation levels, which does not seem all that
> appropriate. Using relevant windows – e.g. all annotated promoters and
> enhancers like geneHancer or something like that is a much more
> reasonable strategy. Even just all promoter regions such as TSS +/- 2
> kbp or something.

In response to this comment, we greatly expanded the scope of our benchmarks by including additional data sets as well as other sets of genomic regions.
Specifically, we added the following data sets: ...
Furthermore, as kindly suggested by this reviewer, we included the set of promoter regions (TSS ± 2 kbp), as well as the latest release of human and mouse candidate cis-regulatory elements (cCREs) published by the ENCODE consortium.

Add a quick summary of our updated conclusions here:
- if you're working on human/mouse, cCREs also work nicely, but lead to hefty resource and time requirements on intermediate to big data sets.
- our proposed approach (VMR detection -> residuals) robustly worked across all tested data sets

> They do not compare cluster resolution with CH methylation windows which
> is generally used for neuron subtype clustering (as often CpG
> methylation will be the same between two subtypes of the same class, eg
> excitatory, but the CH levels will be different). Does the tool work
> with CH methylation? How does the variable window approach for CpG
> methylation look compared to 100 kbp tiling CH methylation (where 100
> kbp tiles make sense due to the structure of the mark in neurons).

Here we have to kindly explain the following:
- to assess performance of a given combination of analysis methods, we assess its ability to distinguish cell types (neighbor score)
- for this, we need ground-truth cell type labels
- in our own dataset, we also have the transcriptome of each cell, which we can use to infer ground-truth cell type labels
- most published data sets only have methylome information though, so we cant use them for benchmarking
- as the cell type labels published by Luo et al. are already based on CH methylation, we cannot use these labels as a ground truth for benchmarking the performance of our tool on CH-data
- nonetheless, to demonstrate that scbs can be used to quantify and explore CH-data, we performed a representative CH-methylation analysis on Luo's data (add a UMAP based on CH-data here)

> These analyses should be carried out on both their own dataset, the Luo
> et al dataset, but also on datasets from other tissue types. For example:

> Argelaguet R, Clark SJ, Mohammed H, Stapel LC, Krueger C, Kapourani CA,
> Imaz-Rosshandler I, Lohoff T, Xiang Y, Hanna CW, Smallwood S,
> Ibarra-Soria X, Buettner F, Sanguinetti G, Xie W, Krueger F, Göttgens B,
> Rugg-Gunn PJ, Kelsey G, Dean W, Nichols J, Stegle O, Marioni JC, Reik W.
> Multi-omics profiling of mouse gastrulation at single-cell resolution.
> Nature. 2019 Dec;576(7787):487-491. doi: 10.1038/s41586-019-1825-8. Epub
> 2019 Dec 11.

> Chatterton, Z., Lamichhane, P., Ahmadi Rastegar, D. et al. Single-cell
> DNA methylation sequencing by combinatorial indexing and enzymatic DNA
> methylation conversion. Cell Biosci 13, 2 (2023).
> https://doi.org/10.1186/s13578-022-00938-9
<https://doi.org/10.1186/s13578-022-00938-9> (Enzymatic conversion workflow)

We were unable to retrieve the cell type annotation and methylation report files for this particular data set, but as mentioned in a previous reply, we included X additional data sets from various tissues in our benchmarks.

> Furthermore, the tool is demonstrated on relatively small datasets on
> the order of ~3000 cells; however, single-cell methylation datasets are
> increasing in size – the Luo et al dataset used is quite old at this
> point, with much larger datasets produced by the Ecker and Luo labs,
> with hundreds-of-thousands of cells. It is likely that such datasets
> will become more common and any tool that will reach wide adoption will
> have to be able to handle large datasets. At least one of these should
> be used as an example with compute times reported – eg:

> Liu, H., Zhou, J., Tian, W. et al. DNA methylation atlas of the mouse
> brain at single-cell resolution. Nature 598, 120–128 (2021).
> https://doi.org/10.1038/s41586-020-03182-8
<https://doi.org/10.1038/s41586-020-03182-8> (~100k cells)
> Or the newer dataset that is ~300-400k cells; however, the 100k cell
> dataset should be sufficient for demonstration purposes. Again comparing
> CH methylation 100 kbp windows (standard for neurons) vs variable
> windows vs promoters & enhancers.

L: I think they are underestimating the size of these data sets, and also how uncommon such huge data sets currently are.
Even the smaller 100k cell data set was stitched together from almost 40 separate experiments, and the analysis likely took months to complete...
Quantifying enhancers in this data set is not feasible for sure, the methylation matrix would be 110k cells x 340k regions.
We can try CpG (not CH) in genomic tiles and/or VMRs just as a proof of concept.
To show that we can work with CH meth, we should use the Luo data set.

### 2023-12-19 paper.tex
update date
label the stepsize benchmark panel 'C'

@@
 
-May 2023
+December 2023, version 3
 \vspace{4.5ex}
@@
 \new{To assess whether our proposed methods are also suitable for the analysis of CH-methylation (mCH), we furthermore performed VMR detection with default options on the mCH-portion of this data set (Fig.~\ref{fig:sweep}C).
-We obtained results that are qualitatively similar to those reported in \citep{luo2017single}, who quantified mCH in 100~kb tiles, suggesting that our methods also function on mCH data.}
+We obtained results that are qualitatively similar to those reported in \citet{luo2017single}, who quantified mCH in 100~kb tiles, suggesting that our methods also function on mCH data.}
 
-In the more challenging data set of \citet{kremer_scnmt} (Fig.~\ref{fig:sweep}A), cell types were less cleanly separated when very large bandwidths, very strict variance thresholds, \new{or a very large step size was} selected.
+In the more challenging data set of \citet{kremer_scnmt}, cell types were less cleanly separated when very large bandwidths, very strict variance thresholds, \new{or a very large step size was} selected (Fig.~\ref{fig:sweep}A,C).
 However, very small bandwidths or very lenient thresholds resulted in a much higher number of VMRs and thus long computing times.
@@
     \textbf{(A-B)} Mean neighbor scores obtained after analyzing single-cell methylomes from \citet{kremer_scnmt} (\textbf{A}) or \citet{luo2017single} (\textbf{B}) with our proposed methods.
-    VMRs were detected with \texttt{scbs} \texttt{scan} using various sliding window bandwidths, variance thresholds, or sliding window step sizes (\textbf{A}, bottom right), and on sub-samples of the full data sets.
-    \textbf{(C)} UMAPs obtained when analysing data from \citep{luo2017single} with three different strategies for producing a methylation matrix:
+    VMRs were detected with \texttt{scbs} \texttt{scan} using various sliding window bandwidths, variance thresholds, and on sub-samples of the full data sets.
+    \textbf{(C)} Mean neighbor scores obtained after analyzing the \citet{kremer_scnmt} data set with various sliding window step sizes.
+    \textbf{(D)} UMAPs obtained when analysing data from \citep{luo2017single} with three different strategies for producing a methylation matrix:
     top: averaging CH-methylation (mCH) in 100~kb genomic tiles \citep[as done in][to obtain the depicted cell type labels]{luo2017single},

### 2023-12-22 paper.tex
add todo
propose 1k cells as target

@@
 
-Therefore, we propose to first obtain, for each CpG position, an average of the methylation across all cells, and then quantify each cell's deviation from this average.
+Therefore, we propose to first obtain, for each CpG position, a \new{smoothed} average of the methylation across all cells, and then quantify each cell's deviation from this average.
 In Figure \ref{fig:smoothres}B, the curved line depicts such an average over all cells, and the red vertical lines show an individual cell's deviation from the ensemble average.
@@
 Filtering regulatory regions by sequencing coverage reduced resource requirements, but at the cost of a reduced ability to distinguish cell types (Fig.~\ref{fig:todo}).
+\todo{simon suggested to stress here that our VMRs are more information-rich than encode regions, because you need fewer VMRs for the same results.
+We could also mention that encode regions perform poorly in combination with MOFA+ for some reason}
 }
@@
 We furthermore included three additional approaches to perform dimensionality reduction in our benchmarks, including PCA with two different pre-processing strategies, as well as MOFA+, a dimensionality reduction technique designed for multi-modal single-cell data that can also process methylation data \citep{argelaguet2020mofa}.
+}
+
+\new{
 These extensive benchmarks confirmed that our proposed combination of methods, i.e.~using the shrunken means of residuals of VMRs for dimensionality reduction, is able to distinguish diverse cellular properties such as cell type, \todo{colorectal cancer sampling region(?)}, embryonic stage, and germ layer (Fig.~\ref{fig:resource_usage}).
+Our benchmarks furthermore reveal that groups of cells are more readily separated in large data sets.
+As the neighbor score started to plateau at approximately 1000 cells, we propose that this represents a reasonable target cell number for future scBS studies.
 }

### 2024-01-10 paper.tex
add 100k cell umap

@@
 
-Lastly, we assessed whether our proposed workflow requires fine-tuning of parameters (Fig.~\ref{fig:sweep}).
+\new{Next}, we assessed whether our proposed workflow requires fine-tuning of parameters (Fig.~\ref{fig:sweep}).
 To this end, we re-analyzed \new{two} data sets, as well as sub-samples of the data with different VMR detection parameters namely the width of the sliding window in bp, (set with option \texttt{--bandwidth} in our \texttt{scbs} software, default 2000), the variance threshold above which windows are merged to VMRs (\texttt{--var-threshold}, default 0.02) \new{and the step size of the sliding window (\texttt{--stepsize}, default 100bp).}
@@
 For the \new{CpG-methylation} data of \citet{luo2017single}, results are nearly independent of the parameters (Fig.~\ref{fig:sweep}B).
-\new{To assess whether our proposed methods are also suitable for the analysis of CH-methylation (mCH), we furthermore performed VMR detection with default options on the mCH-portion of this data set (Fig.~\ref{fig:sweep}C).
-We obtained results that are qualitatively similar to those reported in \citet{luo2017single}, who quantified mCH in 100~kb tiles, suggesting that our methods also function on mCH data.}
-
 In the more challenging data set of \citet{kremer_scnmt}, cell types were less cleanly separated when very large bandwidths, very strict variance thresholds, \new{or a very large step size was} selected (Fig.~\ref{fig:sweep}A,C).
@@
 Overall, our default parameter combination provided good results and fast compute times in both data sets.
+
+\new{
+Lastly, we asked whether our methods are also suitable for the analysis of DNA methylation outside the default CpG context.
+To this end, we revisited the \citet{luo2017single} data set, but this time only considered CH-methylation (mCH).
+VMR detection with default options produced results that were qualitatively similar to those reported in \citet{luo2017single}, suggesting that our methods are also suitable for this data type (Fig.~\ref{fig:umaps}A).
+As single-cell methylome data sets are expected to rapidly grow in size in the coming years, we furthermore performed a stress test on a large data set comprising 100\,350 cells \citep{liu2021dna} (Fig.~\ref{fig:umaps}B).
+}
 
@@
     \begin{center}
-        \includegraphics[width=.85\textwidth]{figures/SFig_parameter_sweep.png}
+        \includegraphics[width=.55\textwidth]{figures/SFig_parameter_sweep.png}
     \end{center}
-    \captionof{figure}{\small \textbf{Effect of VMR detection parameters \new{and methylation context} on separation of cell types.}\\
+    \captionof{figure}{\small \textbf{Effect of VMR detection parameters on separation of cell types.}\\
 	\new{
@@
     \textbf{(C)} Mean neighbor scores obtained after analyzing the \citet{kremer_scnmt} data set with various sliding window step sizes.
-    \textbf{(D)} UMAPs obtained when analysing data from \citep{luo2017single} with three different strategies for producing a methylation matrix:
-    top: averaging CH-methylation (mCH) in 100~kb genomic tiles \citep[as done in][to obtain the depicted cell type labels]{luo2017single},
-    center: using the shrunken mean of residuals to quantify mCH in mCH-VMRs,
-    bottom: shrunken mean of residuals in CpG-VMRs.
     }}
@@
 
+\noindent%
+\begin{minipage}{\linewidth}
+	\begin{center}
+		\includegraphics[width=.85\textwidth]{figures/SFig_100k_and_CH.png}
+	\end{center}
+	\captionof{figure}{\small \new{\textbf{\texttt{scbs} performs well on CH-methylation data and on extremely large data sets.}\\
+    \textbf{(A)} UMAPs obtained when analysing data from \citet{luo2017single} with three different strategies for producing a methylation matrix:
+	top: averaging CH-methylation (mCH) in 100~kb genomic tiles \citep[as done in][to obtain the depicted cell type labels]{luo2017single};
+	center: using the shrunken mean of residuals to quantify mCH in mCH-VMRs;
+	bottom: shrunken mean of residuals to quantify CpG methylation in CpG-VMRs.
+	\textbf{(B)} UMAP obtained when analysing 100\,350 neural single-cell methylomes \citep{liu2021dna} with the default \texttt{scbs} workflow.
+	The depicted cell labels were reported by \citet{liu2021dna}.
+	\todo{Analysing this large data set took approximately XXX hours and required approximately XXX GB of RAM. (Sveta, pls give some rough estimate here)}
+	}}
+	\label{fig:umaps}
+\end{minipage}
+
 \end{document}

### 2024-01-10 paper.tex
update methods section

@@
 
-The starting point in most scRNA-seq analyses is a matrix of UMI counts (i.e. of counts of distinct RNA molecules), one row for each cell and one column for each gene.
+The starting point in most scRNA-seq analyses is a matrix of UMI counts (i.e.\ of counts of distinct RNA molecules), one row for each cell and one column for each gene.
 A first goal is usually to assign cell types or states to cells.
@@
 
-To demonstrate the value of our proposed improvements, we benchmarked various combinations of analysis methods on \new{\todo{N} diverse single-cell methylome data sets.}
+To demonstrate the value of our proposed improvements, we benchmarked various combinations of analysis methods on \new{five diverse single-cell methylome data sets.}
 
@@
 This is likely due to the large number of annotated regulatory features (339\,815 in mouse, 1\,063\,878 in human) as opposed to the 63\,421 detected VMRs.
-Filtering regulatory regions by sequencing coverage reduced resource requirements, but at the cost of a reduced ability to distinguish cell types (Fig.~\ref{fig:todo}).
+Filtering regulatory regions by sequencing coverage reduced resource requirements, but at the cost of a reduced ability to distinguish cell types (Fig.~\ref{fig:resource_usage}).
 \todo{simon suggested to stress here that our VMRs are more information-rich than encode regions, because you need fewer VMRs for the same results.
@@
 \new{
-We repeated this benchmark on an additional \todo{3(?)} published single-cell methylome data sets.
+We repeated this benchmark on an additional three published single-cell methylome data sets (Fig.~\ref{fig:score_luo}).
 These include neuronal sub-types of the murine cortex \citep[][using cell type labels derived from CH-methylation in genomic tiles, instead of CpG-methylation, as ground truth]{luo2017single}, cells isolated from mouse embryos during the onset of gastrulation \citep[][using RNA-derived cell clusters, or alternatively embryonic stage as ground truth]{argelaguet2019gastru} and human colorectal cancer cells \citep[][using \todo{tumor region(?)} as ground truth]{bian2018single}.
@@
 \new{
-These extensive benchmarks confirmed that our proposed combination of methods, i.e.~using the shrunken means of residuals of VMRs for dimensionality reduction, is able to distinguish diverse cellular properties such as cell type, \todo{colorectal cancer sampling region(?)}, embryonic stage, and germ layer (Fig.~\ref{fig:resource_usage}).
+These extensive benchmarks confirmed that our proposed combination of methods, i.e.\ using the shrunken means of residuals of VMRs for dimensionality reduction, is able to distinguish diverse cellular properties such as cell type, \todo{colorectal cancer sampling region(?)}, embryonic stage, and germ layer (Fig.~\ref{fig:resource_usage}).
 Our benchmarks furthermore reveal that groups of cells are more readily separated in large data sets.
@@
 To analyze scBS data from \citet{kremer_scnmt}, single-cell CpG methylation reports from all conditions were first stored with \texttt{scbs prepare} and then smoothed with \texttt{scbs smooth} using the default bandwidth of 1000~bp.
-The mm10 mouse genome was then annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 0.02$, minimum cell requirement $=6$).
-Methylation fractions, commonly represented as percentages, and the shrunken means of residuals were quantified with \texttt{scbs matrix} for both VMRs and tiles.
-The same analysis was repeated for mouse scBS data from \citet{luo2017single}, and for smaller data sets generated by randomly sub-sampling cells separately from both data sets.
-For each data set, we thus obtained four methylation matrices:
+\new{
+Methylation was then quantified in four sets of genomic intervals:
+VMRs, obtained with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 0.02$, minimum cell requirement $=6$);
+non-overlapping tiles of 100~kb width;
+promoter regions as defined by the \textpm2~kb domain around the TSS of coding genes;
+and candidate cis-regulatory regions annotated by the ENCODE consortium \citep{encode2020expanded},
+Averaged m}ethylation fractions and the shrunken means of residuals were quantified with \texttt{scbs matrix} \new{for all four sets of regions}.
+The same analysis was repeated for \new{three additional scBS data sets \citep{luo2017single, bian2018single, argelaguet2019gastru}}, and for smaller data sets generated by randomly sub-sampling cells separately from \new{these} data sets.
+\todo{fix: For each data set, we thus obtained four methylation matrices:
 either 100~kb tiles or VMRs, quantified either by simple averaging (methylation fractions) or with the shrunken mean of the residuals.
@@
 We then imputed missing values by imputing the mean for each tile, centered the values, and performed PCA.
-In the case of VMRs, we used iterative PCA to handle missing values, as described earlier.
-For visualization, the first 15 PCs were subjected to UMAP with parameters min\_dist $=0.2$, init $=$ "spca".
+In the case of VMRs, we used iterative PCA to handle missing values, as described earlier.}
+For visualization, the first 15 PCs were subjected to UMAP with parameters min\_dist $=0.2$, init $=$ ``spca''.
 To flexibly adapt to data sets of different sizes, we set n\_neighbors  $=\frac{\sqrt{n}}{1.5}$ (rounded to the nearest integer), where $n$ is the total cell number.

### 2024-01-19 paper.tex

@@
 \new{
-Methylation was then quantified in four sets of genomic intervals:
+This data was then analysed multiple times with different combinations of analysis methods, namely four ways to divide the genome into intervals, two ways to quantify methylation in these intervals, and four approaches for dimensionality reductions.
+The following four sets of genomic intervals were used:
 VMRs, obtained with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 0.02$, minimum cell requirement $=6$);
-non-overlapping tiles of 100~kb width;
+adjacent tiles of 100~kb width;
 promoter regions as defined by the \textpm2~kb domain around the TSS of coding genes;
-and candidate cis-regulatory regions annotated by the ENCODE consortium \citep{encode2020expanded},
-Averaged m}ethylation fractions and the shrunken means of residuals were quantified with \texttt{scbs matrix} \new{for all four sets of regions}.
+and candidate cis-regulatory regions annotated by the ENCODE consortium \citep{encode2020expanded}.
+Methylation was quantified either by averaging binary methylation values, or by calculating the shrunken mean of the residuals as described earlier.
+}
+\todo{...describe the four dimensionality reduction approaches, see below...}
 The same analysis was repeated for \new{three additional scBS data sets \citep{luo2017single, bian2018single, argelaguet2019gastru}}, and for smaller data sets generated by randomly sub-sampling cells separately from \new{these} data sets.
-\todo{fix: For each data set, we thus obtained four methylation matrices:
+\todo{update this and maybe move it up?: For each data set, we thus obtained four methylation matrices:
 either 100~kb tiles or VMRs, quantified either by simple averaging (methylation fractions) or with the shrunken mean of the residuals.
@@
 We then imputed missing values by imputing the mean for each tile, centered the values, and performed PCA.
-In the case of VMRs, we used iterative PCA to handle missing values, as described earlier.}
+In the case of VMRs, we used iterative PCA to handle missing values, as described earlier.
 For visualization, the first 15 PCs were subjected to UMAP with parameters min\_dist $=0.2$, init $=$ ``spca''.
 To flexibly adapt to data sets of different sizes, we set n\_neighbors  $=\frac{\sqrt{n}}{1.5}$ (rounded to the nearest integer), where $n$ is the total cell number.
+}
 
@@
 	The depicted cell labels were reported by \citet{liu2021dna}.
-	\todo{Analysing this large data set took approximately XXX hours and required approximately XXX GB of RAM. (Sveta, pls give some rough estimate here)}
+	\todo{Analysing this large data set took approximately two days on a cluster with 256 GB RAM and and 48 CPUs.}
 	}}

### 2024-01-30 paper.tex
a few edits

@@
 
-December 2023, version 3
+version 3 /\!/ January 2024
 \vspace{4.5ex}
@@
 
-To demonstrate the value of our proposed improvements, we benchmarked various combinations of analysis methods on \new{five diverse single-cell methylome data sets.}
+To demonstrate the value of our proposed improvements, we benchmarked various combinations of analysis methods on \new{five diverse single-cell methylome data sets, starting with a data set from our own research}.
 
@@
 Specifically, we selected four different sets of genomic intervals at which CpG methylation is to be quantified:
-either VMRs detected with our approach, 100~kb genomic tiles, promoter regions (TSS\textpm2000~bp), or candidate cis-regulatory elements \citep[ENCODE cCREs,][]{encode2020expanded}.
-To quantify methylation at these features, we either averaged in these intervals, obtaining methylation percentages, or we calculated the shrunken mean of residuals.
+either VMRs detected with our approach, 100~kb genomic tiles, promoter regions (i.e., transcription start site (TSS)~\textpm~2000~bp), or candidate cis-regulatory elements \citep[ENCODE cCREs,][]{encode2020expanded}.
+To quantify methylation at these features, we either simply averaged in these intervals, obtaining methylation percentages, or we calculated the shrunken mean of residuals, as described above.
 Finally, the resulting methylation matrices were subjected to iterative PCA (see Methods) and subsequent UMAP for visualization.
@@
 To quantify this performance gain in a more rigorous manner, we used a score that quantifies whether cells were placed, in 15-dimensional PC space, in a neighborhood comprising cells of the same cell type ("neighbor score", see Fig.~\ref{fig:score}E, and Methods for details).
-We then ask how this score depends on the number of available cells.
+\new{The neighborhood relation in PCA space is relevant as it is used in many down-stream analyses, e.g., for clustering.}
+We ask how the neighbor score depends on the number of available cells.
 This is relevant as scBS protocols are costly and labor-intensive, and only few laboratories are currently able to obtain thousands of single cell methylomes.
@@
 The results confirmed that quantifying methylation at VMRs leads to a cleaner separation of cell states than using promoter regions or 100~kb tiles.
-Using the shrunken mean of residuals as a measure of DNA methylation furthermore improved results.
-This effect was most noticeable in promoters and genomic tiles, presumably because an individual promoter region or tile might span genomic regions with varying levels of DNA methylation as depicted in Fig.~\ref{fig:smoothres}B, which our method accounts for.
+Using the shrunken mean of residuals as a measure of DNA methylation improved results further.
+This effect was most noticeable when using promoters or genomic tiles as intervals, presumably because an individual promoter region or tile might span genomic regions with varying levels of DNA methylation as depicted in Fig.~\ref{fig:smoothres}B, which our method accounts for.
 Although cell cluster separation generally becomes more difficult in those, performance gains were observed also in smaller data sets.
@@
 \new{
-Overall, VMR quantification yielded results similar to those obtained when quantifying ENCODE regulatory regions.
-This demonstrates that our VMR detection approach identifies genomic regions in scBS data that are as informative as state-of-the-art epigenomic annotations.
+Overall, VMR quantification yielded results similar to those obtained when quantifying ENCODE regulatory regions, even though the number of detected VMRs (63\,421) is considerably smaller than the number of ENCODE cCREs (339\,815).
+This demonstrates that our de-novo VMR detection approach identifies genomic regions in scBS data that are as informative as state-of-the-art epigenomic annotations, as well as its ability to find a parsimonous set of relevant elements.
+We not that the overlap between the VMRs and the cCREs is quite limited \todo{Give a number} indicating that VMR detection yields information that is complementary to other epigenetic marks.
 An added benefit of VMR detection is that this approach is available even in the absence of such annotations, for instance when studying species other than human or mouse.
 Furthermore, using VMRs over regulatory regions resulted in decreased RAM requirements as well as a shorter runtime, even when accounting for the additional step of VMR detection (Fig.~\ref{fig:score}F).
-This is likely due to the large number of annotated regulatory features (339\,815 in mouse, 1\,063\,878 in human) as opposed to the 63\,421 detected VMRs.
-Filtering regulatory regions by sequencing coverage reduced resource requirements, but at the cost of a reduced ability to distinguish cell types (Fig.~\ref{fig:resource_usage}).
-\todo{simon suggested to stress here that our VMRs are more information-rich than encode regions, because you need fewer VMRs for the same results.
-We could also mention that encode regions perform poorly in combination with MOFA+ for some reason}
+\todo{maybe move panel F to the Supplement}.
 }
+%Filtering regulatory regions by sequencing coverage reduced resource requirements, but at the cost of a reduced ability to distinguish cell types (Fig.~\ref{fig:resource_usage}).
 
@@
 These extensive benchmarks confirmed that our proposed combination of methods, i.e.\ using the shrunken means of residuals of VMRs for dimensionality reduction, is able to distinguish diverse cellular properties such as cell type, \todo{colorectal cancer sampling region(?)}, embryonic stage, and germ layer (Fig.~\ref{fig:resource_usage}).
-Our benchmarks furthermore reveal that groups of cells are more readily separated in large data sets.
+Our benchmarks also quantify how groups of cells become more readily separated in larger data sets.
 As the neighbor score started to plateau at approximately 1000 cells, we propose that this represents a reasonable target cell number for future scBS studies.
+\todo{We may want to omit these last two sentences, as they might invite too much debate.}
 }
@@
 Overall, our default parameter combination provided good results and fast compute times in both data sets.
+
+\subsubsection{Further applications}
 
@@
 \section{Supplementary Figures}
+\vspace{1.5cm}
 
@@
 	iterative PCA as proposed by us, mean-imputed PCA on intervals with high sequencing coverage \citep{luo2017single}, mean-imputed PCA on all intervals, and MOFA+ \citep{argelaguet2020mofa} on all intervals.
-	For each combination of methods, we quantified the ability to separate ground-truth cell labels reported by the authors in 15-dimensional PCA space (neighbor score).
+	For each combination of methods, we quantified the ability to use distance in 15-dimensional PCA space to separate ground-truth cell labels reported by the authors (neighbor score).
 	\textbf{(A)} Separation of neural cell types/states annotated based on scRNA-seq of the same cells \citep{kremer_scnmt}.
@@
 	\label{fig:score_luo}
+
 \end{minipage}
+\vfill
+\newpage
 

### 2024-01-31 paper.tex
Discussion updated

@@
 We repeated this benchmark on an additional three published single-cell methylome data sets (Fig.~\ref{fig:score_luo}).
-These include neuronal sub-types of the murine cortex \citep[][using cell type labels derived from CH-methylation in genomic tiles, instead of CpG-methylation, as ground truth]{luo2017single}, cells isolated from mouse embryos during the onset of gastrulation \citep[][using RNA-derived cell clusters, or alternatively embryonic stage as ground truth]{argelaguet2019gastru} and human colorectal cancer cells \citep[][using \todo{tumor region(?)} as ground truth]{bian2018single}.
+These include neuronal sub-types of the murine cortex \citep[][using cell type labels derived from CH-methylation in genomic tiles, instead of CpG-methylation, as ground truth]{luo2017single}, cells isolated from mouse embryos during the onset of gastrulation \citep[][using RNA-derived cell clusters, or alternatively embryonic stage as ground truth]{argelaguet2019gastru} and human colorectal cancer cells \citep[][using sampling region as ground truth]{bian2018single}.
 Again, we subjected each data set to all possible combinations of genomic feature selection and methylation quantification.
@@
 \new{
-These extensive benchmarks confirmed that our proposed combination of methods, i.e.\ using the shrunken means of residuals of VMRs for dimensionality reduction, is able to distinguish diverse cellular properties such as cell type, \todo{colorectal cancer sampling region(?)}, embryonic stage, and germ layer (Fig.~\ref{fig:resource_usage}).
+These extensive benchmarks confirmed that our proposed combination of methods, i.e.\ using the shrunken means of residuals of VMRs for dimensionality reduction, is able to distinguish diverse cellular properties such as cell type, colorectal cancer stage (normal tissue, primary tumour, metastasis), embryonic stage, and germ layer (Fig.~\ref{fig:resource_usage}).
 Our benchmarks also quantify how groups of cells become more readily separated in larger data sets.
@@
 Based on the observation that incomplete read coverage of a genomic interval can lead to inaccurate methylation estimates, we suggest a scoring scheme that is aware of a read's local context rather than just treating all reads in an interval alike.
+\new{In this way our approach offers an alternative to existing tools that impute missing CpG infomation, such as Melissa \citep{kapourani2019melissa}, DeepCpG \citep{angermueller2017deepcpg} or Epiclomal \citep{pe2020epiclomal}.
+The propper discussion about benefits and drawbacks of data imputation is worth its own paper, but we believe it important to provide the research community with multiple viable options of how to treat their data.}
 
 Furthermore, we show a way to pinpoint minimal regions of high variability across cells, which we call variably methylated regions (VMRs).
+\new{Unlike other tools for scBS data analysis \citep{kapourani2019melissa, kapourani2021scmet}, which rely on the user to manually specify which genomic intervals should be quantified, \texttt{scbs} implements an approach to discover these intervals in the data itself.
+This not only reduces noise and allows to concentrate only on the features that are important for the given dataset, but also provides useful input for genomics-style analyses.
+Depending on the research question at hand, individual VMRs may be related to nearby genomic features such as gene bodies or known regulatory elements, tested for methylation differences between groups of cells, or subjected to gene ontology and motif enrichment.}
+
+\new{To furthermore aid interpretation of scBS data, we developed and implemented the first algorithm for detecting dataset-specific DMRs for scBS data.}
+FDR estimation via permutation allows us to report statistical significance of each DMR.
+\new{In a similar manner as VMRs, pinpointing the regions that differ between groups of cells and hence have a putative regulatory role aids bilogical interpretability.}
+\new{For instance,} applying our approach to NSCs and oligodendrocytes demonstrated that the obtained DMRs locate near meaningful loci associated with cell-type specific functions.
+
+We also presented and published an open source software tool, called \texttt{scbs}, that provides an easy-to-use implementation of the described algorithms.
+\new{It can start directly from the output of methylation callers such as Bismark \citep{bismark}, Biscuit \citep{biscuit} and methylpy \citep{methylpy} and produce a cell\texttimes region matrix.
+We suggest iterative PCA as an approach to map the count matrix to a reduced-dimensional space overcoming the abundance of missing values.
+However, one can also use to the same end established tools based on matrix factorisation such as MOFA+, \citep[][included in our benchmarking]{argelaguet2020mofa} or LIGER \citep{welch2019single}.
+Once a low dimentional emberdding is obtained, one can switch to well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc.
+By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.
+Our implementation can handle datasets of various sizes up to hundred thousand cells (see Supplementary Figure \ref{fig:umaps}B.}
+
 By re-analyzing two published data sets, we could show that these improvements to data preprocessing help to increase signal and decrease noise, resulting in a more informative intermediate-dimensional representation of the data.
 As examples of practical benefits, we demonstrate that our preprocessing allows for better distinction of cell sub-types, especially for challenging data sets comprising cellular sub-states and lineage transitions or for data sets with small cell number.
-
-To furthermore aid interpretation of scBS data, we developed and implemented the first single-cell DMR detection algorithm.
-FDR estimation via permutation allows us to report statistical significance of each DMR.
-Applying our approach to NSCs and oligodendrocytes demonstrated that the obtained DMRs locate near meaningful loci associated with cell-type specific functions.
-Thus, our method aids biological interpretability by pinpointing the regions that differ between groups of cells and hence have a putative regulatory role.
-
-We also presented and published an open source software tool, called \texttt{scbs}, that provides an easy-to-use implementation of the described algorithms.
-\todo{Rewrite this part to discuss other tools: The package also offers functionality for all other steps of data preprocessing that are needed to get from the output of methylation callers such as Bismark and methylpy to a cell\texttimes region matrix that is suitable as input for standard analysis tools that work in a reduced-dimensionality setting.
-This allows the user to use well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc.
-By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.}
-\new{
-Our benchmarks demonstrate that selecting informative genomic features is a crucial aspect of scBS data analysis.
-Unlike other tools for scBS data analysis \citep{}, which rely on the user to manually specify which genomic intervals should be quantified, \texttt{scbs} implements an approach to discover these intervals in the data itself.
-Furthermore, \todo{other tools don't offer DMR detection and our residuals}.
-\todo{scbs is compatible with other tools that are up- or downstream of its capabilities, as we've shown here for MOFA+.}
-}
-In addition, by pinpointing exact regions of interest, the VMR and DMR functionalities provide useful input for genomics-style analyses.
-Depending on the research question at hand, individual DMRs may be related to nearby genomic features such as gene bodies or known regulatory elements, tested for methylation differences between groups of cells, or subjected to gene ontology and motif enrichment.
 

### 2024-01-31 paper.tex
Benchmark in methods section updated

@@
 
+\new{Throughout this paper we refer to this approach as iterative PCA.}
+
 \subsection{Analysis of scBS datasets for benchmarks}
@@
 Methylation was quantified either by averaging binary methylation values, or by calculating the shrunken mean of the residuals as described earlier.
-}
-\todo{...describe the four dimensionality reduction approaches, see below...}
-The same analysis was repeated for \new{three additional scBS data sets \citep{luo2017single, bian2018single, argelaguet2019gastru}}, and for smaller data sets generated by randomly sub-sampling cells separately from \new{these} data sets.
-\todo{update this and maybe move it up?: For each data set, we thus obtained four methylation matrices:
-either 100~kb tiles or VMRs, quantified either by simple averaging (methylation fractions) or with the shrunken mean of the residuals.
-All four matrices were then subjected to PCA.
-In the case of methylation fractions, we closely followed methods described in \citet{luo2017single}:
+For dimensionality reduction we used for different approaches.
+Three of them involved imputing missing values and applying PCA.
+Imputation with iterative PCA is described in the section above.
+We also imputed missing values with mean methylation levels for the given gene (or zeros for shrunken residuals).
+The same process was also combined with the selection of only high-quality features as is described in \citet{luo2017single}:}
 We selected tiles spanning $\ge20$ CpG sites and with sequencing coverage in at least 70\% of cells.
 We then imputed missing values by imputing the mean for each tile, centered the values, and performed PCA.
-In the case of VMRs, we used iterative PCA to handle missing values, as described earlier.
-For visualization, the first 15 PCs were subjected to UMAP with parameters min\_dist $=0.2$, init $=$ ``spca''.
+\new{
+The last approach we tried was using MOFA+ (with default parameters) instead of PCA, which does not require to impute missing values. 
+In all four cases we reduced the dimensionality of the input data to 15 PCs or MOFA factors. In some cases MOFA+ returned smaller number of factors, since some of the requested 15 had zero variance.
+For visualization, these 15 PCs or MOFA factors were subjected to UMAP with parameters min\_dist $=0.2$, init $=$ ``spca''.
 To flexibly adapt to data sets of different sizes, we set n\_neighbors  $=\frac{\sqrt{n}}{1.5}$ (rounded to the nearest integer), where $n$ is the total cell number.
 }
+The same analysis was repeated for \new{three additional scBS data sets \citep{luo2017single, bian2018single, argelaguet2019gastru}}, and for smaller data sets generated by randomly sub-sampling cells separately from \new{these} data sets.
 
@@
 	The depicted cell labels were reported by \citet{liu2021dna}.
-	\todo{Analysing this large data set took approximately two days on a cluster with 256 GB RAM and and 48 CPUs.}
+	\new{Analysing this large data set takes approximately a week on a computer with 256 GB RAM and and 48 CPUs, with almost all of this time (152 hours) spent by the \texttt{prepare} step, which should be run only once for a dataset.
+One can speed up the process considerably by processing each chromosome separately (each takes between 6 and 12 hours depending on its size).}
 	}}

### 2024-01-31 paper.tex
add panel that quantifies overlap with other genomic features

@@
 		\new{
-		\textbf{(F)} Resource and time requirements for dimensionality reduction of VMRs, regulatory elements and genomic tiles.
-		VMR runtime includes VMR detection (light blue).
+		\textbf{(F)} Proportion of VMRs (blue dot) that have at least 1~bp overlap with other genomic features.
+		The black range illustrates the minimum and maximum overlap observed for 100 permutations of randomly shuffled VMRs.
 		}

### 2024-01-31 paper.tex
add panel that quantifies overlap with other genomic features

@@
 Overall, VMR quantification yielded results similar to those obtained when quantifying ENCODE regulatory regions, even though the number of detected VMRs (63\,421) is considerably smaller than the number of ENCODE cCREs (339\,815).
-This demonstrates that our de-novo VMR detection approach identifies genomic regions in scBS data that are as informative as state-of-the-art epigenomic annotations, as well as its ability to find a parsimonous set of relevant elements.
-We not that the overlap between the VMRs and the cCREs is quite limited \todo{Give a number} indicating that VMR detection yields information that is complementary to other epigenetic marks.
+This demonstrates that our \textit{de novo} VMR detection approach identifies genomic regions in scBS data that are as informative as state-of-the-art epigenomic annotations, as well as its ability to find a parsimonous set of relevant elements.
+We detected only limited overlap between VMRs and cCREs (Fig.~\ref{fig:score}F), indicating that VMR detection yields information that is complementary to other epigenetic marks.
 An added benefit of VMR detection is that this approach is available even in the absence of such annotations, for instance when studying species other than human or mouse.
-Furthermore, using VMRs over regulatory regions resulted in decreased RAM requirements as well as a shorter runtime, even when accounting for the additional step of VMR detection (Fig.~\ref{fig:score}F).
-\todo{maybe move panel F to the Supplement}.
-}
+Furthermore, using VMRs over regulatory regions resulted in decreased RAM requirements as well as a shorter runtime, even when accounting for the additional step of VMR detection.
 %Filtering regulatory regions by sequencing coverage reduced resource requirements, but at the cost of a reduced ability to distinguish cell types (Fig.~\ref{fig:resource_usage}).

### 2024-02-01 paper.tex
added missing brace
fix typos, one sentence per line

@@
 %Filtering regulatory regions by sequencing coverage reduced resource requirements, but at the cost of a reduced ability to distinguish cell types (Fig.~\ref{fig:resource_usage}).
+}
 
@@
 \new{In this way our approach offers an alternative to existing tools that impute missing CpG infomation, such as Melissa \citep{kapourani2019melissa}, DeepCpG \citep{angermueller2017deepcpg} or Epiclomal \citep{pe2020epiclomal}.
-The propper discussion about benefits and drawbacks of data imputation is worth its own paper, but we believe it important to provide the research community with multiple viable options of how to treat their data.}
+The proper discussion about benefits and drawbacks of data imputation is worth its own paper, but we believe it important to provide the research community with multiple viable options of how to treat their data.}
 
@@
 FDR estimation via permutation allows us to report statistical significance of each DMR.
-\new{In a similar manner as VMRs, pinpointing the regions that differ between groups of cells and hence have a putative regulatory role aids bilogical interpretability.}
+\new{In a similar manner as VMRs, pinpointing the regions that differ between groups of cells and hence have a putative regulatory role aids biological interpretability.}
 \new{For instance,} applying our approach to NSCs and oligodendrocytes demonstrated that the obtained DMRs locate near meaningful loci associated with cell-type specific functions.
@@
 However, one can also use to the same end established tools based on matrix factorisation such as MOFA+, \citep[][included in our benchmarking]{argelaguet2020mofa} or LIGER \citep{welch2019single}.
-Once a low dimentional emberdding is obtained, one can switch to well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc.
+Once a low dimensional embedding is obtained, one can switch to well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc.
 By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.
-Our implementation can handle datasets of various sizes up to hundred thousand cells (see Supplementary Figure \ref{fig:umaps}B.}
+Our implementation can handle datasets of various sizes up to a hundred thousand cells (see Supplementary Figure \ref{fig:umaps}B.}
 
@@
 
+\new{Throughout this paper we refer to this approach as iterative PCA.}
 We note that similar approaches have also been used elsewhere \citep{josse2012}.
-
-\new{Throughout this paper we refer to this approach as iterative PCA.}
 

### 2024-02-01 paper.tex
minor

@@
 By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.
-Our implementation can handle datasets of various sizes up to a hundred thousand cells (see Supplementary Figure \ref{fig:umaps}B.}
+Our implementation can handle datasets of various sizes up to a hundred thousand cells (see Supplementary Figure \ref{fig:umaps}B).}
 
-By re-analyzing two published data sets, we could show that these improvements to data preprocessing help to increase signal and decrease noise, resulting in a more informative intermediate-dimensional representation of the data.
+By re-analyzing two published data sets, we showed that these improvements to data preprocessing help to increase signal and decrease noise, resulting in a more informative intermediate-dimensional representation of the data.
 As examples of practical benefits, we demonstrate that our preprocessing allows for better distinction of cell sub-types, especially for challenging data sets comprising cellular sub-states and lineage transitions or for data sets with small cell number.

### 2024-02-01 paper.tex
change order, less confusing

@@
 	\textbf{(B)} Separation of neuron sub-types, annotated by averaging CH-methylation (mCH) in 100~kb genomic tiles \citep{luo2017single}.
-	\textbf{(C)} Separation of human colorectal cancer sampling regions (normal adjacent tissue, primary tumor, lymph node metastasis, liver metastasis before and after treatment, omental metastasis) \citep{bian2018single}.
+	\textbf{(C)} Separation of human colorectal cancer sampling regions (primary tumor, normal adjacent tissue, lymph node metastasis, liver metastasis before and after treatment, omental metastasis) \citep{bian2018single}.
 	\textbf{(D-E)} Separation of three properties of mouse embryo cells:

### 2024-02-01 paper.tex
rm last todos

@@
 These extensive benchmarks confirmed that our proposed combination of methods, i.e.\ using the shrunken means of residuals of VMRs for dimensionality reduction, is able to distinguish diverse cellular properties such as cell type, colorectal cancer stage (normal tissue, primary tumour, metastasis), embryonic stage, and germ layer (Fig.~\ref{fig:resource_usage}).
-Our benchmarks also quantify how groups of cells become more readily separated in larger data sets.
-As the neighbor score started to plateau at approximately 1000 cells, we propose that this represents a reasonable target cell number for future scBS studies.
-\todo{We may want to omit these last two sentences, as they might invite too much debate.}
 }
@@
 SA and LPMK acknowledge funding by the Klaus Tschira Foundation (project 00.022.2019).
-LPMK, MMB, LK, SC, and AMV acknowledge funding from the European Commission via ERC grant 771376 \todo{add synergy ERC?}.
+LPMK, MMB, LK, SC, and AMV acknowledge funding from the European Commission via ERC grant 771376.
 We thank Alexey Uvarovskii for code refactoring and helpful suggestions on the source code.

### 2024-02-01 paper.tex
briefly explain overlap analysis

@@
 Methylation was quantified either by averaging binary methylation values, or by calculating the shrunken mean of the residuals as described earlier.
-For dimensionality reduction we used for different approaches.
+For dimensionality reduction we used four different approaches.
 Three of them involved imputing missing values and applying PCA.
@@
 \new{
-The last approach we tried was using MOFA+ (with default parameters) instead of PCA, which does not require to impute missing values. 
-In all four cases we reduced the dimensionality of the input data to 15 PCs or MOFA factors. In some cases MOFA+ returned smaller number of factors, since some of the requested 15 had zero variance.
-For visualization, these 15 PCs or MOFA factors were subjected to UMAP with parameters min\_dist $=0.2$, init $=$ ``spca''.
+Lastly, we used MOFA+ with default parameters instead of PCA, which does not require imputation of missing values. 
+In all four cases we reduced the dimensionality of the input data to 15 PCs or MOFA factors. In some cases MOFA+ returned a smaller number of factors, since some of the requested 15 had zero variance.
+For visualization, these 15 PCs or factors were subjected to UMAP with parameters min\_dist $=0.2$, init $=$ ``spca''.
 To flexibly adapt to data sets of different sizes, we set n\_neighbors  $=\frac{\sqrt{n}}{1.5}$ (rounded to the nearest integer), where $n$ is the total cell number.
@@
 The same analysis was repeated for \new{three additional scBS data sets \citep{luo2017single, bian2018single, argelaguet2019gastru}}, and for smaller data sets generated by randomly sub-sampling cells separately from \new{these} data sets.
+
+\new{VMRs that intersect protein-coding gene bodies, CpG islands or cCREs were quantified by subtracting VMRs with at least one bp of overlap using \texttt{bedtools} \texttt{subtract} \texttt{-A} \citep{quinlan2010bedtools}.
+}
 

### 2024-02-01 paper.tex
fix typo and shorten

@@
 	The depicted cell labels were reported by \citet{liu2021dna}.
-	\new{Analysing this large data set takes approximately a week on a computer with 256 GB RAM and and 48 CPUs, with almost all of this time (152 hours) spent by the \texttt{prepare} step, which should be run only once for a dataset.
-One can speed up the process considerably by processing each chromosome separately (each takes between 6 and 12 hours depending on its size).}
+	\new{Analysing this large data set takes approximately a week on a computer with 256 GB RAM and 48 CPUs, with almost all of this time (152 hours) spent by the \texttt{prepare} step, which needs to be run only once per data set.
+The speed of this step can be increased considerably by processing chromosomes in parallel (each takes 6 to 12 hours depending on its size).}
 	}}

### 2024-02-02 paper.tex
minor

@@
 A simple and common approach to construct a methylation matrix suitable for PCA, used for instance by \citet{luo2017single}, is to divide the genome into tiles of e.g., 100~kb size, and calculate for each cell the average methylation of the DNA within each tile.
-To this end, one identifies in the tile all CpG sites that are covered by at least one read and averages their methylation state, i.e., one denotes as average DNA methylation of the tile in a given cell the proportion of the observed CpG sites in the tile that were found to be methylated (Figure \ref{fig:smoothres}A).
+To this end, one identifies in the tile all CpG sites that are covered by at least one read and averages their methylation state, i.e., one denotes as average DNA methylation of the tile in a given cell the proportion of the observed CpG sites in the tile that were found to be methylated (Fig.~\ref{fig:smoothres}A).
 This yields a matrix, with one row for each cell and one column for each genomic tile, comprising numbers (``methylation fractions'') between 0 and 1.
@@
 This is relevant as scBS protocols are costly and labor-intensive, and only few laboratories are currently able to obtain thousands of single cell methylomes.
-To simulate smaller data sets, we sub-sampled the 1566-cell data set into smaller data sets, analyzed them again with all combinations of analysis methods, and calculated the mean neighbor score for each (Figure \ref{fig:score}D).
+To simulate smaller data sets, we sub-sampled the 1566-cell data set into smaller data sets, analyzed them again with all combinations of analysis methods, and calculated the mean neighbor score for each (Fig.~\ref{fig:score}D).
 \new{
@@
 
-We have implemented the approach just described in a Python package with a command line interface, called \texttt{scbs}, which also offers a number of other functionalities for the analysis of scBS data (Figure \ref{fig:workflow}).
+We have implemented the approach just described in a Python package with a command line interface, called \texttt{scbs}, which also offers a number of other functionalities for the analysis of scBS data (Fig.~\ref{fig:workflow}).
 The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark}, methylpy \citep{methylpy} or BISCUIT \citep{biscuit}.
@@
 By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.
-Our implementation can handle datasets of various sizes up to a hundred thousand cells (see Supplementary Figure \ref{fig:umaps}B).}
+Our implementation can handle datasets of various sizes up to a hundred thousand cells (see Supplementary Fig.~\ref{fig:umaps}B).}
 

### 2024-02-02 paper.tex
one sentence to mention supp fig

@@
 Overall, VMR quantification yielded results similar to those obtained when quantifying ENCODE regulatory regions, even though the number of detected VMRs (63\,421) is considerably smaller than the number of ENCODE cCREs (339\,815).
+Repeating our analysis on the 63\,421 cCREs with the highest quality came at the cost of a reduced ability to distinguish cell types, suggesting that the average VMR is more informative for this task (Fig.~\ref{fig:resource_usage}A,B).
 This demonstrates that our \textit{de novo} VMR detection approach identifies genomic regions in scBS data that are as informative as state-of-the-art epigenomic annotations, as well as its ability to find a parsimonous set of relevant elements.
@@
 An added benefit of VMR detection is that this approach is available even in the absence of such annotations, for instance when studying species other than human or mouse.
-Furthermore, using VMRs over regulatory regions resulted in decreased RAM requirements as well as a shorter runtime, even when accounting for the additional step of VMR detection.
-%Filtering regulatory regions by sequencing coverage reduced resource requirements, but at the cost of a reduced ability to distinguish cell types (Fig.~\ref{fig:resource_usage}).
+Furthermore, using VMRs over regulatory regions resulted in decreased RAM requirements as well as a shorter runtime, even when accounting for the additional step of VMR detection (Fig.~\ref{fig:resource_usage}C,D).
 }
@@
 \new{
-These extensive benchmarks confirmed that our proposed combination of methods, i.e.\ using the shrunken means of residuals of VMRs for dimensionality reduction, is able to distinguish diverse cellular properties such as cell type, colorectal cancer stage (normal tissue, primary tumour, metastasis), embryonic stage, and germ layer (Fig.~\ref{fig:resource_usage}).
+These extensive benchmarks confirmed that our proposed combination of methods, i.e.\ using the shrunken means of residuals of VMRs for dimensionality reduction, is able to distinguish diverse cellular properties such as cell type, colorectal cancer stage (normal tissue, primary tumour, metastasis), embryonic stage, and germ layer.
 }

### 2024-02-02 paper.tex
minor improvements i hope

@@
 As an alternative, we propose improved strategies to identify more informative regions for methylation quantification, and a more accurate quantitation method than simple averaging.
-Our approach enables better discrimination of cell types and reduces the need for large numbers of cells.
+Our approach enables better discrimination of cell types \new{and other features of interest} and reduces the need for large numbers of cells.
 We also present an approach to detect differentially methylated regions (DMRs) between groups of cells, and demonstrate its ability to identify biologically meaningful regions that are associated with genes involved in the core functions of specific cell types.
@@
 In the present paper, we discuss strategies to analyze single-cell bisulfite-sequencing (scBS) data.
-We suggest improvements to current approaches, and demonstrate their value in benchmarks, using two real-world datasets.
+We suggest improvements to current approaches, and demonstrate their value in benchmarks, using \new{four} real-world datasets.
 Furthermore, we discuss how to perform comparative analyses.
@@
 
-The starting point in most scRNA-seq analyses is a matrix of UMI counts (i.e.\ of counts of distinct RNA molecules), one row for each cell and one column for each gene.
+The starting point in most scRNA-seq analyses is a matrix of UMI counts (i.e.\ counts of distinct RNA molecules), one row for each cell and one column for each gene.
 A first goal is usually to assign cell types or states to cells.
 To this end, one needs to establish which cells are similar to each other, i.e., quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile.
-A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows:
+A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or Scanpy \citep{Wolf_2018}, is as follows:
 One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm.
@@
 Here, it is justified to simply put zero into the matrix element, because a shrunken residual average of zero indicates that there is no evidence of the cell deviating from the mean.
-We slightly refine this by using an iterative imputation within the PCA (see Methods for details).
+We slightly refine this by using an iterative imputation within the PCA (\new{"iterative PCA"}, see Methods for details).
 
@@
 Indeed, a gene-wise comparison revealed many genes whose expression is correlated with the nearest VMR, but not with promoter methylation.
-One such example gene, \textit{Htra1}, is depicted in Fig.~\ref{fig:correlation}C.
+One such example gene, \textit{Htra1}, is depicted in Fig.~\ref{fig:correlation}B.
 While the promoter of this gene is lowly methylated regardless of gene expression, a VMR located downstream of the promoter is lowly methylated in cells with high \textit{Htra1} expression.
@@
 either VMRs detected with our approach, 100~kb genomic tiles, promoter regions (i.e., transcription start site (TSS)~\textpm~2000~bp), or candidate cis-regulatory elements \citep[ENCODE cCREs,][]{encode2020expanded}.
-To quantify methylation at these features, we either simply averaged in these intervals, obtaining methylation percentages, or we calculated the shrunken mean of residuals, as described above.
+To quantify methylation at these features, we either simply averaged in these intervals, obtaining methylation percentages, or we calculated the shrunken mean of residuals, as described earlier.
 Finally, the resulting methylation matrices were subjected to iterative PCA (see Methods) and subsequent UMAP for visualization.
@@
 To quantify this performance gain in a more rigorous manner, we used a score that quantifies whether cells were placed, in 15-dimensional PC space, in a neighborhood comprising cells of the same cell type ("neighbor score", see Fig.~\ref{fig:score}E, and Methods for details).
-\new{The neighborhood relation in PCA space is relevant as it is used in many down-stream analyses, e.g., for clustering.}
+\new{The neighborhood relation in PCA space is relevant as it is used in many downstream analyses, e.g., for clustering.}
 We ask how the neighbor score depends on the number of available cells.
@@
 Using the shrunken mean of residuals as a measure of DNA methylation improved results further.
-This effect was most noticeable when using promoters or genomic tiles as intervals, presumably because an individual promoter region or tile might span genomic regions with varying levels of DNA methylation as depicted in Fig.~\ref{fig:smoothres}B, which our method accounts for.
+This effect was most noticeable when quantifying promoters or genomic tiles, presumably because an individual promoter region or tile might span genomic regions with varying levels of DNA methylation as depicted in Fig.~\ref{fig:smoothres}B, which our method accounts for.
 Although cell cluster separation generally becomes more difficult in those, performance gains were observed also in smaller data sets.
@@
 We repeated this benchmark on an additional three published single-cell methylome data sets (Fig.~\ref{fig:score_luo}).
-These include neuronal sub-types of the murine cortex \citep[][using cell type labels derived from CH-methylation in genomic tiles, instead of CpG-methylation, as ground truth]{luo2017single}, cells isolated from mouse embryos during the onset of gastrulation \citep[][using RNA-derived cell clusters, or alternatively embryonic stage as ground truth]{argelaguet2019gastru} and human colorectal cancer cells \citep[][using sampling region as ground truth]{bian2018single}.
+These include neuronal sub-types of the murine cortex \citep[][using cell type labels derived from CH-methylation in genomic tiles instead of CpG-methylation as ground truth]{luo2017single}, cells isolated from mouse embryos during the onset of gastrulation \citep[][using RNA-derived cell clusters or alternatively embryonic stage as ground truth]{argelaguet2019gastru} and human colorectal cancer cells \citep[][using sampling region as ground truth]{bian2018single}.
 Again, we subjected each data set to all possible combinations of genomic feature selection and methylation quantification.
@@
 Repeating this after permuting the cell-type labels, in order to obtain a null distribution of the t statistics, yielded DMRs with much weaker methylation differences.
-Consequently, we could assign to many of the DMRs detected in the unpermuted data a low adjusted p-values (colors in Fig.~\ref{fig:dmr}A).
+Consequently, we could assign to many of the DMRs detected in the unpermuted data a low adjusted p-value (colors in Fig.~\ref{fig:dmr}A).
 
@@
 
+\section{Discussion and Conclusion}
+
+Here, we have proposed an improved strategy to preprocess single-cell bisulfite sequencing data.
+Based on the observation that incomplete read coverage of a genomic interval can lead to inaccurate methylation estimates, we suggest a scoring scheme that is aware of a read's local context rather than just treating all reads in an interval alike.
+\new{In this way our approach offers an alternative to existing tools that impute missing CpG infomation, such as Melissa \citep{kapourani2019melissa}, DeepCpG \citep{angermueller2017deepcpg} or Epiclomal \citep{pe2020epiclomal}.}
+
+Furthermore, we show a way to pinpoint minimal regions of high variability across cells, which we call variably methylated regions (VMRs).
+\new{Unlike other tools for scBS data analysis \citep{kapourani2019melissa, kapourani2021scmet}, which rely on the user to manually specify which genomic intervals should be quantified, \texttt{scbs} implements an approach to discover these intervals in the data itself.
+This not only reduces noise and allows to focus only on the features that are important for the given dataset, but also provides useful input for genomics-style analyses.
+Depending on the research question at hand, individual VMRs may be related to nearby genomic features such as gene bodies or known regulatory elements, or subjected to gene ontology and motif enrichment.}
+
+\new{To furthermore aid interpretation of scBS data, we developed and implemented the first algorithm for genome-wide detection of DMRs in single-cell methylomes.}
+FDR estimation via permutation allows us to report statistical significance of each DMR.
+\new{In a similar manner as VMRs, pinpointing the regions that differ between groups of cells and hence have a putative regulatory role aids biological interpretability.}
+\new{For instance,} applying our approach to NSCs and oligodendrocytes demonstrated that the obtained DMRs locate near meaningful loci associated with cell-type specific functions.
+
+We also presented and published an open source software tool, called \texttt{scbs}, that provides an easy-to-use implementation of the described algorithms.
+\new{It can start directly from the output of methylation callers such as Bismark \citep{bismark}, Biscuit \citep{biscuit} and methylpy \citep{methylpy} and produce a cell\texttimes region matrix.
+We suggest iterative PCA as an approach to map the count matrix to a reduced-dimensional space overcoming the abundance of missing values.
+However, one can also use to the same end established tools based on matrix factorisation such as MOFA+, \citep[][included in our benchmarks]{argelaguet2020mofa} or LIGER \citep{welch2019single}.
+Once a low dimensional embedding is obtained, one can switch to well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc.
+By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.
+Our implementation can handle datasets of various sizes up to a hundred thousand cells (see Supplementary Fig.~\ref{fig:umaps}B).}
+
+By re-analyzing \new{four} published data sets, we showed that these improvements to data preprocessing help to increase signal and decrease noise, resulting in a more informative intermediate-dimensional representation of the data.
+As examples of practical benefits, we demonstrate that our preprocessing allows for better distinction of cell sub-types, especially for challenging data sets comprising cellular sub-states and lineage transitions or for data sets with small cell number.
+
+In conclusion, we have presented powerful improvements to scBS data preprocessing and a software tool that implements these.
+
+\vspace{1.4ex}
+\noindent\hfil\rule{.6\columnwidth}{.2pt}\hfil
+
 \subsection{Data Availability}
@@
 The source code is hosted at \href{https://github.com/LKremer/scbs}{https://github.com/LKremer/scbs}, where we also provide detailed documentation including a tutorial that demonstrates a complete \texttt{scbs} analysis on an example data set.
-
-\section{Discussion and Conclusion}
-
-Here, we have proposed an improved strategy to preprocess single-cell bisulfite sequencing data.
-Based on the observation that incomplete read coverage of a genomic interval can lead to inaccurate methylation estimates, we suggest a scoring scheme that is aware of a read's local context rather than just treating all reads in an interval alike.
-\new{In this way our approach offers an alternative to existing tools that impute missing CpG infomation, such as Melissa \citep{kapourani2019melissa}, DeepCpG \citep{angermueller2017deepcpg} or Epiclomal \citep{pe2020epiclomal}.
-The proper discussion about benefits and drawbacks of data imputation is worth its own paper, but we believe it important to provide the research community with multiple viable options of how to treat their data.}
-
-Furthermore, we show a way to pinpoint minimal regions of high variability across cells, which we call variably methylated regions (VMRs).
-\new{Unlike other tools for scBS data analysis \citep{kapourani2019melissa, kapourani2021scmet}, which rely on the user to manually specify which genomic intervals should be quantified, \texttt{scbs} implements an approach to discover these intervals in the data itself.
-This not only reduces noise and allows to concentrate only on the features that are important for the given dataset, but also provides useful input for genomics-style analyses.
-Depending on the research question at hand, individual VMRs may be related to nearby genomic features such as gene bodies or known regulatory elements, tested for methylation differences between groups of cells, or subjected to gene ontology and motif enrichment.}
-
-\new{To furthermore aid interpretation of scBS data, we developed and implemented the first algorithm for detecting dataset-specific DMRs for scBS data.}
-FDR estimation via permutation allows us to report statistical significance of each DMR.
-\new{In a similar manner as VMRs, pinpointing the regions that differ between groups of cells and hence have a putative regulatory role aids biological interpretability.}
-\new{For instance,} applying our approach to NSCs and oligodendrocytes demonstrated that the obtained DMRs locate near meaningful loci associated with cell-type specific functions.
-
-We also presented and published an open source software tool, called \texttt{scbs}, that provides an easy-to-use implementation of the described algorithms.
-\new{It can start directly from the output of methylation callers such as Bismark \citep{bismark}, Biscuit \citep{biscuit} and methylpy \citep{methylpy} and produce a cell\texttimes region matrix.
-We suggest iterative PCA as an approach to map the count matrix to a reduced-dimensional space overcoming the abundance of missing values.
-However, one can also use to the same end established tools based on matrix factorisation such as MOFA+, \citep[][included in our benchmarking]{argelaguet2020mofa} or LIGER \citep{welch2019single}.
-Once a low dimensional embedding is obtained, one can switch to well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc.
-By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.
-Our implementation can handle datasets of various sizes up to a hundred thousand cells (see Supplementary Fig.~\ref{fig:umaps}B).}
-
-By re-analyzing two published data sets, we showed that these improvements to data preprocessing help to increase signal and decrease noise, resulting in a more informative intermediate-dimensional representation of the data.
-As examples of practical benefits, we demonstrate that our preprocessing allows for better distinction of cell sub-types, especially for challenging data sets comprising cellular sub-states and lineage transitions or for data sets with small cell number.
-
-In conclusion, we have presented powerful improvements to scBS data preprocessing and a software tool that implements these.
-
-\vspace{1.4ex}
-\noindent\hfil\rule{.6\columnwidth}{.2pt}\hfil
 
@@
 
-\subsection{PCA with imputation}
+\subsection{PCA with iterative imputation}
 
@@
 The standard approach to calculate PCA, commonly done using the IRLBA algorithm \citep{Baglama2005}, is not suited to deal with missing data.
-We circumvent this issue by simply using the PCA itself to impute the missing value.
+We circumvent this issue by simply using the PCA itself to impute the missing value \new{in an approach that we call "iterative PCA").}
 
@@
 
-\new{Throughout this paper we refer to this approach as iterative PCA.}
 We note that similar approaches have also been used elsewhere \citep{josse2012}.
@@
 Methylation was quantified either by averaging binary methylation values, or by calculating the shrunken mean of the residuals as described earlier.
-For dimensionality reduction we used four different approaches.
-Three of them involved imputing missing values and applying PCA.
-Imputation with iterative PCA is described in the section above.
-We also imputed missing values with mean methylation levels for the given gene (or zeros for shrunken residuals).
-The same process was also combined with the selection of only high-quality features as is described in \citet{luo2017single}:}
+}
+
+\new{ 
+We used four different approaches for dimensionality reduction. Three of them involve imputation of missing values followed by PCA:
+The first approach, iterative PCA, was described earlier.
+Second, "PCA on high-quality features" imputes missing values with the mean methylation level of a the given interval, while only retaining high-quality features selected as in \citet{luo2017single}:}
 We selected tiles spanning $\ge20$ CpG sites and with sequencing coverage in at least 70\% of cells.
-We then imputed missing values by imputing the mean for each tile, centered the values, and performed PCA.
+We then imputed missing values with the mean of each tile, centered the values, and performed PCA.
 \new{
+The third approach, "mean-imputed PCA" is identical to approach two, but without the quality-filtering step.
 Lastly, we used MOFA+ with default parameters instead of PCA, which does not require imputation of missing values. 
@@
 
-\new{VMRs that intersect protein-coding gene bodies, CpG islands or cCREs were quantified by subtracting VMRs with at least one bp of overlap using \texttt{bedtools} \texttt{subtract} \texttt{-A} \citep{quinlan2010bedtools}.
+\new{VMRs that intersect protein-coding gene bodies, CpG islands (from the UCSC genome browser) or cCREs were quantified by subtracting VMRs with at least one bp of overlap using \texttt{bedtools} \texttt{subtract} \texttt{-A} \citep{quinlan2010bedtools} and counting the remaining VMRs.
 }
@@
 To test our DMR detection approach, we selected oligodendrocytes and NSCs from healthy wild type mice of \citet{kremer_scnmt} and ran \texttt{scbs diff} with default parameters.
-For GO enrichment analysis, DMRs with $p^\text{adj}<0.01$ were uploaded to GREAT \citep{mclean2010great} with the association rule "basal plus extension, 0~bp upstream, 20~kb downstream, 1~Mbp max extension, curated regulatory domains included".
+For GO enrichment analysis, DMRs with adjusted $p<0.01$ were uploaded to GREAT \citep{mclean2010great} with the association rule "basal plus extension, 0~bp upstream, 20~kb downstream, 1~Mbp max extension, curated regulatory domains included".
 

### 2024-02-02 paper.tex
very minor

@@
 
-To assess statistical significance of DMRs, we repeat the same procedure on permutations of the scBS data, i.e. the same data set with randomly shuffled cell labels.
+To assess statistical significance of DMRs, we repeat the same procedure on permutations of the scBS data, i.e.\ the same data set with randomly shuffled cell labels.
 The DMR t statistics obtained from permuted data are then used to estimate the false-discovery rate (FDR), yielding an adjusted p-value for each DMR.
@@
 
-\texttt{scbs prepare} also computes a number of summary statistics for each cell, including the mean genome-wide methylation level and the number of observed CpG sites, i.e.
-the number of CpG sites that have sequencing coverage.
+\texttt{scbs prepare} also computes a number of summary statistics for each cell, including the mean genome-wide methylation level and the number of observed CpG sites, i.e.\ the number of CpG sites that have sequencing coverage.
 These summary statistics can be used to detect cells with poor quality.
@@
 By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.
-Our implementation can handle datasets of various sizes up to a hundred thousand cells (see Supplementary Fig.~\ref{fig:umaps}B).}
+Our implementation can handle datasets of various sizes up to a hundred thousand cells (see Fig.~\ref{fig:umaps}B).}
 
@@
 
-It may seem that we could use the standard procedure for the Welch t-test here, i.e.
-use the Welch-Satterthwaite formula to get an approximate degree of freedom and then calculate the tail probability of the corresponding t distribution.
+It may seem that we could use the standard procedure for the Welch t-test here, i.e.\ use the Welch-Satterthwaite formula to get an approximate degree of freedom and then calculate the tail probability of the corresponding t distribution.
 However, this is unlikely to hold for two reasons:

### 2024-02-02 paper.tex
final touches

@@
 		\new{
-		\textbf{(F)} Proportion of VMRs (blue dot) that have at least 1~bp overlap with other genomic features.
+		\textbf{(F)} Proportion of VMRs (blue dots) that have at least 1~bp overlap with other genomic features.
 		The black range illustrates the minimum and maximum overlap observed for 100 permutations of randomly shuffled VMRs.
@@
 
-\section{Finding differentially methylated regions}
+\section{Finding differentially methylated regions (DMRs)}
 
@@
 	\end{center}
-	\caption{\small \textbf{Identification of differentially methylated regions.}\\
+	\caption{\small \textbf{Identification of differentially methylated regions (DMRs).}\\
 		The chromosome is divided into windows overlapping by a small and fixed size.
@@
 	The developmental stage (E4.5, E5.5, E6.5, E7.5), the germ layer or extra-embryonic tissue of origin, or the cell lineage.
-	Germ layer and lineage (\textbf{E}, only E7.5-cells) were annotated by \citet{argelaguet2019gastru} based on scRNA-seq of the same cells.
+	Germ layer and lineage (E, only E7.5-cells) were annotated by \citet{argelaguet2019gastru} based on scRNA-seq of the same cells.
 	}}
@@
 	\end{center}
-	\captionof{figure}{\small \new{\textbf{VMRs are as informative as regulatory regions, but more resource-efficient.}\\
-	\textbf{(A-B)} Mean neighbor score obtained when quantifying CpG methylation at VMRs, all ENCODE regulatory regions, or a subset of regulatory regions for the data sets of \citep{kremer_scnmt} and \citet{luo2017single}.
+	\captionof{figure}{\small \new{\textbf{VMRs are more informative than an equal number of regulatory regions with high sequencing coverage.}\\
+	\textbf{(A-B)} Mean neighbor score obtained when quantifying CpG methylation at VMRs, all ENCODE regulatory regions, or a subset of regulatory regions for the data sets of \citet{kremer_scnmt} and \citet{luo2017single}.
 	The subset of high-quality (HQ) regulatory regions was selected in such a way that it matches the number of VMRs and contains those regulatory regions with the highest coverage.
@@
 	The depicted cell labels were reported by \citet{liu2021dna}.
-	\new{Analysing this large data set takes approximately a week on a computer with 256 GB RAM and 48 CPUs, with almost all of this time (152 hours) spent by the \texttt{prepare} step, which needs to be run only once per data set.
-The speed of this step can be increased considerably by processing chromosomes in parallel (each takes 6 to 12 hours depending on its size).}
+	\new{Analysing this large data set takes approximately a week on a computer with 256 GB RAM and 48 CPUs, with almost all of this time (152 hours) spent by the \texttt{prepare} step, which is run only once per data set.
+The speed of this step can be increased considerably by processing chromosomes in parallel, which reduced the total runtime to approximately two days in our case.}
 	}}

### 2024-02-04 paper.tex
last changes

@@
 Overall, VMR quantification yielded results similar to those obtained when quantifying ENCODE regulatory regions, even though the number of detected VMRs (63\,421) is considerably smaller than the number of ENCODE cCREs (339\,815).
-Repeating our analysis on the 63\,421 cCREs with the highest quality came at the cost of a reduced ability to distinguish cell types, suggesting that the average VMR is more informative for this task (Fig.~\ref{fig:resource_usage}A,B).
-This demonstrates that our \textit{de novo} VMR detection approach identifies genomic regions in scBS data that are as informative as state-of-the-art epigenomic annotations, as well as its ability to find a parsimonous set of relevant elements.
-We detected only limited overlap between VMRs and cCREs (Fig.~\ref{fig:score}F), indicating that VMR detection yields information that is complementary to other epigenetic marks.
-An added benefit of VMR detection is that this approach is available even in the absence of such annotations, for instance when studying species other than human or mouse.
-Furthermore, using VMRs over regulatory regions resulted in decreased RAM requirements as well as a shorter runtime, even when accounting for the additional step of VMR detection (Fig.~\ref{fig:resource_usage}C,D).
+When we repeated our analysis using only the 63\,421 cCREs with the highest coverage the ability to distinguish cell types was diminished, suggesting that the average VMR is more informative for this task then the average cCRE (Fig.~\ref{fig:resource_usage}A,B).
+This demonstrates our \textit{de novo} VMR detection approach's ability to identify in scBS data a parsimonous set of relevant elements.
+The overlap between VMRs and cCREs is limited (Fig.~\ref{fig:score}F), indicating that VMR detection yields information that is complementary to other epigenetic marks.
+An further benefit of VMR detection is that this approach is available even in the absence of such annotations, for instance when studying species other than human or mouse.
+Lastly, using VMRs over regulatory regions resulted in decreased RAM requirements as well as a shorter runtime, even when accounting for the additional step of VMR detection (Fig.~\ref{fig:resource_usage}C,D).
 }
@@
 VMR detection with default options produced results that were qualitatively similar to those reported in \citet{luo2017single}, suggesting that our methods are also suitable for this data type (Fig.~\ref{fig:umaps}A).
-As single-cell methylome data sets are expected to rapidly grow in size in the coming years, we furthermore performed a stress test on a large data set comprising 100\,350 cells \citep{liu2021dna} (Fig.~\ref{fig:umaps}B).
+Finally, as single-cell methylome data sets are expected to rapidly grow in size in the coming years, we furthermore performed a stress test on a large data set comprising 100\,350 cells \citep{liu2021dna} (Fig.~\ref{fig:umaps}B).
 }
@@
 		\textbf{(F)} Proportion of VMRs (blue dots) that have at least 1~bp overlap with other genomic features.
-		The black range illustrates the minimum and maximum overlap observed for 100 permutations of randomly shuffled VMRs.
+		The black range illustrates the minimum and maximum overlap observed for 100 reshufflings of the VMRs to randomly chosen positions.
 		}
@@
 Based on the observation that incomplete read coverage of a genomic interval can lead to inaccurate methylation estimates, we suggest a scoring scheme that is aware of a read's local context rather than just treating all reads in an interval alike.
-\new{In this way our approach offers an alternative to existing tools that impute missing CpG infomation, such as Melissa \citep{kapourani2019melissa}, DeepCpG \citep{angermueller2017deepcpg} or Epiclomal \citep{pe2020epiclomal}.}
+%\new{In this way our approach offers an alternative to existing tools that impute missing CpG infomation, such as Melissa \citep{kapourani2019melissa}, DeepCpG \citep{angermueller2017deepcpg} or Epiclomal \citep{pe2020epiclomal}.}
 
 Furthermore, we show a way to pinpoint minimal regions of high variability across cells, which we call variably methylated regions (VMRs).
-\new{Unlike other tools for scBS data analysis \citep{kapourani2019melissa, kapourani2021scmet}, which rely on the user to manually specify which genomic intervals should be quantified, \texttt{scbs} implements an approach to discover these intervals in the data itself.
+\new{Unlike other tools for scBS data analysis \citep{kapourani2019melissa, kapourani2021scmet, danese2021episcanpy}, which rely on the user to manually specify which genomic intervals should be quantified, \texttt{scbs} implements an approach to discover these intervals in the data itself.
 This not only reduces noise and allows to focus only on the features that are important for the given dataset, but also provides useful input for genomics-style analyses.
@@
 
-\new{To furthermore aid interpretation of scBS data, we developed and implemented the first algorithm for genome-wide detection of DMRs in single-cell methylomes.}
+\new{To furthermore aid interpretation of scBS data, we developed and implemented an algorithm for genome-wide detection of DMRs in single-cell methylomes.}
 FDR estimation via permutation allows us to report statistical significance of each DMR.
@@
 We suggest iterative PCA as an approach to map the count matrix to a reduced-dimensional space overcoming the abundance of missing values.
-However, one can also use to the same end established tools based on matrix factorisation such as MOFA+, \citep[][included in our benchmarks]{argelaguet2020mofa} or LIGER \citep{welch2019single}.
+Alternatively, one may use for this purpose established tools based on matrix factorisation such as MOFA+, \citep[][included in our benchmarks]{argelaguet2020mofa} or LIGER \citep{welch2019single}.
 Once a low dimensional embedding is obtained, one can switch to well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc.

### 2024-02-04 paper.tex
a few final chanegs to bib etc

@@
 
-version 3 /\!/ January 2024
+version 2 /\!/ January 2024
 \vspace{4.5ex}
@@
 Alternatively, one may use for this purpose established tools based on matrix factorisation such as MOFA+, \citep[][included in our benchmarks]{argelaguet2020mofa} or LIGER \citep{welch2019single}.
-Once a low dimensional embedding is obtained, one can switch to well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc.
+Once a low dimensional embedding is obtained, one can switch to well established methods from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc., e.g., by using either overall toolkits like Seurat \citep{seurat5}, ScanPy/EpiScanPy \citep{Wolf_2018,danese2021episcanpy} or any of the many available tools for specific tasks.
 By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.

### 2024-03-21 paper.tex
changed name from 'scbs' to 'MethSCAn'

@@
 \centering\phantomsection\addcontentsline{toc}{section}{Title}
-\textbf{\Large Analyzing single-cell bisulfite sequencing data with \texttt{scbs}}\\[1.5ex]
+\textbf{\Large Analyzing single-cell bisulfite sequencing data with \textit{MethSCAn}}\\[1.5ex]
 
@@
 Santiago Cerrizuela\textsuperscript{2},
-Ana Martin-Villalba\textsuperscript{2},
+Ana Martin-Villalba\textsuperscript{2,*},
 Simon Anders\textsuperscript{1,*}\\[1ex]
@@
 
-To facilitate the analysis of scBS data, we have developed a software tool called ``\texttt{scbs}'' that implements these methods and provides additional functionality.
+To facilitate the analysis of scBS data, we have developed a software tool called \textit{MethSCAn} that implements these methods and provides additional functionality.
 
@@
 Furthermore, we discuss how to perform comparative analyses.
-Finally, we present ``\texttt{scbs}'', a comprehensive software toolkit to perform scBS data analysis, using the described methods.
+Finally, we present \textit{MethSCAn}, a comprehensive software toolkit to perform scBS data analysis, using the described methods.
 \blfootnote{\hspace{-.5cm}\raggedright\textsuperscript{*} e-mail:  \texttt{l.kremer@dkfz-heidelberg.de} and\\ \hspace{1cm}\texttt{simon.anders@bioquant.uni-heidelberg.de}}
@@
 Finally, we describe and demonstrate an approach to detect differentially methylated regions (DMRs) in scBS data.
-We also describe our software toolkit, \texttt{scbs}, that facilitates all these analyses.
+We also describe our software toolkit, \textit{MethSCAn}, that facilitates all these analyses.
 
@@
 \new{Next}, we assessed whether our proposed workflow requires fine-tuning of parameters (Fig.~\ref{fig:sweep}).
-To this end, we re-analyzed \new{two} data sets, as well as sub-samples of the data with different VMR detection parameters namely the width of the sliding window in bp, (set with option \texttt{--bandwidth} in our \texttt{scbs} software, default 2000), the variance threshold above which windows are merged to VMRs (\texttt{--var-threshold}, default 0.02) \new{and the step size of the sliding window (\texttt{--stepsize}, default 100bp).}
+To this end, we re-analyzed \new{two} data sets, as well as sub-samples of the data with different VMR detection parameters namely the width of the sliding window in bp, (set with option \texttt{--bandwidth} in our \textit{MethSCAn} software, default 2000), the variance threshold above which windows are merged to VMRs (\texttt{--var-threshold}, default 0.02) \new{and the step size of the sliding window (\texttt{--stepsize}, default 100bp).}
 This parameter sweep showed that our workflow gives good results over a wide range of parameter values.
@@
 
-\section{The \texttt{scbs} software toolkit}
+\section{The \textit{MethSCAn} software toolkit}
 
@@
     \end{center}
-    \caption{\small \textbf{Overview of the functionalities implemented in the \texttt{scbs} package.}\\
-        \texttt{scbs prepare} parses methylation files produced by common bisulfite sequencing mappers and stores their contents in a compressed format optimized for efficient access to genomic intervals.
-        \texttt{scbs} also produces cell-wise summary statistics and quality plots (here: average methylation around the transcription start site) that are used to detect low-quality cells.
-        These cells can be discarded with \texttt{scbs filter}.
+    \caption{\small \textbf{Overview of the functionalities implemented in the \textit{MethSCAn} package.}\\
+        \texttt{methscan prepare} parses methylation files produced by common bisulfite sequencing mappers and stores their contents in a compressed format optimized for efficient access to genomic intervals.
+        \textit{MethSCAn} also produces cell-wise summary statistics and quality plots (here: average methylation around the transcription start site) that are used to detect low-quality cells.
+        These cells can be discarded with \texttt{methscan filter}.
         To obtain a methylation matrix, similar to the count matrices used in scRNA-seq, the user must first decide in which genomic intervals methylation should be quantified.
-        The user may either provide genomic regions of \emph{a priori} interest, or they may choose to discover VMRs (variably methylated regions) in the data with \texttt{scbs scan}.
+        The user may either provide genomic regions of \emph{a priori} interest, or they may choose to discover VMRs (variably methylated regions) in the data with \texttt{methscan scan}.
         The resulting methylation matrix can then be used for downstream analysis such as cell clustering and dimensionality reduction.
-        Identified groups of cells (e.g. clusters) can then be scanned for DMRs with \texttt{scbs diff}.
+        Identified groups of cells (e.g. clusters) can then be scanned for DMRs with \texttt{methscan diff}.
     }
@@
 
-We have implemented the approach just described in a Python package with a command line interface, called \texttt{scbs}, which also offers a number of other functionalities for the analysis of scBS data (Fig.~\ref{fig:workflow}).
+We have implemented the approach just described in a Python package with a command line interface, \textit{MethSCAn} (ANalysis of Single-Cell METHylation data), which also offers a number of other functionalities for the analysis of scBS data (Fig.~\ref{fig:workflow}).
 The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark}, methylpy \citep{methylpy} or BISCUIT \citep{biscuit}.
 For each cell, these tools produce a text file that lists the methylation status of all CpG sites.
-Since it is inconvenient to work with hundreds or thousands of text files, \texttt{scbs} provides the \texttt{prepare} command which parses these methylation files and stores their content in a compressed format that enables efficient access to all CpG sites of the genome.
-\texttt{scbs prepare} is flexible and accepts all tabular input formats including Bismark .cov-files, methylpy .allc-files, or custom user-defined formats.
+Since it is inconvenient to work with hundreds or thousands of text files, \textit{MethSCAn} provides the \texttt{prepare} command which parses these methylation files and stores their content in a compressed format that enables efficient access to all CpG sites of the genome.
+\texttt{methscan prepare} is flexible and accepts all tabular input formats including Bismark .cov-files, methylpy .allc-files, or custom user-defined formats.
 
@@
 
-\texttt{scbs prepare} also computes a number of summary statistics for each cell, including the mean genome-wide methylation level and the number of observed CpG sites, i.e.\ the number of CpG sites that have sequencing coverage.
+\texttt{methscan prepare} also computes a number of summary statistics for each cell, including the mean genome-wide methylation level and the number of observed CpG sites, i.e.\ the number of CpG sites that have sequencing coverage.
 These summary statistics can be used to detect cells with poor quality.
-The quality of each single cell methylome can furthermore be inspected with \texttt{scbs profile}, which computes the average methylation profile of a set of user-defined genomic regions such as transcription start sites (TSS) at single-cell resolution.
+The quality of each single cell methylome can furthermore be inspected with \texttt{methscan profile}, which computes the average methylation profile of a set of user-defined genomic regions such as transcription start sites (TSS) at single-cell resolution.
 The TSS profile is a useful quality control plot since methylation shows a characteristic dip roughly \textpm1~kb around the transcription start site in mammalian genomes.
-Cells that do not show this pattern, or cells with few observed CpG sites, may then be discarded with \texttt{scbs filter}.
+Cells that do not show this pattern, or cells with few observed CpG sites, may then be discarded with \texttt{methscan filter}.
 
-After quality control, the user has access to the genome-wide VMR detection approach described earlier, using the \texttt{scbs scan} command.
+After quality control, the user has access to the genome-wide VMR detection approach described earlier, using the \texttt{methscan scan} command.
 This produces a BED file that lists the genome coordinates of VMRs.
-To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED file can be used as input for \texttt{scbs matrix}, which quantifies methylation at genomic intervals in all cells.
+To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED file can be used as input for \texttt{methscan matrix}, which quantifies methylation at genomic intervals in all cells.
 The command produces both, the simple percentage of methylated CpG sites, as well as our proposed methylation measure, the shrunken mean of the residuals, which is more robust to variation in sequencing coverage or stochastic differences in read position between cells.
@@
 
-Finally, after annotation and exploration of the data set, the user may specify two groups of cells for DMR detection with \texttt{scbs diff}.
+Finally, after annotation and exploration of the data set, the user may specify two groups of cells for DMR detection with \texttt{methscan diff}.
 This command produces a BED file listing the genome coordinates, adjusted p-values, as well as several other metrics associated with DMRs.
@@
 Furthermore, we show a way to pinpoint minimal regions of high variability across cells, which we call variably methylated regions (VMRs).
-\new{Unlike other tools for scBS data analysis \citep{kapourani2019melissa, kapourani2021scmet, danese2021episcanpy}, which rely on the user to manually specify which genomic intervals should be quantified, \texttt{scbs} implements an approach to discover these intervals in the data itself.
+\new{Unlike other tools for scBS data analysis \citep{kapourani2019melissa, kapourani2021scmet, danese2021episcanpy}, which rely on the user to manually specify which genomic intervals should be quantified, \textit{MethSCAn} implements an approach to discover these intervals in the data itself.
 This not only reduces noise and allows to focus only on the features that are important for the given dataset, but also provides useful input for genomics-style analyses.
@@
 
-We also presented and published an open source software tool, called \texttt{scbs}, that provides an easy-to-use implementation of the described algorithms.
+We also presented and published an open source software tool, called \textit{MethSCAn}, that provides an easy-to-use implementation of the described algorithms.
 \new{It can start directly from the output of methylation callers such as Bismark \citep{bismark}, Biscuit \citep{biscuit} and methylpy \citep{methylpy} and produce a cell\texttimes region matrix.
@@
 Once a low dimensional embedding is obtained, one can switch to well established methods from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc., e.g., by using either overall toolkits like Seurat \citep{seurat5}, ScanPy/EpiScanPy \citep{Wolf_2018,danese2021episcanpy} or any of the many available tools for specific tasks.
-By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.
+By offering these functions, \textit{MethSCAn} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.
 Our implementation can handle datasets of various sizes up to a hundred thousand cells (see Fig.~\ref{fig:umaps}B).}
@@
 
-\texttt{scbs} is free and open source.
+\textit{MethSCAn} is free and open source.
 The package is available on the Python Package Index (PyPI) and can be installed with the Python package installer \texttt{pip}.
-The source code is hosted at \href{https://github.com/LKremer/scbs}{https://github.com/LKremer/scbs}, where we also provide detailed documentation including a tutorial that demonstrates a complete \texttt{scbs} analysis on an example data set.
+The source code is hosted at \href{https://github.com/anders-biostat/MethSCAn}{https://github.com/anders-biostat/MethSCAn}, where we also provide detailed documentation including a tutorial that demonstrates a complete \textit{MethSCAn} analysis on an example data set.
 
@@
 
-The function \texttt{scbs prepare} reads a set of methylation files (e.g. produced by Bismark) and produces a file that stores the matrix $x$ in a space-efficient format, as follows:
+The function \texttt{methscan prepare} reads a set of methylation files (e.g. produced by Bismark) and produces a file that stores the matrix $x$ in a space-efficient format, as follows:
 $x$ is represented as a SciPy sparse matrix \citep{SciPy}, encoding the actual values 0, 1, and NA as -1, 1, and 0, respectively.
@@
 \subsection{Analysis of scBS datasets for benchmarks}
-To analyze scBS data from \citet{kremer_scnmt}, single-cell CpG methylation reports from all conditions were first stored with \texttt{scbs prepare} and then smoothed with \texttt{scbs smooth} using the default bandwidth of 1000~bp.
+To analyze scBS data from \citet{kremer_scnmt}, single-cell CpG methylation reports from all conditions were first stored with \texttt{methscan prepare} and then smoothed with \texttt{methscan smooth} using the default bandwidth of 1000~bp.
 \new{
@@
 The following four sets of genomic intervals were used:
-VMRs, obtained with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 0.02$, minimum cell requirement $=6$);
+VMRs, obtained with \texttt{methscan scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 0.02$, minimum cell requirement $=6$);
 adjacent tiles of 100~kb width;
@@
 
-To test our DMR detection approach, we selected oligodendrocytes and NSCs from healthy wild type mice of \citet{kremer_scnmt} and ran \texttt{scbs diff} with default parameters.
+To test our DMR detection approach, we selected oligodendrocytes and NSCs from healthy wild type mice of \citet{kremer_scnmt} and ran \texttt{methscan diff} with default parameters.
 For GO enrichment analysis, DMRs with adjusted $p<0.01$ were uploaded to GREAT \citep{mclean2010great} with the association rule "basal plus extension, 0~bp upstream, 20~kb downstream, 1~Mbp max extension, curated regulatory domains included".
@@
 \subsection{Correlation of DNA methylation and gene expression}
-To assess correlation between gene expression and the methylation status of VMRs or promoters, we first detected VMRs with \texttt{scbs scan --bandwidth 1000 --var-threshold 0.05}.
-We then quantified DNA methylation at VMRs and promoters with \texttt{scbs matrix}.
+To assess correlation between gene expression and the methylation status of VMRs or promoters, we first detected VMRs with \texttt{methscan scan --bandwidth 1000 --var-threshold 0.05}.
+We then quantified DNA methylation at VMRs and promoters with \texttt{methscan matrix}.
 We defined promoters as $\pm2000$~bp regions centered on a genes transcription start site (TSS).
@@
     \textbf{(A-B)} Mean neighbor scores obtained after analyzing single-cell methylomes from \citet{kremer_scnmt} (\textbf{A}) or \citet{luo2017single} (\textbf{B}) with our proposed methods.
-    VMRs were detected with \texttt{scbs} \texttt{scan} using various sliding window bandwidths, variance thresholds, and on sub-samples of the full data sets.
+    VMRs were detected with \texttt{methscan} \texttt{scan} using various sliding window bandwidths, variance thresholds, and on sub-samples of the full data sets.
     \textbf{(C)} Mean neighbor scores obtained after analyzing the \citet{kremer_scnmt} data set with various sliding window step sizes.
@@
 	\end{center}
-	\captionof{figure}{\small \new{\textbf{\texttt{scbs} performs well on CH-methylation data and on extremely large data sets.}\\
+	\captionof{figure}{\small \new{\textbf{\textit{MethSCAn} performs well on CH-methylation data and on extremely large data sets.}\\
     \textbf{(A)} UMAPs obtained when analysing data from \citet{luo2017single} with three different strategies for producing a methylation matrix:
@@
 	bottom: shrunken mean of residuals to quantify CpG methylation in CpG-VMRs.
-	\textbf{(B)} UMAP obtained when analysing 100\,350 neural single-cell methylomes \citep{liu2021dna} with the default \texttt{scbs} workflow.
+	\textbf{(B)} UMAP obtained when analysing 100\,350 neural single-cell methylomes \citep{liu2021dna} with the default \textit{MethSCAn} workflow.
 	The depicted cell labels were reported by \citet{liu2021dna}.

### 2024-03-21 paper.tex
removed \scriptsize
added Ana's e-mail address

@@
 Finally, we present \textit{MethSCAn}, a comprehensive software toolkit to perform scBS data analysis, using the described methods.
-\blfootnote{\hspace{-.5cm}\raggedright\textsuperscript{*} e-mail:  \texttt{l.kremer@dkfz-heidelberg.de} and\\ \hspace{1cm}\texttt{simon.anders@bioquant.uni-heidelberg.de}}
+\blfootnote{\hspace{-.5cm}\raggedright\textsuperscript{*} e-mails:  \texttt{l.kremer@dkfz-heidelberg.de}, \hspace{1cm}\texttt{a.martin-villalba@dkfz-heidelberg.de}, \hspace{1cm}\texttt{simon.anders@bioquant.uni-heidelberg.de}}
 
@@
 \begin{equation} 
-r_{Ij} = \frac{\scriptsize{1}}{\scriptsize{|I\cap C_i|+1}}\sum_{i\in I\cap C_i}\left( x_{ij} - \tilde x_i \right).
+r_{Ij} = \frac{1}{|I\cap C_i|+1}\sum_{i\in I\cap C_i}\left( x_{ij} - \tilde x_i \right).
 \label{avgres}
@@
 \begin{equation} 
-v_I = \frac{\scriptsize{1}}{\scriptsize{|C_I|}}\sum_{j\in C_I}\left( r_{Ij} - \langle r_{Ij'}\rangle_{j'\in C_I} \right)^2, \label{var}
+v_I = \frac{1}{|C_I|}\sum_{j\in C_I}\left( r_{Ij} - \langle r_{Ij'}\rangle_{j'\in C_I} \right)^2, \label{var}
 \end{equation}

### 2024-04-10 paper.tex
removed highlighting of new text
comply with reporting summary

@@
 \newcommand{\todo}[1]{[\textcolor{orange}{#1}]}
-\newcommand{\new}[1]{\textcolor{blue}{#1}}
+\newcommand{\new}[1]{#1} %\textcolor{blue}{#1}}
 
@@
 Martina M. Braun\textsuperscript{2},
-\new{Svetlana Ovchinnikova\textsuperscript{1},}  % feel free to adjust the author order, Simon.
+Svetlana Ovchinnikova\textsuperscript{1},
 Leonie Küchenhoff\textsuperscript{1,2},\\
@@
 As an alternative, we propose improved strategies to identify more informative regions for methylation quantification, and a more accurate quantitation method than simple averaging.
-Our approach enables better discrimination of cell types \new{and other features of interest} and reduces the need for large numbers of cells.
+Our approach enables better discrimination of cell types and other features of interest and reduces the need for large numbers of cells.
 We also present an approach to detect differentially methylated regions (DMRs) between groups of cells, and demonstrate its ability to identify biologically meaningful regions that are associated with genes involved in the core functions of specific cell types.
@@
 To facilitate the analysis of scBS data, we have developed a software tool called \textit{MethSCAn} that implements these methods and provides additional functionality.
-
-%In summary, our work provides a more robust and accurate approach to analyze scBS data, which has the potential to enhance our understanding of DNA methylation dynamics in single cells and improve the identification of epigenetic changes associated with different cell states or diseases.
 \end{minipage}
@@
 In the present paper, we discuss strategies to analyze single-cell bisulfite-sequencing (scBS) data.
-We suggest improvements to current approaches, and demonstrate their value in benchmarks, using \new{four} real-world datasets.
+We suggest improvements to current approaches, and demonstrate their value in benchmarks, using four real-world datasets.
 Furthermore, we discuss how to perform comparative analyses.
@@
 
-Therefore, we propose to first obtain, for each CpG position, a \new{smoothed} average of the methylation across all cells, and then quantify each cell's deviation from this average.
+Therefore, we propose to first obtain, for each CpG position, a smoothed average of the methylation across all cells, and then quantify each cell's deviation from this average.
 In Figure \ref{fig:smoothres}B, the curved line depicts such an average over all cells, and the red vertical lines show an individual cell's deviation from the ensemble average.
@@
 Here, it is justified to simply put zero into the matrix element, because a shrunken residual average of zero indicates that there is no evidence of the cell deviating from the mean.
-We slightly refine this by using an iterative imputation within the PCA (\new{"iterative PCA"}, see Methods for details).
+We slightly refine this by using an iterative imputation within the PCA ("iterative PCA", see Methods for details).
 
@@
 Only the latter regions are of value for our goal of quantitating dissimilarity between cells.
-We call these the variably methylated regions (VMRs). % maybe mention the parallel to highly variable genes in scRNAseq?
+We call these the variably methylated regions (VMRs).
 
@@
 
-To demonstrate the value of our proposed improvements, we benchmarked various combinations of analysis methods on \new{five diverse single-cell methylome data sets, starting with a data set from our own research}.
+To demonstrate the value of our proposed improvements, we benchmarked various combinations of analysis methods on five diverse single-cell methylome data sets, starting with a data set from our own research.
 
@@
 We consider these cell labels as ground truth and tested whether we are able to distinguish the same groups of cells based on their methylomes.
-\new{
 To do this, we subjected the methylomes to various combinations of analysis methods including our own proposed methods and others that are commonly used.
@@
 Finally, the resulting methylation matrices were subjected to iterative PCA (see Methods) and subsequent UMAP for visualization.
-}
 
@@
 To quantify this performance gain in a more rigorous manner, we used a score that quantifies whether cells were placed, in 15-dimensional PC space, in a neighborhood comprising cells of the same cell type ("neighbor score", see Fig.~\ref{fig:score}E, and Methods for details).
-\new{The neighborhood relation in PCA space is relevant as it is used in many downstream analyses, e.g., for clustering.}
+The neighborhood relation in PCA space is relevant as it is used in many downstream analyses, e.g., for clustering.
 We ask how the neighbor score depends on the number of available cells.
@@
 To simulate smaller data sets, we sub-sampled the 1566-cell data set into smaller data sets, analyzed them again with all combinations of analysis methods, and calculated the mean neighbor score for each (Fig.~\ref{fig:score}D).
-\new{
 The results confirmed that quantifying methylation at VMRs leads to a cleaner separation of cell states than using promoter regions or 100~kb tiles.
@@
 Although cell cluster separation generally becomes more difficult in those, performance gains were observed also in smaller data sets.
-}
 
-\new{
 Overall, VMR quantification yielded results similar to those obtained when quantifying ENCODE regulatory regions, even though the number of detected VMRs (63\,421) is considerably smaller than the number of ENCODE cCREs (339\,815).
@@
 Lastly, using VMRs over regulatory regions resulted in decreased RAM requirements as well as a shorter runtime, even when accounting for the additional step of VMR detection (Fig.~\ref{fig:resource_usage}C,D).
-}
 
-\new{
 We repeated this benchmark on an additional three published single-cell methylome data sets (Fig.~\ref{fig:score_luo}).
@@
 We furthermore included three additional approaches to perform dimensionality reduction in our benchmarks, including PCA with two different pre-processing strategies, as well as MOFA+, a dimensionality reduction technique designed for multi-modal single-cell data that can also process methylation data \citep{argelaguet2020mofa}.
-}
 
-\new{
 These extensive benchmarks confirmed that our proposed combination of methods, i.e.\ using the shrunken means of residuals of VMRs for dimensionality reduction, is able to distinguish diverse cellular properties such as cell type, colorectal cancer stage (normal tissue, primary tumour, metastasis), embryonic stage, and germ layer.
-}
 
@@
 
-\new{Next}, we assessed whether our proposed workflow requires fine-tuning of parameters (Fig.~\ref{fig:sweep}).
-To this end, we re-analyzed \new{two} data sets, as well as sub-samples of the data with different VMR detection parameters namely the width of the sliding window in bp, (set with option \texttt{--bandwidth} in our \textit{MethSCAn} software, default 2000), the variance threshold above which windows are merged to VMRs (\texttt{--var-threshold}, default 0.02) \new{and the step size of the sliding window (\texttt{--stepsize}, default 100bp).}
+Next, we assessed whether our proposed workflow requires fine-tuning of parameters (Fig.~\ref{fig:sweep}).
+To this end, we re-analyzed two data sets, as well as sub-samples of the data with different VMR detection parameters namely the width of the sliding window in bp, (set with option \texttt{--bandwidth} in our \textit{MethSCAn} software, default 2000), the variance threshold above which windows are merged to VMRs (\texttt{--var-threshold}, default 0.02) and the step size of the sliding window (\texttt{--stepsize}, default 100bp).
 This parameter sweep showed that our workflow gives good results over a wide range of parameter values.
-For the \new{CpG-methylation} data of \citet{luo2017single}, results are nearly independent of the parameters (Fig.~\ref{fig:sweep}B).
-In the more challenging data set of \citet{kremer_scnmt}, cell types were less cleanly separated when very large bandwidths, very strict variance thresholds, \new{or a very large step size was} selected (Fig.~\ref{fig:sweep}A,C).
+For the CpG-methylation data of \citet{luo2017single}, results are nearly independent of the parameters (Fig.~\ref{fig:sweep}B).
+In the more challenging data set of \citet{kremer_scnmt}, cell types were less cleanly separated when very large bandwidths, very strict variance thresholds, or a very large step size was selected (Fig.~\ref{fig:sweep}A,C).
 However, very small bandwidths or very lenient thresholds resulted in a much higher number of VMRs and thus long computing times.
@@
 
-\new{
 Lastly, we asked whether our methods are also suitable for the analysis of DNA methylation outside the default CpG context.
@@
 Finally, as single-cell methylome data sets are expected to rapidly grow in size in the coming years, we furthermore performed a stress test on a large data set comprising 100\,350 cells \citep{liu2021dna} (Fig.~\ref{fig:umaps}B).
-}
 
@@
 		\textbf{(A)} Cell labels based on clustering of single-cell transcriptomes from \citet{kremer_scnmt}.
-		\textbf{(B, C)} Exemplary UMAPs obtained when analyzing the data set with \new{conventional methods based on genome tiling (B) or our proposed methods (C)}.
-		\new{
+		\textbf{(B, C)} Exemplary UMAPs obtained when analyzing the data set with conventional methods based on genome tiling (B) or our proposed methods (C).
 		\textbf{(D, E)} Mean degree of cell type separation (neighbor score, E, computed in 15-dimensional PC space) obtained when analyzing single-cell methylomes with different combinations of methods.
@@
 		CpG methylation in these intervals was either quantified by averaging (dotted lines) or using the shrunken mean of the residuals as proposed in this work (solid lines).
-	    }
 		The full 1566-cell data set was sub-sampled to simulate smaller data sets (x axis).
 		A higher neighbor score implies better separation of cell types inferred from single-cell transcriptomes of the same cells (A).
-		\new{
 		\textbf{(F)} Proportion of VMRs (blue dots) that have at least 1~bp overlap with other genomic features.
 		The black range illustrates the minimum and maximum overlap observed for 100 reshufflings of the VMRs to randomly chosen positions.
-		}
 	}
@@
 Similar to the previously described approach for VMR identification (Fig.~\ref{fig:vmr}), we divide each chromosome into overlapping windows shifted by a small and fixed size (step size) and quantify the methylation of each cell in each window.
-Next, instead of the variance, we obtain the t statistic % Welch t-test
-as a measure of differential methylation between the two cell groups.
+Next, instead of the variance, we obtain the t statistic as a measure of differential methylation between the two cell groups.
 We identify the windows with the most extreme t statistics, e.g. windows in the 2\% upper and lower tails.
@@
 We note that both \texttt{matrix} and \texttt{profile} accept any valid BED file as input, which means that the user can quantify and visualize methylation at any set of genomic features of interest, such as promoters, enhancers or transcription factor binding sites, obtaining one profile plot per cell.
-% This is not really true cause you still need to deal with missing values and all the default transformations don't make sense when you're not dealing with count data: "The methylation matrix produced by the \texttt{matrix} function can then serve as input for standard single-cell analysis tools like Seurat \citep{seurat} or ScanPy \citep{Wolf_2018}, in order to perform, e.g., dimensionality reduction and cell clustering.
-For this, the methylation matrix should be provided in lieu of the log-normalized expression matrix usually expected by these tools.
 The methylation matrix produced by the \texttt{matrix} function can then serve as input for established methods used for the exploration of single-cell data, such as dimensionality reduction and cell clustering.
@@
 Based on the observation that incomplete read coverage of a genomic interval can lead to inaccurate methylation estimates, we suggest a scoring scheme that is aware of a read's local context rather than just treating all reads in an interval alike.
-%\new{In this way our approach offers an alternative to existing tools that impute missing CpG infomation, such as Melissa \citep{kapourani2019melissa}, DeepCpG \citep{angermueller2017deepcpg} or Epiclomal \citep{pe2020epiclomal}.}
 
 Furthermore, we show a way to pinpoint minimal regions of high variability across cells, which we call variably methylated regions (VMRs).
-\new{Unlike other tools for scBS data analysis \citep{kapourani2019melissa, kapourani2021scmet, danese2021episcanpy}, which rely on the user to manually specify which genomic intervals should be quantified, \textit{MethSCAn} implements an approach to discover these intervals in the data itself.
+Unlike other tools for scBS data analysis \citep{kapourani2019melissa, kapourani2021scmet, danese2021episcanpy}, which rely on the user to manually specify which genomic intervals should be quantified, \textit{MethSCAn} implements an approach to discover these intervals in the data itself.
 This not only reduces noise and allows to focus only on the features that are important for the given dataset, but also provides useful input for genomics-style analyses.
-Depending on the research question at hand, individual VMRs may be related to nearby genomic features such as gene bodies or known regulatory elements, or subjected to gene ontology and motif enrichment.}
+Depending on the research question at hand, individual VMRs may be related to nearby genomic features such as gene bodies or known regulatory elements, or subjected to gene ontology and motif enrichment.
 
-\new{To furthermore aid interpretation of scBS data, we developed and implemented an algorithm for genome-wide detection of DMRs in single-cell methylomes.}
+To furthermore aid interpretation of scBS data, we developed and implemented an algorithm for genome-wide detection of DMRs in single-cell methylomes.
 FDR estimation via permutation allows us to report statistical significance of each DMR.
-\new{In a similar manner as VMRs, pinpointing the regions that differ between groups of cells and hence have a putative regulatory role aids biological interpretability.}
-\new{For instance,} applying our approach to NSCs and oligodendrocytes demonstrated that the obtained DMRs locate near meaningful loci associated with cell-type specific functions.
+In a similar manner as VMRs, pinpointing the regions that differ between groups of cells and hence have a putative regulatory role aids biological interpretability.
+For instance, applying our approach to NSCs and oligodendrocytes demonstrated that the obtained DMRs locate near meaningful loci associated with cell-type specific functions.
 
 We also presented and published an open source software tool, called \textit{MethSCAn}, that provides an easy-to-use implementation of the described algorithms.
-\new{It can start directly from the output of methylation callers such as Bismark \citep{bismark}, Biscuit \citep{biscuit} and methylpy \citep{methylpy} and produce a cell\texttimes region matrix.
+It can start directly from the output of methylation callers such as Bismark \citep{bismark}, Biscuit \citep{biscuit} and methylpy \citep{methylpy} and produce a cell\texttimes region matrix.
 We suggest iterative PCA as an approach to map the count matrix to a reduced-dimensional space overcoming the abundance of missing values.
@@
 By offering these functions, \textit{MethSCAn} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.
-Our implementation can handle datasets of various sizes up to a hundred thousand cells (see Fig.~\ref{fig:umaps}B).}
+Our implementation can handle datasets of various sizes up to a hundred thousand cells (see Fig.~\ref{fig:umaps}B).
 
-By re-analyzing \new{four} published data sets, we showed that these improvements to data preprocessing help to increase signal and decrease noise, resulting in a more informative intermediate-dimensional representation of the data.
+By re-analyzing four published data sets, we showed that these improvements to data preprocessing help to increase signal and decrease noise, resulting in a more informative intermediate-dimensional representation of the data.
 As examples of practical benefits, we demonstrate that our preprocessing allows for better distinction of cell sub-types, especially for challenging data sets comprising cellular sub-states and lineage transitions or for data sets with small cell number.
@@
 
-Single-cell multi-omic data of cells of the murine forebrain \citep{kremer_scnmt} is available at the NCBI Gene Expression Omnibus (GEO) under the accession \texttt{GSE210806}.
+All data used to benchmark and showcase the \textit{MethSCAn} software is publicly available under the following NCBI Gene Expression Omnibus accessions:
+Single-cell multi-omics of the murine forebrain \citep{kremer_scnmt}: \texttt{GSE210806};
+mouse neurons \citep{luo2017single}: \texttt{GSE97179};
+colorectal cancer \citep{bian2018single}: \texttt{GSE97693};
+mouse gastrulation \citep{argelaguet2019gastru}: \texttt{GSE121708};
+100k brain cells \citep{liu2021dna}: \texttt{GSE132489}.
 
@@
 The standard approach to calculate PCA, commonly done using the IRLBA algorithm \citep{Baglama2005}, is not suited to deal with missing data.
-We circumvent this issue by simply using the PCA itself to impute the missing value \new{in an approach that we call "iterative PCA").}
+We circumvent this issue by simply using the PCA itself to impute the missing value in an approach that we call "iterative PCA").
 
@@
 To analyze scBS data from \citet{kremer_scnmt}, single-cell CpG methylation reports from all conditions were first stored with \texttt{methscan prepare} and then smoothed with \texttt{methscan smooth} using the default bandwidth of 1000~bp.
-\new{
 This data was then analysed multiple times with different combinations of analysis methods, namely four ways to divide the genome into intervals, two ways to quantify methylation in these intervals, and four approaches for dimensionality reductions.
@@
 Methylation was quantified either by averaging binary methylation values, or by calculating the shrunken mean of the residuals as described earlier.
-}
 
-\new{ 
 We used four different approaches for dimensionality reduction. Three of them involve imputation of missing values followed by PCA:
 The first approach, iterative PCA, was described earlier.
-Second, "PCA on high-quality features" imputes missing values with the mean methylation level of a the given interval, while only retaining high-quality features selected as in \citet{luo2017single}:}
+Second, "PCA on high-quality features" imputes missing values with the mean methylation level of a the given interval, while only retaining high-quality features selected as in \citet{luo2017single}:
 We selected tiles spanning $\ge20$ CpG sites and with sequencing coverage in at least 70\% of cells.
 We then imputed missing values with the mean of each tile, centered the values, and performed PCA.
-\new{
 The third approach, "mean-imputed PCA" is identical to approach two, but without the quality-filtering step.
@@
 To flexibly adapt to data sets of different sizes, we set n\_neighbors  $=\frac{\sqrt{n}}{1.5}$ (rounded to the nearest integer), where $n$ is the total cell number.
-}
-The same analysis was repeated for \new{three additional scBS data sets \citep{luo2017single, bian2018single, argelaguet2019gastru}}, and for smaller data sets generated by randomly sub-sampling cells separately from \new{these} data sets.
 
-\new{VMRs that intersect protein-coding gene bodies, CpG islands (from the UCSC genome browser) or cCREs were quantified by subtracting VMRs with at least one bp of overlap using \texttt{bedtools} \texttt{subtract} \texttt{-A} \citep{quinlan2010bedtools} and counting the remaining VMRs.
-}
+The same analysis was repeated for three additional scBS data sets \citep{luo2017single, bian2018single, argelaguet2019gastru}, and for smaller data sets generated by randomly sub-sampling cells separately from these data sets.
+
+VMRs that intersect protein-coding gene bodies, CpG islands (from the UCSC genome browser) or cCREs were quantified by subtracting VMRs with at least one bp of overlap using \texttt{bedtools} \texttt{subtract} \texttt{-A} \citep{quinlan2010bedtools} and counting the remaining VMRs.
 
 To test our DMR detection approach, we selected oligodendrocytes and NSCs from healthy wild type mice of \citet{kremer_scnmt} and ran \texttt{methscan diff} with default parameters.
-For GO enrichment analysis, DMRs with adjusted $p<0.01$ were uploaded to GREAT \citep{mclean2010great} with the association rule "basal plus extension, 0~bp upstream, 20~kb downstream, 1~Mbp max extension, curated regulatory domains included".
+For GO enrichment analysis, DMRs with adjusted $p<0.01$ were uploaded to GREAT 4.0.4 \citep{mclean2010great} with the association rule "basal plus extension, 0~bp upstream, 20~kb downstream, 1~Mbp max extension, curated regulatory domains included".
 
@@
 
+MethSCAn was implemented in Python 3.8 using the packages NumPy 1.20.1, SciPy 1.6.1, numba 0.53.0 and Pandas 1.2.3.
+Benchmarks were performed on MethSCAn version 0.6.2 using Snakemake 7.26 and analyzed/visualized with tidyverse 1.3.1 packages.
 \vspace{1.4ex}
@@
 SA and LPMK acknowledge funding by the Klaus Tschira Foundation (project 00.022.2019).
-LPMK, MMB, LK, SC, and AMV acknowledge funding from the European Commission via ERC grant 771376.
+LPMK, MMB, LK, SC, and AMV acknowledge funding from the European Commission via ERC grant 771376, from the DFG via SFB grant 873, and from the DKFZ.
 We thank Alexey Uvarovskii for code refactoring and helpful suggestions on the source code.
@@
 	\end{center}
-	\captionof{figure}{\small \new{ \textbf{Benchmark of several combinations of analysis methods on four single-cell methylation data sets.}\\
+	\captionof{figure}{\small \textbf{Benchmark of several combinations of analysis methods on four single-cell methylation data sets.}\\
 	\textbf{(A-E)} Benchmarking the ability to separate groups of cells based on CpG methylation data, using different analysis approaches.
@@
 	Germ layer and lineage (E, only E7.5-cells) were annotated by \citet{argelaguet2019gastru} based on scRNA-seq of the same cells.
-	}}
+	}
 	\label{fig:score_luo}
@@
     \captionof{figure}{\small \textbf{Effect of VMR detection parameters on separation of cell types.}\\
-	\new{
     \textbf{(A-B)} Mean neighbor scores obtained after analyzing single-cell methylomes from \citet{kremer_scnmt} (\textbf{A}) or \citet{luo2017single} (\textbf{B}) with our proposed methods.
@@
     \textbf{(C)} Mean neighbor scores obtained after analyzing the \citet{kremer_scnmt} data set with various sliding window step sizes.
-    }}
+    }
     \label{fig:sweep}
@@
 	\end{center}
-	\captionof{figure}{\small \new{\textbf{VMRs are more informative than an equal number of regulatory regions with high sequencing coverage.}\\
+	\captionof{figure}{\small \textbf{VMRs are more informative than an equal number of regulatory regions with high sequencing coverage.}\\
 	\textbf{(A-B)} Mean neighbor score obtained when quantifying CpG methylation at VMRs, all ENCODE regulatory regions, or a subset of regulatory regions for the data sets of \citet{kremer_scnmt} and \citet{luo2017single}.
@@
 	\textbf{(C-D)} Runtime (top) and RAM usage (bottom) of the two dimensionality reduction techniques shown in A and B.
-	}}
+	}
 	\label{fig:resource_usage}
@@
 	\end{center}
-	\captionof{figure}{\small \new{\textbf{\textit{MethSCAn} performs well on CH-methylation data and on extremely large data sets.}\\
+	\captionof{figure}{\small \textbf{\textit{MethSCAn} performs well on CH-methylation data and on extremely large data sets.}\\
     \textbf{(A)} UMAPs obtained when analysing data from \citet{luo2017single} with three different strategies for producing a methylation matrix:
@@
 	The depicted cell labels were reported by \citet{liu2021dna}.
-	\new{Analysing this large data set takes approximately a week on a computer with 256 GB RAM and 48 CPUs, with almost all of this time (152 hours) spent by the \texttt{prepare} step, which is run only once per data set.
-The speed of this step can be increased considerably by processing chromosomes in parallel, which reduced the total runtime to approximately two days in our case.}
-	}}
+	Analysing this large data set takes approximately a week on a computer with 256 GB RAM and 48 CPUs, with almost all of this time (152 hours) spent by the \texttt{prepare} step, which is run only once per data set.
+The speed of this step can be increased considerably by processing chromosomes in parallel, which reduced the total runtime to approximately two days in our case.
+	}
 	\label{fig:umaps}

### 2024-04-10 paper.tex
supp fig -> extended data fig

@@
 Overall, VMR quantification yielded results similar to those obtained when quantifying ENCODE regulatory regions, even though the number of detected VMRs (63\,421) is considerably smaller than the number of ENCODE cCREs (339\,815).
-When we repeated our analysis using only the 63\,421 cCREs with the highest coverage the ability to distinguish cell types was diminished, suggesting that the average VMR is more informative for this task then the average cCRE (Fig.~\ref{fig:resource_usage}A,B).
+When we repeated our analysis using only the 63\,421 cCREs with the highest coverage the ability to distinguish cell types was diminished, suggesting that the average VMR is more informative for this task then the average cCRE (Extended Data Fig.~\ref{fig:resource_usage}A,B).
 This demonstrates our \textit{de novo} VMR detection approach's ability to identify in scBS data a parsimonous set of relevant elements.
@@
 An further benefit of VMR detection is that this approach is available even in the absence of such annotations, for instance when studying species other than human or mouse.
-Lastly, using VMRs over regulatory regions resulted in decreased RAM requirements as well as a shorter runtime, even when accounting for the additional step of VMR detection (Fig.~\ref{fig:resource_usage}C,D).
+Lastly, using VMRs over regulatory regions resulted in decreased RAM requirements as well as a shorter runtime, even when accounting for the additional step of VMR detection (Extended Data Fig.~\ref{fig:resource_usage}C,D).
 
-We repeated this benchmark on an additional three published single-cell methylome data sets (Fig.~\ref{fig:score_luo}).
+We repeated this benchmark on an additional three published single-cell methylome data sets (Extended Data Fig.~\ref{fig:score_luo}).
 These include neuronal sub-types of the murine cortex \citep[][using cell type labels derived from CH-methylation in genomic tiles instead of CpG-methylation as ground truth]{luo2017single}, cells isolated from mouse embryos during the onset of gastrulation \citep[][using RNA-derived cell clusters or alternatively embryonic stage as ground truth]{argelaguet2019gastru} and human colorectal cancer cells \citep[][using sampling region as ground truth]{bian2018single}.
@@
 
-Next, we assessed whether our proposed workflow requires fine-tuning of parameters (Fig.~\ref{fig:sweep}).
+Next, we assessed whether our proposed workflow requires fine-tuning of parameters (Extended Data Fig.~\ref{fig:sweep}).
 To this end, we re-analyzed two data sets, as well as sub-samples of the data with different VMR detection parameters namely the width of the sliding window in bp, (set with option \texttt{--bandwidth} in our \textit{MethSCAn} software, default 2000), the variance threshold above which windows are merged to VMRs (\texttt{--var-threshold}, default 0.02) and the step size of the sliding window (\texttt{--stepsize}, default 100bp).
 This parameter sweep showed that our workflow gives good results over a wide range of parameter values.
-For the CpG-methylation data of \citet{luo2017single}, results are nearly independent of the parameters (Fig.~\ref{fig:sweep}B).
-In the more challenging data set of \citet{kremer_scnmt}, cell types were less cleanly separated when very large bandwidths, very strict variance thresholds, or a very large step size was selected (Fig.~\ref{fig:sweep}A,C).
+For the CpG-methylation data of \citet{luo2017single}, results are nearly independent of the parameters (Extended Data Fig.~\ref{fig:sweep}B).
+In the more challenging data set of \citet{kremer_scnmt}, cell types were less cleanly separated when very large bandwidths, very strict variance thresholds, or a very large step size was selected (Extended Data Fig.~\ref{fig:sweep}A,C).
 However, very small bandwidths or very lenient thresholds resulted in a much higher number of VMRs and thus long computing times.
@@
 To this end, we revisited the \citet{luo2017single} data set, but this time only considered CH-methylation (mCH).
-VMR detection with default options produced results that were qualitatively similar to those reported in \citet{luo2017single}, suggesting that our methods are also suitable for this data type (Fig.~\ref{fig:umaps}A).
-Finally, as single-cell methylome data sets are expected to rapidly grow in size in the coming years, we furthermore performed a stress test on a large data set comprising 100\,350 cells \citep{liu2021dna} (Fig.~\ref{fig:umaps}B).
+VMR detection with default options produced results that were qualitatively similar to those reported in \citet{luo2017single}, suggesting that our methods are also suitable for this data type (Extended Data Fig.~\ref{fig:umaps}A).
+Finally, as single-cell methylome data sets are expected to rapidly grow in size in the coming years, we furthermore performed a stress test on a large data set comprising 100\,350 cells \citep{liu2021dna} (Extended Data Fig.~\ref{fig:umaps}B).
 
@@
 By offering these functions, \textit{MethSCAn} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.
-Our implementation can handle datasets of various sizes up to a hundred thousand cells (see Fig.~\ref{fig:umaps}B).
+Our implementation can handle datasets of various sizes up to a hundred thousand cells (see Extended Data Fig.~\ref{fig:umaps}B).
 
@@
 \newpage
-\section{Supplementary Figures}
+\section{Extended Data Figures}
 \vspace{1.5cm}
@@
 \setcounter{figure}{0}
-\renewcommand{\thefigure}{S\arabic{figure}}
+%\renewcommand{\thefigure}{ \arabic{figure}}
+\renewcommand{\figurename}{Extended Data Figure}
 

### 2024-04-29 paper.tex
minor adjustments

@@
-\documentclass[twocolumn,10pt]{article}
+\documentclass[10pt]{article}
 
 \usepackage[a4paper,hmargin=1.5cm,vmargin=2cm]{geometry}
-\setlength{\columnsep}{0.7cm}
+%\setlength{\columnsep}{0.7cm}
 %\usepackage{microtype} % nice typesetting  % doesn't work for me :-( -S
@@
 \usepackage{balance}
-\usepackage{palatino}
 \usepackage[varqu]{inconsolata}
@@
 
-\newcommand{\todo}[1]{[\textcolor{orange}{#1}]}
-\newcommand{\new}[1]{#1} %\textcolor{blue}{#1}}
-
 \begin{document}
 
-\twocolumn[{%
-\centering\phantomsection\addcontentsline{toc}{section}{Title}
+{\centering\phantomsection\addcontentsline{toc}{section}{Title}
 \textbf{\Large Analyzing single-cell bisulfite sequencing data with \textit{MethSCAn}}\\[1.5ex]
@@
 \textsuperscript{2} Division of Molecular Neurobiology, German Cancer Research Center (DKFZ), Heidelberg, Germany}
-\vspace{1.5ex}
-
-version 2 /\!/ January 2024
+}
 \vspace{4.5ex}
 
-\begin{minipage}{0.7\textwidth}
-\small \setlength{\parindent}{1em}
-\phantomsection\addcontentsline{toc}{subsection}{Abstract}
-\noindent \textit{Abstract} -- Single-cell bisulfite sequencing (scBS) is a technique that enables the assessment of DNA methylation at single-base pair and single-cell resolution.
+\section*{Abstract}
+
+\noindent Single-cell bisulfite sequencing (scBS) is a technique that enables the assessment of DNA methylation at single-base pair and single-cell resolution.
 The analysis of large datasets obtained from scBS requires preprocessing to reduce data size, improve signal-to-noise ratio, and provide interpretability.
@@
 To facilitate the analysis of scBS data, we have developed a software tool called \textit{MethSCAn} that implements these methods and provides additional functionality.
-\end{minipage}
-\vspace{6.5ex}
-}]
-\setcounter{secnumdepth}{0}
 
-\section{Introduction}
+\section*{Introduction}
 
@@
 
-%\phantomsection  % idk why we need this to silence hyperref's warning
-\subsection{Standard approach}
+\subsection*{Standard approach}
 
@@
 
-\section{Improved preprocessing of scBS data}
+\section*{Results}
 
-\subsection{Read-position aware quantitation} \label{residuals}
+\subsection*{Read-position aware quantitation} \label{residuals}
 
@@
 
-\subsection{Finding variably methylated regions}
+\subsection*{Finding variably methylated regions}
 
@@
 
-\subsection{Application and benchmarks}
+\subsection*{Application and benchmarks}
 
@@
 
-%\phantomsection
 \subsubsection{Correlating VMR methylation with gene expression}
@@
 
-\section{Finding differentially methylated regions (DMRs)}
+\section*{Finding differentially methylated regions (DMRs)}
 
@@
 
-%\phantomsection
-\subsection{Detecting DMRs between oligodendrocytes and neural stem cells}
+\subsection*{Detecting DMRs between oligodendrocytes and neural stem cells}
 
@@
 
-\section{The \textit{MethSCAn} software toolkit}
+\section*{The \textit{MethSCAn} software toolkit}
 
@@
 
-\section{Discussion and Conclusion}
+\section*{Discussion}
 
@@
 
-\subsection{Data Availability}
+\subsection*{Data Availability}
 
@@
 
-\subsection{Code Availability}
+\subsection*{Code Availability}
 
@@
 %\phantomsection
-\section{Methods}
+\section*{Methods}
 
-\subsection{Raw data}
+\subsection*{Raw data}
 
@@
 
-\subsection{Data storage}
+\subsection*{Data storage}
 
@@
 
-\subsection{Smoothing}
+\subsection*{Smoothing}
 
@@
 
-\subsection{Methylation for an interval}
+\subsection*{Methylation for an interval}
 
@@
 
-\subsection{Finding differentially methylated regions}
+\subsection*{Finding differentially methylated regions}
 
@@
 
-\subsection{Calculating cell to cell distances}
+\subsection*{Calculating cell to cell distances}
 
@@
 
-\subsection{PCA with iterative imputation}
+\subsection*{PCA with iterative imputation}
 
@@
 
-\subsection{Analysis of scBS datasets for benchmarks}
+\subsection*{Analysis of scBS datasets for benchmarks}
 To analyze scBS data from \citet{kremer_scnmt}, single-cell CpG methylation reports from all conditions were first stored with \texttt{methscan prepare} and then smoothed with \texttt{methscan smooth} using the default bandwidth of 1000~bp.
@@
 
-\subsection{Mean neighbor score} \label{methods:score}
+\subsection*{Mean neighbor score} \label{methods:score}
 To assess performance of our methods, we employed a score that quantifies how well cell types (or cell states) are separated in 15-dimensional PCA space.
@@
 
-\subsection{Correlation of DNA methylation and gene expression}
+\subsection*{Correlation of DNA methylation and gene expression}
 To assess correlation between gene expression and the methylation status of VMRs or promoters, we first detected VMRs with \texttt{methscan scan --bandwidth 1000 --var-threshold 0.05}.
@@
 \phantomsection\addcontentsline{toc}{section}{Back Matter}
-\subsection{Acknowledgments}
+\subsection*{Acknowledgments}
 
@@
 \newpage
-\section{Extended Data Figures}
+\section*{Extended Data Figures}
 \vspace{1.5cm}

### 2024-04-29 paper.tex
add competing interests, author contribs (WIP), simplify format

@@
 %\setlength{\columnsep}{0.7cm}
-%\usepackage{microtype} % nice typesetting  % doesn't work for me :-( -S
 \usepackage{amsmath}
@@
 \usepackage[utf8]{inputenc}
-\usepackage{natbib}
+\usepackage[numbers]{natbib}
 \bibliographystyle{simon}
@@
 Simon Anders\textsuperscript{1,*}\\[1ex]
-{\footnotesize \textsuperscript{1} BioQuant Centre, University of Heidelberg, Heidelberg, Germany\\[-.5ex] 
+{\footnotesize
+\textsuperscript{1} BioQuant Centre, University of Heidelberg, Heidelberg, Germany\\[-.5ex] 
 \textsuperscript{2} Division of Molecular Neurobiology, German Cancer Research Center (DKFZ), Heidelberg, Germany}
@@
 \subsection*{Standard approach}
-
-\begin{figure*}[t]
-    \begin{center}
-        \includegraphics[width=0.7\textwidth]{figures/Fig_residuals_AB.png}\\
-    \end{center}
-    \caption{\small \textbf{Improved quantification of DNA methylation in a given genomic interval.}\\
-    \textbf{(A)} Depicted is a genomic interval (between vertical blue lines) along a chromosome, for which DNA methylation is to be quantified.
-    Two cells cover differing parts of the interval with one read each.
-    If one simply counts for each cell which fraction of the covered CpG sites are methylated, one obtains very different values for the two cells.
-    \textbf{(B)} By averaging each CpG site's methylation over all cells and subsequent smoothing, the thick black ``average methylation curve'' is obtained.
-    To quantify the methylation of cell 1 from (A) relative to this average over all cells, we propose to use the cell's residuals to the smoothed curve (lengths of the vertical red lines) and take their average, counting residuals of methylated CpGs positive and residuals of unmethylated CpGs negative.}
-    \label{fig:smoothres}
-\end{figure*}
 
@@
 \subsection*{Finding variably methylated regions}
-
-\begin{figure}
-    \begin{center}
-    \includegraphics[width=\columnwidth]{figures/Fig_sliding.png}
-    \end{center}
-    \caption{\small \textbf{Finding variably methylated regions.}\\
-    The chromosomes are divided up into overlapping windows (first five shown at the bottom), and for each window, the cells' methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
-    Then, the variance of these values is calculated (each point represents one of the overlapping windows), a threshold (dashed line) is chosen such that a chosen quantile of windows have a variance exceeding the threshold.
-    Windows with above-threshold variance are merged if they overlap, yielding the ``variably methylated regions'' (VMRs).}
-    \label{fig:vmr}
-\end{figure}
 
@@
 
-\begin{figure*}
-	\begin{center}
-		\includegraphics[width=\textwidth]{figures/Fig_correlation.png}
-	\end{center}
-	\caption{\small \textbf{Correlation of DNA methylation and gene expression}.\\
-		\textbf{(A)} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red): for most promoters, correlation is weak.
-        Promoters are defined as intervals \textpm2~kb around the TSS.
-		\textbf{(B)} Mean methylation near the gene \textit{Htra1}.
-		Cells are assigned to 4 groups based on \textit{Htra1} expression (group 0: cells with no \textit{Htra1} expression, groups 1-3: cells that express \textit{Htra1}, divided into three equally large groups with group 3 having the highest expression).
-		Data from \citet{kremer_scnmt}.}
-	\label{fig:correlation}
-\end{figure*}
-
-\begin{figure*}
-	\begin{center}
-		\includegraphics[width=0.9\textwidth]{figures/Fig_benchmark.png}
-	\end{center}
-	\caption{\small \textbf{Benchmark of our methods on single-cell multi-omic data of cells of the murine forebrain}.\\
-		\textbf{(A)} Cell labels based on clustering of single-cell transcriptomes from \citet{kremer_scnmt}.
-		\textbf{(B, C)} Exemplary UMAPs obtained when analyzing the data set with conventional methods based on genome tiling (B) or our proposed methods (C).
-		\textbf{(D, E)} Mean degree of cell type separation (neighbor score, E, computed in 15-dimensional PC space) obtained when analyzing single-cell methylomes with different combinations of methods.
-		Either VMRs, ENCODE regulatory elements, 100~kb genomic tiles or promoter regions (TSS\textpm2kb) were subjected to iterative PCA and UMAP.
-		CpG methylation in these intervals was either quantified by averaging (dotted lines) or using the shrunken mean of the residuals as proposed in this work (solid lines).
-		The full 1566-cell data set was sub-sampled to simulate smaller data sets (x axis).
-		A higher neighbor score implies better separation of cell types inferred from single-cell transcriptomes of the same cells (A).
-		\textbf{(F)} Proportion of VMRs (blue dots) that have at least 1~bp overlap with other genomic features.
-		The black range illustrates the minimum and maximum overlap observed for 100 reshufflings of the VMRs to randomly chosen positions.
-	}
-	\label{fig:score}
-\end{figure*}
-
 \section*{Finding differentially methylated regions (DMRs)}
@@
 To enable DMR detection in scBS data, we thus developed an approach that detects DMRs of variable size between two groups of cells, and controls the false discovery rate (Fig.~\ref{fig:dmrscan}):
-
-\begin{figure*}
-	\begin{center}
-		\includegraphics[width=.75\textwidth]{figures/Fig_DMRdetection.png}
-	\end{center}
-	\caption{\small \textbf{Identification of differentially methylated regions (DMRs).}\\
-		The chromosome is divided into windows overlapping by a small and fixed size.
-		For each window, the cell’s methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
-		Then the t statistic of these values is obtained as a measure of methylation difference between two groups of cells (each point represents one of the windows).
-		Upper and lower thresholds (dashed lines) are chosen such that a defined quantile of windows have a t statistic above or below the threshold.
-		Windows exceeding either threshold are merged if they overlap, yielding the differentially methylated regions (DMRs).}
-	\label{fig:dmrscan}
-\end{figure*}
 
@@
 While the primary purpose of VMRs is to provide better input for PCA and distance calculations, DMR detection facilitates the discovery of epigenetic differences between conditions or cell types, as we demonstrate next.
-
-\begin{figure*}
-	\begin{center}
-		\includegraphics[width=.7\textwidth]{figures/Fig_DMRs.png}
-	\end{center}
-	\caption{\small \textbf{Detection of differentially methylated regions between oligodendrocytes and neural stem cells.}\\
-		\textbf{(A)} DMRs detected between 58 oligodendrocytes (ODs) and 130 neural stem cells (NSCs) from \citet{kremer_scnmt} (top) and DMRs detected in the same data with randomly permuted cell labels (bottom, used to estimate FDR and determine adjusted p-values).
-		\textbf{(B)} Enrichment of GO terms associated with DMRs lowly methylated in ODs (top) or NSCs (bottom).
-		Depicted are the top 5 GO terms of the "Biological Process" (GO:BP) and "Cellular Component" (GO:CC) GO category, and their binomial test p-value and enrichment, as reported by GREAT \citep{mclean2010great}.
-		\textbf{(C)} Mean methylation of NSCs and ODs at an exemplary DMR.
-		Each point corresponds to a cell, black lines denote the median.
-		\textbf{(D)} Detailed view of the DMR (yellow dashed rectangle) from C in pseudo-bulk samples consisting of NSCs or ODs.
-		Vertical bars represent CpG sites.
-	}
-	\label{fig:dmr}
-\end{figure*}
 
@@
 \section*{The \textit{MethSCAn} software toolkit}
-
-\begin{figure*}[t]
-    \begin{center}
-        \includegraphics[width=.8\textwidth]{figures/Fig_workflow.png}
-    \end{center}
-    \caption{\small \textbf{Overview of the functionalities implemented in the \textit{MethSCAn} package.}\\
-        \texttt{methscan prepare} parses methylation files produced by common bisulfite sequencing mappers and stores their contents in a compressed format optimized for efficient access to genomic intervals.
-        \textit{MethSCAn} also produces cell-wise summary statistics and quality plots (here: average methylation around the transcription start site) that are used to detect low-quality cells.
-        These cells can be discarded with \texttt{methscan filter}.
-        To obtain a methylation matrix, similar to the count matrices used in scRNA-seq, the user must first decide in which genomic intervals methylation should be quantified.
-        The user may either provide genomic regions of \emph{a priori} interest, or they may choose to discover VMRs (variably methylated regions) in the data with \texttt{methscan scan}.
-        The resulting methylation matrix can then be used for downstream analysis such as cell clustering and dimensionality reduction.
-        Identified groups of cells (e.g. clusters) can then be scanned for DMRs with \texttt{methscan diff}.
-    }
-    \label{fig:workflow}
-\end{figure*}
 
@@
 
-\subsection*{Data Availability}
+\subsection*{Acknowledgments}
+SA and LPMK acknowledge funding by the Klaus Tschira Foundation (project 00.022.2019).
+LPMK, MMB, LK, SC, and AMV acknowledge funding from the European Commission via ERC grant 771376, from the DFG via SFB grant 873, and from the DKFZ.
+We thank Alexey Uvarovskii for code refactoring and helpful suggestions on the source code.
 
-All data used to benchmark and showcase the \textit{MethSCAn} software is publicly available under the following NCBI Gene Expression Omnibus accessions:
-Single-cell multi-omics of the murine forebrain \citep{kremer_scnmt}: \texttt{GSE210806};
-mouse neurons \citep{luo2017single}: \texttt{GSE97179};
-colorectal cancer \citep{bian2018single}: \texttt{GSE97693};
-mouse gastrulation \citep{argelaguet2019gastru}: \texttt{GSE121708};
-100k brain cells \citep{liu2021dna}: \texttt{GSE132489}.
+\subsection*{Author Contributions}
+MMB developed and implemented DMR detection under supervision of LPMK and SA.
+SO, LK and LPMK performed benchmarks.
+LPMK and SA wrote the manuscript.
+% TO DO: complete this
 
-\subsection*{Code Availability}
+\subsection*{Competing Interests}
+The authors declare no competing interests.
 
-\textit{MethSCAn} is free and open source.
-The package is available on the Python Package Index (PyPI) and can be installed with the Python package installer \texttt{pip}.
-The source code is hosted at \href{https://github.com/anders-biostat/MethSCAn}{https://github.com/anders-biostat/MethSCAn}, where we also provide detailed documentation including a tutorial that demonstrates a complete \textit{MethSCAn} analysis on an example data set.
+\section*{Main Figures and Captions}
+
+\begin{figure*}[p]
+	\begin{center}
+		\includegraphics[width=0.7\textwidth]{figures/Fig_residuals_AB.png}\\
+	\end{center}
+	\caption{\small \textbf{Improved quantification of DNA methylation in a given genomic interval.}\\
+		\textbf{(A)} Depicted is a genomic interval (between vertical blue lines) along a chromosome, for which DNA methylation is to be quantified.
+		Two cells cover differing parts of the interval with one read each.
+		If one simply counts for each cell which fraction of the covered CpG sites are methylated, one obtains very different values for the two cells.
+		\textbf{(B)} By averaging each CpG site's methylation over all cells and subsequent smoothing, the thick black ``average methylation curve'' is obtained.
+		To quantify the methylation of cell 1 from (A) relative to this average over all cells, we propose to use the cell's residuals to the smoothed curve (lengths of the vertical red lines) and take their average, counting residuals of methylated CpGs positive and residuals of unmethylated CpGs negative.}
+	\label{fig:smoothres}
+\end{figure*}
+
+\begin{figure}[p]
+	\begin{center}
+		\includegraphics[width=\columnwidth]{figures/Fig_sliding.png}
+	\end{center}
+	\caption{\small \textbf{Finding variably methylated regions.}\\
+		The chromosomes are divided up into overlapping windows (first five shown at the bottom), and for each window, the cells' methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
+		Then, the variance of these values is calculated (each point represents one of the overlapping windows), a threshold (dashed line) is chosen such that a chosen quantile of windows have a variance exceeding the threshold.
+		Windows with above-threshold variance are merged if they overlap, yielding the ``variably methylated regions'' (VMRs).}
+	\label{fig:vmr}
+\end{figure}
+
+\begin{figure*}[p]
+	\begin{center}
+		\includegraphics[width=\textwidth]{figures/Fig_correlation.png}
+	\end{center}
+	\caption{\small \textbf{Correlation of DNA methylation and gene expression}.\\
+		\textbf{(A)} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red): for most promoters, correlation is weak.
+		Promoters are defined as intervals \textpm2~kb around the TSS.
+		\textbf{(B)} Mean methylation near the gene \textit{Htra1}.
+		Cells are assigned to 4 groups based on \textit{Htra1} expression (group 0: cells with no \textit{Htra1} expression, groups 1-3: cells that express \textit{Htra1}, divided into three equally large groups with group 3 having the highest expression).
+		Data from \citet{kremer_scnmt}.}
+	\label{fig:correlation}
+\end{figure*}
+
+\begin{figure*}[p]
+	\begin{center}
+		\includegraphics[width=0.9\textwidth]{figures/Fig_benchmark.png}
+	\end{center}
+	\caption{\small \textbf{Benchmark of our methods on single-cell multi-omic data of cells of the murine forebrain}.\\
+		\textbf{(A)} Cell labels based on clustering of single-cell transcriptomes from \citet{kremer_scnmt}.
+		\textbf{(B, C)} Exemplary UMAPs obtained when analyzing the data set with conventional methods based on genome tiling (B) or our proposed methods (C).
+		\textbf{(D, E)} Mean degree of cell type separation (neighbor score, E, computed in 15-dimensional PC space) obtained when analyzing single-cell methylomes with different combinations of methods.
+		Either VMRs, ENCODE regulatory elements, 100~kb genomic tiles or promoter regions (TSS\textpm2kb) were subjected to iterative PCA and UMAP.
+		CpG methylation in these intervals was either quantified by averaging (dotted lines) or using the shrunken mean of the residuals as proposed in this work (solid lines).
+		The full 1566-cell data set was sub-sampled to simulate smaller data sets (x axis).
+		A higher neighbor score implies better separation of cell types inferred from single-cell transcriptomes of the same cells (A).
+		\textbf{(F)} Proportion of VMRs (blue dots) that have at least 1~bp overlap with other genomic features.
+		The black range illustrates the minimum and maximum overlap observed for 100 reshufflings of the VMRs to randomly chosen positions.
+	}
+	\label{fig:score}
+\end{figure*}
+
+\begin{figure*}[p]
+	\begin{center}
+		\includegraphics[width=.75\textwidth]{figures/Fig_DMRdetection.png}
+	\end{center}
+	\caption{\small \textbf{Identification of differentially methylated regions (DMRs).}\\
+		The chromosome is divided into windows overlapping by a small and fixed size.
+		For each window, the cell’s methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
+		Then the t statistic of these values is obtained as a measure of methylation difference between two groups of cells (each point represents one of the windows).
+		Upper and lower thresholds (dashed lines) are chosen such that a defined quantile of windows have a t statistic above or below the threshold.
+		Windows exceeding either threshold are merged if they overlap, yielding the differentially methylated regions (DMRs).}
+	\label{fig:dmrscan}
+\end{figure*}
+
+\begin{figure*}[p]
+	\begin{center}
+		\includegraphics[width=.7\textwidth]{figures/Fig_DMRs.png}
+	\end{center}
+	\caption{\small \textbf{Detection of differentially methylated regions between oligodendrocytes and neural stem cells.}\\
+		\textbf{(A)} DMRs detected between 58 oligodendrocytes (ODs) and 130 neural stem cells (NSCs) from \citet{kremer_scnmt} (top) and DMRs detected in the same data with randomly permuted cell labels (bottom, used to estimate FDR and determine adjusted p-values).
+		\textbf{(B)} Enrichment of GO terms associated with DMRs lowly methylated in ODs (top) or NSCs (bottom).
+		Depicted are the top 5 GO terms of the "Biological Process" (GO:BP) and "Cellular Component" (GO:CC) GO category, and their binomial test p-value and enrichment, as reported by GREAT \citep{mclean2010great}.
+		\textbf{(C)} Mean methylation of NSCs and ODs at an exemplary DMR.
+		Each point corresponds to a cell, black lines denote the median.
+		\textbf{(D)} Detailed view of the DMR (yellow dashed rectangle) from C in pseudo-bulk samples consisting of NSCs or ODs.
+		Vertical bars represent CpG sites.
+	}
+	\label{fig:dmr}
+\end{figure*}
+
+\begin{figure*}[p]
+	\begin{center}
+		\includegraphics[width=.8\textwidth]{figures/Fig_workflow.png}
+	\end{center}
+	\caption{\small \textbf{Overview of the functionalities implemented in the \textit{MethSCAn} package.}\\
+		\texttt{methscan prepare} parses methylation files produced by common bisulfite sequencing mappers and stores their contents in a compressed format optimized for efficient access to genomic intervals.
+		\textit{MethSCAn} also produces cell-wise summary statistics and quality plots (here: average methylation around the transcription start site) that are used to detect low-quality cells.
+		These cells can be discarded with \texttt{methscan filter}.
+		To obtain a methylation matrix, similar to the count matrices used in scRNA-seq, the user must first decide in which genomic intervals methylation should be quantified.
+		The user may either provide genomic regions of \emph{a priori} interest, or they may choose to discover VMRs (variably methylated regions) in the data with \texttt{methscan scan}.
+		The resulting methylation matrix can then be used for downstream analysis such as cell clustering and dimensionality reduction.
+		Identified groups of cells (e.g. clusters) can then be scanned for DMRs with \texttt{methscan diff}.
+	}
+	\label{fig:workflow}
+\end{figure*}
 
 %\phantomsection
+\clearpage
 \section*{Methods}
@@
 
-\phantomsection\addcontentsline{toc}{section}{Back Matter}
-\subsection*{Acknowledgments}
+\subsection*{Data Availability}
 
-SA and LPMK acknowledge funding by the Klaus Tschira Foundation (project 00.022.2019).
-LPMK, MMB, LK, SC, and AMV acknowledge funding from the European Commission via ERC grant 771376, from the DFG via SFB grant 873, and from the DKFZ.
-We thank Alexey Uvarovskii for code refactoring and helpful suggestions on the source code.
-\bigskip
+All data used to benchmark and showcase the \textit{MethSCAn} software is publicly available under the following NCBI Gene Expression Omnibus accessions:
+Single-cell multi-omics of the murine forebrain \citep{kremer_scnmt}: \texttt{GSE210806};
+mouse neurons \citep{luo2017single}: \texttt{GSE97179};
+colorectal cancer \citep{bian2018single}: \texttt{GSE97693};
+mouse gastrulation \citep{argelaguet2019gastru}: \texttt{GSE121708};
+100k brain cells \citep{liu2021dna}: \texttt{GSE132489}.
+
+\subsection*{Code Availability}
+
+\textit{MethSCAn} is free and open source.
+The package is available on the Python Package Index (PyPI) and can be installed with the Python package installer \texttt{pip}.
+The source code is hosted at \href{https://github.com/anders-biostat/MethSCAn}{https://github.com/anders-biostat/MethSCAn}, where we also provide detailed documentation including a tutorial that demonstrates a complete \textit{MethSCAn} analysis on an example data set.
 

### 2024-04-29 paper.tex
minor
shorten paper

@@
 
-\subsubsection{Correlating VMR methylation with gene expression}
+\subsubsection*{Correlating VMR methylation with gene expression}
 
@@
 
-\subsubsection{Improved identification of cell types}
+\subsubsection*{Improved identification of cell types}
 
@@
 
-\subsubsection{Robustness to parameter changes}
+\subsubsection*{Robustness to parameter changes}
 
@@
 
-\subsubsection{Further applications}
+\subsubsection*{Further applications}
 
@@
 
-We have implemented the approach just described in a Python package with a command line interface, \textit{MethSCAn} (ANalysis of Single-Cell METHylation data), which also offers a number of other functionalities for the analysis of scBS data (Fig.~\ref{fig:workflow}).
-The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark}, methylpy \citep{methylpy} or BISCUIT \citep{biscuit}.
-For each cell, these tools produce a text file that lists the methylation status of all CpG sites.
-Since it is inconvenient to work with hundreds or thousands of text files, \textit{MethSCAn} provides the \texttt{prepare} command which parses these methylation files and stores their content in a compressed format that enables efficient access to all CpG sites of the genome.
-\texttt{methscan prepare} is flexible and accepts all tabular input formats including Bismark .cov-files, methylpy .allc-files, or custom user-defined formats.
-
-In brief, we store methylation of each chromosome in a matrix where each column represents a cell and each row represents a base pair.
-Methylated and unmethylated sites are coded as $+1$ and $-1$, respectively.
-Since the vast majority of values in this matrix are missing due to the sparsity of scBS data (and because rows for base-pairs not corresponding to a CpG site contain no data), we encode missing values as zero and then store the data in Compressed Sparse Row (CSR) format.
-This format does not explicitly store zeroes (here: missing values) and is optimized for row-wise access, which results in significant compression and allows fast access to the methylation status of genomic intervals.
-
-\texttt{methscan prepare} also computes a number of summary statistics for each cell, including the mean genome-wide methylation level and the number of observed CpG sites, i.e.\ the number of CpG sites that have sequencing coverage.
-These summary statistics can be used to detect cells with poor quality.
-The quality of each single cell methylome can furthermore be inspected with \texttt{methscan profile}, which computes the average methylation profile of a set of user-defined genomic regions such as transcription start sites (TSS) at single-cell resolution.
-The TSS profile is a useful quality control plot since methylation shows a characteristic dip roughly \textpm1~kb around the transcription start site in mammalian genomes.
-Cells that do not show this pattern, or cells with few observed CpG sites, may then be discarded with \texttt{methscan filter}.
-
-After quality control, the user has access to the genome-wide VMR detection approach described earlier, using the \texttt{methscan scan} command.
-This produces a BED file that lists the genome coordinates of VMRs.
-To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED file can be used as input for \texttt{methscan matrix}, which quantifies methylation at genomic intervals in all cells.
-The command produces both, the simple percentage of methylated CpG sites, as well as our proposed methylation measure, the shrunken mean of the residuals, which is more robust to variation in sequencing coverage or stochastic differences in read position between cells.
-We note that both \texttt{matrix} and \texttt{profile} accept any valid BED file as input, which means that the user can quantify and visualize methylation at any set of genomic features of interest, such as promoters, enhancers or transcription factor binding sites, obtaining one profile plot per cell.
-The methylation matrix produced by the \texttt{matrix} function can then serve as input for established methods used for the exploration of single-cell data, such as dimensionality reduction and cell clustering.
-
-Finally, after annotation and exploration of the data set, the user may specify two groups of cells for DMR detection with \texttt{methscan diff}.
-This command produces a BED file listing the genome coordinates, adjusted p-values, as well as several other metrics associated with DMRs.
-These DMRs may then be associated with nearby genes, or used for GO enrichment with tools such as GREAT \citep{mclean2010great} to enable functional interpretation.
+We have implemented the methods just described in a Python package with a command line interface, \textit{MethSCAn} (Single-Cell ANalysis of METHylation data), which also offers a number of other functionalities for the analysis of scBS data.
+Fig.~\ref{fig:workflow} illustrates a typical scBS data analysis with \texttt{MethSCAn} and provides an overview of the implemented core functionalities.
+The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark}, methylpy \citep{methylpy} or BISCUIT \citep{biscuit}, which are first stored in a compressed format that enables efficient access to all CpG sites of the genome (see Methods).
 
@@
 \subsection*{Author Contributions}
+
 MMB developed and implemented DMR detection under supervision of LPMK and SA.
@@
 	\caption{\small \textbf{Overview of the functionalities implemented in the \textit{MethSCAn} package.}\\
-		\texttt{methscan prepare} parses methylation files produced by common bisulfite sequencing mappers and stores their contents in a compressed format optimized for efficient access to genomic intervals.
-		\textit{MethSCAn} also produces cell-wise summary statistics and quality plots (here: average methylation around the transcription start site) that are used to detect low-quality cells.
-		These cells can be discarded with \texttt{methscan filter}.
-		To obtain a methylation matrix, similar to the count matrices used in scRNA-seq, the user must first decide in which genomic intervals methylation should be quantified.
-		The user may either provide genomic regions of \emph{a priori} interest, or they may choose to discover VMRs (variably methylated regions) in the data with \texttt{methscan scan}.
-		The resulting methylation matrix can then be used for downstream analysis such as cell clustering and dimensionality reduction.
-		Identified groups of cells (e.g. clusters) can then be scanned for DMRs with \texttt{methscan diff}.
+		Common bisulfite sequencing mappers produce one methylation report file per cell.
+		As it is inconvenient to work with thousands of text files, \texttt{methscan prepare} first stores their contents in a compressed format optimized for efficient access to genomic intervals.
+		This command also produces cell-wise summary statistics that can be used to discard low-quality cells with \texttt{methscan filter}.
+		The quality of each cell can furthermore be assessed by visualizing the average methylation around transcription start sites, which are known to show a characteristic dip in mammalian genomes, using \texttt{methscan profile}.
+		This command can also visualize methylation at other genomic features of interest, such as enhancers or transcription factor binding sites.
+		To obtain a methylation matrix analogous to scRNA-seq count matrices, the user must first decide in which genomic intervals methylation should be quantified.
+		The user may either provide genomic regions of \emph{a priori} interest as a BED file, or they may choose to discover VMRs (variably methylated regions) in the data with \texttt{methscan scan}.
+		The \texttt{methscan matrix} command then quantifies methylation at these genomic intervals in all cells.
+		It provides two metrics:
+		the simple percentage of methylated CpG sites and our proposed measure, the shrunken mean of the residuals, which is more robust to variation in sequencing coverage and stochastic differences in read position between cells.
+		The resulting methylation matrix can then serve as input for established methods used for the exploration of single-cell data, such as dimensionality reduction and cell clustering.
+		Finally, after annotation and exploration of the data set, the user may specify two groups of cells for DMR (differentially methylated region) detection with \texttt{methscan diff}.
+		This command produces a BED file listing the genome coordinates, adjusted p-values, as well as several other metrics associated with DMRs.
+		These DMRs may then be associated with nearby genes, or used for GO enrichment with tools such as GREAT \citep{mclean2010great} to enable functional interpretation.
 	}
@@
 
-The function \texttt{methscan prepare} reads a set of methylation files (e.g. produced by Bismark) and produces a file that stores the matrix $x$ in a space-efficient format, as follows:
+The function \texttt{methscan prepare} reads a set of methylation files (e.g. produced by Bismark) and produces one file per chromosome.
+These files store the matrix $x$, where each column represents a cell and each row represents a base pair, in a space-efficient format as follows:
 $x$ is represented as a SciPy sparse matrix \citep{SciPy}, encoding the actual values 0, 1, and NA as -1, 1, and 0, respectively.
-Coding NA (the most common value) as 0 leverages SciPy's sparse matrix storage.
+Since the vast majority of values in this matrix are missing due to the sparsity of scBS data (and because rows for base-pairs not corresponding to a CpG site contain no data), we encode missing values as zero and then store the data in Compressed Sparse Row (CSR) format.
+This format does not explicitly store zeroes (here: missing values) and is optimized for row-wise access, which results in significant compression and allows fast access to the methylation status of genomic intervals.
 In all that follows here, any mention of $x$ will, however, always mean the encoding as $x_{ij}\in\{0,1,\text{NA}\}$.

### 2024-04-30 paper.tex
merge Fig5+6, mention tutorial

@@
 \usepackage[utf8]{inputenc}
-\usepackage[numbers]{natbib}
+\usepackage[super,sort&compress]{natbib}
 \bibliographystyle{simon}
@@
 
-\newcommand\blfootnote[1]{%
-    \begingroup
-    \renewcommand\thefootnote{}\footnote{#1}%
-    \addtocounter{footnote}{-1}%
-    \endgroup
-}
-
 \begin{document}
@@
 Ana Martin-Villalba\textsuperscript{2,*},
-Simon Anders\textsuperscript{1,*}\\[1ex]
-{\footnotesize
-\textsuperscript{1} BioQuant Centre, University of Heidelberg, Heidelberg, Germany\\[-.5ex] 
-\textsuperscript{2} Division of Molecular Neurobiology, German Cancer Research Center (DKFZ), Heidelberg, Germany}
+Simon Anders\textsuperscript{1,*}\\[2em]
 }
-\vspace{4.5ex}
+\noindent\textsuperscript{1} BioQuant Centre, University of Heidelberg, Heidelberg, Germany\\
+\textsuperscript{2} Division of Molecular Neurobiology, German Cancer Research Center (DKFZ), Heidelberg, Germany\\
+\textsuperscript{*} e-mails:\\
+\texttt{l.kremer@dkfz-heidelberg.de}\\
+\texttt{a.martin-villalba@dkfz-heidelberg.de}\\
+\texttt{simon.anders@bioquant.uni-heidelberg.de}
 
@@
 Finally, we present \textit{MethSCAn}, a comprehensive software toolkit to perform scBS data analysis, using the described methods.
-\blfootnote{\hspace{-.5cm}\raggedright\textsuperscript{*} e-mails:  \texttt{l.kremer@dkfz-heidelberg.de}, \hspace{1cm}\texttt{a.martin-villalba@dkfz-heidelberg.de}, \hspace{1cm}\texttt{simon.anders@bioquant.uni-heidelberg.de}}
 
@@
 For instance, it is known  since long that CpG-rich promoters of housekeeping genes are unmethylated, and that a large proportion of the remaining genome is highly methylated regardless of cell type \citep{bird1986cpg}.
-In contrast, DNA methylation at certain genomic features such as enhancers is more dynamic \citep[e.g.,][]{argelaguet2019gastru}, and thus more variable across cells.
+In contrast, DNA methylation at certain genomic features such as enhancers is more dynamic \citep{argelaguet2019gastru}, and thus more variable across cells.
 Only the latter regions are of value for our goal of quantitating dissimilarity between cells.
@@
 Specifically, we selected four different sets of genomic intervals at which CpG methylation is to be quantified:
-either VMRs detected with our approach, 100~kb genomic tiles, promoter regions (i.e., transcription start site (TSS)~\textpm~2000~bp), or candidate cis-regulatory elements \citep[ENCODE cCREs,][]{encode2020expanded}.
+either VMRs detected with our approach, 100~kb genomic tiles, promoter regions (i.e., transcription start site (TSS)~\textpm~2000~bp), or candidate cis-regulatory elements (ENCODE cCREs\citep{encode2020expanded}).
 To quantify methylation at these features, we either simply averaged in these intervals, obtaining methylation percentages, or we calculated the shrunken mean of residuals, as described earlier.
@@
 We repeated this benchmark on an additional three published single-cell methylome data sets (Extended Data Fig.~\ref{fig:score_luo}).
-These include neuronal sub-types of the murine cortex \citep[][using cell type labels derived from CH-methylation in genomic tiles instead of CpG-methylation as ground truth]{luo2017single}, cells isolated from mouse embryos during the onset of gastrulation \citep[][using RNA-derived cell clusters or alternatively embryonic stage as ground truth]{argelaguet2019gastru} and human colorectal cancer cells \citep[][using sampling region as ground truth]{bian2018single}.
+These include neuronal sub-types of the murine cortex \citep{luo2017single} (using cell type labels derived from CH-methylation in genomic tiles instead of CpG-methylation as ground truth), cells isolated from mouse embryos during the onset of gastrulation \citep{argelaguet2019gastru} (using RNA-derived cell clusters or alternatively embryonic stage as ground truth) and human colorectal cancer cells \citep{bian2018single} (using sampling region as ground truth).
 Again, we subjected each data set to all possible combinations of genomic feature selection and methylation quantification.
@@
 However, to date no approach to detect DMRs in scBS data has been reported.
-To enable DMR detection in scBS data, we thus developed an approach that detects DMRs of variable size between two groups of cells, and controls the false discovery rate (Fig.~\ref{fig:dmrscan}):
+To enable DMR detection in scBS data, we thus developed an approach that detects DMRs of variable size between two groups of cells, and controls the false discovery rate (Fig.~\ref{fig:dmr}A):
 
@@
 We first selected two cell populations from the healthy murine ventricular-subventricular zone: neural stem cells (NSCs, 130 cells) and oligodendrocytes (58 cells).
-Then we used the method just described to identify DMRs between these two cell types (Fig.~\ref{fig:dmr}A).
+Then we used the method just described to identify DMRs between these two cell types (Fig.~\ref{fig:dmr}B).
 Repeating this after permuting the cell-type labels, in order to obtain a null distribution of the t statistics, yielded DMRs with much weaker methylation differences.
-Consequently, we could assign to many of the DMRs detected in the unpermuted data a low adjusted p-value (colors in Fig.~\ref{fig:dmr}A).
+Consequently, we could assign to many of the DMRs detected in the unpermuted data a low adjusted p-value (colors in Fig.~\ref{fig:dmr}B).
 
@@
 
-Fig.~\ref{fig:dmr}C-D depicts an exemplary DMR, located at the gene body of myelin-basic protein (\textit{Mbp}), the major component of myelin that is essential for myelination of neuronal axons \citep{mbp}.
+Fig.~\ref{fig:dmr}D-E depicts an exemplary DMR, located at the gene body of myelin-basic protein (\textit{Mbp}), the major component of myelin that is essential for myelination of neuronal axons \citep{mbp}.
 Our results suggest that oligodendrocyte-specific gene expression of \textit{Mbp} is supported by low methylation at the detected DMR.
@@
 The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark}, methylpy \citep{methylpy} or BISCUIT \citep{biscuit}, which are first stored in a compressed format that enables efficient access to all CpG sites of the genome (see Methods).
+A tutorial that showcases the analysis of a small example data set can be found at \url{https://github.com/anders-biostat/MethSCAn}.
 
@@
 We suggest iterative PCA as an approach to map the count matrix to a reduced-dimensional space overcoming the abundance of missing values.
-Alternatively, one may use for this purpose established tools based on matrix factorisation such as MOFA+, \citep[][included in our benchmarks]{argelaguet2020mofa} or LIGER \citep{welch2019single}.
+Alternatively, one may use for this purpose established tools based on matrix factorisation such as MOFA+, \citep{argelaguet2020mofa} (included in our benchmarks) or LIGER \citep{welch2019single}.
 Once a low dimensional embedding is obtained, one can switch to well established methods from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc., e.g., by using either overall toolkits like Seurat \citep{seurat5}, ScanPy/EpiScanPy \citep{Wolf_2018,danese2021episcanpy} or any of the many available tools for specific tasks.
@@
 	\begin{center}
-		\includegraphics[width=\columnwidth]{figures/Fig_sliding.png}
+		\includegraphics[width=.7\columnwidth]{figures/Fig_sliding.png}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[width=.75\textwidth]{figures/Fig_DMRdetection.png}
+		\includegraphics[width=.8\textwidth]{figures/Fig_DMRs.png}
 	\end{center}
-	\caption{\small \textbf{Identification of differentially methylated regions (DMRs).}\\
-		The chromosome is divided into windows overlapping by a small and fixed size.
+	\caption{\small \textbf{Detection of differentially methylated regions (DMRs).}\\
+		\textbf{(A)} To identify DMRs, the chromosome is first divided into windows overlapping by a small and fixed size.
 		For each window, the cell’s methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
-		Then the t statistic of these values is obtained as a measure of methylation difference between two groups of cells (each point represents one of the windows).
+		Then the t statistic of these values is obtained as a measure of methylation difference between two groups of cells that were selected by the user (each point represents one of the windows).
 		Upper and lower thresholds (dashed lines) are chosen such that a defined quantile of windows have a t statistic above or below the threshold.
-		Windows exceeding either threshold are merged if they overlap, yielding the differentially methylated regions (DMRs).}
-	\label{fig:dmrscan}
-\end{figure*}
-
-\begin{figure*}[p]
-	\begin{center}
-		\includegraphics[width=.7\textwidth]{figures/Fig_DMRs.png}
-	\end{center}
-	\caption{\small \textbf{Detection of differentially methylated regions between oligodendrocytes and neural stem cells.}\\
-		\textbf{(A)} DMRs detected between 58 oligodendrocytes (ODs) and 130 neural stem cells (NSCs) from \citet{kremer_scnmt} (top) and DMRs detected in the same data with randomly permuted cell labels (bottom, used to estimate FDR and determine adjusted p-values).
-		\textbf{(B)} Enrichment of GO terms associated with DMRs lowly methylated in ODs (top) or NSCs (bottom).
+		Windows exceeding either threshold are merged if they overlap, yielding DMRs that are lowly methylated in either group of cells.
+		\textbf{(B)} DMRs detected between 58 oligodendrocytes (ODs) and 130 neural stem cells (NSCs) from \citet{kremer_scnmt} (top) and DMRs detected in the same data with randomly permuted cell labels (bottom, used to estimate FDR and determine adjusted p-values).
+		\textbf{(C)} Enrichment of GO terms associated with DMRs lowly methylated in ODs (top) or NSCs (bottom).
 		Depicted are the top 5 GO terms of the "Biological Process" (GO:BP) and "Cellular Component" (GO:CC) GO category, and their binomial test p-value and enrichment, as reported by GREAT \citep{mclean2010great}.
-		\textbf{(C)} Mean methylation of NSCs and ODs at an exemplary DMR.
+		\textbf{(D)} Mean methylation of NSCs and ODs at an exemplary DMR.
 		Each point corresponds to a cell, black lines denote the median.
-		\textbf{(D)} Detailed view of the DMR (yellow dashed rectangle) from C in pseudo-bulk samples consisting of NSCs or ODs.
+		\textbf{(E)} Detailed view of the DMR (yellow dashed rectangle) from C in pseudo-bulk samples consisting of NSCs or ODs.
 		Vertical bars represent CpG sites.
@@
     \textbf{(A)} UMAPs obtained when analysing data from \citet{luo2017single} with three different strategies for producing a methylation matrix:
-	top: averaging CH-methylation (mCH) in 100~kb genomic tiles \citep[as done in][to obtain the depicted cell type labels]{luo2017single};
+	top: averaging CH-methylation (mCH) in 100~kb genomic tiles (as done in \citep{luo2017single} to obtain the depicted cell type labels);
 	center: using the shrunken mean of residuals to quantify mCH in mCH-VMRs;

### 2024-04-30 paper.tex
update contributions, fix references

@@
 \usepackage[utf8]{inputenc}
-\usepackage[super,sort&compress]{natbib}
-\bibliographystyle{simon}
+\usepackage[super,sort]{natbib}
+%\bibliographystyle{simon}
+\bibliographystyle{unsrtnat}
 \setlength{\bibsep}{0.3ex plus 0.2ex}
@@
 
+\section*{Main Text}
 \section*{Introduction}
@@
 
-Fig.~\ref{fig:dmr}D-E depicts an exemplary DMR, located at the gene body of myelin-basic protein (\textit{Mbp}), the major component of myelin that is essential for myelination of neuronal axons \citep{mbp}.
+Fig.~\ref{fig:dmr}D-E depicts an exemplary DMR, located at the gene encoding myelin-basic protein (\textit{Mbp}), the major component of myelin that is essential for myelination of neuronal axons \citep{mbp}.
 Our results suggest that oligodendrocyte-specific gene expression of \textit{Mbp} is supported by low methylation at the detected DMR.
@@
 
-\subsection*{Acknowledgments}
+\subsection*{Acknowledgements}
 SA and LPMK acknowledge funding by the Klaus Tschira Foundation (project 00.022.2019).
-LPMK, MMB, LK, SC, and AMV acknowledge funding from the European Commission via ERC grant 771376, from the DFG via SFB grant 873, and from the DKFZ.
+LPMK, MMB, LK, SC, and AMV acknowledge funding from the European Commission via ERC grant 771376 (to AMV), from the DFG via SFB 873, and from the DKFZ.
 We thank Alexey Uvarovskii for code refactoring and helpful suggestions on the source code.
@@
 
-MMB developed and implemented DMR detection under supervision of LPMK and SA.
-SO, LK and LPMK performed benchmarks.
+LPMK and SA conceived and worked out the method.
+LPKM wrote the software package, with contributions from MMB (differential methylation functionality) and LKü.
+LPMK, SO and LKü prepared the use-case demonstrations and performed the benchmarks.
+SC and AMV contributed experimental data and biological expertise.
+LPMK, AMV and SA contributed supervision and project management.
 LPMK and SA wrote the manuscript.
-% TO DO: complete this
 
@@
 		\textbf{(C)} Enrichment of GO terms associated with DMRs lowly methylated in ODs (top) or NSCs (bottom).
-		Depicted are the top 5 GO terms of the "Biological Process" (GO:BP) and "Cellular Component" (GO:CC) GO category, and their binomial test p-value and enrichment, as reported by GREAT \citep{mclean2010great}.
+		Depicted are the top 5 GO terms of the "Biological Process" (GO:BP) and "Cellular Component" (GO:CC) GO category, and their binomial test p-value (two-sided, not adjusted for multiple comparisons) and enrichment, as reported by GREAT \citep{mclean2010great}.
 		\textbf{(D)} Mean methylation of NSCs and ODs at an exemplary DMR.

### 2024-04-30 figures/convert_to_pdf.sh
add figures as pdf

@@ -1,0 +1,4 @@
+#!/bin/bash
+
+# sudo apt install librsvg2-bin
+for file in *svg; do rsvg-convert -f pdf -o ${file%.*}.pdf $file; done

### 2024-04-30 figures/convert_to_pdf.sh
fix pdf

@@ -1,4 +1,8 @@
 #!/bin/bash
 
 # sudo apt install librsvg2-bin
-for file in *svg; do rsvg-convert -f pdf -o ${file%.*}.pdf $file; done
+#for file in *svg; do rsvg-convert -f pdf -o ${file%.*}.pdf $file; done
+
+for file in *svg; do inkscape $file --export-pdf=${file%.*}.pdf; done
+
+

### 2024-04-30 paper.tex
two minor changes
maybe we dont need 3 references for seurat :D

@@
 \textsuperscript{2} Division of Molecular Neurobiology, German Cancer Research Center (DKFZ), Heidelberg, Germany\\
-\textsuperscript{*} e-mails:\\
+\textsuperscript{*} corresponding authors. e-mails:\\
 \texttt{l.kremer@dkfz-heidelberg.de}\\
@@
 To this end, one needs to establish which cells are similar to each other, i.e., quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile.
-A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or Scanpy \citep{Wolf_2018}, is as follows:
+A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{seurat5} or Scanpy \citep{Wolf_2018}, is as follows:
 One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm.
@@
 LPMK, AMV and SA contributed supervision and project management.
-LPMK and SA wrote the manuscript.
+LPMK and SA wrote the paper.
 
@@
 
-%\phantomsection
 \clearpage
@@
 For the multi-omic data set, we repeated the single-cell transcriptomics analysis described in \citet{kremer_scnmt} with two adjustments:
-We did not filter off-target cells such as endothelial cells, and we annotated cell types using Leiden clustering with the Seurat \citep{seurat} function \texttt{FindClusters(resolution = 0.1)}.
+We did not filter off-target cells such as endothelial cells, and we annotated cell types using Leiden clustering with the Seurat \citep{seurat5} function \texttt{FindClusters(resolution = 0.1)}.
 The score, based on the $\Gamma$-score \citep{Kireeva_2014}, varies between 0 and 1, where higher scores reflect better separation of cell types.

### 2024-04-30 figures/convert_to_pdf.sh

@@ -1,8 +1,5 @@
 #!/bin/bash
 
-# sudo apt install librsvg2-bin
-#for file in *svg; do rsvg-convert -f pdf -o ${file%.*}.pdf $file; done
-
 for file in *svg; do inkscape $file --export-pdf=${file%.*}.pdf; done
 
 

### 2024-04-30 paper.tex
use pdf, include references

@@
 \usepackage[super,sort]{natbib}
-%\bibliographystyle{simon}
 \bibliographystyle{unsrtnat}
 \setlength{\bibsep}{0.3ex plus 0.2ex}
-%\setcitestyle{aysep={}} 
 \usepackage[dvipsnames]{xcolor}
@@
 	\begin{center}
-		\includegraphics[width=0.7\textwidth]{figures/Fig_residuals_AB.png}\\
+		\includegraphics[width=0.7\textwidth]{figures/Fig_residuals_AB.pdf}\\
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[width=.7\columnwidth]{figures/Fig_sliding.png}
+		\includegraphics[width=.7\columnwidth]{figures/Fig_sliding.pdf}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[width=\textwidth]{figures/Fig_correlation.png}
+		\includegraphics[width=\textwidth]{figures/Fig_correlation.pdf}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[width=0.9\textwidth]{figures/Fig_benchmark.png}
+		\includegraphics[width=0.9\textwidth]{figures/Fig_benchmark.pdf}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[width=.8\textwidth]{figures/Fig_DMRs.png}
+		\includegraphics[width=.8\textwidth]{figures/Fig_DMRs.pdf}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[width=.8\textwidth]{figures/Fig_workflow.png}
+		\includegraphics[width=.8\textwidth]{figures/Fig_workflow.pdf}
 	\end{center}
@@
 
-{\phantomsection\addcontentsline{toc}{subsection}{References}
-\small \balance \bibliography{scbs} \balance}
+\begin{thebibliography}{31}
+	\providecommand{\natexlab}[1]{#1}
+	\providecommand{\url}[1]{\texttt{#1}}
+	\expandafter\ifx\csname urlstyle\endcsname\relax
+	\providecommand{\doi}[1]{doi: #1}\else
+	\providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi
 
-\onecolumn 
+	\bibitem[Frommer et~al.(1992)Frommer, McDonald, Millar, Collis, Watt, Grigg,
+	Molloy, and Paul]{Frommer_1992}
+	M.~Frommer, L.~E. McDonald, D.~S. Millar, C.~M. Collis, F.~Watt, G.~W. Grigg,
+	P.~L. Molloy, and C.~L. Paul.
+	\newblock A genomic sequencing protocol that yields a positive display of
+	5-methylcytosine residues in individual {DNA} strands.
+	\newblock \emph{Proceedings of the National Academy of Sciences}, 89\penalty0
+	(5):\penalty0 1827--1831, mar 1992.
+	\newblock \doi{10.1073/pnas.89.5.1827}.
+	\newblock URL \url{https://doi.org/10.1073/pnas.89.5.1827}.
+
+	\bibitem[Smallwood et~al.(2014)Smallwood, Lee, Angermueller, Krueger, Saadeh,
+	Peat, Andrews, Stegle, Reik, and Kelsey]{Smallwood_2014}
+	S{\'{e}}bastien~A Smallwood, Heather~J Lee, Christof Angermueller, Felix
+	Krueger, Heba Saadeh, Julian Peat, Simon~R Andrews, Oliver Stegle, Wolf Reik,
+	and Gavin Kelsey.
+	\newblock Single-cell genome-wide bisulfite sequencing for assessing epigenetic
+	heterogeneity.
+	\newblock \emph{Nature Methods}, 11\penalty0 (8):\penalty0 817--820, jul 2014.
+	\newblock \doi{10.1038/nmeth.3035}.
+	\newblock URL \url{https://doi.org/10.1038/nmeth.3035}.
+
+	\bibitem[Hu et~al.(2016)Hu, Huang, An, Du, Hu, Xue, Zhu, Wang, Xue, and
+	Fan]{scMTseq}
+	Youjin Hu, Kevin Huang, Qin An, Guizhen Du, Ganlu Hu, Jinfeng Xue, Xianmin Zhu,
+	Cun-Yu Wang, Zhigang Xue, and Guoping Fan.
+	\newblock Simultaneous profiling of transcriptome and {DNA} methylome from a
+	single cell.
+	\newblock \emph{Genome Biology}, 17\penalty0 (1):\penalty0 1--11, 2016.
+	\newblock \doi{10.1007/978-1-4939-9240-9_21}.
+
+	\bibitem[Clark et~al.(2018)Clark, Argelaguet, Kapourani, Stubbs, Lee,
+	Alda-Catalinas, Krueger, Sanguinetti, Kelsey, Marioni, Stegle, and
+	Reik]{Clark2018}
+	Stephen~J. Clark, Ricard Argelaguet, Chantriolnt-Andreas Kapourani, Thomas~M.
+	Stubbs, Heather~J. Lee, Celia Alda-Catalinas, Felix Krueger, Guido
+	Sanguinetti, Gavin Kelsey, John~C. Marioni, Oliver Stegle, and Wolf Reik.
+	\newblock {scNMT}-seq enables joint profiling of chromatin accessibility {DNA}
+	methylation and transcription in single cells.
+	\newblock \emph{Nature Communications}, 9\penalty0 (1), February 2018.
+	\newblock \doi{10.1038/s41467-018-03149-4}.
+	\newblock URL \url{https://doi.org/10.1038/s41467-018-03149-4}.
+
+	\bibitem[Hao et~al.(2023)Hao, Stuart, Kowalski, Choudhary, Hoffman, Hartman,
+	Srivastava, Molla, Madad, Fernandez-Granda, and Satija]{seurat5}
+	Yuhan Hao, Tim Stuart, Madeline~H Kowalski, Saket Choudhary, Paul Hoffman,
+	Austin Hartman, Avi Srivastava, Gesmira Molla, Shaista Madad, Carlos
+	Fernandez-Granda, and Rahul Satija.
+	\newblock Dictionary learning for integrative, multimodal and scalable
+	single-cell analysis.
+	\newblock \emph{Nature Biotechnology}, 2023.
+	\newblock \doi{10.1038/s41587-023-01767-y}.
+
+	\bibitem[Wolf et~al.(2018)Wolf, Angerer, and Theis]{Wolf_2018}
+	F.~Alexander Wolf, Philipp Angerer, and Fabian~J. Theis.
+	\newblock {SCANPY}: large-scale single-cell gene expression data analysis.
+	\newblock \emph{Genome Biology}, 19:\penalty0 15, 2018.
+	\newblock \doi{10.1186/s13059-017-1382-0}.
+
+	\bibitem[Luo et~al.(2017)Luo, Keown, Kurihara, Zhou, He, Li, Castanon, Lucero,
+	Nery, Sandoval, et~al.]{luo2017single}
+	Chongyuan Luo, Christopher~L Keown, Laurie Kurihara, Jingtian Zhou, Yupeng He,
+	Junhao Li, Rosa Castanon, Jacinta Lucero, Joseph~R Nery, Justin~P Sandoval,
+	et~al.
+	\newblock Single-cell methylomes identify neuronal subtypes and regulatory
+	elements in mammalian cortex.
+	\newblock \emph{Science}, 357\penalty0 (6351):\penalty0 600--604, 2017.
+	\newblock \doi{10.1126/science.aan3351}.
+
+	\bibitem[Bird(1986)]{bird1986cpg}
+	Adrian~P Bird.
+	\newblock {CpG-rich} islands and the function of {DNA} methylation.
+	\newblock \emph{Nature}, 321\penalty0 (6067):\penalty0 209--213, 1986.
+	\newblock \doi{10.1038/321209a0}.
+
+	\bibitem[Argelaguet et~al.(2019)Argelaguet, Clark, Mohammed, Stapel, Krueger,
+	Kapourani, Imaz-Rosshandler, Lohoff, Xiang, Hanna,
+	et~al.]{argelaguet2019gastru}
+	Ricard Argelaguet, Stephen~J Clark, Hisham Mohammed, L~Carine Stapel, Christel
+	Krueger, Chantriolnt-Andreas Kapourani, Ivan Imaz-Rosshandler, Tim Lohoff,
+	Yunlong Xiang, Courtney~W Hanna, et~al.
+	\newblock Multi-omics profiling of mouse gastrulation at single-cell
+	resolution.
+	\newblock \emph{Nature}, 576\penalty0 (7787):\penalty0 487--491, 2019.
+	\newblock \doi{10.1038/s41586-019-1825-8}.
+
+	\bibitem[Kremer et~al.(2022)Kremer, Cerrizuela, Al~Shukairi, Ellinger, Straub,
+	Dehler, Korkmaz, Weichenhan, Plass, Anders, and
+	Martin-Villalba]{kremer_scnmt}
+	Lukas P.~M. Kremer, Santiago Cerrizuela, Mohammad~E Al~Shukairi, Tobias
+	Ellinger, Jannes Straub, Sascha Dehler, Aylin Korkmaz, Dieter Weichenhan,
+	Christoph Plass, Simon Anders, and Ana Martin-Villalba.
+	\newblock Single-cell triple-omics uncovers {DNA} methylation as key feature of
+	stemness in the healthy and ischemic adult brain.
+	\newblock \emph{bioRxiv preprint}, 2022.
+	\newblock \doi{10.1101/2022.07.13.499860}.
+
+	\bibitem[Moore et~al.(2020)Moore, Purcaro, Pratt, Epstein, Shoresh, Adrian,
+	Kawli, Davis, Dobin, et~al.]{encode2020expanded}
+	Jill~E Moore, Michael~J Purcaro, Henry~E Pratt, Charles~B Epstein, Noam
+	Shoresh, Jessika Adrian, Trupti Kawli, Carrie~A Davis, Alexander Dobin,
+	et~al.
+	\newblock Expanded encyclopaedias of {DNA} elements in the human and mouse
+	genomes.
+	\newblock \emph{Nature}, 583\penalty0 (7818):\penalty0 699--710, 2020.
+	\newblock \doi{10.1038/s41586-020-2493-4}.
+
+	\bibitem[Bian et~al.(2018)Bian, Hou, Zhou, Li, Yong, Wang, Wang, Yan, Hu, Guo,
+	et~al.]{bian2018single}
+	Shuhui Bian, Yu~Hou, Xin Zhou, Xianlong Li, Jun Yong, Yicheng Wang, Wendong
+	Wang, Jia Yan, Boqiang Hu, Hongshan Guo, et~al.
+	\newblock Single-cell multiomics sequencing and analyses of human colorectal
+	cancer.
+	\newblock \emph{Science}, 362\penalty0 (6418):\penalty0 1060--1063, 2018.
+	\newblock \doi{10.1126/science.aao3791}.
+
+	\bibitem[Argelaguet et~al.(2020)Argelaguet, Arnol, Bredikhin, Deloro, Velten,
+	Marioni, and Stegle]{argelaguet2020mofa}
+	Ricard Argelaguet, Damien Arnol, Danila Bredikhin, Yonatan Deloro, Britta
+	Velten, John~C Marioni, and Oliver Stegle.
+	\newblock {MOFA+}: a statistical framework for comprehensive integration of
+	multi-modal single-cell data.
+	\newblock \emph{Genome Biology}, 21\penalty0 (1):\penalty0 1--17, 2020.
+	\newblock \doi{10.1186/s13059-020-02015-1}.
+
+	\bibitem[Liu et~al.(2021)Liu, Zhou, Tian, Luo, Bartlett, Aldridge, Lucero,
+	Osteen, Nery, Chen, et~al.]{liu2021dna}
+	Hanqing Liu, Jingtian Zhou, Wei Tian, Chongyuan Luo, Anna Bartlett, Andrew
+	Aldridge, Jacinta Lucero, Julia~K Osteen, Joseph~R Nery, Huaming Chen, et~al.
+	\newblock {DNA} methylation atlas of the mouse brain at single-cell resolution.
+	\newblock \emph{Nature}, 598\penalty0 (7879):\penalty0 120--128, 2021.
+	\newblock \doi{10.1038/s41586-020-03182-8}.
+
+	\bibitem[Hebestreit et~al.(2013)Hebestreit, Dugas, and Klein]{Hebestreit2013}
+	Katja Hebestreit, Martin Dugas, and Hans-Ulrich Klein.
+	\newblock Detection of significantly differentially methylated regions in
+	targeted bisulfite sequencing data.
+	\newblock \emph{Bioinformatics}, 29\penalty0 (13):\penalty0 1647--1653, May
+	2013.
+	\newblock \doi{10.1093/bioinformatics/btt263}.
+	\newblock URL \url{https://doi.org/10.1093/bioinformatics/btt263}.
+
+	\bibitem[Korthauer et~al.(2018)Korthauer, Chakraborty, Benjamini, and
+	Irizarry]{dmrseq}
+	Keegan Korthauer, Sutirtha Chakraborty, Yuval Benjamini, and Rafael~A Irizarry.
+	\newblock Detection and accurate false discovery rate control of differentially
+	methylated regions from whole genome bisulfite sequencing.
+	\newblock \emph{Biostatistics}, 20\penalty0 (3):\penalty0 367--383, February
+	2018.
+	\newblock \doi{10.1093/biostatistics/kxy007}.
+	\newblock URL \url{https://doi.org/10.1093/biostatistics/kxy007}.
+
+	\bibitem[McLean et~al.(2010)McLean, Bristor, Hiller, Clarke, Schaar, Lowe,
+	Wenger, and Bejerano]{mclean2010great}
+	Cory~Y McLean, Dave Bristor, Michael Hiller, Shoa~L Clarke, Bruce~T Schaar,
+	Craig~B Lowe, Aaron~M Wenger, and Gill Bejerano.
+	\newblock {GREAT} improves functional interpretation of cis-regulatory regions.
+	\newblock \emph{Nature Biotechnology}, 28\penalty0 (5):\penalty0 495--501,
+	2010.
+	\newblock \doi{10.1038/nbt.1630}.
+
+	\bibitem[Boggs(2006)]{mbp}
+	JM~Boggs.
+	\newblock Myelin basic protein: a multifunctional protein.
+	\newblock \emph{Cellular and Molecular Life Sciences CMLS}, 63:\penalty0
+	1945--1961, 2006.
+	\newblock \doi{10.1007/s00018-006-6094-7}.
+
+	\bibitem[Krueger and Andrews(2011)]{bismark}
+	Felix Krueger and Simon~R Andrews.
+	\newblock Bismark: a flexible aligner and methylation caller for
+	{Bisulfite-Seq} applications.
+	\newblock \emph{Bioinformatics}, 27\penalty0 (11):\penalty0 1571--1572, 2011.
+	\newblock \doi{10.1093/bioinformatics/btr167}.
+
+	\bibitem[Schultz et~al.(2015)Schultz, He, Whitaker, Hariharan, Mukamel, Leung,
+	Rajagopal, Nery, Urich, Chen, et~al.]{methylpy}
+	Matthew~D Schultz, Yupeng He, John~W Whitaker, Manoj Hariharan, Eran~A Mukamel,
+	Danny Leung, Nisha Rajagopal, Joseph~R Nery, Mark~A Urich, Huaming Chen,
+	et~al.
+	\newblock Human body epigenome maps reveal noncanonical {DNA} methylation
+	variation.
+	\newblock \emph{Nature}, 523\penalty0 (7559):\penalty0 212--216, 2015.
+	\newblock \doi{10.1038/nature14465}.
+
+	\bibitem[Zhou et~al.(2021)]{biscuit}
+	Wanding Zhou et~al.
+	\newblock {BISCUIT} -- understand sequencing data with bisulfite conversion.
+	\newblock Software, \url{https://huishenlab.github.io/biscuit}, 2021.
+
+	\bibitem[Kapourani and Sanguinetti(2019)]{kapourani2019melissa}
+	Chantriolnt-Andreas Kapourani and Guido Sanguinetti.
+	\newblock Melissa: Bayesian clustering and imputation of single-cell
+	methylomes.
+	\newblock \emph{Genome Biology}, 20\penalty0 (1):\penalty0 61, 2019.
+	\newblock \doi{10.1186/s13059-019-1665-8}.
+
+	\bibitem[Kapourani et~al.(2021)Kapourani, Argelaguet, Sanguinetti, and
+	Vallejos]{kapourani2021scmet}
+	Chantriolnt-Andreas Kapourani, Ricard Argelaguet, Guido Sanguinetti, and
+	Catalina~A Vallejos.
+	\newblock {scMET}: Bayesian modeling of {DNA} methylation heterogeneity at
+	single-cell resolution.
+	\newblock \emph{Genome Biology}, 22:\penalty0 1--21, 2021.
+	\newblock \doi{10.1186/s13059-021-02329-8}.
+
+	\bibitem[Danese et~al.(2021)Danese, Richter, Chaichoompu, Fischer, Theis, and
+	Colomé-Tatché]{danese2021episcanpy}
+	Anna Danese, Maria~L. Richter, Kridsadakorn Chaichoompu, David~S. Fischer,
+	Fabian~J. Theis, and Maria Colomé-Tatché.
+	\newblock {EpiScanpy}: integrated single-cell epigenomic analysis.
+	\newblock \emph{Nature Communications}, 12:\penalty0 5228, 2021.
+	\newblock \doi{10.1038/s41467-021-25131-3}.
+
+	\bibitem[Welch et~al.(2019)Welch, Kozareva, Ferreira, Vanderburg, Martin, and
+	Macosko]{welch2019single}
+	Joshua~D Welch, Velina Kozareva, Ashley Ferreira, Charles Vanderburg, Carly
+	Martin, and Evan~Z Macosko.
+	\newblock Single-cell multi-omic integration compares and contrasts features of
+	brain cell identity.
+	\newblock \emph{Cell}, 177\penalty0 (7):\penalty0 1873--1887, 2019.
+	\newblock \doi{10.1016/j.cell.2019.05.006}.
+
+	\bibitem[Virtanen et~al.(2020)Virtanen, Gommers, Oliphant, Haberland, Reddy,
+	et~al.]{SciPy}
+	Pauli Virtanen, Ralf Gommers, Travis~E. Oliphant, Matt Haberland, Tyler Reddy,
+	et~al.
+	\newblock {SciPy 1.0: fundamental algorithms for scientific computing in
+		Python}.
+	\newblock \emph{Nature Methods}, 17:\penalty0 261--272, 2020.
+	\newblock \doi{10.1038/s41592-019-0686-2}.
+
+	\bibitem[Baglama and Reichel(2005)]{Baglama2005}
+	James Baglama and Lothar Reichel.
+	\newblock Augmented implicitly restarted {Lanczos} bidiagonalization methods.
+	\newblock \emph{{SIAM} Journal on Scientific Computing}, 27\penalty0
+	(1):\penalty0 19--42, January 2005.
+	\newblock \doi{10.1137/04060593x}.
+	\newblock URL \url{https://doi.org/10.1137/04060593x}.
+
+	\bibitem[Josse and Husson(2012)]{josse2012}
+	Julie Josse and Fran{\c c}ois Husson.
+	\newblock {Handling missing values in exploratory multivariate data analysis
+		methods}.
+	\newblock \emph{{Journal de la Societe Fran{\c c}aise de Statistique}},
+	153\penalty0 (2):\penalty0 79--99, 2012.
+	\newblock URL \url{https://hal.science/hal-00811888}.
+
+	\bibitem[Quinlan and Hall(2010)]{quinlan2010bedtools}
+	Aaron~R Quinlan and Ira~M Hall.
+	\newblock {BEDTools}: a flexible suite of utilities for comparing genomic
+	features.
+	\newblock \emph{Bioinformatics}, 26\penalty0 (6):\penalty0 841--842, 2010.
+	\newblock \doi{10.1093/bioinformatics/btq033}.
+
+	\bibitem[Kireeva et~al.(2014)Kireeva, Ovchinnikova, Tetko, Asiri, Balakin, and
+	Tsivadze]{Kireeva_2014}
+	Natalia~V. Kireeva, Svetlana~I. Ovchinnikova, Igor~V. Tetko, Abdullah~M. Asiri,
+	Konstantin~V. Balakin, and Aslan~Yu Tsivadze.
+	\newblock {Nonlinear dimensionality reduction for visualizing toxicity data:
+		Distance-based versus topology-based approaches}.
+	\newblock \emph{ChemMedChem}, 9\penalty0 (5):\penalty0 1047--1059, 2014.
+	\newblock ISSN 18607187.
+	\newblock \doi{10.1002/cmdc.201400027}.
+
+	\bibitem[Rodriguez et~al.(2013)Rodriguez, Maietta, Ezkurdia, Pietrelli,
+	Wesselink, Lopez, Valencia, and Tress]{appris}
+	Jose~Manuel Rodriguez, Paolo Maietta, Iakes Ezkurdia, Alessandro Pietrelli,
+	Jan-Jaap Wesselink, Gonzalo Lopez, Alfonso Valencia, and Michael~L Tress.
+	\newblock {APPRIS}: annotation of principal and alternative splice isoforms.
+	\newblock \emph{Nucleic Acids Research}, 41\penalty0 (D1):\penalty0 D110--D117,
+	2013.
+	\newblock \doi{10.1093/nar/gks1058}.
+\end{thebibliography}
+
 \newpage
 \section*{Extended Data Figures}
-\vspace{1.5cm}
 
 \setcounter{figure}{0}
-%\renewcommand{\thefigure}{ \arabic{figure}}
 \renewcommand{\figurename}{Extended Data Figure}
@@
 	\begin{center}
-	    \includegraphics[width=\textwidth]{figures/SFig_benchmark.png}
+	    \includegraphics[width=\textwidth]{figures/SFig_benchmark.pdf}
 	\end{center}
@@
     \begin{center}
-        \includegraphics[width=.55\textwidth]{figures/SFig_parameter_sweep.png}
+        \includegraphics[width=.55\textwidth]{figures/SFig_parameter_sweep.pdf}
     \end{center}
@@
 	\begin{center}
-		\includegraphics[width=.85\textwidth]{figures/SFig_benchmark_resources.png}
+		\includegraphics[width=.85\textwidth]{figures/SFig_benchmark_resources.pdf}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[width=.85\textwidth]{figures/SFig_100k_and_CH.png}
+		\includegraphics[width=.85\textwidth]{figures/SFig_100k_and_CH.pdf}
 	\end{center}

### 2024-06-05 paper.tex
editorial main text changes

@@
 Here, we demonstrate that this coarse-graining approach can lead to signal dilution.
-As an alternative, we propose improved strategies to identify more informative regions for methylation quantification, and a more accurate quantitation method than simple averaging.
+We propose improved strategies to identify more informative regions for methylation quantification, and a more accurate quantitation method than simple averaging.
 Our approach enables better discrimination of cell types and other features of interest and reduces the need for large numbers of cells.
@@
 
-To facilitate the analysis of scBS data, we have developed a software tool called \textit{MethSCAn} that implements these methods and provides additional functionality.
+Finally, we present the software tool \textit{MethSCAn} for scBS data analysis: \url{https://anders-biostat.github.io/MethSCAn}
 
@@
 
-Here, we discuss bisulfite sequencing, a method to study DNA methylation.
-In brief, DNA is treated with bisulfite, which converts unmethylated cytosines to uracils which are read as thymine in subsequent PCR, while methylated cytosines are protected from conversion.
+Briefly, in a bisulfite sequencing assay DNA is treated with bisulfite, which converts unmethylated cytosines to uracils which are read as thymine in subsequent PCR, while methylated cytosines are protected from conversion.
 After sequencing, these conversions allow for the determination of the methylation status of all cytosines covered by reads \citep{Frommer_1992}.
@@
 Furthermore, we discuss how to perform comparative analyses.
-Finally, we present \textit{MethSCAn}, a comprehensive software toolkit to perform scBS data analysis, using the described methods.
-
-\subsection*{Standard approach}
+Finally, we present \textit{MethSCAn}, a comprehensive software toolkit to perform scBS data analysis.
 
@@
 The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark}, methylpy \citep{methylpy} or BISCUIT \citep{biscuit}, which are first stored in a compressed format that enables efficient access to all CpG sites of the genome (see Methods).
-A tutorial that showcases the analysis of a small example data set can be found at \url{https://github.com/anders-biostat/MethSCAn}.
+A tutorial that showcases the analysis of a small example data set can be found at \url{https://anders-biostat.github.io/MethSCAn}.
 
@@
 
-We also presented and published an open source software tool, called \textit{MethSCAn}, that provides an easy-to-use implementation of the described algorithms.
+We also presented an open source software tool, called \textit{MethSCAn}, that provides an easy-to-use implementation of the described algorithms.
 It can start directly from the output of methylation callers such as Bismark \citep{bismark}, Biscuit \citep{biscuit} and methylpy \citep{methylpy} and produce a cell\texttimes region matrix.

### 2024-06-05 paper.tex
minor edit

@@
 
-\textit{MethSCAn} is free and open source.
-The package is available on the Python Package Index (PyPI) and can be installed with the Python package installer \texttt{pip}.
-The source code is hosted at \href{https://github.com/anders-biostat/MethSCAn}{https://github.com/anders-biostat/MethSCAn}, where we also provide detailed documentation including a tutorial that demonstrates a complete \textit{MethSCAn} analysis on an example data set.
+\textit{MethSCAn} is available on the Python Package Index (PyPI) and can be installed with the Python package installer \texttt{pip}.
+The source code is hosted at \url{https://github.com/anders-biostat/MethSCAn}, where we also provide detailed documentation including a tutorial that demonstrates a complete \textit{MethSCAn} analysis on an example data set.
 

### 2024-06-05 paper.tex
mv Fig6 caption to main text

@@
 Fig.~\ref{fig:workflow} illustrates a typical scBS data analysis with \texttt{MethSCAn} and provides an overview of the implemented core functionalities.
-The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark}, methylpy \citep{methylpy} or BISCUIT \citep{biscuit}, which are first stored in a compressed format that enables efficient access to all CpG sites of the genome (see Methods).
 A tutorial that showcases the analysis of a small example data set can be found at \url{https://anders-biostat.github.io/MethSCAn}.
+
+The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark}, methylpy \citep{methylpy} or BISCUIT \citep{biscuit}.
+Since it is inconvenient to work with hundreds or thousands of text files, \textit{MethSCAn} provides the \texttt{prepare} command which parses these methylation files and stores their content in a compressed format that enables efficient access to all CpG sites of the genome (see Methods).
+\texttt{methscan prepare} also computes a number of summary statistics for each cell, including the mean genome-wide methylation level and the number of observed CpG sites, i.e.\ the number of CpG sites that have sequencing coverage.
+These summary statistics can be used to detect cells with poor quality.
+The quality of each single cell methylome can furthermore be inspected with \texttt{methscan profile}, which computes the average methylation profile of a set of user-defined genomic regions such as transcription start sites (TSS) at single-cell resolution.
+The TSS profile is a useful quality control plot since methylation shows a characteristic dip roughly \textpm1~kb around the transcription start site in mammalian genomes.
+Cells that do not show this pattern, or cells with few observed CpG sites, may then be discarded with \texttt{methscan filter}.
+
+After quality control, the user has access to the genome-wide VMR detection approach described earlier, using the \texttt{methscan scan} command.
+This produces a BED file that lists the genome coordinates of VMRs.
+To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED file can be used as input for \texttt{methscan matrix}, which quantifies methylation at genomic intervals in all cells.
+The command produces both, the simple percentage of methylated CpG sites, as well as our proposed methylation measure, the shrunken mean of the residuals, which is more robust to variation in sequencing coverage or stochastic differences in read position between cells.
+We note that both \texttt{matrix} and \texttt{profile} accept any valid BED file as input, which means that the user can quantify and visualize methylation at any set of genomic features of interest, such as promoters, enhancers or transcription factor binding sites, obtaining one profile plot per cell.
+The methylation matrix produced by the \texttt{matrix} function can then serve as input for established methods used for the exploration of single-cell data, such as dimensionality reduction and cell clustering.
+
+Finally, after annotation and exploration of the data set, the user may specify two groups of cells for DMR detection with \texttt{methscan diff}.
+This command produces a BED file listing the genome coordinates, adjusted p-values, as well as several other metrics associated with DMRs.
+These DMRs may then be associated with nearby genes, or used for GO enrichment with tools such as GREAT \citep{mclean2010great} to enable functional interpretation.
 
@@
 	\caption{\small \textbf{Overview of the functionalities implemented in the \textit{MethSCAn} package.}\\
-		Common bisulfite sequencing mappers produce one methylation report file per cell.
-		As it is inconvenient to work with thousands of text files, \texttt{methscan prepare} first stores their contents in a compressed format optimized for efficient access to genomic intervals.
-		This command also produces cell-wise summary statistics that can be used to discard low-quality cells with \texttt{methscan filter}.
-		The quality of each cell can furthermore be assessed by visualizing the average methylation around transcription start sites, which are known to show a characteristic dip in mammalian genomes, using \texttt{methscan profile}.
-		This command can also visualize methylation at other genomic features of interest, such as enhancers or transcription factor binding sites.
-		To obtain a methylation matrix analogous to scRNA-seq count matrices, the user must first decide in which genomic intervals methylation should be quantified.
-		The user may either provide genomic regions of \emph{a priori} interest as a BED file, or they may choose to discover VMRs (variably methylated regions) in the data with \texttt{methscan scan}.
-		The \texttt{methscan matrix} command then quantifies methylation at these genomic intervals in all cells.
-		It provides two metrics:
-		the simple percentage of methylated CpG sites and our proposed measure, the shrunken mean of the residuals, which is more robust to variation in sequencing coverage and stochastic differences in read position between cells.
-		The resulting methylation matrix can then serve as input for established methods used for the exploration of single-cell data, such as dimensionality reduction and cell clustering.
-		Finally, after annotation and exploration of the data set, the user may specify two groups of cells for DMR (differentially methylated region) detection with \texttt{methscan diff}.
-		This command produces a BED file listing the genome coordinates, adjusted p-values, as well as several other metrics associated with DMRs.
-		These DMRs may then be associated with nearby genes, or used for GO enrichment with tools such as GREAT \citep{mclean2010great} to enable functional interpretation.
+		Depicted is an scBS data analysis workflow that starts from methylation report files generated by mappers designed for bisulfite-converted sequencing reads.
+		\textit{MethSCAn} facilitates exploration of this data by constructing a methylation matrix suitable for dimensionality reduction, and by enabling the detection of DMRs between groups of cells.
 	}

### 2024-06-05 paper.tex
shorten fig captions

@@
 Furthermore, even cellular sub-states of cells in the continuous neural stem cell lineage were partially separated.
-
 To quantify this performance gain in a more rigorous manner, we used a score that quantifies whether cells were placed, in 15-dimensional PC space, in a neighborhood comprising cells of the same cell type ("neighbor score", see Fig.~\ref{fig:score}E, and Methods for details).
+A higher neighbor score implies better separation of cell types inferred from single-cell transcriptomes of the same cells.
+
 The neighborhood relation in PCA space is relevant as it is used in many downstream analyses, e.g., for clustering.
@@
 		\textbf{(A)} Depicted is a genomic interval (between vertical blue lines) along a chromosome, for which DNA methylation is to be quantified.
-		Two cells cover differing parts of the interval with one read each.
-		If one simply counts for each cell which fraction of the covered CpG sites are methylated, one obtains very different values for the two cells.
 		\textbf{(B)} By averaging each CpG site's methylation over all cells and subsequent smoothing, the thick black ``average methylation curve'' is obtained.
@@
 	\caption{\small \textbf{Correlation of DNA methylation and gene expression}.\\
-		\textbf{(A)} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red): for most promoters, correlation is weak.
+		\textbf{(A)} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red).
 		Promoters are defined as intervals \textpm2~kb around the TSS.
@@
 		The full 1566-cell data set was sub-sampled to simulate smaller data sets (x axis).
-		A higher neighbor score implies better separation of cell types inferred from single-cell transcriptomes of the same cells (A).
 		\textbf{(F)} Proportion of VMRs (blue dots) that have at least 1~bp overlap with other genomic features.
@@
 	\caption{\small \textbf{Detection of differentially methylated regions (DMRs).}\\
-		\textbf{(A)} To identify DMRs, the chromosome is first divided into windows overlapping by a small and fixed size.
-		For each window, the cell’s methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
-		Then the t statistic of these values is obtained as a measure of methylation difference between two groups of cells that were selected by the user (each point represents one of the windows).
-		Upper and lower thresholds (dashed lines) are chosen such that a defined quantile of windows have a t statistic above or below the threshold.
+		\textbf{(A)} Schematic depiction of our DMR detection algorithm.
+		Points represent windows for which the t statistic is obtained, dashed lines represent upper and lower t statistic thresholds.
 		Windows exceeding either threshold are merged if they overlap, yielding DMRs that are lowly methylated in either group of cells.
@@
 \textit{MethSCAn} is available on the Python Package Index (PyPI) and can be installed with the Python package installer \texttt{pip}.
-The source code is hosted at \url{https://github.com/anders-biostat/MethSCAn}, where we also provide detailed documentation including a tutorial that demonstrates a complete \textit{MethSCAn} analysis on an example data set.
+At \url{https://anders-biostat.github.io/MethSCAn} we provide detailed documentation including a link to the source code and a tutorial that demonstrates a complete \textit{MethSCAn} analysis on an example data set.
 

### 2024-06-10 paper.tex
add erc

@@
 SA and LPMK acknowledge funding by the Klaus Tschira Foundation (project 00.022.2019).
-LPMK, MMB, LK, SC, and AMV acknowledge funding from the European Commission via ERC grant 771376 (to AMV), from the DFG via SFB 873, and from the DKFZ.
+LPMK, MMB, LK, SC, and AMV acknowledge funding from the European Commission via ERC grants 771376 (to AMV) and 101071786 (to AMV and SA), from the DFG via SFB 873, and from the DKFZ.
 We thank Alexey Uvarovskii for code refactoring and helpful suggestions on the source code.
@@
 
-LPMK and SA conceived and worked out the method.
+LPMK, SA and AMV conceived the project and LPKM and SA worked out the method.
 LPKM wrote the software package, with contributions from MMB (differential methylation functionality) and LKü.
-LPMK, SO and LKü prepared the use-case demonstrations and performed the benchmarks.
+LPMK, SO, LKü and MMB prepared the use-case demonstrations and performed the benchmarks.
 SC and AMV contributed experimental data and biological expertise.

### 2024-06-10 paper.tex
minor

@@
 SA and LPMK acknowledge funding by the Klaus Tschira Foundation (project 00.022.2019).
-LPMK, MMB, LK, SC, and AMV acknowledge funding from the European Commission via ERC grants 771376 (to AMV) and 101071786 (to AMV and SA), from the DFG via SFB 873, and from the DKFZ.
+LPMK, MMB, LK, SC, and AMV acknowledge funding from the European Commission via ERC grants 771376 (to AMV) and 101071786 (to AMV and SA), from the DFG via SFB 873 (to AMV), and from the DKFZ.
 We thank Alexey Uvarovskii for code refactoring and helpful suggestions on the source code.

### 2024-06-10 paper.tex
update acknowledgments

--- 2024-06-10 paper.tex
@@
 \subsection*{Acknowledgements}
-SA and LPMK acknowledge funding by the Klaus Tschira Foundation (project 00.022.2019).
-LPMK, MMB, LK, SC, and AMV acknowledge funding from the European Commission via ERC grants 771376 (to AMV) and 101071786 (to AMV and SA), from the DFG via SFB 873, and from the DKFZ.
+
+SA acknowledges funding by the Klaus Tschira Foundation (project 00.022.2019).
+AMV acknowledges funding from the European Commission via ERC grant 771376, from the DFG via SFB 873, and from the DKFZ.
 We thank Alexey Uvarovskii for code refactoring and helpful suggestions on the source code.
